### 2021-10-01 ecoop-2022/paper.tex
ecoop: paper skeleton

\documentclass[a4paper,english,cleveref,autoref,thm-restate,,]{lipics-v2021}
\bibliographystyle{plainurl}

%% Thesis:
%%  the addition of sound nominal types to an optionally-typed codebase
%%  can effectively improve peformance

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.

%% TODO before camera-ready
%% -

\title{Gradual Soundness: Lessons from Static Python}
%% ... types make things fast?
%% ... gradual migratory progressive soundness

%\titlerunning{short title}

% > Here’s the thing: I think this is the sort of thing an academic might say “oh,
% > this is too much work, it’d never work” But in fact it looks like they seem to
% > be making it work
% -- Shriram 2021-09-30

\author{Kuang-Chen Lu}{Department of Computer Science, Brown University, USA}{LuKuangchen1024@gmail.com}{}{}
\author{Ben Greenman}{Department of Computer Science, Brown University, USA}{benjamin.l.greenman@gmail.com}{0000-0001-7078-9287}{}
\author{Carl Meyer}{Facebook, Inc.}{carljm@fb.com}{}{}
\author{Dino Viehland}{Facebook, Inc.}{dinoviehland@fb.com}{}{}
\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{joanrpublic@dummycollege.org}{0000-0001-5184-1975}{}

\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Kuang-Chen Lu, Ben Greenman, Carl Meyer, Dino Viehland, and Shriram Krishnamurthi}
%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}}
%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{gradual typing, migratory typing}

\category{} %optional, e.g. invited paper

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{
  Thanks to
   Guido van Rossom for provocative tweets.
  This work was partly supported by the US National Science Foundation.
  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
}

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
  Soundness, migratory typing.
  Novel run-time check strategy, combines concrete and transient for ergonomics.
\end{abstract}

\section{Typing for Performance}
\label{s:intro}

Go go gradual~\cite{tf-dls-2006,st-sfp-2006}.

\paragraph*{Contributions}

\begin{itemize}
\item
  Evidence that gradual soundness works. Performance rises incrementally from runtime engineering and programmer buy-in.
\item
  Informal description of the programming methods that enabled good performance.
  % migration strategy, use of dyn., use of handwritten annotations
\item Formal description of the core language, to promote a reproduction.
\end{itemize}

The language design is a novel point in the space, combining concrete and transient checking strategies.
Concrete has the best performance, but imposes an annotation burden.
Transient mitigates the burden.

\section{A Tour of Static Python}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

\subsection{Language Pipeline}

\begin{itemize}
  \item Pyre as a recommended, optional pre-check
  \item SP type checker
  \item custom bytecode
\end{itemize}

\subsection{Type Dynamic}

The dynamic type is an important part of Static Python.
Dynamic avoids the need for whole-sale migration,
lets programmers work around dynamic features like splatting,
and is (proof?) in some cases necessary for optimal performance.

\subsection{Migration Issues}

%% more to say about the style of migration in SP?

Enabling sound types on a function has a ripple effect on callers.
All untyped callers are now forced to supply type-correct inputs; may call for
refactoring.

Enabling checked types has a huge ripple effect.
The worst case is changing an annotation to a expect a checked data structure:
\begin{enumerate}
  \item all callers must create a checked value by invoking the right constructor, and
  \item all clients of those callers can no longer expect an unchecked type.
\end{enumerate}
Programmers have to trace the annotation back to value-creation points, and then
ensure that all uses of those values are compatible.

\section{Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

\subsection{Surface Language}

\subsection{Evaluation Language, Optimizations}

\subsection{Bug Reports}

While formalizing Static Python, we submitted N bug reports to the language developers.
The developers acknowledged M of these bugs as issues to fix.

\section{Under the Hood: Static Python Specifics}
\label{s:impl}

Purpose: explain significant gaps between the model and the implementation.
E.g. what would the TypeScript team need to know before using our model
to add soundness?

\section{Evaluation}
\label{s:eval}

\subsection{Benchmarks}

\subsection{Production Experience}

\section{Related Work}

Nom~\cite{mt-oopsla-2017} is foundational.
Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.

PyPy is another runtime for Python
Pycket is built on PyPy and improves both Typed Racket~\cite{bbst-oopsla-2017}
and Reticulated Python~\cite{vsc-dls-2019}.

\section{Future Work}

Adapt SP ideas to a new setting, test performance takeaways.

Improve SP with generic types, structural types (lambda), \ldots.

Build an automatic migration tool for Checked data.

Build a static analysis that hoists transient annotations to an early, shared point.
Take care to give quality error messages.

\section{Conclusion}
\label{s:conclusion}

Sound types in Static Python give programmers a way to improve performance.
The change is not forced, not intrusive;
programmers can keep using unsound types when working toward a deadline.

Conjecture that soundness will find bugs too, and lead to more reliable products.
Too soon to say.

\bibliography{bg}

%\appendix

\end{document}

### 2021-10-05 ecoop-2022/paper.tex
ecoop: intro

@@
-\documentclass[a4paper,english,cleveref,autoref,thm-restate,,]{lipics-v2021}
+\documentclass[a4paper,english,cleveref,autoref,thm-restate,anonymous,]{lipics-v2021}
 \bibliographystyle{plainurl}
@@
 
+% > Here’s the thing: I think this is the sort of thing an academic might say “oh,
+% > this is too much work, it’d never work” But in fact it looks like they seem to
+% > be making it work -- Shriram 2021-09-30
+
 %% TODO before submission
@@
 
-%% TODO before camera-ready
+%% TODO after submission, before camera-ready
 %% -
@@
 
-% > Here’s the thing: I think this is the sort of thing an academic might say “oh,
-% > this is too much work, it’d never work” But in fact it looks like they seem to
-% > be making it work
-% -- Shriram 2021-09-30
-
 \author{Kuang-Chen Lu}{Department of Computer Science, Brown University, USA}{LuKuangchen1024@gmail.com}{}{}
@@
 \author{Dino Viehland}{Facebook, Inc.}{dinoviehland@fb.com}{}{}
-\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{joanrpublic@dummycollege.org}{0000-0001-5184-1975}{}
+\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{shriram@brown.edu}{0000-0001-5184-1975}{}
 
@@
 
+\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
+\newcommand{\SP}{Static Python}
+
 \begin{abstract}
@@
 
-Go go gradual~\cite{tf-dls-2006,st-sfp-2006}.
+Gradual typing has attracted significant interest as a solution to
+the old tradeoff between static and dynamic typing.
+The premise is simple: let programmers apply types to part of the
+codebase and leave the rest un-analyzed.
+Run-time checks can in principle enforce the assumptions that
+typed code makes about untyped components, thereby ensuring sound types.
+
+Unfortunately, the high run-time cost of sound types has split
+the community.
+Industry teams have developed innovative type systems that accommodate
+untyped designs, but these systems are all unsound [CITE].
+They intentionally check nothing at run-time when untyped values enter
+typed code.
+Academic teams have primarily focused on the theory of sound
+graduality and its extension to new types and correctness theorems [CITE].
+A few academics have studied the cost of run-time checks
+in detail [CITE] and proposed implementation methods [CITE],
+compiler technology [CITE],
+and even weakened semantics [CITE], but none of these efforts
+decisively close the performance gap.
+The most promising attempt is the \emph{concrete} semantics
+for gradual types [CITE kafka nom etc], but its low performance
+overhead stems from severe restrictions on untyped code.
+These restrictions may not be acceptable to the maintainers of a
+working untyped codebase.
+Indeed, the maintainers might prefer to re-implement their system
+in a typed language as Dropbox recently did to its core sync
+engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}
+
+In short, academic researchers are working to close the performance gap
+without overly restricting the promise of gradual typing.
+Industry researchers are, for the most part, sidestepping the problem with unsound types.
+
+This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
+The \SP{} team at Instagram has developed a sound type system for a subset of
+Python along with a runtime system designed to lower the cost of run-time checks
+and capitalize on the benefits of sound types.
+The language design is a novel twist on concrete types that lets programmers
+begin with a shallow claims about data structures and gradually opt in to deep soundness.
+Remarkably, the performance of Python is an upper bound on their observed performance of
+Static Python.
 
@@
   \item
-  Evidence that gradual soundness works. Performance rises incrementally from runtime engineering and programmer buy-in.
+    Evidence that gradual soundness can work.
+    \SP{} performs well on a microbenchmark suite and, more importantly, in production.
+    The effort needed to convert code to \SP{} is low thanks to its novel application
+    of concrete types.
   \item
-  Informal description of the programming methods that enabled good performance.
-  % migration strategy, use of dyn., use of handwritten annotations
-\item Formal description of the core language, to promote a reproduction.
+    A formal description and analysis of the \SP{} core language.
+    The analysis includes a proof of soundness and (perhaps) a metric for
+    the relative cost of a type migration.
+  \item
+    A full list of the limitations in \SP{} relative to idealized gradual type
+    systems and an informal description of the programming discipline that Instagram
+    follows in their gradually sound codebase.
 \end{itemize}
 
-The language design is a novel point in the space, combining concrete and transient checking strategies.
-Concrete has the best performance, but imposes an annotation burden.
-Transient mitigates the burden.
+\paragraph*{Significance}
+
+We have written this paper to promote two future directions.
+First, we want to encourage system-builders to reproduce the essentials of
+\SP{} in another context and measure the net performance.
+Second, we want to lend focus to theoretical research on gradual typing.
+On one hand, the restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
+may be worthwhile assumptions for theory going forward.
+On the flip side, there are issues in today's \SP{} that might
+be addressed by future research.
 
@@
 lets programmers work around dynamic features like splatting,
-and is (proof?) in some cases necessary for optimal performance.
+and is in some cases necessary for optimal performance (evidence?).
 

### 2021-10-12 ecoop-2022/paper.tex
ecoop: quick intro pass

@@
   Thanks to
-   Guido van Rossom for provocative tweets.
+   Guido van Rossom for stimulating tweets.
   This work was partly supported by the US National Science Foundation.
@@
 Gradual typing has attracted significant interest as a solution to
-the old tradeoff between static and dynamic typing.
-The premise is simple: let programmers apply types to part of the
-codebase and leave the rest un-analyzed.
+the impasse between static and dynamic typing.
+The premise is simple: let programmers introduce types in part of a
+codebase and while leaving the rest untyped.
 Run-time checks can in principle enforce the assumptions that
-typed code makes about untyped components, thereby ensuring sound types.
+typed code makes about untyped components, thereby ensuring that
+the types are sound no matter how the untyped code behaves.
 
 Unfortunately, the high run-time cost of sound types has split
-the community.
+the gradual typing community.
 Industry teams have developed innovative type systems that accommodate
 untyped designs, but these systems are all unsound [CITE].
-They intentionally check nothing at run-time when untyped values enter
+These systems intentionally check nothing at run-time when untyped values enter
 typed code.
 Academic teams have primarily focused on the theory of sound
-graduality and its extension to new types and correctness theorems [CITE].
+gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
 A few academics have studied the cost of run-time checks
@@
 Indeed, the maintainers might prefer to re-implement their system
-in a typed language as Dropbox recently did to its core sync
+in a typed language, as Dropbox recently did with its core sync
 engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}
@@
 The \SP{} team at Instagram has developed a sound type system for a subset of
-Python along with a runtime system designed to lower the cost of run-time checks
-and capitalize on the benefits of sound types.
-The language design is a novel twist on concrete types that lets programmers
-begin with a shallow claims about data structures and gradually opt in to deep soundness.
-Remarkably, the performance of Python is an upper bound on their observed performance of
+Python along with a runtime system designed to
+capitalize on the benefits of sound types.
+The language design is a variation of concrete types that lets programmers
+begin with shallow claims about data structures and gradually opt in to deeper soundness.
+Remarkably, the performance of Python is an upper bound on the observed performance of
 Static Python.
@@
 Second, we want to lend focus to theoretical research on gradual typing.
-On one hand, the restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
+The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
 may be worthwhile assumptions for theory going forward.

### 2021-10-14 ecoop-2022/paper.tex
ecoop: quick notes on eval and communication

@@
 
+\subsection{Goals and Limitations}
+
+The designers of Static Python hope to eventually build a system that is sound for arbitrary Python code,
+but are not there yet.
+The known side channels include:
+eval, the locals and globals dicts (the latter can be accessed through frame objects at the moment),
+and mutable closure cells.
+Our model does not cover these either.
+We essentially ban all these features, rejecting any program in which they appear.
+Code mining in JS (and perhaps the newer R work?) suggests that at least eval
+uses can be rewritten without too much pain. The few exceptions would point
+to language weaknesses / call for new semantically-descriptive features.
+
+FILL do we stop at strict modules?\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
+A strict module is 100\% repeatable and isolated from the side effects of other modules.
+
 \subsection{Surface Language}
@@
 
+\subsection{Communication}
+
+How did we communicate effectively?
+Starting point was GT survey [CITE].
+Moved forward with example programs, see emails and github issues.
+Our formalism had to cover their unit tests.
+FILL
+
 \bibliography{bg}

### 2021-10-19 ecoop-2022/paper.tex
ecoop: contributions pass

@@
 without overly restricting the promise of gradual typing.
-Industry researchers are, for the most part, sidestepping the problem with unsound types.
+Industry researchers are sidestepping the problem with unsound types---for the most part.
 
@@
   \item
-    Evidence that gradual soundness can work.
-    \SP{} performs well on a microbenchmark suite and, more importantly, in production.
-    The effort needed to convert code to \SP{} is low thanks to its novel application
-    of concrete types.
+    Evidence that gradual soundness delivers fast performance incrementally,
+    as programmers migrate to sound types.
+    \SP{} can be applied to a codebase with no refactoring and no performance overhead.
+    After refactoring to sound types, \SP{} performs well on a microbenchmark
+    suite and in production.
   \item
     A formal description and analysis of the \SP{} core language.
-    The analysis includes a proof of soundness and (perhaps) a metric for
-    the relative cost of a type migration.
+    The analysis includes a proof of type soundness,
+    a proof that optimization soundly improves code,
+    and [FILL perhaps] a metric to quantify the refactoring cost and optimization benefit
+    of a type migration.
   \item
-    A full list of the limitations in \SP{} relative to idealized gradual type
-    systems and an informal description of the programming discipline that Instagram
-    follows in their gradually sound codebase.
+    A list of the expressiveness limitations in \SP{} relative to
+    idealized gradual type systems,
+    and an informal description of the programming discipline that Instagram
+    follows in their gradually-sound codebase.
 \end{itemize}
@@
 
-We have written this paper to promote two future directions.
-First, we want to encourage system-builders to reproduce the essentials of
-\SP{} in another context and measure the net performance.
+We have written this paper with two audiences in mind.
+First, we want to encourage system-builders to reproduce the compelling
+\SP{} language design in another context and measure performance.
 Second, we want to lend focus to theoretical research on gradual typing.

### 2021-10-20 ecoop-2022/paper.tex
ecoop: clarify 'flip' research direction
ecoop: sec2 intro

@@
 \newcommand{\SP}{Static Python}
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 may be worthwhile assumptions for theory going forward.
-On the flip side, there are issues in today's \SP{} that might
-be addressed by future research.
+On the flip side, future research may find a way to lift restrictions
+that \SP{} currently adopts.
 
-\section{A Tour of Static Python}
+\section{\SP{} and the Cinder JIT}
 \label{s:tour}
@@
 
+\SP{} is a type checker and bytecode compiler.
+It validates programs written in ordinary Python syntax
+and generates type-specific bytecode for the Cinder JIT compiler.
+Cinder is a fork of CPython 3.8 that introduces a method-based
+JIT and several bytecode instructions.
+
+Running \SP{} code is no different than standard Python.
+The main Cinder executable generates bytecode and performs
+optimizations on the fly, same as CPython.
+
+There are a few restrictions on classes
+and a few optional types that programmers may wish to use,
+but otherwise Python programs can run via \SP{}.
+%% TODO awk
+
+\subsection{Example Program}
+
+\begin{figure}
+  %% TODO replace with a real program
+%  \begin{minipage}[t]{0.5\columnwidth}
+  \begin{verbatim}
+    from __static__ import PyDict
+
+    def f(x: PyDict[str, int]):
+      return x["A"]
+
+    f({"A": 1}) # ==> 1
+  \end{verbatim}
+%  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
+%  \begin{verbatim}
+%    from __static__ import PyDict
+%
+%    def f(x: PyDict[str, int]):
+%      check_args(x, PyDict)
+%      return cast(x["A"], int)
+%
+%    invoke_function(f, {"A": 1}) # ==> 1
+%  \end{verbatim}
+%  %  ....
+%  %  CHECK_ARGS
+%  %  ....
+%  %  INVOKE_FUNCTION
+%  \end{minipage}
+%  \begin{verbatim}
+%    from __static__ import CheckedDict
+%
+%    def f(x: CheckedDict[str, int]):
+%      return x["A"]
+%
+%    f(CheckedDict({"A": 1}))
+%  \end{verbatim}
+  \caption{A first \SP{} program}
+  \label{fig:sp-example}
+\end{figure}
+
+\Cref{fig:sp-example} (left) presents a first example program
+that defines a function \code{f} and calls it from typed code.
+The syntax is normal Python with type annotations.
+Only the import statement makes special use of \SP{};
+it imports a type \code{PyDict} that describes Python dictionaries.
+
+Although this program is typed, a different untyped module is free
+to import the function \code{f} and invoke it with any sort of
+argument.
+For this reason, the compiled version of \cref{fig:sp-example} must
+include a few run-time casts:
+\begin{itemize}
+  \item
+    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
+    Concretely, this check accepts any dictionary.
+  \item
+    Because the domain check does not validate the contents of an incoming dictionary,
+    the dict access must check that its result matches the function codomain.
+\end{itemize}
+These checks ensure a kind of type soundness.
+
+\SP{} also uses types to drive optimizations.
+Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
+there is no need to check that it sends type-correct inputs.
+Thus, the compiled call uses a Cinder bytecode instruction to skip past the
+domain check and immediately enter the function body.
+
+The tension between checks that enforce soundness and optimizations allowed by soundness
+is a central theme of \SP{}.
+
 \subsection{Language Pipeline}
 
-\begin{itemize}
-  \item Pyre as a recommended, optional pre-check
+\begin{enumerate}
+  \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
   \item SP type checker
-  \item custom bytecode
-\end{itemize}
+  \item Custom bytecode, Cinder JIT
+\end{enumerate}
 
@@
 
-\subsection{Migration Issues}
+\subsection{Checked Types}
 
@@
 ensure that all uses of those values are compatible.
+
+\subsection{Restrictions}
+
+No overloading.
+At most one method per class and name.
+Ensures that receiver type does not affect dispatch.
+
+Overrides must be compatible, if done in static code.
+What's the reason?
+
+\subsection{Additional Options}
+
+strict modules
+
+decorator to force inlining, for 1-line functions
 

### 2021-10-20 ecoop-2022/paper.tex
ecoop: more sec2 filling

@@
 
-There are a few restrictions on classes
-and a few optional types that programmers may wish to use,
-but otherwise Python programs can run via \SP{}.
-%% TODO awk
+Almost all Python programs can run under \SP{} with no changes.
+The few exceptions are due to restrictions that \SP{} puts
+on classes to promote fast-running bytecode.
 
@@
 
-\Cref{fig:sp-example} (left) presents a first example program
+\Cref{fig:sp-example} presents a first example program
 that defines a function \code{f} and calls it from typed code.
@@
     The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
-    Concretely, this check accepts any dictionary.
+    Concretely, this check accepts any Python dictionary.
   \item
@@
 
+[FILL]
+
 \begin{enumerate}
@@
 \subsection{Type Dynamic}
+
+[FILL]
 
@@
 %% more to say about the style of migration in SP?
+[FILL]
 
@@
 
-\subsection{Restrictions}
+\subsection{Virtual Method Tables}
 
-No overloading.
-At most one method per class and name.
-Ensures that receiver type does not affect dispatch.
+\SP{} adds vtables to Python.
+[FILL]
 
-Overrides must be compatible, if done in static code.
-What's the reason?
+\subsection{Code Restrictions}
 
-\subsection{Additional Options}
+\SP{} imposes several restrictions on classes and objects
+to support efficient compilation:
+
+\begin{itemize}
+  \item \textbf{First-order Classes}.
+    In order to benefit from types, class definitions must be defined at the top level of a module.
+    Nested classes are supported by the language, but have type \code{Dynamic} according
+    to the type checker.
+
+  \item \textbf{No overloading}.
+    Methods cannot be overloaded; each combination of a class
+    and a name can refer to at most one method.
+    This ensures that method resolution depends only on the
+    type of the receiver object.
+    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
+    % Q. does resolution depend on args?
+
+  %\item \textbf{Single Inheritance}.
+  %  Multiple inheritance is not supported.
+  %  But, it might someday be supported.
+
+  \item \textbf{Compatible method overrides}.
+    Methods may be overridden, either by \SP{} classes or
+    untyped Python classes.
+    Overrides that appear in \emph{the same} \SP{} file as the class definition, however,
+    must supply a method whose type is a static subtype of the previous method.
+    In particular, a parent method that returns type \code{int} cannot be
+    overridden by a child method that returns type \code{Dynamic}.
+    %% TODO what does this enable ... optimizations in the same file?
+
+  % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
+  % but the normal field access syntax skips any overrides (o.f)
+
+  \item \textbf{Compatible field overrides}.
+    Similarly, object fields and class variables may be overridden
+    only with subtype-compatible values.
+    In fact, \SP{} compiles fields using Python slots
+    declarations (\code{\_\_slots\_\_}).
+    This declaration prevent new fields from arising at runtime,
+    but saves space and speeds up field accesses.
+
+\end{itemize}
+
+\subsection{Type Restrictions}
+
+Certain types are not supported:
+
+\begin{itemize}
+  \item \textbf{Parameterized Classes}.
+    At present, \SP{} treats all class parameters as type \code{Dynamic}.
+    The team plans to lift this restriction (\cref{s:future}).
+
+  \item \textbf{Callable}.
+    The type system cannot yet express first-class functions.
+    The team plans to lift this restriction~(\cref{s:future}),
+    but there is little internal pressure to do so.
+
+  \item \textbf{Dynamic Checked Types}.
+    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
+    keys or values.
+    [FILL] no desire, and makes runtime checks much simpler.
+
+  \item \textbf{Recursive Types}.
+    Sadly this is normal for Python type checkers.
+
+\end{itemize}
+
+\subsection{Additional Features}
 
@@
 decorator to force inlining, for 1-line functions
+
+ctypes for primitives, unboxed representation
 
@@
 \section{Future Work}
+\label{s:future}
 

### 2021-10-22 ecoop-2022/paper.tex
ecoop: temporary note about lack of Py Int optimizations
ecoop: sec2 intro edits

@@
 
+% {Language Pipeline}
+% \begin{enumerate}
+%   \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
+%   \item SP type checker
+%   \item Custom bytecode, Cinder JIT
+% \end{enumerate}
+
 \SP{} is a type checker and bytecode compiler.
 It validates programs written in ordinary Python syntax
-and generates type-specific bytecode for the Cinder JIT compiler.
-Cinder is a fork of CPython 3.8 that introduces a method-based
-JIT and several bytecode instructions.
+and generates type-specific bytecode for the Cinder runtime system.
+Cinder is an extension of CPython 3.8 that adds a method-based
+JIT compiler, virtual method tables, and several bytecode instructions.
+In sum, the projects consists of both a new type system (\SP{})
+and a new language (Cinder = Python extended to leverage types).
 
-Running \SP{} code is no different than standard Python.
+Running \SP{} code is no different than standard Python,
+though it takes place on a different platform.
 The main Cinder executable generates bytecode and performs
 optimizations on the fly, same as CPython.
+This is one of the main advantages of \SP{} over mypyc,
+which implements type-directed optimizations for CPython
+by compiling source code to C extension modules.
 
@@
 on classes to promote fast-running bytecode.
-
-\subsection{Example Program}
+In general, switching to \SP{} requires minimal effort.
+Any additional work that developers are willing to invest
+in refining their types is rewarded by performance gains.
 
@@
 
-\subsection{Language Pipeline}
-
-[FILL]
-
-\begin{enumerate}
-  \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
-  \item SP type checker
-  \item Custom bytecode, Cinder JIT
-\end{enumerate}
-
 \subsection{Type Dynamic}
@@
 
-\subsection{Virtual Method Tables}
+\subsection{Runtime Changes}
 
-\SP{} adds vtables to Python.
-[FILL]
+\begin{itemize}
+  \item
+    \textbf{VTables}
+    Cinder adds virtual method tables to all classes.
+
+  \item
+    \textbf{Bytecode Optimization}
+    (could use a full section)
+
+  %\item
+  %  \textbf{Shadow Frames}
+  %  (CinderDoc/static_python.rst)
+  %  magic import that signals to the Static
+  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
+  %  improves performance of function calls by avoiding the creation of full
+  %  Python frame objects until they are definitely needed (e.g. if an
+  %  exception is raised.) In the future this should become default.
+
+\end{itemize}
 
@@
     [FILL] no desire, and makes runtime checks much simpler.
+    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now
 
@@
 
+Optimize Python integer operations (in addition to the ctypes).
+Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
+or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
+
 \section{Conclusion}

### 2021-10-25 SP_tests_to_our_python_tests.py
pass all

@@ -167,6 +167,7 @@ for test in read_tests(input_file):
         assert not 'box' in test
         assert not 'cbool' in test
         assert not 'Array' in test
+        assert not '*args' in test
     except AssertionError:
         continue
     except ValueError:

### 2021-10-25 conformance_suite/test_incompat_override_method_starargs.py

@@ -1,9 +1,0 @@
-# test_incompat_override_method_starargs.py
-# This should fail.
-
-class A:
-    def m(self) -> int:
-        return 42
-class B(A):
-    def m(self, *args) -> int:
-        return 0

### 2021-10-25 conformance_suite/test_incompat_override_method_starkwargs.py

@@ -1,9 +1,0 @@
-# test_incompat_override_method_starkwargs.py
-# This should fail.
-
-class A:
-    def m(self) -> int:
-        return 42
-class B(A):
-    def m(self, **args) -> int:
-        return 0

### 2021-10-25 ecoop-2022/paper.tex

@@
 \begin{abstract}
-  Soundness, migratory typing.
-  Novel run-time check strategy, combines concrete and transient for ergonomics.
+  TBD
+%  Soundness, migratory typing.
+%  Novel run-time check strategy, combines concrete and transient for ergonomics.
 \end{abstract}
@@
 We have written this paper with two audiences in mind.
-First, we want to encourage system-builders to reproduce the compelling
+First, we want to encourage system-builders to reproduce the
 \SP{} language design in another context and measure performance.
 Second, we want to lend focus to theoretical research on gradual typing.
-The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
+The restrictions that \SP{} adopts
 may be worthwhile assumptions for theory going forward.
-On the flip side, future research may find a way to lift restrictions
-that \SP{} currently adopts.
+On the flip side, future research may find a way to lift some of
+the restrictions currently in \SP{}.
 
@@
 It validates programs written in ordinary Python syntax
-and generates type-specific bytecode for the Cinder runtime system.
+and generates type-specific instructions for the Cinder runtime system.
+The type system can express common Python idioms and is guaranteed
+sound through run-time checks, which in turn enable optimizations.
+The type system also includes a dynamic type as an escape hatch to
+untyped behavior.
+An unannotated variable behaves exactly as it would in Python.
+
 Cinder is an extension of CPython 3.8 that adds a method-based
-JIT compiler, virtual method tables, and several bytecode instructions.
-In sum, the projects consists of both a new type system (\SP{})
-and a new language (Cinder = Python extended to leverage types).
+JIT compiler, virtual method tables, and several bytecode instructions
+to directly express type checks and type-based optimizations.
+These enhancements take full advantage of \SP{} type information;
+they are built on top of CPython so that untyped code runs on
+the platform.
 
-Running \SP{} code is no different than standard Python,
-though it takes place on a different platform.
-The main Cinder executable generates bytecode and performs
-optimizations on the fly, same as CPython.
-This is one of the main advantages of \SP{} over mypyc,
-which implements type-directed optimizations for CPython
-by compiling source code to C extension modules.
-
-Almost all Python programs can run under \SP{} with no changes.
-The few exceptions are due to restrictions that \SP{} puts
-on classes to promote fast-running bytecode.
-In general, switching to \SP{} requires minimal effort.
-Any additional work that developers are willing to invest
-in refining their types is rewarded by performance gains.
+Using \SP{} and Cinder is as easy as installing a new version
+of CPython.
+The main executable compiles programs to bytecode on the fly,
+same as CPython, allowing standard developer tools to work.
+By contrast, the only effective way to get type-directed optimizations
+without changing CPython is to use C extension modules,
+which introduce an extra compilation step and get in the way
+of tooling~(\cref{s:related}).
 
@@
 
-\Cref{fig:sp-example} presents a first example program
-that defines a function \code{f} and calls it from typed code.
+\Cref{fig:sp-example} presents a first example program.
+It defines a function \code{f} and calls it from typed code.
 The syntax is normal Python with type annotations.
@@
     Because the domain check does not validate the contents of an incoming dictionary,
-    the dict access must check that its result matches the function codomain.
+    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
 \end{itemize}
-These checks ensure a kind of type soundness.
+These checks ensure a \emph{shape-level} notion of type soundness~(\cite{s:model}).
+In short: if an expression has type \code{T} and reduces to a value, then the value
+matches the top-level constructor, or shape, of the type.
+The shape for a Python dictionary is merely \code{dict}.
+Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
+type that predicts the keys and values in a special kind of dictionary.
 
-\SP{} also uses types to drive optimizations.
+Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
 Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
@@
 
-The tension between checks that enforce soundness and optimizations allowed by soundness
-is a central theme of \SP{}.
+The checks and optimizations in this example illustrate the general balancing act
+of \SP{}.
+Type soundness requires run-time checks, but also enables optimizations.
+The gamble is that the optizations help more that the checks hurt.
 
-\subsection{Type Dynamic}
+\subsection{Type System Highlights}
+
+\subsubsection{Type Dynamic}
 
@@
 
-\subsection{Checked Types}
+\subsubsection{Checked Types}
 
@@
 
-\subsection{Runtime Changes}
+\subsubsection{CTypes}
+
+[FILL]
+CTypes for primitives, unboxed representation.
+Casts and unboxes happen at the boundary to Python.
+Straightforward, not included in model.
+Important for optimizations.
+
+% https://github.com/facebookincubator/cinder/issues/52
+
+\subsubsection{Restrictions}
+
+OO restrictions:
 
 \begin{itemize}
-  \item
-    \textbf{VTables}
-    Cinder adds virtual method tables to all classes.
-
-  \item
-    \textbf{Bytecode Optimization}
-    (could use a full section)
-
-  %\item
-  %  \textbf{Shadow Frames}
-  %  (CinderDoc/static_python.rst)
-  %  magic import that signals to the Static
-  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
-  %  improves performance of function calls by avoiding the creation of full
-  %  Python frame objects until they are definitely needed (e.g. if an
-  %  exception is raised.) In the future this should become default.
-
-\end{itemize}
-
-\subsection{Code Restrictions}
-
-\SP{} imposes several restrictions on classes and objects
-to support efficient compilation:
-
-\begin{itemize}
-  \item \textbf{First-order Classes}.
+  \item \textbf{First-order Typed Classes}.
     In order to benefit from types, class definitions must be defined at the top level of a module.
@@
 
-  \item \textbf{No overloading}.
+  \item \textbf{No Overloading}.
     Methods cannot be overloaded; each combination of a class
@@
 
-\subsection{Type Restrictions}
-
-Certain types are not supported:
+Type restrictions:
 
@@
 
+\subsection{Runtime System Highlights}
+
+\begin{itemize}
+  \item
+    \textbf{VTables}
+    Cinder adds virtual method tables to typed classes.
+    ``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
+    the static compiler will only emit against a statically known static type.'' -Carl
+
+%    No, there’s no overload resolution by arguments in Python. This difference
+%    in behavior isn’t so much desired as just a consequence of making
+%    `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
+%    the callable is placed on the stack, then the arguments, then there is a
+%    `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
+%    means that first the callable is resolved, then the arguments. But with
+%    `INVOKE_METHOD` we want to resolve the callable as part of the invoke
+%    itself, since this gives us opportunity to inline-cache the target of the
+%    call instead of always having to call something dynamic and unknown that’s
+%    on the stack. So that necessarily implies that first the arguments are
+%    resolved and placed on the stack, then the callable is resolved as part of
+%    the `INVOKE_METHOD`. -Carl
+
+  \item
+    \textbf{Bytecode Optimization}
+    (could use a full section)
+
+  %\item
+  %  \textbf{Shadow Frames}
+  %  (CinderDoc/static_python.rst)
+  %  magic import that signals to the Static
+  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
+  %  improves performance of function calls by avoiding the creation of full
+  %  Python frame objects until they are definitely needed (e.g. if an
+  %  exception is raised.) In the future this should become default.
+
+\end{itemize}
+
 \subsection{Additional Features}
+
+% things we did not model ... is that worth making a distinction here?
 
@@
 decorator to force inlining, for 1-line functions
-
-ctypes for primitives, unboxed representation
 
@@
 \section{Related Work}
+\label{s:related}
 
@@
 and Reticulated Python~\cite{vsc-dls-2019}.
+
+This is one of the main advantages of \SP{} over mypyc,
+which implements type-directed optimizations for CPython
+by compiling source code to C extension modules.
 

### 2021-10-26 ecoop-2022/paper.tex
ecoop: ctypes overview

@@
 It validates programs written in ordinary Python syntax
-and generates type-specific instructions for the Cinder runtime system.
+and generates type-specific instructions for the Cinder runtime.
 The type system can express common Python idioms and is guaranteed
@@
 
-\subsubsection{CTypes}
-
-[FILL]
-CTypes for primitives, unboxed representation.
-Casts and unboxes happen at the boundary to Python.
-Straightforward, not included in model.
-Important for optimizations.
-
-% https://github.com/facebookincubator/cinder/issues/52
-
 \subsubsection{Restrictions}
@@
 
+  \item
+    \textbf{Checked Dict}
+    Implementation too, to support the types.
+
 \end{itemize}
@@
 decorator to force inlining, for 1-line functions
+
+\subsubsection{Primitive Types}
+
+For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types.
+The basic primitives describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
+Two additional primitives, \code{Array} and \code{Vector}, describe collections.
+Cinder implements these types with unboxed C numbers and C arrays, which cost
+(and do!) much less than their Python counterparts.
+
+% For now, programmers have to write and manage primitive types.
+% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.
+
+Primitives are carefully limited by the type system.
+Neither a module-level nor a closure-level variable may have a primitive
+type---because untyped code can mutate such variables.
+Primitive types are incompatible with the dynamic type,
+and there are no implicit upcasts from primitives
+to a matching Python type.
+Even a conjunction disallows mixing; for example, the \code{and} operator
+requires either two Python booleans or two primitive booleans.
+
+The only way for Python values and primitive values to interact
+is through dedicated conversion functions.
+Cinder handles conversions at the boundaries between \SP{} code
+and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
+Within typed code, programmers must satisfy the type checker with
+appropriate conversions.
+
+% https://github.com/facebookincubator/cinder/issues/52
 

### 2021-10-27 ecoop-2022/paper.tex

@@
 Nom~\cite{mt-oopsla-2017} is foundational.
+
+\citet{pmmwplck-oopsla-2013} core calculus for Python language, foundational.
+
 Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.
@@
 
-This is one of the main advantages of \SP{} over mypyc,
+Type systems for Python ...
+mypy pyre pytype;
+Reticulated;
+
+Optimizing type systems for Python ...
+Reticulated;
+mypyc;
+Developer experience is one of the main advantages of \SP{} over mypyc,
 which implements type-directed optimizations for CPython

### 2021-10-27 ecoop-2022/paper.tex
ecoop: refine Array prose

@@
 
-For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types.
-The basic primitives describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
-Two additional primitives, \code{Array} and \code{Vector}, describe collections.
-Cinder implements these types with unboxed C numbers and C arrays, which cost
-(and do!) much less than their Python counterparts.
-
-% For now, programmers have to write and manage primitive types.
-% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.
+For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
+that describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
+Cinder implements these types with unboxed C values, which are much simpler and cheaper
+than their Python counterparts.
+\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
+These are not themselves primitive.
 
@@
 requires either two Python booleans or two primitive booleans.
-
 The only way for Python values and primitive values to interact
@@
 appropriate conversions.
+
+% For now, programmers have to write and manage primitive types.
+% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.
 
@@
 
-\citet{pmmwplck-oopsla-2013} core calculus for Python language, foundational.
+Full Monty(?)~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational.
 

### 2021-10-27 ecoop-2022/paper.tex
ecoop: type Dyn notes

@@
   \item
-    A list of the expressiveness limitations in \SP{} relative to
-    idealized gradual type systems,
-    and an informal description of the programming discipline that Instagram
+    In-depth discussions of the expressiveness limitations in \SP{} relative to
+    idealized gradual type systems, and the programming discipline that Instagram
     follows in their gradually-sound codebase.
@@
 
+No way to write the dynamic type.
+Instead, it fills in for absent annotations.
+Either totally missing or a missing generic parameter.
+
+%% TODO are there any types C[T0, T1] where you might want C[dyn, T1]? how to express?
+
+Cannot appear everywhere ... ChkDict can't hold a Dynamic,
+and primitives are incompatible with Dynamic.
+
+Dynamic behavior is always (?) the same as normal Python.
+
 \subsubsection{Checked Types}
@@
   \item \textbf{No Overloading}.
-    Methods cannot be overloaded; each combination of a class
+    Methods cannot be overloaded. Each combination of a class
     and a name can refer to at most one method.

### 2021-10-28 SP_tests_to_our_python_tests.py
push

@@ -116,10 +116,13 @@ def translate_simple_compile_test(test):
         ]) + code
     else:
         assert False
+
+    commented_src = '\n' + '\n'.join('# ' + line for line in test.splitlines())
+    content += commented_src + '\n'
     return (name, content)
 
 
-def translate_less_simple_compile_test(test):
+def translate_less_simple_compile_test(test: str):
     # Capture tests that look like
     #     def test_name(self) -> None:
     #         codestr = """
@@ -150,6 +153,8 @@ def translate_less_simple_compile_test(test):
         '',
         ''
     ]) + code
+    commented_src = '\n' + '\n'.join('# ' + line for line in test.splitlines())
+    content += commented_src + '\n'
     return (name, content)
 
 
@@ -168,9 +173,9 @@ for test in read_tests(input_file):
         # ⬆️ This test uses an unbound identifier
         # Skip more
         assert not 'while' in test # TODO
-        assert not 'reveal_type' in test # TODO
         assert not '@staticmethod' in test # TODO
         assert not '@final' in test # TODO
+        assert not 'reveal_type' in test # TODO
         assert not 'nonlocal' in test
         assert not 'global' in test
         assert not '...' in test
@@ -192,7 +197,7 @@ for test in read_tests(input_file):
         try:
             (name, content) = translate_less_simple_compile_test(test)
         except AssertionError:
-            print(name)
+            print("SKIPPED", name)
             # exit(1)
             continue
 

### 2021-10-28 conformance_suite/test_assign_generic_optional_2.py

@@ -4,3 +4,11 @@
 from typing import Optional
 def f():
     x: Optional = 42 + 1
+# def test_assign_generic_optional_2(self):
+#     codestr = """
+#         from typing import Optional
+#         def f():
+#             x: Optional = 42 + 1
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_assign_test_var.py

@@ -6,3 +6,12 @@ def f(x: Optional[int]) -> int:
     if x is None:
         x = 1
     return x
+# def test_assign_test_var(self):
+#     codestr = """
+#         from typing import Optional
+#         def f(x: Optional[int]) -> int:
+#             if x is None:
+#                 x = 1
+#             return x
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_assign_type_propagation.py

@@ -4,3 +4,10 @@
 def test() -> int:
     x = 5
     return x
+# def test_assign_type_propagation(self):
+#     codestr = """
+#         def test() -> int:
+#             x = 5
+#             return x
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_bool_int.py

@@ -4,3 +4,10 @@
 def f():
     x: int = True
     return x
+# def test_bool_int(self):
+#     codestr = """
+#         def f():
+#             x: int = True
+#             return x
+#     """
+#     self.compile(codestr)

### 2021-10-28 conformance_suite/test_cast_wrong_args.py

@@ -4,3 +4,11 @@
 from __static__ import cast
 def f():
     cast(42)
+# def test_cast_wrong_args(self):
+#     codestr = """
+#         from __static__ import cast
+#         def f():
+#             cast(42)
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_clen_bad_arg.py

@@ -4,3 +4,13 @@
 from __static__ import clen
 def f(l):
     clen(l)
+# def test_clen_bad_arg(self):
+#     codestr = """
+#         from __static__ import clen
+#         def f(l):
+#             clen(l)
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError, "bad argument type 'dynamic' for clen()"
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_compile_dict_get_typed.py

@@ -5,3 +5,11 @@ from __static__ import CheckedDict
 def testfunc():
     x = CheckedDict[int, str]({42: 'abc', })
     y: str | None = x.get(42)
+# def test_compile_dict_get_typed(self):
+#     codestr = """
+#         from __static__ import CheckedDict
+#         def testfunc():
+#             x = CheckedDict[int, str]({42: 'abc', })
+#             y: str | None = x.get(42)
+#     """
+#     self.compile(codestr)

### 2021-10-28 conformance_suite/test_duplicate_function_replaces_class.py

@@ -3,3 +3,12 @@
 
 class X: pass
 def X(): pass
+# def test_duplicate_function_replaces_class(self) -> None:
+#     codestr = """
+#         class X: pass
+#         def X(): pass
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError, "function conflicts with other member X in <module>"
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_duplicate_function_replaces_function.py

@@ -3,3 +3,13 @@
 
 def f(): pass
 def f(): pass
+# def test_duplicate_function_replaces_function(self) -> None:
+#     codestr = """
+#         def f(): pass
+#         def f(): pass
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         "function 'function <module>.f' conflicts with other member",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_final.py

@@ -3,3 +3,9 @@
 
 from typing import Final
 x: Final[int] = 0xdeadbeef
+# def test_final(self):
+#     codestr = """
+#     from typing import Final
+#     x: Final[int] = 0xdeadbeef
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_frozenset_constant.py

@@ -7,3 +7,13 @@ def i(s: str) -> bool:
     return i in {"a", "b"}
 def t() -> bool:
     return i("p")
+# def test_frozenset_constant(self):
+#     codestr = """
+#     from __static__ import inline
+#     @inline
+#     def i(s: str) -> bool:
+#         return i in {"a", "b"}
+#     def t() -> bool:
+#         return i("p")
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_if_else_optional.py

@@ -17,3 +17,23 @@ def f(x: Optional[C], y: Optional[C]):
     else:
         return g(x)
     return None
+# def test_if_else_optional(self):
+#     codestr = """
+#         from typing import Optional
+#         class C:
+#             def __init__(self):
+#                 self.field = self
+#         def g(x: C):
+#             pass
+#         def f(x: Optional[C], y: Optional[C]):
+#             if x is None:
+#                 x = y
+#                 if x is None:
+#                     return None
+#                 else:
+#                     return g(x)
+#             else:
+#                 return g(x)
+#             return None
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_if_else_optional_return.py

@@ -9,3 +9,15 @@ def f(x: Optional[C]):
     if x is None:
         return 0
     return x.field
+# def test_if_else_optional_return(self):
+#     codestr = """
+#         from typing import Optional
+#         class C:
+#             def __init__(self):
+#                 self.field = self
+#         def f(x: Optional[C]):
+#             if x is None:
+#                 return 0
+#             return x.field
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_if_else_optional_return_in_else_assignment_in_if.py

@@ -8,3 +8,14 @@ def f(x: Optional[int]) -> int:
     else:
         return 2
     return x
+# def test_if_else_optional_return_in_else_assignment_in_if(self):
+#     codestr = """
+#         from typing import Optional
+#         def f(x: Optional[int]) -> int:
+#             if x is None:
+#                 x = 1
+#             else:
+#                 return 2
+#             return x
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_if_optional_cond.py

@@ -7,3 +7,13 @@ class C:
         self.field = 42
 def f(x: Optional[C]):
     return x.field if x is not None else None
+# def test_if_optional_cond(self):
+#     codestr = """
+#         from typing import Optional
+#         class C:
+#             def __init__(self):
+#                 self.field = 42
+#         def f(x: Optional[C]):
+#             return x.field if x is not None else None
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_incompat_override.py

@@ -5,3 +5,12 @@ class C:
     x: int
 class D(C):
     def x(self): pass
+# def test_incompat_override(self):
+#     codestr = """
+#     class C:
+#         x: int
+#     class D(C):
+#         def x(self): pass
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_incompat_override_method_arg_name.py

@@ -7,3 +7,18 @@ class A:
 class B(A):
     def m(self, y: str) -> int:
         return 0
+# def test_incompat_override_method_arg_name(self):
+#     codestr = """
+#         class A:
+#             def m(self, x: str) -> int:
+#                 return 42
+#         class B(A):
+#             def m(self, y: str) -> int:
+#                 return 0
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         "<module>.B.m overrides <module>.A.m inconsistently. "
+#         "Positional argument 2 named `x` is overridden as `y`",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_incompat_override_method_arg_type.py

@@ -7,3 +7,18 @@ class A:
 class B(A):
     def m(self, x: int) -> int:
         return 0
+# def test_incompat_override_method_arg_type(self):
+#     codestr = """
+#         class A:
+#             def m(self, x: str) -> int:
+#                 return 42
+#         class B(A):
+#             def m(self, x: int) -> int:
+#                 return 0
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         "<module>.B.m overrides <module>.A.m inconsistently. "
+#         "Parameter x of type `int` is not a subtype of the overridden parameter `str`",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_incompat_override_method_arg_type_okay.py

@@ -7,3 +7,13 @@ class A:
 class B(A):
     def m(self, x: object) -> int:
         return 0
+# def test_incompat_override_method_arg_type_okay(self):
+#     codestr = """
+#         class A:
+#             def m(self, x: str) -> int:
+#                 return 42
+#         class B(A):
+#             def m(self, x: object) -> int:
+#                 return 0
+#     """
+#     self.compile(codestr)

### 2021-10-28 conformance_suite/test_incompat_override_method_ret_type.py

@@ -7,3 +7,18 @@ class A:
 class B(A):
     def m(self) -> int:
         return 0
+# def test_incompat_override_method_ret_type(self):
+#     codestr = """
+#         class A:
+#             def m(self) -> str:
+#                 return "hello"
+#         class B(A):
+#             def m(self) -> int:
+#                 return 0
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         "<module>.B.m overrides <module>.A.m inconsistently. "
+#         "Returned type `int` is not a subtype of the overridden return `str`",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_inline_arg_type_mismatch.py

@@ -7,3 +7,16 @@ def f(x: int) -> bool:
     return x == 1
 def g(arg: str) -> bool:
     return f(arg)
+# def test_inline_arg_type_mismatch(self):
+#     codestr = """
+#         from __static__ import inline
+#         @inline
+#         def f(x: int) -> bool:
+#             return x == 1
+#         def g(arg: str) -> bool:
+#             return f(arg)
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError, r"str received for positional arg 'x', expected int"
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_none_compare.py

@@ -5,3 +5,15 @@ def f(x: int | None):
     if x > 1:
         x = 1
     return x
+# def test_none_compare(self):
+#     codestr = """
+#         def f(x: int | None):
+#             if x > 1:
+#                 x = 1
+#             return x
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         r"'>' not supported between 'Optional\[int\]' and 'Literal\[1\]'",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 conformance_suite/test_optional_assign_none.py

@@ -5,3 +5,11 @@ from typing import Optional
 class B: pass
 def f(x: Optional[B]):
     a: Optional[B] = None
+# def test_optional_assign_none(self):
+#     codestr = """
+#         from typing import Optional
+#         class B: pass
+#         def f(x: Optional[B]):
+#             a: Optional[B] = None
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_optional_assign_subclass.py

@@ -6,3 +6,12 @@ class B: pass
 class D(B): pass
 def f(x: D):
     a: Optional[B] = x
+# def test_optional_assign_subclass(self):
+#     codestr = """
+#         from typing import Optional
+#         class B: pass
+#         class D(B): pass
+#         def f(x: D):
+#             a: Optional[B] = x
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_optional_assign_subclass_opt.py

@@ -6,3 +6,12 @@ class B: pass
 class D(B): pass
 def f(x: Optional[D]):
     a: Optional[B] = x
+# def test_optional_assign_subclass_opt(self):
+#     codestr = """
+#         from typing import Optional
+#         class B: pass
+#         class D(B): pass
+#         def f(x: Optional[D]):
+#             a: Optional[B] = x
+#     """
+#     self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_redefine_type.py

@@ -6,3 +6,13 @@ class D: pass
 def f(a):
     x: C = C()
     x: D = D()
+# def test_redefine_type(self):
+#     codestr = """
+#         class C: pass
+#         class D: pass
+#         def f(a):
+#             x: C = C()
+#             x: D = D()
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_static_import_star.py

@@ -2,3 +2,9 @@
 # This should fail.
 
 from __static__ import *
+# def test_static_import_star(self) -> None:
+#     codestr = """
+#         from __static__ import *
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_static_import_unknown.py

@@ -2,3 +2,9 @@
 # This should fail.
 
 from __static__ import does_not_exist
+# def test_static_import_unknown(self) -> None:
+#     codestr = """
+#         from __static__ import does_not_exist
+#     """
+#     with self.assertRaises(TypedSyntaxError):
+#         self.compile(codestr, modname="foo")

### 2021-10-28 conformance_suite/test_type_type_final.py

@@ -3,3 +3,9 @@
 
 class A(type):
     pass
+# def test_type_type_final(self):
+#     codestr = """
+#     class A(type):
+#         pass
+#     """
+#     self.compile(codestr)

### 2021-10-28 conformance_suite/test_verify_too_many_args.py

@@ -4,3 +4,14 @@
 def x():
     return 42
 x(1)
+# def test_verify_too_many_args(self):
+#     codestr = """
+#         def x():
+#             return 42
+#         x(1)
+#     """
+#     with self.assertRaisesRegex(
+#         TypedSyntaxError,
+#         r"Mismatched number of args for function <module>.x. Expected 0, got 1",
+#     ):
+#         self.compile(codestr)

### 2021-10-28 ecoop-2022/paper.tex

@@
   \item
-    A formal description and analysis of the \SP{} core language.
-    The analysis includes a proof of type soundness,
-    a proof that optimization soundly improves code,
-    and [FILL perhaps] a metric to quantify the refactoring cost and optimization benefit
-    of a type migration.
+    A formal description of the \SP{} core language and an analysis of its soundness
+    and optimizations.
+    Soundness is not uniform across all types.
+    Some are fully sound while others are constructor sound.
   \item
@@
 
-No way to write the dynamic type.
-Instead, it fills in for absent annotations.
-Either totally missing or a missing generic parameter.
+Can't really write the dynamic type.
+But ANY unbound id. or non-type object gets interpreted as the dynamic type for now.
 
-%% TODO are there any types C[T0, T1] where you might want C[dyn, T1]? how to express?
-
-Cannot appear everywhere ... ChkDict can't hold a Dynamic,
+Cannot appear everywhere ... ChkDict can't hold a Dynamic (or rather, you can declare this type and cannot instantiate it),
 and primitives are incompatible with Dynamic.
@@
 
-\subsection{Benchmarks}
+\subsection{Microbenchmarks}
+
+richards etc.
 
 \subsection{Production Experience}
+
+\section{Design Principles}
+
+\subsection{Nominal, Checked Types}
+
+CheckedDict enables strong type checks and optimizations, but is incredibly painful to use.
+It is painful even though ChkDict supports the full dict API
+(any context that uses a Python dict can use a ChkDict without code changes).
+
+The problem is that the ChkDict type rejects all Python values.
+It accepts only ChkDict values, built through a special constructor
+that installs a tag for type tests and guards writes.
+Suppose that \code{f} is an untyped dict function.
+As is, it can process ChkDict but gets no benefit from type checks nor from optimizations.
+Adding a ChkDict domain type enables optimizations within \code{f} but raises a non-local
+problem: all callers of \code{f} must be sure to create a ChkDict.
+These ChkDicts must also be monomorphic to match whatever type \code{f} chose.
+It can take many edits to get a program running again after adding a ChkDict type.
+
+Lessons:
+\begin{itemize}
+  \item
+    Nominal types are not compatible with structual values, such as Python values.
+    Programmers are forced to edit old code to use such types ... unless the recent Nom work has a better idea~\cite{mt-oopsla-2021}
+  \item
+    Monomorphic nominal types are even worse.
+    Editing old code to fit their rigid constraints may not be feasible.
+
+\end{itemize}
 
@@
 Improve SP with generic types, structural types (lambda), \ldots.
+
+Use confined GT to relax ChkDict ... let programmers decide whether
+the type should reject or convert untyped data.
+Maybe a ChkDict function could compile to an "overloaded" version with
+fast and slow paths for ChkDict vs normal dict.
 

### 2021-11-01 ecoop-2022/paper.tex
ecoop: first draft of theorems and types

@@
 \newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\defeq}{=}
+\newcommand{\mfeq}{=}
+\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
+\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
+
+\newcommand{\typefont}[1]{\mathsf{#1}}
+\newcommand{\paramtype}[2]{#1[#2]}
+\newcommand{\sptype}{\typefont{T}}
+\newcommand{\sptclass}{\typefont{C}}
+\newcommand{\sptint}{\typefont{Int}}
+\newcommand{\sptfloat}{\typefont{Float}}
+\newcommand{\sptdyn}{\typefont{Dynamic}}
+\newcommand{\sptobject}{\typefont{Object}}
+\newcommand{\sptnone}{\typefont{NoneType}}
+\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
+\newcommand{\sptfun}[2]{#1 \rightarrow #2}
+\newcommand{\sptunion}[2]{#1 \cup #2}
+\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
+\newcommand{\sptpydict}[2]{\paramtype{\typefont{PyDict}}{#1, #2}}
+\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
+
+\newcommand{\spexpr}{e}
+\newcommand{\spvalue}{v}
+
+\newcommand{\sprred}{\rightarrow^*}
+
+\newcommand{\mfapply}[2]{#1\,(#2)}
+\newcommand{\mffont}[1]{\mathit{#1}}
+\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
+\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}
+
+\newcommand{\sperror}{\mathrm{Error}}
 
@@
 
+\begin{figure}[t]
+  %% ported types from redex model
+  \begin{langarray}
+    \sptype & \defeq &
+      \sptdyn \langmid
+      \sptnone \langmid
+      \sptclass \langmid
+      \sptinstanceof{\sptclass} \langmid
+      \sptfun{\sptype}{\sptype} \langmid
+      \sptunion{\sptclass}{\sptclass}
+  \\
+    \sptclass & \defeq &
+      \sptint \langmid
+      \sptfloat \langmid
+      \sptobject \langmid
+      \sptpydict{\sptclass}{\sptclass} \langmid
+      \sptchkdict{\sptclass}{\sptclass} \langmid
+      \mbox{(user defined class)}
+  \end{langarray}
+
+  \begin{center}\parbox{0.8\columnwidth}{
+    Abbreviation: $\sptoptional{\sptclass} \defeq \sptunion{\sptclass}{\sptnone}$
+
+    EXCEPT THAT $\sptoptional{\sptclass}$ may be a generic parameter just like any other class $\sptclass$
+
+    Function types cannot expect or return other function types.
+  }\end{center}
+
+  \caption{Surface Types}
+  \label{f:surface-types}
+\end{figure}
+
 \subsection{Evaluation Language, Optimizations}
+
+\begin{figure}[t]
+  \(
+    \mftypeF{\sptype_0}
+    \mfeq
+    \left\{\begin{array}{ll}
+      \sptpydict{}{} & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
+    \\
+      \sptype_0 & \mbox{otherwise}
+    \end{array}\right.
+  \)
+
+  \caption{Surface Types to Evaluation Types}
+  \label{f:surface-to-eval-types}
+\end{figure}
+
+\subsection{Properties}
+
+\begin{theorem}[Type Soundness]
+  If\ \(\vdash \spexpr_0 : \sptype_0\)
+  then one of the following holds:
+  \begin{itemize}
+    \item
+      \(\spexpr_0 \sprred \spvalue_0
+        \mbox{ and }
+        \vdash \spvalue_0 : \mftypeF{\sptype_0}
+      \)
+    \item
+      \(\spexpr_0\) diverges
+    \item
+      \(\spexpr_0 \sprred \sperror\)
+  \end{itemize}
+\end{theorem}
+
+\begin{theorem}[Optimization Soundness]
+  If\ \(\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
+  then\ \(\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
+\end{theorem}
+
+\begin{theorem}[Optimization Effectiveness]
+  FILL how do we know optimizations are useful rather than trivial?
+  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
+  %%   for general heuristic opts.
+  %% - is [opt(e) != e] true for interesting programs?
+\end{theorem}
 

### 2021-11-29 pj-7.1/paper.tex
pj: init

--- 2021-11-01 ecoop-2022/paper.tex
@@
-\documentclass[a4paper,english,cleveref,autoref,thm-restate,anonymous,]{lipics-v2021}
-\bibliographystyle{plainurl}
+\documentclass[english,submission]{programming}
 
@@
 %%   helps the static checker find new errors. ChkDict is one example.
+%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
 
@@
 %% -
-
-\title{Gradual Soundness: Lessons from Static Python}
-%% ... types make things fast?
-%% ... gradual migratory progressive soundness
-
-%\titlerunning{short title}
-
-\author{Kuang-Chen Lu}{Department of Computer Science, Brown University, USA}{LuKuangchen1024@gmail.com}{}{}
-\author{Ben Greenman}{Department of Computer Science, Brown University, USA}{benjamin.l.greenman@gmail.com}{0000-0001-7078-9287}{}
-\author{Carl Meyer}{Facebook, Inc.}{carljm@fb.com}{}{}
-\author{Dino Viehland}{Facebook, Inc.}{dinoviehland@fb.com}{}{}
-\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{shriram@brown.edu}{0000-0001-5184-1975}{}
-
-\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
-%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Kuang-Chen Lu, Ben Greenman, Carl Meyer, Dino Viehland, and Shriram Krishnamurthi}
-%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}}
-%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{gradual typing, migratory typing}
-
-\category{} %optional, e.g. invited paper
-
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{
-  Thanks to
-   Guido van Rossom for stimulating tweets.
-  This work was partly supported by the US National Science Foundation.
-  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
-  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
-  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
 
@@
 
+%\overfullrule=1mm
+
+%\usepackage{draftwatermark}
+%\SetWatermarkText{DRAFT}
+%\SetWatermarkScale{1}
+
+%% BEGIN tobias pape 2021-11-06
+\makeatletter
+\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
+\makeatother
+%% END
+
+\usepackage{cleveref}
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
+\usepackage[backend=biber]{biblatex}
+\addbibresource{bg.bib}
+
+\begin{document}
+
+\title{Gradual Soundness: Lessons from Static Python}
+%% ... types make things fast?
+%% ... gradual migratory progressive soundness
+
+%\titlerunning{short title}
+
+\author[a]{Kuang-Chen Lu}
+\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
+\affiliation[a]{Brown University, Providence, RI, USA}
+\author[a]{Ben Greenman}
+\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
+\author[b]{Carl Meyer}
+\authorinfo{(\email{carljm@fb.com}) FILL}
+\affiliation[b]{Facebook, Inc.}
+\author[b]{Dino Viehland}
+\authorinfo{(\email{dinoviehland@fb.com}) FILL}
+\author[a]{Shriram Krishnamurthi}
+\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}
+
+\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
+%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\keywords{gradual typing, migratory typing}
+
+\begin{CCSXML}
+\end{CCSXML}
+% \ccsdesc[100]{FILL}
+%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\paperdetails{
+  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
+  perspective=scienceempirical,
+  %% State one or more areas, separated by a comma. (see 2.2)
+  %% Please see list of areas in http://programming-journal.org/cfp/
+  %% The list is open-ended, so use other areas if yours is/are not listed.
+  area={Database programming, General-purpose programming, Program
+    verification, Programming education},
+  %% License options include: cc-by (default), cc-by-nc
+  % license=cc-by,
+}
+
+\maketitle
+
 \begin{abstract}
-  TBD
 %  Soundness, migratory typing.
 %  Novel run-time check strategy, combines concrete and transient for ergonomics.
+
+  \let\paragraph\abstractpart
+
+  \paragraph{Context}
+  % What is the broad context of the work? What is
+  % the importance of the general research area?
+  FILL
+
+  \paragraph{Inquiry}
+  % What problem or question does the paper
+  % address? How has this problem or question been
+  % addressed by others (if at all)?
+  FILL
+
+  \paragraph{Approach}
+  % What was done that unveiled new knowledge?
+  FILL
+
+  \paragraph{Knowledge}
+  % What new facts were uncovered? If the
+  % research was not results oriented, what new
+  % capabilities are enabled by the work?
+  FILL
+
+  \paragraph{Grounding}
+  % What argument, feasibility proof, artifacts,
+  % or results and evaluation support this work?
+  FILL
+
+  \paragraph{Importance}
+  % Why does this work matter?
+  FILL
+
 \end{abstract}
@@
 
+%    2021-11-15: Dino says the JIT will check/upgrade INVOKE_METHOD to INVOKE_FUNCTION
+%    2021-07-23: final classes get INVOKE_FUNCTION, not INVOKE_METHOD
+
   \item
@@
 
-\bibliography{bg}
+\acks{
+  Thanks to
+   Guido van Rossom for stimulating tweets.
+  This work was partly supported by the US National Science Foundation.
+  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
+  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
+  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+}
+
+{\sloppy
+\printbibliography
+}
 

### 2021-12-01 pj-7.1/paper.tex
pj: intro cleanup, sec 2 thoughts

@@
-\documentclass[english,submission]{programming}
+\documentclass[english,cleveref,submission]{programming}
 
@@
 \newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
-\newcommand{\sptpydict}[2]{\paramtype{\typefont{PyDict}}{#1, #2}}
+\newcommand{\sptrawpydict}{\typefont{PyDict}}
+\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
 \newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
@@
 
-\usepackage{cleveref}
+%\usepackage{cleveref}
 \usepackage{amsthm}
@@
 
-\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
-%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+%\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
 
@@
 Industry teams have developed innovative type systems that accommodate
-untyped designs, but these systems are all unsound [CITE].
+untyped designs, but are unsound [CITE].
 These systems intentionally check nothing at run-time when untyped values enter
@@
 compiler technology [CITE],
-and even weakened semantics [CITE], but none of these efforts
-decisively close the performance gap.
+and even weakened semantics [CITE], but these efforts have not yet
+decisively closed the performance gap.
 The most promising attempt is the \emph{concrete} semantics
@@
 overhead stems from severe restrictions on untyped code.
-These restrictions may not be acceptable to the maintainers of a
-working untyped codebase.
-Indeed, the maintainers might prefer to re-implement their system
-in a typed language, as Dropbox recently did with its core sync
-engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}
+%% In other words, concrete types impose low performance costs
+%% but high migration costs.
+%% ... you can have any color as long as it's black -h.ford
+%% ... concrete expresses few programs, but runs them all nicely
+Whether developers can accept these restrictions is unclear;
+indeed, they may prefer a whole-sale migration over an awkward
+gradual transition.\footnote{Successful migrations of untyped codebases
+to a typed language are rare, but not unheard of.
+Twitter ported its server-side code from Ruby to Scala
+and Dropbox moved its core sync engine from Python mypy to Rust: see
+\shorturl{http://www.}{artima.com/scalazine/articles/twitter\_on\_scala.html} and
+\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}.}
 
@@
 This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
-The \SP{} team at Instagram has developed a sound type system for a subset of
-Python along with a runtime system designed to
-capitalize on the benefits of sound types.
-The language design is a variation of concrete types that lets programmers
-begin with shallow claims about data structures and gradually opt in to deeper soundness.
-Remarkably, the performance of Python is an upper bound on the observed performance of
-Static Python.
+The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
+Python along with a runtime system that uses soundness to drive optimizations.
+The language design is a variation of concrete types that eases its restrictions.
+Instead of asking programmers to migrate from untyped code to sound concrete types
+in one leap, \SP{} offers an intermediate step via shallow types that enforce
+only the top-level shape of values.
+Programmers can begin with shallow claims and gradually dial up soundness.
+On a suite of benchmarks and production modules, Static Python out-performs untyped
+Python across the board.
 
@@
   \item
-    Evidence that gradual soundness delivers fast performance incrementally,
-    as programmers migrate to sound types.
-    \SP{} can be applied to a codebase with no refactoring and no performance overhead.
+    Evidence that gradual soundness delivers Python-level performance at a low migration cost
+    and faster performance as programmers gradually incorporate sound types.
+    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
     After refactoring to sound types, \SP{} performs well on a microbenchmark
@@
   \item
-    In-depth discussions of the expressiveness limitations in \SP{} relative to
+    Discussions of the expressiveness limitations in \SP{} relative to
     idealized gradual type systems, and the programming discipline that Instagram
@@
 First, we want to encourage system-builders to reproduce the
-\SP{} language design in another context and measure performance.
-Second, we want to lend focus to theoretical research on gradual typing.
-The restrictions that \SP{} adopts
-may be worthwhile assumptions for theory going forward.
-On the flip side, future research may find a way to lift some of
-the restrictions currently in \SP{}.
+\SP{} language design in another context and measure its performance.
+Second, we want to lend focus to researchers.
+Some of the restrictions that \SP{} adopts may be useful for
+other languages.
+Other restrictions might be lifted by future research.
 
@@
     The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
-    Concretely, this check accepts any Python dictionary.
+    \SP{} inserts a check that accepts any Python dictionary.
   \item
-    Because the domain check does not validate the contents of an incoming dictionary,
+    Because the domain check does not validate the elements of an incoming dictionary,
     the dict access (\code{x["A"]}) must check that its result matches the function codomain.
 \end{itemize}
-These checks ensure a \emph{shape-level} notion of type soundness~(\cite{s:model}).
+These checks ensure a \emph{shape-level} notion of type soundness~(\cref{s:model}).
 In short: if an expression has type \code{T} and reduces to a value, then the value
@@
 there is no need to check that it sends type-correct inputs.
-Thus, the compiled call uses a Cinder bytecode instruction to skip past the
-domain check and immediately enter the function body.
+Thus, the compiled call is optimized to use a Cinder bytecode instruction
+that skips the domain check and immediately enters the function body.
 
@@
 \subsection{Type System Highlights}
+
+[FILL]
 
@@
 
-\subsubsection{Checked Types}
+\subsubsection{Checked Types and Shallow Types}
 
@@
 
+\subsubsection{Mixed-Typed Class Hierarchies}
+
+[FILL]
+
+Typed classes can inherit from untyped classes.
+Untyped classes can inherit from typed classes.
+
+Interesting, because StrongScript disallowed.
+(But Nom allowed right?! Maybe allowed in a funny way, possibly to support their Dyn.)
+
+\subsubsection{Types for Python}
+
+[FILL]
+
+Which types accommodate Python code? Which from PEP typing etc.
+
+Surprisingly Unions are static only.
+Except for unions to false because those are so common.
+
+\subsubsection{Other}
+
+More notes and surprises:
+
+\begin{itemize}
+  \item
+    Optional(T) is the only union. Others turn into Dynamic.
+  \item
+    TBD
+\end{itemize}
+
 \subsubsection{Restrictions}
-
-OO restrictions:
 
@@
     \left\{\begin{array}{ll}
-      \sptpydict{}{} & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
+      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
     \\
@@
 \begin{theorem}[Type Soundness]
-  If\ \(\vdash \spexpr_0 : \sptype_0\)
+  If\ \(~\vdash \spexpr_0 : \sptype_0\)
   then one of the following holds:
@@
 \begin{theorem}[Optimization Soundness]
-  If\ \(\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
-  then\ \(\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
+  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
+  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
 \end{theorem}
 
-\begin{theorem}[Optimization Effectiveness]
-  FILL how do we know optimizations are useful rather than trivial?
-  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
-  %%   for general heuristic opts.
-  %% - is [opt(e) != e] true for interesting programs?
-\end{theorem}
+%\begin{theorem}[Optimization Effectiveness]
+%  FILL how do we know optimizations are useful rather than trivial?
+%  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
+%  %%   for general heuristic opts.
+%  %% - is [opt(e) != e] true for interesting programs?
+%\end{theorem}
 
@@
 The developers acknowledged M of these bugs as issues to fix.
+
+\begin{center}
+  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
+\end{center}
 
@@
 
-Nom~\cite{mt-oopsla-2017} is foundational.
+Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
 
-Full Monty(?)~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational.
+Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.
 
@@
 
-\acks{
-  Thanks to
-   Guido van Rossom for stimulating tweets.
-  This work was partly supported by the US National Science Foundation.
-  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
-  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
-  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
+%\acks{
+%  Thanks to
+%  Guido van Rossum for stimulating tweets.
+%  This work was partly supported by the US National Science Foundation.
+%  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
+%  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
+%  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+%}
 

### 2021-12-06 conformance_suite/method_from_def.py
more test

@@ -1,0 +1,11 @@
+# method_from_def.py
+# This should pass.
+# This should terminate.
+
+class C:
+    def m(self):
+        return 2
+
+
+obj = C()
+assert obj.m() is 2

### 2021-12-06 conformance_suite/method_from_lambda.py

@@ -1,0 +1,10 @@
+# method_from_lambda.py
+# This should pass.
+# This should terminate.
+
+class C:
+    m = lambda self: 2
+
+
+obj = C()
+assert obj.m() is 2

### 2021-12-06 conformance_suite/method_generative.py

@@ -1,0 +1,13 @@
+# method_generative.py
+# This should pass.
+# This should terminate.
+
+class C:
+    m1 = lambda self: 2
+    def m2(self):
+        return 3
+
+
+obj = C()
+assert obj.m1 is not obj.m1
+assert obj.m2 is not obj.m2

### 2021-12-06 pj-7.1/paper.tex

@@
     and optimizations.
-    Soundness is not uniform across all types.
-    Some are fully sound while others are constructor sound.
+    Soundness is gradual in the sense that some types are fully sound and others validate
+    only top-level shapes.
   \item
@@
 
-The designers of Static Python hope to eventually build a system that is sound for arbitrary Python code,
-but are not there yet.
-The known side channels include:
-eval, the locals and globals dicts (the latter can be accessed through frame objects at the moment),
-and mutable closure cells.
-Our model does not cover these either.
-We essentially ban all these features, rejecting any program in which they appear.
-Code mining in JS (and perhaps the newer R work?) suggests that at least eval
-uses can be rewritten without too much pain. The few exceptions would point
-to language weaknesses / call for new semantically-descriptive features.
+Our model of \SP{} includes the key aspects of its types,
+statements, and values.
+There are two main goals:
+illustrate the boundaries where typed and untyped code can mix,
+and to show how to soundly protect these boundaries.
+Compared to other gradual languages, these boundaries are quite restrictive---which
+partly explains \SP{} performance~(\cref{s:eval}).
+Indeed the model is nearly a subset of model for Nom~\cite{mt-oopsla-2021},
+hence we focus on explaining the gaps and omit formal proofs.
 
-FILL do we stop at strict modules?\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
-A strict module is 100\% repeatable and isolated from the side effects of other modules.
+The model intentionally does not cover all of Python.
+Some aspects of \SP{} are left out because they are straightforward to
+handle soundly (FILL examples ... while loop?);
+omitting these lets us focus on the difficult corners.
+Other Python features are left out because \SP{} does not ascribe types
+to them.
+These include \code{eval}, first-class classes, and module-level cells;
+see \cref{s:impl} for further discussion.
 
@@
 
+%% SP side channels:
+%% eval,
+%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
+%% and mutable closure cells.
+
 Purpose: explain significant gaps between the model and the implementation.
@@
 to add soundness?
+
+\subsection{Dynamic Python Features}
+
+\SP{} does not ascribe types to the following Python features.
+These are not covered in our model because the implementation simply
+gives them the dynamic type and lets the runtime treat them as untyped
+Python code.
+For each one, the dynamic type is a reasonable choice; static types are
+likely to cause more trouble than helpful structure.
+
+\paragraph{First-Class Classes}
+
+\SP{} does not attempt to type first-class classes:
+partly because they see little use at Instagram (FILL how much?)
+and partly because it is unclear how to incorporate them into
+the nominal type system.
+The straightforward but restrictive approach is to force code
+that uses a first-class class to expect instances of a particular
+static class.
+FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
+\cite{fkf-popl-1998} propose a different method that accommodates mixins,
+but the method calls a second layer of interface types.
+\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
+Adding these layers to \SP{} is an open question; in particular,
+prior work does not handle multiple inheritance.
+
+%First-class classes are untyped
+%(good discussion here:
+%- mixin based tends to structrural types
+%- nominality overly restrictive, get forced into <: hierarchy
+%- but don't have NO IDEA for wat to do
+%  + classes and mixins paper shows one idea,
+%    ... need complicated mixin study??? (many sub objects)
+%    built new type system to be mixin-aware, layered atop java
+%    unsure if such extension works for SP
+%    furthermor, class-and-mixins depends on Java single inheritance
+%  + bracha cook mixin pattern
+%  + tate muehlboeck structural object + interfaces
+
+\paragraph{Dynamic Execution}
+%% TODO is eval a side channel, or safe?
+%% TODO check that eval/exec args really are unoptimized / sound etc.
+Results computed by calls to \code{eval} and \code{exec}
+have the dynamic type.
+Their inputs run without being rewritten
+by the optimizer.
+
+Studies of JavaScript and R have shown that many uses of dynamic execution
+can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
+Assuming these findings carry over to Python, we recommend either similar
+adjustments or the introduction of semantically-descriptive replacements
+over attempting to type eval.
+
+\paragraph{Module-Level Variables}
+
+Any Python module can read and write to the module-level variables of another module.
+When compiling a module, \SP{} therefore assumes that its module-level variables
+may be modified by untyped code and assigns the dynamic type.
+
+Cinder offers \emph{strict modules} as an alternative to the Python
+semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
+If a programmer chooses to declare a module as strict, then its module-level varibles
+are immutable and thus typeable.
+
+Another potential solution is for the Cinder runtime to check that writes to module-level variables
+preserve their types.
+There are two downsides to this idea:
+it will add some performance overhead,
+and it will force developers to rewrite untyped code in order to fix any errors that arise.
+%% Because of the latter concern, type dynamic is a very reasonable default.
 

### 2021-12-07 conformance_suite/classes_work.py
adjust tests

@@ -1,8 +1,8 @@
-# slots_is_defined.py
+# classes_work.py
 # This should pass.
 # This should terminate.
 
 class C:
     pass
 
-assert C.__slots__ == ()
+o = C()

### 2021-12-07 conformance_suite/slots_contain_object_fields_but_not_class_fields.py

@@ -1,13 +1,0 @@
-# slots_contains_object_fields_but_not_class_fields.py
-# This should pass.
-# This should terminate.
-
-from typing import ClassVar
-
-
-class C:
-    x: ClassVar[int] = 2
-    y: str
-
-
-assert C.__slots__ == ("y",)

### 2021-12-07 conformance_suite/slots_do_not_accumulate.py

@@ -1,11 +1,0 @@
-# slots_do_not_accumulate.py
-# This should pass.
-# This should terminate.
-
-class C1:
-    x: str
-
-class C2(C1):
-    y: int
-
-assert C2.__slots__ == ('y',)

### 2021-12-07 conformance_suite/slots_do_not_contain_methods.py

@@ -1,9 +1,0 @@
-# slots_do_not_contain_methods.py
-# This should pass.
-# This should terminate.
-
-class C:
-    def mth1(self):
-        pass
-
-assert C.__slots__ == ()

### 2021-12-07 scripts/import_SP_tests.py

@@ -12,6 +12,11 @@ output_path_prefix = "./conformance_suite/"
 skipped_tests_path_prefix = "./skipped_tests/"
 # Ignore a test if it contains one of the following word
 ban_anywhere_in_test = [
+    # float are fairly broken
+    #   it is not a super class of int
+    #   many systems consider float a super type, but in fact it is not
+    #   because .is_integer is not in int.
+    'float',
     # Weird python scope
     'nonlocal',
     'global',
@@ -65,11 +70,6 @@ ban_in_test_name = [
     # This test uses string literal to write Optional type...
 ]
 skip_anywhere_in_test = [
-    # float are fairly broken
-    #   it is not a super class of int
-    #   many systems consider float a super type, but in fact it is not
-    #   because .is_integer is not in int.
-    'float',
     # To confirm
     'Protocol',
     'NamedTuple',

### 2021-12-07 pj-7.1/paper.tex

@@
 omitting these lets us focus on the difficult corners.
-Other Python features are left out because \SP{} does not ascribe types
-to them.
-These include \code{eval}, first-class classes, and module-level cells;
+Other Python features are left out because their \SP{} semantics is identical
+to Python.
+These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
+\SP{} does not rely on types for these features to guide program transformations;
 see \cref{s:impl} for further discussion.
@@
 
-%% SP side channels:
-%% eval,
+%% TODO discuss optimizations here, not in sec 2
+
+%% TODO SP side channels:
+%% ?? eval,
 %% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
@@
 FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
-\cite{fkf-popl-1998} propose a different method that accommodates mixins,
+Flatt et al.~\cite{fkf-popl-1998} propose a different method that accommodates mixins,
 but the method calls a second layer of interface types.
-\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
+Muehlboeck and Tate~\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
 Adding these layers to \SP{} is an open question; in particular,
@@
 %  + tate muehlboeck structural object + interfaces
+
+\paragraph{Multiple Inheritance}
+
+%% https://docs.python.org/3/reference/datamodel.html
+
+Python allows multiple inheritance.
+Classes can inherit from any number of other classes.
+To resolve method calls, Python traverses the class hierarchy
+to find the first matching receiver (MRO).
+
+[FILL]
+\SP{} lets most programs use multiple inheritance.
+The one constraint is that a class cannot have two
+values bound to the same member name, to ensure v-table
+integrity.
 

### 2021-12-10 pj-7.1/paper.tex
pj: cleanup

@@
 %% Thesis:
-%%  the addition of sound nominal types to an optionally-typed codebase
-%%  can effectively improve peformance
+%%  1. Static Python is sound, fast, and practical
+%%  2. SP has unique features and restrictions for speed
+%%  3. Gradual soundness is a feature: start with weak types and eventually migrate to checked types
 
@@
     Discussions of the expressiveness limitations in \SP{} relative to
-    idealized gradual type systems, and the programming discipline that Instagram
+    idealized gradual type systems, and of the programming discipline that Instagram
     follows in their gradually-sound codebase.
@@
 
-Interesting, because StrongScript disallowed.
+Interesting because StrongScript disallowed.
 (But Nom allowed right?! Maybe allowed in a funny way, possibly to support their Dyn.)
@@
 
-Surprisingly Unions are static only.
-Except for unions to false because those are so common.
+Unions are static only,
+except for unions to false because those are common and can be implemented efficiently.
 
@@
 %    the `INVOKE_METHOD`. -Carl
-
+%
 %    2021-11-15: Dino says the JIT will check/upgrade INVOKE_METHOD to INVOKE_FUNCTION
@@
 \begin{center}
+  %% most filed by KC, a few by Ben
   \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
@@
 
-\section{Under the Hood: Static Python Specifics}
+\section{Scaling to Python}
 \label{s:impl}
@@
 
-%First-class classes are untyped
-%(good discussion here:
-%- mixin based tends to structrural types
+%Notes on first-class classes being untyped:
+%- mixin-based code tends to structrural types
 %- nominality overly restrictive, get forced into <: hierarchy
@@
 %    unsure if such extension works for SP
-%    furthermor, class-and-mixins depends on Java single inheritance
+%    furthermore, class-and-mixins depends on Java single inheritance
 %  + bracha cook mixin pattern
@@
 
-CheckedDict enables strong type checks and optimizations, but is incredibly painful to use.
+CheckedDict enables strong type checks and optimizations, but is painful to use.
 It is painful even though ChkDict supports the full dict API

### 2021-12-11 pj-7.1/paper.tex
pj: multiple inheritance, sp has no vtable check

@@
 statements, and values.
-There are two main goals:
+The model has two main goals:
 illustrate the boundaries where typed and untyped code can mix,
-and to show how to soundly protect these boundaries.
+and show how to soundly protect these boundaries.
 Compared to other gradual languages, these boundaries are quite restrictive---which
 partly explains \SP{} performance~(\cref{s:eval}).
-Indeed the model is nearly a subset of model for Nom~\cite{mt-oopsla-2021},
+Indeed, the model is nearly a subset of the model for Nom~\cite{mt-oopsla-2021},
 hence we focus on explaining the gaps and omit formal proofs.
@@
 Some aspects of \SP{} are left out because they are straightforward to
-handle soundly (FILL examples ... while loop?);
-omitting these lets us focus on the difficult corners.
+handle soundly (FILL examples ... while loop?).
+Omitting these lets us focus on the difficult corners.
 Other Python features are left out because their \SP{} semantics is identical
@@
 
+[FILL]
+
 Purpose: explain significant gaps between the model and the implementation.
@@
 These are not covered in our model because the implementation simply
-gives them the dynamic type and lets the runtime treat them as untyped
+assigns the dynamic type and lets the runtime treat them as untyped
 Python code.
-For each one, the dynamic type is a reasonable choice; static types are
-likely to cause more trouble than helpful structure.
+For each, we claim that the dynamic type is a reasonable choice;
+accurate static types would be difficult to maintain.
 
@@
 Adding these layers to \SP{} is an open question; in particular,
-prior work does not handle multiple inheritance.
+Python's multiple inheritance would complicate the mixin story.
 
@@
 
-Python allows multiple inheritance.
-Classes can inherit from any number of other classes.
-To resolve method calls, Python traverses the class hierarchy
-to find the first matching receiver (MRO).
+Python allows classes to inherit from a list of parents.
+When resolving a method call to such a class, Python
+dynamically traverses the parent list in a fixed order
+seeking a first match.
 
-[FILL]
-\SP{} lets most programs use multiple inheritance.
-The one constraint is that a class cannot have two
-values bound to the same member name, to ensure v-table
-integrity.
+Due to the dynamic method resolution order (MRO),
+\SP{} does not track types for classes with multiple parents.
+Method calls to these classes execute the same way as
+in Python, with no type-directed optimizations to speed up dispatch.
 

### 2021-12-11 pj-7.1/paper.tex
pj: reword first-class / mixin discussion

@@
 partly because they see little use at Instagram (FILL how much?)
-and partly because it is unclear how to incorporate them into
-the nominal type system.
-The straightforward but restrictive approach is to force code
-that uses a first-class class to expect instances of a particular
-static class.
-FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
-Flatt et al.~\cite{fkf-popl-1998} propose a different method that accommodates mixins,
-but the method calls a second layer of interface types.
-Muehlboeck and Tate~\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
-Adding these layers to \SP{} is an open question; in particular,
-Python's multiple inheritance would complicate the mixin story.
+and partly because it is unclear how to incorporate them into the
+nominal type system.
+The straightforward but restrictive approach is to force code that uses a
+first-class class to expect subtypes of a particular named static class.
+Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
+but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
+MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
+structural objects.
+Adding an interface layer to \SP{} is an open question.
+
+% in particular , Python's multiple inheritance would complicate the mixin story.
+% Classic mixins rely on structural types~\cite{bc-oopsla-1990}.
 
@@
 richards etc.
+
+%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
+%%
+%% deltablue
+%% fannkuch
+%% nbody
+%% nqueens
+%% pystone
+%% richards
+%% richards_static
+%%
+%% UHOH except for richards_static, none of these have types!!!
+%% Need to ask if C&D have other typed benchmarks around
+%% ... if not, ask them to port these and measure and tune if needed.
 

### 2021-12-14 pj-7.1/paper.tex
pj: abstract

@@
   % the importance of the general research area?
-  FILL
+  Gradually-typed languages allow typed and untyped code to interoperate,
+  but typically come with some kind of drawback.
+  In some languages, the types are unreliable;
+  in others, communication across type boundaries can be extremely expensive;
+  and still others allow only limited forms of interoperability.
+  The research community is actively seeking a sound, fast, and expressive
+  approach to gradual typing.
+
+  %% TODO acknowlegde the success of new language / tracing JIT approaches
 
@@
   % addressed by others (if at all)?
-  FILL
+  This paper describes \SP{}, a language developed by engineers at Instagram
+  that has proven itself sound, fast, and reasonably expressive in production.
+  \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
+  the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
+  Concrete types provide soundness in an efficient but difficult-to-use way.
+  Transient types are sound in a shallow sense and easier to use; they help
+  to bridge the gap between untyped code and typed concrete code.
 
@@
   % What was done that unveiled new knowledge?
-  FILL
+  In collaboration with the \SP{} team, we have evaluated the language
+  as it exists today and developed a model to capture the essence of its
+  approach to gradual types.
+  We drew upon personal communications, bug reports, and the \SP{}
+  regression test suite.
 
@@
   % capabilities are enabled by the work?
-  FILL
+  Our main finding is that the \emph{gradual soundness}\/ that
+  arises from a mix of concrete and transient types is an effective
+  way to lower the maintenance cost of the concrete approach.
+  On a more technical level, this paper contributes two artifacts:
+  a model of \SP{} that passes all relevant tests from the \SP{} codebase
+  and a performance evaluation of \SP{}.
 
@@
   % or results and evaluation support this work?
-  FILL
+  Our model of \SP{} is implemented in PLT Redex and tested against
+  the \SP{} regression suite.
+  This paper includes a small core of the model to convey the main ideas
+  of the \SP{} approach and its soundness.
+  Our performance claims are supported by data from both
+  a public suite a microbenchmarks and the proprietary Instagram codebase.
 
@@
   % Why does this work matter?
-  FILL
+  \SP{} is the first language that improves performance across the board through
+  the addition of sound types to an untyped language.
+  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
+  Other language designers may wish to replicate its approach,
+  especially those who maintain unsound type checkers for untyped languages,
+  e.g., JavaScript.
 
@@
 %% https://docs.python.org/3/reference/datamodel.html
+%% TODO check Dino email, confirm with C&D that layout conflict is unrelated to vtables
 

### 2021-12-15 pj-7.1/paper.tex
pj: sec 2.0

@@
   Other language designers may wish to replicate its approach,
-  especially those who maintain unsound type checkers for untyped languages,
-  e.g., JavaScript.
+  especially those who maintain unsound type checkers for untyped languages
+  such as JavaScript.
 
@@
 First, we want to encourage system-builders to reproduce the
-\SP{} language design in another context and measure its performance.
+\SP{} language design.
+In particular, the maintainers of optionally-typed languages
+may find that adding transient and concrete types is a low-overhead
+way to turn types into reliable claims.
 Second, we want to lend focus to researchers.
-Some of the restrictions that \SP{} adopts may be useful for
-other languages.
+Some of the restrictions that \SP{} adopts may be useful to
+ground theoretical work.
 Other restrictions might be lifted by future research.
@@
 
-[FILL]
+The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
+and prior work on types for Python.
+Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
+Relative to these sources, the \SP{} takes some interesting turns
+because of the engineering context at Instagram.
+The following contextual points are important to bear in mind as we present
+notable aspects of the type system:
+\begin{itemize}
+  \item
+    Performance is the bottom line.
+    Unless \SP{} helps production code run faster, the team will be disbanded
+    and reallocated.
+    %% 2021-12-14: too extreme? and what about soundness?
+  \item
+    Instagram is / had been using Cython for performance.
+    Cython compiles Python source code to C extension modules for
+    a faster Python [CITE].
+    Instagram had converted several performance-critical modules to Cython prior
+    to starting work on \SP{}.
+    \SP{} replaced them.
+    (How does this help understanding type system highlights?)
+  \item
+    The Instagram codebase is already typed.
+    Developers use Pyre [CITE], an optional type checker for Python,
+    to write type annotations and to drive IDE tools (e.g., autocompletion).
+  \item
+    Q. why is the python so nominal? style guide? java background?
+\end{itemize}
 
 \subsubsection{Type Dynamic}
+
+%% 2021-12-14:
+%% - anything to say about Any being a "meta" type in Python typing module?
+
+\SP{} includes a dynamic type to accommodate untyped Python code.
+Any un-annotated value or expression gets the dynamic type by default.
+\SP{} does not make any predicitions / optimizations about dynamic code;
+it assumes nothing and lets the Python runtime handle.
+
+- lack of ann
+- or, incorrect ann
+- at toplevel, everything compatible
+- diffirent rules inside data structures
 
@@
 
+    Optional(T) is the only union. Others turn into Dynamic.
+
 \subsubsection{Other}
 
-More notes and surprises:
-
 \begin{itemize}
-  \item
-    Optional(T) is the only union. Others turn into Dynamic.
-  \item
-    TBD
+  \item \textbf{Recursive Types}.
+    Sadly this is normal for Python type checkers.
 \end{itemize}
 
-\subsubsection{Restrictions}
+\subsubsection{First-Order Functions and Classes}
 
-\begin{itemize}
+Incidentally, \SP{} is a first-order type system.
+
+... no user-facing type for functions
+ imports / exports follow transient
+
+... ditto for classes (see below)
+
   \item \textbf{First-order Typed Classes}.
@@
 
+  \item \textbf{Callable}.
+    The type system cannot yet express first-class functions.
+    The team plans to lift this restriction~(\cref{s:future}),
+    but there is little internal pressure to do so.
+
+\subsubsection{Restrictions}
+
+\begin{itemize}
   \item \textbf{No Overloading}.
@@
 
-  \item \textbf{Callable}.
-    The type system cannot yet express first-class functions.
-    The team plans to lift this restriction~(\cref{s:future}),
-    but there is little internal pressure to do so.
-
-  \item \textbf{Dynamic Checked Types}.
-    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
-    keys or values.
-    [FILL] no desire, and makes runtime checks much simpler.
-    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now
-
-  \item \textbf{Recursive Types}.
-    Sadly this is normal for Python type checkers.
+%  \item \textbf{Dynamic Checked Types}.
+%    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
+%    keys or values.
+%    [FILL] no desire, and makes runtime checks much simpler.
+%    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now
 

### 2021-12-16 pj-7.1/paper.tex
pj: type dynamic

@@
 %% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
+%% - what are the idiomatic base SP types ... int or Int ?
 
@@
 Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
-Relative to these sources, the \SP{} takes some interesting turns
+Relative to these, the \SP{} takes some interesting turns
 because of the engineering context at Instagram.
@@
   \item
-    Instagram is / had been using Cython for performance.
+    %% Q. why is the python so nominal? style guide? java background?
+    The codebase makes heavy use of first-order classes and objects.
+    First-class classes, objects, and even functions are rare.
+  \item
+    Instagram is / had been using Cython to improve the performance of select modules.
     Cython compiles Python source code to C extension modules for
@@
     to write type annotations and to drive IDE tools (e.g., autocompletion).
-  \item
-    Q. why is the python so nominal? style guide? java background?
 \end{itemize}
@@
 \subsubsection{Type Dynamic}
+\label{s:type-dynamic}
 
@@
 %% - anything to say about Any being a "meta" type in Python typing module?
-
-\SP{} includes a dynamic type to accommodate untyped Python code.
-Any un-annotated value or expression gets the dynamic type by default.
-\SP{} does not make any predicitions / optimizations about dynamic code;
-it assumes nothing and lets the Python runtime handle.
-
-- lack of ann
-- or, incorrect ann
-- at toplevel, everything compatible
-- diffirent rules inside data structures
-
-[FILL]
-
-The dynamic type is an important part of Static Python.
-Dynamic avoids the need for whole-sale migration,
-lets programmers work around dynamic features like splatting,
-and is in some cases necessary for optimal performance (evidence?).
-
-Can't really write the dynamic type.
-But ANY unbound id. or non-type object gets interpreted as the dynamic type for now.
-
-Cannot appear everywhere ... ChkDict can't hold a Dynamic (or rather, you can declare this type and cannot instantiate it),
-and primitives are incompatible with Dynamic.
-
-Dynamic behavior is always (?) the same as normal Python.
-
-\subsubsection{Checked Types and Shallow Types}
+%% - 3 classes of code ... how does dynamic differ from untyped?
+
+Like most gradual languages, \SP{} includes a dynamic type
+that allows untyped expressions within a statically-typed context.
+Whenever an expression or variable lacks a type annotation,
+\SP{} uses the dynamic type as a default and skips most compile-time checks.
+Every untyped Python program is thus a well-formed \SP{} program.
+Furthermore, programmers can omit the annotations for specific
+function arguments without losing the benefits of other type checks.
+Freedom to omit annotations helps programmers convert modules to \SP{}
+with little initial refactoring.
+
+In addition to serving as a default, the dynamic type also replaces any undefined types.
+For example, suppose that the type \code{X} is undefined in \SP{} and that a
+program contains the annotation \code{ChkDict[int, X]}, which describes a
+dictionary with integer keys and \code{X}-typed values.
+\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
+of values inside the dictionary at runtime.
+To an outsider, this behavior might seem odd: why not reject undefined types with an error?
+The reason is to reduce friction with the Pyre annotations that are already
+widely-used at Instagram.
+Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
+efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions.
+By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.
+
+The dynamic type is not, however, the flexible dynamic type provided
+by \emph{true} gradual languages~\cite{svcb-snapl-2015}.
+Replacing part of a type with dynamic can lead to both compile-time errors
+and run-time errors.
+In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee.\footnote{We
+assume a standard type precision relation. Of course, one could argue that these programs
+do not break the gradual guarantees if type precision does not allow our uses of the dynamic type.}
+\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
+\begin{itemize}
+  \item
+    \cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
+    The subclass definition raises a compile-time error because it attempts to override
+    the typed \code{m1} method with another method that returns the dynamic type.
+  \item
+    \cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
+    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
+    It sends a dictionary with integer values to a function that expects
+    dictionaries with dynamic values.
+    At runtime, the function rejects this argument because its type is not an
+    exact match.
+\end{itemize}
+In sum, these behaviors show that performance is paramount in \SP{}.
+Static errors in dynamic-typed code are designed to protect the integrity of
+type-based optimizations.
+%% other examples?
+The dynamic errors are a consequence of strict run-time constraints, which provide stronger guarantees
+and are simpler to validate than lax constraints.
+%% hmph, unclear. Spell it out!
+
+\begin{figure}
+  \begin{subfigure}[t]{0.5\columnwidth}
+  \begin{verbatim}
+    class A:
+      def m1(self)->int:
+        return 0
+
+    class B(A):
+      def m1(self):
+        return 0
+    # Error: type dynamic does not match int
+  \end{verbatim}
+    \caption{Removing a type in \code{B} breaks the static gradual guarantee}
+    \label{f:gg-failure-stat}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[t]{0.5\columnwidth}
+  \begin{verbatim}
+    from __static__ import ChkDict
+
+    def f(x: ChkDict[str, dyn]):
+      return x["A"]
+
+    d = ChkDict[str, int]({"A": 1})
+    f(d)
+    # Error: f expected ChkDict[str, dyn]
+  \end{verbatim}
+    \caption{Removing part of the type for \code{x} breaks the dynamic gradual guarantee}
+    \label{f:gg-failure-dyn}
+  \end{subfigure}
+  \caption{\SP{} values performance over the gradual guarantees.}
+  \label{fig:gg-failure}
+\end{figure}
+
+\subsubsection{Concrete Types and Shallow Types}
+\label{s:checked-type}
 
@@
 
+%% TODO very important, one of the main things IG does use!!!
+
 [FILL]
@@
 
-  \item \textbf{First-order Typed Classes}.
-    In order to benefit from types, class definitions must be defined at the top level of a module.
-    Nested classes are supported by the language, but have type \code{Dynamic} according
-    to the type checker.
-
-  \item \textbf{Callable}.
-    The type system cannot yet express first-class functions.
-    The team plans to lift this restriction~(\cref{s:future}),
-    but there is little internal pressure to do so.
+%  \item \textbf{First-order Typed Classes}.
+%    In order to benefit from types, class definitions must be defined at the top level of a module.
+%    Nested classes are supported by the language, but have type \code{Dynamic} according
+%    to the type checker.
+%
+%  \item \textbf{Callable}.
+%    The type system cannot yet express first-class functions.
+%    The team plans to lift this restriction~(\cref{s:future}),
+%    but there is little internal pressure to do so.
 
@@
 
+%% TODO built-in vs. user defined ... 
+%% Carl: I don't think this is quite true the way it's phrased here? in general
+%%   for SP classes we wrap non-static subclass override methods at runtime (in our
+%%   vtable implementation) to enforce correct return type, so that we don't need
+%%   exact types, we can trust that Liskov is not broken.
+%%     The cases where we tend to be limited to exact types is in optimizing
+%%   operations on builtin types, where they aren't really static types and
+%%   methods called on them aren't going through SP vtables, so we don't have
+%%   the vtable wrappers to enforce LSP, we just know the behavior of the exact
+%%   builtin type
+
+%% TODO wherever we discuss optimizations, also talk about de-slowdowns
+%% Carl: not sure where this belongs exactly, but we go to a lot of work in the
+%%   compiler to intelligently minimize runtime checks by only inserting them when
+%%   the source value is of dynamic type (this includes also skipping CHECK_ARGS if
+%%   the function was statically invoked by another static function and thus we
+%%   know the compiler checked argument types)
+
 [FILL]
@@
 
-\section{Design Principles}
+%% 2021-12-15: C&D may have an SP off/on switch soon
+%% - accumulated CPU savings since July 1 = ~2% total IG production CPU
+%%   ... how does that number work?
+%% - in the beginning (?) H1 got 1% from first conversions + 0.7% from all Cython (higher bar)
+%%   - no Cython regressions, SP + JIT is good enough
+%%   - some Cython still exists, imported as library code from another codebase
+%% - rough measure: 0 to 25% per module
+%%   ... but depends on nature of module
+%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls
+
+%% - 530 SP modules
+%%   - 400 from a codegen tool
+%%   - 130 by hand
+%%   - currently +~3/4 per week by hand
+
+%% - pushback one time about removing MI
+%% - otherwise, people happy with low-cost improvements
+
+%% - review of 30 winning diffs "provided the bulk of our wins this half"
+%%   most common changes, not in careful order, are:
+%%   1) adjustments to tests to pass correct types instead of mock objects, as as
+%%      not to fail type checks in SP code. This is an inherent and desired
+%%      incompatibility; next half we plan to work on improving the Python mock
+%%      framework to automatically create mocks of the correct types.
+%%      - bg: lookup the dino email about mocks
+%%   2) adjustments to mock assertions in tests about calls to adjust for the
+%%      fact that SP currently converts all call arguments to be positional (we plan
+%%      to fix this incompatibility by falling back to the original call arguments
+%%      when we detect the target function has been patched.)
+%%      - bg: ??? bit fuzzy about positional and others
+%%   3) adjustments around class attributes vs instance attributes (Python allows
+%%      this to be a bit fuzzy, where a class attribute can be a fallback default
+%%      for an instance attribute if not set, but SP currently requires an attribute
+%%      to be clearly one or the other, and we have more efficient access for
+%%      instance attributes via LOAD_FIELD that we don’t have for class attributes.)
+%%      - related to github #37 ?
+%%   4) convert @classmethod that don’t actually need the class to @staticmethod,
+%%      allowing for emitting direct INVOKE_FUNCTION instead of INVOKE_METHOD
+%%   5) extract special types that our compiler doesn’t yet support/understand
+%%      (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
+%%      the module we are converting to static. We have work in progress on
+%%      supporting some Enum types, and plan to add a @dataclass intrinsic soon.
+%%   6) Removing the use of super-dynamic Python features like `__setattr__` that
+%%      we don’t support. So far in the cases where we’ve done this it hasn’t
+%%      required a major rewrite.
+%%      - bg: more details pls ,,, sketch the rewrite
+%%   7) Removing the use of keyword-only arguments (bare `*`) in function
+%%      signatures, since we don’t yet support calling such functions. We plan to
+%%      support this next half.
+%%   8) Convert Python integers and bools used in very hot code paths into SP
+%%      machine primitives.
+%%      - bg: focus on block, convert, no problem?
+%%   9) Occasionally converting a regular Python list or dict to a CheckedList or
+%%      CheckedDict.
+%%      - bg: how occasional? worried about power of the "gradual soundness" claim
+%%
+%% common theme: gotta change more tests than code
+%%
+%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
+%%  removes an extra call layer
+%%  helps with (timing [ context managers / decorators ])
+%%
+%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
+%%  because the pay / payoff isn't yet attractive
+
+\section{Related Work}
+\label{s:related}
+
+Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
+
+Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.
+
+Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.
+
+PyPy is another runtime for Python
+Pycket is built on PyPy and improves both Typed Racket~\cite{bbst-oopsla-2017}
+and Reticulated Python~\cite{vsc-dls-2019}.
+
+Type systems for Python ...
+mypy pyre pytype;
+Reticulated;
+
+Optimizing type systems for Python ...
+Reticulated;
+mypyc;
+Developer experience is one of the main advantages of \SP{} over mypyc,
+which implements type-directed optimizations for CPython
+by compiling source code to C extension modules.
+
+\section{Future Work}
+\label{s:future}
+
+%% Immediate future: fix adoption friction points
+%% 2021-12-15: WHAT ARE THESE?
+
+%% engineering: JIT profiling, instead of hand-requested JIT list
+%% model: metaclasses
+
+Adapt SP ideas to a new setting, test performance takeaways.
+
+Improve SP with generic types, structural types (lambda), \ldots.
+
+Use confined GT to relax ChkDict ... let programmers decide whether
+the type should reject or convert untyped data.
+Maybe a ChkDict function could compile to an "overloaded" version with
+fast and slow paths for ChkDict vs normal dict.
+
+Build an automatic migration tool for Checked data.
+
+Build a static analysis that hoists transient annotations to an early, shared point.
+Take care to give quality error messages.
+
+Optimize Python integer operations (in addition to the ctypes).
+Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
+or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
+
+\section{Conclusion / Lessons / Design Principles}
+\label{s:conclusion}
 
@@
 
-\section{Related Work}
-\label{s:related}
-
-Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
-
-Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.
-
-Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.
-
-PyPy is another runtime for Python
-Pycket is built on PyPy and improves both Typed Racket~\cite{bbst-oopsla-2017}
-and Reticulated Python~\cite{vsc-dls-2019}.
-
-Type systems for Python ...
-mypy pyre pytype;
-Reticulated;
-
-Optimizing type systems for Python ...
-Reticulated;
-mypyc;
-Developer experience is one of the main advantages of \SP{} over mypyc,
-which implements type-directed optimizations for CPython
-by compiling source code to C extension modules.
-
-\section{Future Work}
-\label{s:future}
-
-Adapt SP ideas to a new setting, test performance takeaways.
-
-Improve SP with generic types, structural types (lambda), \ldots.
-
-Use confined GT to relax ChkDict ... let programmers decide whether
-the type should reject or convert untyped data.
-Maybe a ChkDict function could compile to an "overloaded" version with
-fast and slow paths for ChkDict vs normal dict.
-
-Build an automatic migration tool for Checked data.
-
-Build a static analysis that hoists transient annotations to an early, shared point.
-Take care to give quality error messages.
-
-Optimize Python integer operations (in addition to the ctypes).
-Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
-or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
-
-\section{Conclusion}
-\label{s:conclusion}
-
 Sound types in Static Python give programmers a way to improve performance.

### 2021-12-17 pj-7.1/paper.tex
pj: 2.2 outlining

@@
   \item
-    \cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
+    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
     The subclass definition raises a compile-time error because it attempts to override
@@
   \item
-    \cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
+    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
     uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
@@
 \end{itemize}
-In sum, these behaviors show that performance is paramount in \SP{}.
-Static errors in dynamic-typed code are designed to protect the integrity of
-type-based optimizations.
-%% other examples?
-The dynamic errors are a consequence of strict run-time constraints, which provide stronger guarantees
-and are simpler to validate than lax constraints.
-%% hmph, unclear. Spell it out!
+These behaviors are a consequence of \SP{} being driven by the needs of an
+existing codebase.
+The literature on gradual types has served only as a guide, not a mandate.
 
@@
 
-%% more to say about the style of migration in SP?
-[FILL]
-
-Enabling sound types on a function has a ripple effect on callers.
-All untyped callers are now forced to supply type-correct inputs; may call for
-refactoring.
-
-Enabling checked types has a huge ripple effect.
-The worst case is changing an annotation to a expect a checked data structure:
+Every \SP{} type corresponds to a simple run-time check.
+When an unknown value reaches a boundary to typed code at run-time,
+\SP{} uses these checks to quickly decide whether or not the value
+matches the type.\footnote{In other words, \SP{} provides \emph{immediate accountability}~\cite{mt-oopsla-2017}.}
+Unlike structural gradual type systems, \SP{} does not traverse
+incoming values nor does it create wrappers to enforce behavior.
+
+These checks try to faithfully enforce the full type, but some are incomplete.
+We call the incomplete ones \emph{shallow} types; examples follow.
+
+First off, for a basic nominally-typed language (think Java 1.4, before
+generics), simple and full checks are easy to implement.
+Every type has a name and every value can carry a tag that corresponds to the
+name of its type.
+
+Parameterized types (generics) are more difficult to fully enforce without
+changing the language.
+Consider a list datatype \code{List[T]} that is parameterized by the
+e type of its elements.
+If a function expects lists of integers (\code{List[int]}), then a simple check
+for lists is not strong enough to decide whether an incoming value matches the
+type~(\cref{f:list-boundary}).
+A deeper check is needed to validate the list elements.
+But, the default \code{List} type cannot support such a check because untyped
+code is free to create lists.
+
+The \SP{} solution is to provide two versions of every parameterized type,
+both of which support efficient run-time checks.
+One version is shallow.
+The other is \emph{concrete}: complete, but new.
+Consider \code{List[T]} as an example:
 \begin{enumerate}
-  \item all callers must create a checked value by invoking the right constructor, and
-  \item all clients of those callers can no longer expect an unchecked type.
+  \item
+    \emph{Shallow type \code{List}}.
+    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
+    the elements.
+    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
+    values at run-time.
+  \item
+    \emph{Concrete type \code{ChkList[T]}}.
+    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
+    for concrete types.
+    The data structure implements the list API and comes with a checked constructor that
+    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
+    \cref{s:chkdict-impl} discusses the implementation further.
 \end{enumerate}
-Programmers have to trace the annotation back to value-creation points, and then
-ensure that all uses of those values are compatible.
+
+\begin{figure}[t]
+  How to enforce type \code{List[int]} simply?
+  \begin{itemize}
+    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
+    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
+    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
+    \item Show venn diagram of the options?
+  \end{itemize}
+  \caption{FILL illustrate list boundary and checks}
+  \label{f:list-boundary}
+\end{figure}
+
+These options let programmers decide whether to change their code to benefit from
+fully-sound types, or leave it as-is for a lower payoff.
+Often, the shallow types are more attractive because concrete types impose
+a ripple effect throughout the codebase.
+If one type changes to concrete, then all data constructors that reach this type need
+to change \emph{and then} all their clients need to use the concrete type as well.
+\Cref{s:concrete-migration} discusses this migration cost of concrete types further.
 
@@
 
-[FILL]
-
-Typed classes can inherit from untyped classes.
-Untyped classes can inherit from typed classes.
-
-Interesting because StrongScript disallowed.
-(But Nom allowed right?! Maybe allowed in a funny way, possibly to support their Dyn.)
+The Instagram codebase uses lots of classes and objects.
+Most of these are simple, Java-style: classes declared on
+the top level and single inheritance.
+It was therefore essential to get inheritance working across type boundaries.
+Typed classes must be able to inherit from untyped classes,
+and vice-versa untyped classes must be able to inherit from typed ones.
+
+FILL literature? StrongScript Nom classic GT?
+
+FILL either way, spell out what happens for these interactions.
+Where are the checks needed, checks skipped.
 
@@
 
-\subsubsection{Other}
+\subsubsection{Non-Highlights / Restrictions}
+
+The \SP{} type system excludes include some features.
+Each of the following is low priority for the team; that said, they are
+ordered below rougly from most important (soonest to be implemented) to least.
 
 \begin{itemize}
-  \item \textbf{Recursive Types}.
-    Sadly this is normal for Python type checkers.
-\end{itemize}
-
-\subsubsection{First-Order Functions and Classes}
-
-Incidentally, \SP{} is a first-order type system.
-
-... no user-facing type for functions
- imports / exports follow transient
-
-... ditto for classes (see below)
-
-%  \item \textbf{First-order Typed Classes}.
-%    In order to benefit from types, class definitions must be defined at the top level of a module.
-%    Nested classes are supported by the language, but have type \code{Dynamic} according
-%    to the type checker.
-%
-%  \item \textbf{Callable}.
-%    The type system cannot yet express first-class functions.
-%    The team plans to lift this restriction~(\cref{s:future}),
-%    but there is little internal pressure to do so.
-
-\subsubsection{Restrictions}
-
-\begin{itemize}
-  \item \textbf{No Overloading}.
-    Methods cannot be overloaded. Each combination of a class
-    and a name can refer to at most one method.
-    This ensures that method resolution depends only on the
-    type of the receiver object.
-    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
-    % Q. does resolution depend on args?
-
-  %\item \textbf{Single Inheritance}.
-  %  Multiple inheritance is not supported.
-  %  But, it might someday be supported.
-
-  \item \textbf{Compatible method overrides}.
-    Methods may be overridden, either by \SP{} classes or
-    untyped Python classes.
-    Overrides that appear in \emph{the same} \SP{} file as the class definition, however,
-    must supply a method whose type is a static subtype of the previous method.
-    In particular, a parent method that returns type \code{int} cannot be
-    overridden by a child method that returns type \code{Dynamic}.
-    %% TODO what does this enable ... optimizations in the same file?
-
+  \item \emph{Concrete Parameterized Classes}.
+    Except for a handful of built-in types,
+    parameterized classes are enforced only with shallow checks.
+    In the near future, these classes will gain concrete versions
+    for programmers to opt-in to.
+
+  \item \emph{Callable Types, First-Class Classes}.
+    The Instagram codebase rarely uses first-class functions, classes, or objects.
+    Thus there is little need for either a \code{Callable} type to describe functions
+    or first-class types for objects.
+    When such values do appear, \SP{} assigns the dynamic type and spot-checks their
+    uses.
+    For example, every application of a first-class function gets wrapped in a check
+    if the context expects a certain type.
+
+  \item \emph{Method Overloading}.
+    Methods cannot be overloaded.
+    Each combination of a class and a name can refer to at most one method.
+    This behavior comes from Python.
+
+  \item \emph{Multiple Inheritance}.
+    Multiple inheritance is not supported in a typed way.
+    \Cref{s:impl} discusses this point further.
+
+  \item \emph{Strong overrides}.
+    Methods and fields may be overridden, but only with values of the same type.
+    % In fact, \SP{} compiles fields using Python slots
+    % declarations (\code{\_\_slots\_\_}).
+    %
     % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
@@
 
-  \item \textbf{Compatible field overrides}.
-    Similarly, object fields and class variables may be overridden
-    only with subtype-compatible values.
-    In fact, \SP{} compiles fields using Python slots
-    declarations (\code{\_\_slots\_\_}).
-    This declaration prevent new fields from arising at runtime,
-    but saves space and speeds up field accesses.
-
+  \item \emph{Recursive Types}.
+    \SP{} does not support recursive types.
+    Neither does Pyre or mypy.
+    %% demand seems low ... baffling
+    %% mypy = https://github.com/python/mypy/issues/731
+    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
 \end{itemize}
 
-Type restrictions:
-
-\begin{itemize}
-  \item \textbf{Parameterized Classes}.
-    At present, \SP{} treats all class parameters as type \code{Dynamic}.
-    The team plans to lift this restriction (\cref{s:future}).
-
-%  \item \textbf{Dynamic Checked Types}.
-%    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
-%    keys or values.
-%    [FILL] no desire, and makes runtime checks much simpler.
-%    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now
-
-\end{itemize}
-
 \subsection{Runtime System Highlights}
-
-\begin{itemize}
-  \item
-    \textbf{VTables}
+%% Non-Highlights:
+%% - shadow frames = JIT feature, lite array to track Python call stack
+%% - strict modules ... avoids toplevel function wrapper
+
+\subsubsection{Bytecode Optimization}
+
+optimizations and bytecodes for: length, subscripting, function dispatch, method dispatch, field load/store
+
+table much later in paper
+
+IN ADDITION to optimizations ...
+not sure where this belongs exactly, but we go to a lot of work in the compiler
+to intelligently minimize runtime checks by only inserting them when the source
+value is of dynamic type (this includes also skipping CHECK\_ARGS if the
+function was statically invoked by another static function and thus we know the
+compiler checked argument types)
+
+\subsubsection{Virtual Method Tables}
+
 Cinder adds virtual method tables to typed classes.
@@
 the static compiler will only emit against a statically known static type.'' -Carl
+
+with vtables, method resolution happens differently than Python: args. get
+resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
+
+vtables are built on Python's vectorcall signature; though new, they're not
+totally from scratch
 
@@
 
-  \item
-    \textbf{Bytecode Optimization}
-    (could use a full section)
-
-  %\item
-  %  \textbf{Shadow Frames}
-  %  (CinderDoc/static_python.rst)
-  %  magic import that signals to the Static
-  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
-  %  improves performance of function calls by avoiding the creation of full
-  %  Python frame objects until they are definitely needed (e.g. if an
-  %  exception is raised.) In the future this should become default.
-
-  \item
-    \textbf{Checked Dict}
-    Implementation too, to support the types.
-
+\subsubsection{Checked Data Structures}
+\label{s:chkdict-impl}
+
+As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
+come with both a type and an implementation.
+The implementation provides the same interface as the built-in;
+the main difference is that it adds type checks.
+
+For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
+with keys of type \code{K} and values of type \code{V}.
+The implementation has three main components:
+\begin{itemize}
+  \item
+    Constructor, basically from dict to CD, checks integrity.
+    All CD valuess have to go through this, whether types or untyped.
+  \item
+    Instances keep a tag, used as exact type in boundary check.
+    (Tag may not be CD[i i] but rather a sentinel generated from it.)
+  \item
+    Writes are protected in every context.
 \end{itemize}
 
-\subsection{Additional Features}
-
-% things we did not model ... is that worth making a distinction here?
-
-strict modules
-
-decorator to force inlining, for 1-line functions
-
-\subsubsection{Primitive Types}
+%    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
+%    run-time checks inspect this tag for an exact match.\footnote{Immutable
+%    data structures could use a subtype check. \SP{} does not have any of these
+%    because the common Python data structures are all mutable.}
+
+\subsubsection{Unboxed Primitive}
 
@@
 %% purpose = types, checks, and optimizations can be formalized and analyzed
+
+%% NOTE
+%% - remark: model gives errors of the same kind in SP, but not always the same
+%%   message, after all its not our goal to reproduce SP exactly
 
@@
 \subsection{Nominal, Checked Types}
+\label{s:concrete-migration}
 
@@
 
+Enabling checked types has a huge ripple effect.
+The worst case is changing an annotation to a expect a checked data structure:
+\begin{enumerate}
+  \item all callers must create a checked value by invoking the right constructor, and
+  \item all clients of those callers can no longer expect an unchecked type.
+\end{enumerate}
+Programmers have to trace the annotation back to value-creation points, and then
+ensure that all uses of those values are compatible.
+
+\subsection{Pyre Compatibility}
+
+Many decisions in \SP{} are influenced by the widespread use of Pyre at Instagram.
+\begin{itemize}
+  \item
+    To minimize changes needed to adapt Pyre-annotated programs
+    to \SP{}, the syntax of \SP{} is compatible with Pyre.
+    Pyre can run useful checks on every \SP{} program.
+  \item
+    Because Pyre is a mature type checker, the \SP{} team has taken a
+    depth-first approach to development.
+    The language implements a small set of sound types efficiently and ignores
+    other types.
+    Of course, programmers who use advanced Pyre types can still use Pyre to
+    get compile-time feedback.
+  \item
+    The fact that \SP{} interprets unknown Pyre types as type dynamic changes
+    users' expectations about what the dynamic type should mean.
+    It is not necessarily a label for untyped Python code; dynamic code
+    can be analyzed to catch static bugs.
+    %% FILL what are the checks?
+\end{itemize}
+
+\subsection{Conclusion}
+
 Sound types in Static Python give programmers a way to improve performance.
@@
 
-%\appendix
+\appendix
+
+\section{Pyre Limitations}
+
+FILL condense the email thread in comment (see source)
+
+% Pyre limitations
+% 4 messages
+% Ben Greenman <benjaminlgreenman@gmail.com>	Fri, Sep 24, 2021 at 11:18 AM
+% To: Carl Meyer <carljm@fb.com>
+% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
+% >>> When we “add types” to a function (by adopting its module into Static
+% >>> Python, meaning Static Python now cares about the type annotations on that
+% >>> function), there may be other code calling that function with types that
+% >>> don’t match what it is annotated to receive. This then causes new runtime
+% >>> type errors for those calls once the module is made static. So sometimes we
+% >>> do have to fix those – we consider that to be fixing pre-existing type
+% >>> errors that Pyre missed. We probably see a lot less of that because the
+% >>> codebase is already type annotated and statically checked by Pyre, than we
+% >>> would if we were adopting Static Python into a previously-not-type-checked
+% >>> codebase.
+% >>
+% >> I'm surprised that you see any errors like this. Is Pyre unsound at
+% >> these pre-existing spots?
+% >
+% > Yes, Pyre is unsound in various ways (as a purely static checker of a
+% > gradually typed language must inherently be):
+% >
+% > 1. Some code just isn’t covered or checked by Pyre, because Pyre’s gradual
+% > adoption strategy is to only check functions that have annotations in their
+% > signature, or in extreme cases because the “code” that’s calling into Python
+% > isn’t Python code at all, but rather Python C extension code or Cython code.
+% > So Pyre has no visibility into calls from code it isn’t even checking.
+% > 2. Any code with annotations involving the `Any` type (or missing
+% > annotations) will have values that Pyre types as `Any`, and Pyre will
+% > silently allow you to do anything with those values. (It has no other
+% > choice, this is the only way that static-only gradual typing can work and
+% > actually be gradual.) So anything typed as `Any` by Pyre is a potential
+% > unsoundness. And although I say our codebase is largely type annotated,
+% > given that lots of it is pre-typing code that can’t be accurately statically
+% > annotated without rewriting it, there are plenty of Any annotations
+% > (especially `Dict[str, Any]` annotations) floating around.
+% > 3. People can (and do) just silence Pyre by adding a `# pyre-fixme` comment
+% > that silences a type error on the following line, if they are doing some
+% > mass adoption of types or otherwise just don’t feel up to fixing a type
+% > error for whatever reason. This tends to happen most often in cases where an
+% > annotation is just overly restrictive (e.g. claims to require `str` but in
+% > fact the method handles `None` just fine, and the argument really should be
+% > annotated as `str | None`.) So we have to fix the annotation to actually
+% > match the runtime behavior, and then we also are able to remove the
+% > pyre-fixme comments; everyone wins :)
+% > 4. If the entire codebase and all its type annotations and type stubs agree
+% > to lie about some dynamic wrinkle, Pyre will be satisfied that everything is
+% > good (and will have no way to know otherwise). A particular case of this we
+% > hit recently is that lots of values annotated as `str` in our codebase can
+% > actually at runtime be `Promise` objects that dynamically behave like a
+% > `str` if you treat them as one, and implement lazy translation of that
+% > string to the current user’s language. So as we adopt such code into Static
+% > Python, we have to correct those annotations from `str` to `str | Promise`
+% > and handle the fallout. A similar case is the `weakref` module in the Python
+% > standard library, which allows you to create “weak references” to objects
+% > that will not keep them alive in the garbage collector, but will just
+% > resolve to None if the object dies. There’s a `weakref.proxy` that creates a
+% > transparent weak-reference proxy to any object, and sometimes methods will
+% > return one of those but be annotated as returning the proxied type. This is
+% > fine for Pyre, as long as the proxy is sufficiently transparent at the level
+% > of Python semantics, but it doesn’t work for Static Python.
+% 
+% Points 1, 2, and 3 make sense.
+% 
+% I'm wondering if there are ways to write a fully-typed Pyre program (that
+% doesn't use those escape hatches) but is still unsound.* It sounds like
+% point 4 might be on this track ... but also that unchecked code (points 1,2)
+% might be involved.
+% 
+% For the weakrefs case, would Pyre approve a function like this one:
+% 
+% ```
+%   def f()->C:
+%     return weakref.proxy(C())
+% ```
+% 
+% If so, is that because Pyre comes with a builtin annotation for
+% `weakrefs.proxy` that lets it return the Any type?
+% 
+% 
+% * For comparison, TypeScript is unsound by design in a few ways. Any computed
+% string can index any object. And all object properties are covariant, despite
+% being mutable. I wonder if Pyre has similar "static unsoundness" by design.
+% Lu, Kuang-Chen <kuang-chen_lu@brown.edu>	Fri, Sep 24, 2021 at 11:50 AM
+% To: Ben Greenman <benjaminlgreenman@gmail.com>
+% Cc: Shriram Krishnamurthi <shriram@gmail.com>
+% (Internal discussion)
+% 
+% I think point 4, especially the `str` example, comes from the lack of structural typing.
+Interfaces can partially solve this problem.
+But with Nom-style interfaces, they have to define an interface and annotate the classes involved.
+% [Quoted text hidden]
+% Carl Meyer <carljm@fb.com>	Fri, Sep 24, 2021 at 12:37 PM
+% To: Ben Greenman <benjaminlgreenman@gmail.com>
+% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
+% [Quoted text hidden]
+% I think it could be reasonable to define “fully-typed” as to inherently exclude all four points above!
+This would also exclude most real-world Python programs, but might still be an interesting definition, because even then there are some unsoundnesses remaining (that I didn’t cover above because they are not usually a problem for us in practice; in most cases I think because they involve parts of the type system we don’t even pretend to support yet in Static Python.) I’ll list some of these known unsoundnesses within the type system below.
+% 
+% In practice the Python language and standard library include features that are considered too painful to use if fully typed, generally because sound typing would require typing things as `object` (the top type), requiring `isinstance` checks everywhere to keep the static checker happy, so an unsound `Any` is preferred for usability reasons.
+Weakref proxies are just one example of this; there are many others.
+% 
+% https://github.com/python/typeshed is the shared repository of type stubs for the Python standard library and popular third-party packages (used by all the static checkers: Pyre, mypy, pyright, pytype…).
+You can search it for `Any` to find many many more examples.
+% 
+% 
+% > For the weakrefs case, would Pyre approve a function like this one:
+% >
+% > ```
+% >  def f()->C:
+% >    return weakref.proxy(C())
+% > ```
+% >
+% > If so, is that because Pyre comes with a builtin annotation for
+% > `weakrefs.proxy` that lets it return the Any type?
+% 
+% Yes, exactly, this is in the shared “typeshed” repository: https://github.com/python/typeshed/blob/master/stdlib/_weakref.pyi#L33
+% 
+% >
+% > * For comparison, TypeScript is unsound by design in a few ways. Any computed
+% > string can index any object. And all object properties are covariant, despite
+% > being mutable. I wonder if Pyre has similar "static unsoundness" by design.
+% 
+% Right, and the Python type system has some similar unsound-by-design choices. A few that I’m aware of:
+% 
+% - Mutable attributes ought to be invariant (because they are in effect both getters and setters, thus appearing in both covariant and contravariant position), thus subclasses should not be allowed to change the type of a base class attribute.
+But in practice all Python type checkers permit a subclass to narrow the type of a base class attribute.
+Now that `Final[]` annotations are available on attributes (which would make them soundly covariant), it might be more reasonable to close this hole, here’s discussion on the mypy bug tracker: https://github.com/python/mypy/issues/3208.
+We are currently sound here in Static Python, but in an odd way; we should error on the subclass attribute type override itself, but instead we ignore it and continue to use the base class attribute type for the subclass also.
+% 
+% - There are some issues with `typing.Type[]` types (these are the types of class objects themselves) and the fact that they are covariant (which is widely relied on.) This is unsound because incompatible constructor signature overriding is allowed in subclasses (which is also widely used).
+It’s also unsound because the type of the `self` argument to any overridden method will be Parent on the parent method and Child on the child method, which is the opposite of the normal requirement that method arguments (appearing in a contravariant position) must only be widened on subclass overrides.
+This means that if some function accepts a type and calls an unbound method on it, providing the `self` arg explicitly, the covariance of `Type` is unsound.
+On these we are currently OK only because we don’t yet support `Type[]` annotations.
+% 
+% - Type checking argument splatting in calls is unsound.
+If we have a dynamic list `l` or dict `d` and we have a call like `foo(*l)` or `foo(**d)`, unless foo accepts `(*args, **kwds)` it is impossible to statically type check the call.
+Making this sound would have to either require the callee to accept (*args, **kwds) or require that the list or dict is a statically known literal; the latter totally defeats the purpose of argument splatting, the former is more conceivable but still makes it a lot less useful, so in practice type checkers just allow these calls.
+This one is OK in Static Python because we simply fallback to dynamic calls (and thus runtime argument type checks in the callee) for these calls (I mean it’s OK in that it doesn’t result in cascading unsoundness — it still means runtime error rather than static one for the call itself.)
+% 
+% There are more cases than just these, but I think the others are more esoteric.
+% 
+% Carl
 

### 2021-12-18 conformance_suite/methods_mutable.py
more tests

@@ -1,0 +1,9 @@
+# methods_mutable.py
+# This should pass.
+# This should terminate.
+
+class C:
+    def m(self):
+        return 2
+
+C.m = lambda self: 3

### 2021-12-18 conformance_suite/test_final_constant_folding_disabled_on_nonfinals.py

@@ -1,21 +1,0 @@
-# test_final_constant_folding_disabled_on_nonfinals.py
-# This should pass.
-# This should terminate.
-# This should be optimized.
-
-from typing import Final
-X: str = "omg"
-def f() -> str:
-    return X[1]
-# def test_final_constant_folding_disabled_on_nonfinals(self):
-#     codestr = """
-#     from typing import Final
-#     X: str = "omg"
-#     def f() -> str:
-#         return X[1]
-#     """
-#     with self.in_module(codestr) as mod:
-#         f = mod.f
-#         self.assertNotInBytecode(f, "LOAD_CONST", "omg")
-#         self.assertInBytecode(f, "LOAD_GLOBAL", "X")
-#         self.assertEqual(f(), "m")

### 2021-12-18 conformance_suite/test_final_in_args.py

@@ -1,17 +1,0 @@
-# test_final_in_args.py
-# This should fail.
-
-from typing import Final
-def f(a: Final) -> None:
-    pass
-# def test_final_in_args(self):
-#     codestr = """
-#     from typing import Final
-#     def f(a: Final) -> None:
-#         pass
-#     """
-#     with self.assertRaisesRegex(
-#         TypedSyntaxError,
-#         "Final annotation is only valid in initial declaration",
-#     ):
-#         self.compile(codestr, modname="foo")

### 2021-12-18 conformance_suite/test_incompat_override_method_arg_name.py

@@ -1,24 +1,0 @@
-# test_incompat_override_method_arg_name.py
-# This should fail.
-
-class A:
-    def m(self, x: str) -> int:
-        return 42
-class B(A):
-    def m(self, y: str) -> int:
-        return 0
-# def test_incompat_override_method_arg_name(self):
-#     codestr = """
-#         class A:
-#             def m(self, x: str) -> int:
-#                 return 42
-#         class B(A):
-#             def m(self, y: str) -> int:
-#                 return 0
-#     """
-#     with self.assertRaisesRegex(
-#         TypedSyntaxError,
-#         "<module>.B.m overrides <module>.A.m inconsistently. "
-#         "Positional argument 2 named `x` is overridden as `y`",
-#     ):
-#         self.compile(codestr)

### 2021-12-18 conformance_suite/test_module_level_final_decl.py

@@ -1,14 +1,0 @@
-# test_module_level_final_decl.py
-# This should fail.
-
-from typing import Final
-x: Final
-# def test_module_level_final_decl(self):
-#     codestr = """
-#     from typing import Final
-#     x: Final
-#     """
-#     with self.assertRaisesRegex(
-#         TypedSyntaxError, "Must assign a value when declaring a Final"
-#     ):
-#         self.compile(codestr, modname="foo")

### 2021-12-18 conformance_suite/test_type_type_final.py

@@ -1,11 +1,0 @@
-# test_type_type_final.py
-# This should pass.
-
-class A(type):
-    pass
-# def test_type_type_final(self):
-#     codestr = """
-#     class A(type):
-#         pass
-#     """
-#     self.compile(codestr)

### 2021-12-18 skipped_tests/test_assign_try_assign_in_finally.py

@@ -1,4 +1,4 @@
-# Reason: Format too complicated
+# Reason: Test hitted some skipped words
 def test_assign_try_assign_in_finally(self):
     codestr = """
         class B:

### 2021-12-18 skipped_tests/test_final_constant_folding_disabled_on_nonfinals.py

@@ -1,0 +1,13 @@
+# Reason: Test hitted some skipped words
+def test_final_constant_folding_disabled_on_nonfinals(self):
+    codestr = """
+    from typing import Final
+    X: str = "omg"
+    def f() -> str:
+        return X[1]
+    """
+    with self.in_module(codestr) as mod:
+        f = mod.f
+        self.assertNotInBytecode(f, "LOAD_CONST", "omg")
+        self.assertInBytecode(f, "LOAD_GLOBAL", "X")
+        self.assertEqual(f(), "m")

### 2021-12-18 skipped_tests/test_final_in_args.py

@@ -1,0 +1,12 @@
+# Reason: Test hitted some skipped words
+def test_final_in_args(self):
+    codestr = """
+    from typing import Final
+    def f(a: Final) -> None:
+        pass
+    """
+    with self.assertRaisesRegex(
+        TypedSyntaxError,
+        "Final annotation is only valid in initial declaration",
+    ):
+        self.compile(codestr, modname="foo")

### 2021-12-18 skipped_tests/test_module_level_final_decl.py

@@ -1,0 +1,10 @@
+# Reason: Test hitted some skipped words
+def test_module_level_final_decl(self):
+    codestr = """
+    from typing import Final
+    x: Final
+    """
+    with self.assertRaisesRegex(
+        TypedSyntaxError, "Must assign a value when declaring a Final"
+    ):
+        self.compile(codestr, modname="foo")

### 2021-12-18 skipped_tests/test_try_except_finally.py

@@ -1,4 +1,4 @@
-# Reason: Format too complicated
+# Reason: Test hitted some skipped words
 def test_try_except_finally(self):
     codestr = """
         class B:

### 2021-12-18 skipped_tests/test_try_return_finally.py

@@ -1,4 +1,4 @@
-# Reason: Code hitted some skipped words
+# Reason: Test hitted some skipped words
 def test_try_return_finally(self):
     codestr = """
     from typing import List

### 2021-12-18 skipped_tests/test_type_type_final.py

@@ -1,0 +1,7 @@
+# Reason: Test hitted some skipped words
+def test_type_type_final(self):
+    codestr = """
+    class A(type):
+        pass
+    """
+    self.compile(codestr)

### 2021-12-18 pj-7.1/paper.tex

@@
     The data structure implements the list API and comes with a checked constructor that
-    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
-    \cref{s:chkdict-impl} discusses the implementation further.
+    makes it impossible to create a \code{ChkList[int]} with non-integer elements.
+    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
+    run-time checks inspect this tag for an exact match.\footnote{Immutable
+    data structures could use a subtype check. \SP{} does not have any of these
+    because the common Python data structures are all mutable.}
 \end{enumerate}
@@
 
-\subsubsection{Non-Highlights / Restrictions}
-
-The \SP{} type system excludes include some features.
-Each of the following is low priority for the team; that said, they are
-ordered below rougly from most important (soonest to be implemented) to least.
+\subsubsection{Non-Highlights, Restrictions}
+
+The \SP{} type system does not include some features.
+Each of the following is low priority, but they are rougly ordered from most
+important (soonest to be implemented) to least.
 
@@
     %% demand seems low ... baffling
-    %% mypy = https://github.com/python/mypy/issues/731
-    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
 \end{itemize}
@@
 \subsection{Runtime System Highlights}
-%% Non-Highlights:
-%% - shadow frames = JIT feature, lite array to track Python call stack
-%% - strict modules ... avoids toplevel function wrapper
-
-\subsubsection{Bytecode Optimization}
-
-optimizations and bytecodes for: length, subscripting, function dispatch, method dispatch, field load/store
-
-table much later in paper
-
-IN ADDITION to optimizations ...
-not sure where this belongs exactly, but we go to a lot of work in the compiler
-to intelligently minimize runtime checks by only inserting them when the source
-value is of dynamic type (this includes also skipping CHECK\_ARGS if the
-function was statically invoked by another static function and thus we know the
-compiler checked argument types)
-
-\subsubsection{Virtual Method Tables}
-
+
+\begin{itemize}
+  \item
+    \textbf{VTables}
     Cinder adds virtual method tables to typed classes.
@@
     the static compiler will only emit against a statically known static type.'' -Carl
-
-with vtables, method resolution happens differently than Python: args. get
-resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
-
-vtables are built on Python's vectorcall signature; though new, they're not
-totally from scratch
 
@@
 
-\subsubsection{Checked Data Structures}
-\label{s:chkdict-impl}
-
-As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
-come with both a type and an implementation.
-The implementation provides the same interface as the built-in;
-the main difference is that it adds type checks.
-
-For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
-with keys of type \code{K} and values of type \code{V}.
-The implementation has three main components:
-\begin{itemize}
-  \item
-    Constructor, basically from dict to CD, checks integrity.
-    All CD valuess have to go through this, whether types or untyped.
-  \item
-    Instances keep a tag, used as exact type in boundary check.
-    (Tag may not be CD[i i] but rather a sentinel generated from it.)
-  \item
-    Writes are protected in every context.
+  \item
+    \textbf{Bytecode Optimization}
+    (could use a full section)
+
+  %\item
+  %  \textbf{Shadow Frames}
+  %  (CinderDoc/static_python.rst)
+  %  magic import that signals to the Static
+  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
+  %  improves performance of function calls by avoiding the creation of full
+  %  Python frame objects until they are definitely needed (e.g. if an
+  %  exception is raised.) In the future this should become default.
+
+  \item
+    \textbf{Checked Dict}
+    Implementation too, to support the types.
+
 \end{itemize}
 
-%    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
-%    run-time checks inspect this tag for an exact match.\footnote{Immutable
-%    data structures could use a subtype check. \SP{} does not have any of these
-%    because the common Python data structures are all mutable.}
-
-\subsubsection{Unboxed Primitive}
+\subsection{Additional Features}
+
+% things we did not model ... is that worth making a distinction here?
+
+strict modules
+
+decorator to force inlining, for 1-line functions
+
+\subsubsection{Primitive Types}
 

### 2021-12-20 pj-7.1/paper.tex
pj: checkpoint, split non-highlights into other secs

--- 2021-12-17 pj-7.1/paper.tex
@@
 
+%% TODO
+%  \item \emph{Concrete Parameterized Classes}.
+%    Except for a handful of built-in types,
+%    parameterized classes are enforced only with shallow checks.
+%    In the near future, these classes will gain concrete versions
+%    for programmers to opt-in to.
+
 \subsubsection{Mixed-Typed Class Hierarchies}
@@
 
+- static -> static overrides restricted, must be compatible
+- static -> dynamic overrides wrapped, in vtable
+- dyn -> stat ditto?
+
+XX 
+%  \item \emph{Method Overloading}.
+%    Methods cannot be overloaded.
+%    Each combination of a class and a name can refer to at most one method.
+%    This behavior comes from Python.
+%
+%  \item \emph{Multiple Inheritance}.
+%    Multiple inheritance is not supported in a typed way.
+%    \Cref{s:impl} discusses this point further.
+%
+%  \item \emph{Strong overrides}.
+%    Methods and fields may be overridden, but only with values of the same type.
+%    % In fact, \SP{} compiles fields using Python slots
+%    % declarations (\code{\_\_slots\_\_}).
+%    %
+%    % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
+%    % but the normal field access syntax skips any overrides (o.f)
+
 \subsubsection{Types for Python}
@@
 
-\subsubsection{Non-Highlights / Restrictions}
-
-The \SP{} type system excludes include some features.
-Each of the following is low priority for the team; that said, they are
-ordered below rougly from most important (soonest to be implemented) to least.
-
-\begin{itemize}
-  \item \emph{Concrete Parameterized Classes}.
-    Except for a handful of built-in types,
-    parameterized classes are enforced only with shallow checks.
-    In the near future, these classes will gain concrete versions
-    for programmers to opt-in to.
-
-  \item \emph{Callable Types, First-Class Classes}.
-    The Instagram codebase rarely uses first-class functions, classes, or objects.
-    Thus there is little need for either a \code{Callable} type to describe functions
-    or first-class types for objects.
-    When such values do appear, \SP{} assigns the dynamic type and spot-checks their
-    uses.
-    For example, every application of a first-class function gets wrapped in a check
-    if the context expects a certain type.
-
-  \item \emph{Method Overloading}.
-    Methods cannot be overloaded.
-    Each combination of a class and a name can refer to at most one method.
-    This behavior comes from Python.
-
-  \item \emph{Multiple Inheritance}.
-    Multiple inheritance is not supported in a typed way.
-    \Cref{s:impl} discusses this point further.
-
-  \item \emph{Strong overrides}.
-    Methods and fields may be overridden, but only with values of the same type.
-    % In fact, \SP{} compiles fields using Python slots
-    % declarations (\code{\_\_slots\_\_}).
-    %
-    % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
-    % but the normal field access syntax skips any overrides (o.f)
-
-  \item \emph{Recursive Types}.
-    \SP{} does not support recursive types.
-    Neither does Pyre or mypy.
-    %% demand seems low ... baffling
-    %% mypy = https://github.com/python/mypy/issues/731
-    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
-\end{itemize}
+%  \item \emph{Callable Types, First-Class Classes}.
+%    The Instagram codebase rarely uses first-class functions, classes, or objects.
+%    Thus there is little need for either a \code{Callable} type to describe functions
+%    or first-class types for objects.
+%    When such values do appear, \SP{} assigns the dynamic type and spot-checks their
+%    uses.
+%    For example, every application of a first-class function gets wrapped in a check
+%    if the context expects a certain type.
+%
+%  \item \emph{Recursive Types}.
+%    \SP{} does not support recursive types.
+%    Neither does Pyre or mypy.
+%    %% demand seems low ... baffling
+%    %% mypy = https://github.com/python/mypy/issues/731
+%    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
 

### 2021-12-20 pj-7.1/paper.tex
pj: sec 2.1 full draft

--- 2021-12-17 pj-7.1/paper.tex
@@
   \item
-    Instagram is / had been using Cython to improve the performance of select modules.
-    Cython compiles Python source code to C extension modules for
-    a faster Python [CITE].
-    Instagram had converted several performance-critical modules to Cython prior
-    to starting work on \SP{}.
-    \SP{} replaced them.
-    (How does this help understanding type system highlights?)
+    Prior to investing in \SP{}, Instagram used Cython to compile
+    performance-critical code to C extension modules.
+    Cython performed well, but mixing pure Python with a few compiled modules
+    made for an awkward developer experience.
+    \SP{} aims to match the Python workflow.
   \item
@@
 The \SP{} solution is to provide two versions of every parameterized type,
-both of which support efficient run-time checks.
+both of which support efficient run-time checks.\footnote{Currently, only a handful
+of built-in data structures provide shallow and concrete versions. User-defined
+classes will soon follow this approach.}
 One version is shallow.
@@
 \end{enumerate}
+%
+These options let programmers decide whether to change their code to benefit from
+fully-sound types, or leave it as-is for a lower payoff.
 
@@
 
-These options let programmers decide whether to change their code to benefit from
-fully-sound types, or leave it as-is for a lower payoff.
 Often, the shallow types are more attractive because concrete types impose
@@
 
-\subsubsection{Mixed-Typed Class Hierarchies}
-
-%% TODO very important, one of the main things IG does use!!!
-
-The Instagram codebase uses lots of classes and objects.
-Most of these are simple, Java-style: classes declared on
-the top level and single inheritance.
-It was therefore essential to get inheritance working across type boundaries.
-Typed classes must be able to inherit from untyped classes,
-and vice-versa untyped classes must be able to inherit from typed ones.
-
-FILL literature? StrongScript Nom classic GT?
-
-FILL either way, spell out what happens for these interactions.
-Where are the checks needed, checks skipped.
-
-\subsubsection{Types for Python}
-
-[FILL]
-
-Which types accommodate Python code? Which from PEP typing etc.
-
-Unions are static only,
-except for unions to false because those are common and can be implemented efficiently.
-
-    Optional(T) is the only union. Others turn into Dynamic.
-
-\subsubsection{Non-Highlights / Restrictions}
-
-The \SP{} type system excludes include some features.
-Each of the following is low priority for the team; that said, they are
-ordered below rougly from most important (soonest to be implemented) to least.
-
-\begin{itemize}
-  \item \emph{Concrete Parameterized Classes}.
-    Except for a handful of built-in types,
-    parameterized classes are enforced only with shallow checks.
-    In the near future, these classes will gain concrete versions
-    for programmers to opt-in to.
-
-  \item \emph{Callable Types, First-Class Classes}.
-    The Instagram codebase rarely uses first-class functions, classes, or objects.
-    Thus there is little need for either a \code{Callable} type to describe functions
-    or first-class types for objects.
-    When such values do appear, \SP{} assigns the dynamic type and spot-checks their
-    uses.
-    For example, every application of a first-class function gets wrapped in a check
-    if the context expects a certain type.
-
-  \item \emph{Method Overloading}.
+\subsubsection{Gradual Class Hierarchies}
+\label{s:inheritance}
+
+One important feature of \SP{} is that class hierarchies can
+mix typed and untyped code.
+An untyped class may inherit from a typed one and vice-versa.
+The point of this feature is to let programmers add types to a single class
+independently of its ancestors and children.
+
+% st-ecoop-2007 has no inheritance, object is collection of members simple (Abadi--Cardelli)
+
+Gradual class hierarchies are rarely studied in the literature, especially
+for a language where truly untyped classes can enter the mix.\footnote{Two
+languages that provide separate hierarchies for untyped and (gradually) typed
+classes are Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}.}
+The closest related work, for Nom~\cite{mt-oopsla-2017}, is much more flexible.
+\SP{} explores a simpler point in the design space, starting with
+a few restrictions on types and inheritance:
+
+\begin{enumerate}
+  \item
+    To benefit from types, classes must be declared on the module top level
+    and may have at most one parent.
+    Nested class declarations, first-class classes, and classes with
+    multiple parents default to un-optimized Python behavior~(\cref{s:impl}).
+
+  \item
     Methods cannot be overloaded.
-    Each combination of a class and a name can refer to at most one method.
-    This behavior comes from Python.
-
-  \item \emph{Multiple Inheritance}.
-    Multiple inheritance is not supported in a typed way.
-    \Cref{s:impl} discusses this point further.
-
-  \item \emph{Strong overrides}.
-    Methods and fields may be overridden, but only with values of the same type.
+    This restriction comes from Python.
+
+  \item
+    Methods and fields may be overridden in arbitrary ways by untyped code.
+    In typed code, however, overrides must be compatible subtypes.
+    For example, a method that returns a number cannot be overridden by a method
+    that returns the dynamic type~(example: \cref{f:gg-failure-stat}).
+
     % In fact, \SP{} compiles fields using Python slots
@@
 
-  \item \emph{Recursive Types}.
-    \SP{} does not support recursive types.
-    Neither does Pyre or mypy.
+\end{enumerate}
+%
+With this context in mind, every class member is either \emph{reliably} typed or untyped.
+In particular, a typed member cannot be overridden by a less-precise one that uses
+type dynamic.
+\SP{} can therefore optimize dispatch from typed callers to typed methods
+and use checked dispatch for other combinations.
+One extra step arises when an untyped class overrides a typed method.
+In this case, \SP{} creates a wrapper around the overriding method to check
+that it computes type-correct results.
+% Because these wrappers perform a simple first-order check, they
+% are supported by an efficiently implementation that cooperates with
+% \SP{}
+The wrappers are handled efficiently by \SP{}'s vtables~(\cref{s:vtable}).
+
+\subsubsection{Tailored Types}
+
+\SP{} is selective about which types to support and which to omit.
+In general, supported types are widely-used and come with an efficient
+implementation.
+Other widely-used types are deferred to Pyre for static checks until
+the day when \SP{} adds support.\footnote{At present, the \SP{} team
+has no plans to support all Pyre types. The top priority is to implement
+core types efficiently.}
+
+The core supported types describe basic data (integers, booleans, strings),
+data structures (lists, dicts, promises),
+and nominal classes.
+Additional ``C types'' enable aggressive compilation of primitive
+values~(\cref{s:ctypes}).
+Union types are well-supported statically and narrow down in accordance
+with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
+but hardly supported at run-time.
+Only binary unions of a type and false get enforced with a run-time check
+(\code{Union[T, False]}, or \code{Optional[t0]}).
+
+Three unsupported types bear special mention.
+There are no first-class class or object types, as noted in \cref{s:inheritance}.
+There are also no types for first-class functions (callables);
+functions are common at Instagram, but first-class functions are rare.
+Finally, \SP{} does not support recursive types due to low demand.
+Neither Pyre nor mypy have recursive types either.
 %% demand seems low ... baffling
@@
 %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
-\end{itemize}
 
@@
 \subsubsection{Virtual Method Tables}
+\label{s:vtable}
 
@@
 \subsubsection{Unboxed Primitive}
+\label{s:ctypes}
 

### 2021-12-21 pj-7.1/paper.tex
pj: quick note (in comment) about ctypes and dyn

@@
 that allows untyped expressions within a statically-typed context.
-Whenever an expression or variable lacks a type annotation,
-\SP{} uses the dynamic type as a default and skips most compile-time checks.
+Whenever an expression or variable lacks a type annotation, \SP{} uses the
+dynamic type as a default and skips most (but not all!) compile-time checks.
 Every untyped Python program is thus a well-formed \SP{} program.
@@
 of values inside the dictionary at runtime.
-To an outsider, this behavior might seem odd: why not reject undefined types with an error?
+
+At first glance, this behavior might seem odd: why not reject undefined types with an error?
 The reason is to reduce friction with the Pyre annotations that are already
@@
 Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
-efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions.
+efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions~(\cref{s:idiomatic-types}).
 By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.
 
-The dynamic type is not, however, the flexible dynamic type provided
-by \emph{true} gradual languages~\cite{svcb-snapl-2015}.
+%% intro sentence
+The dynamic type is also different than the flexible dynamic type provided
+by true gradual languages~\cite{svcb-snapl-2015}.
 Replacing part of a type with dynamic can lead to both compile-time errors
 and run-time errors.
-In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee.\footnote{We
-assume a standard type precision relation. Of course, one could argue that these programs
-do not break the gradual guarantees if type precision does not allow our uses of the dynamic type.}
+In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
+for a standard type precision relation.
+%% \footnote{We
+%% assume a standard type precision relation. Of course, one could argue that these programs
+%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
+%%   %% tailored = describes exactly the migrations that SP allows
 \Cref{fig:gg-failure} presents two examples, one for each kind of failure:
+%% \footnote{In addition, C types~(\cref{s:ctypes}) are incompatible with type dynamic.}
 \begin{itemize}
@@
 \subsubsection{Tailored Types}
+\label{s:idiomatic-types}
 

### 2021-12-21 pj-7.1/paper.tex
pj: checkpoint, 2.2 intro + vtable + chk

@@
 %% - what are the idiomatic base SP types ... int or Int ?
+%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
 
@@
 to directly express type checks and type-based optimizations.
-These enhancements take full advantage of \SP{} type information;
-they are built on top of CPython so that untyped code runs on
+These enhancements take full advantage of \SP{} type information.
+They are built on top of CPython so that untyped code runs on
 the platform.
@@
 
+Cinder, the \SP{} runtime system, extends the standard Python runtime in
+several ways.
+Some of these ensure the integrity of types.
+Others use type soundness to generate efficient code.
+
 \subsubsection{Bytecode Optimization}
+
+TODO
 
@@
 
-Cinder adds virtual method tables to typed classes.
-``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
-the static compiler will only emit against a statically known static type.'' -Carl
-
-with vtables, method resolution happens differently than Python: args. get
-resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
-
-vtables are built on Python's vectorcall signature; though new, they're not
-totally from scratch
-
-%    No, there’s no overload resolution by arguments in Python. This difference
-%    in behavior isn’t so much desired as just a consequence of making
-%    `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
-%    the callable is placed on the stack, then the arguments, then there is a
-%    `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
-%    means that first the callable is resolved, then the arguments. But with
-%    `INVOKE_METHOD` we want to resolve the callable as part of the invoke
-%    itself, since this gives us opportunity to inline-cache the target of the
-%    call instead of always having to call something dynamic and unknown that’s
-%    on the stack. So that necessarily implies that first the arguments are
-%    resolved and placed on the stack, then the callable is resolved as part of
-%    the `INVOKE_METHOD`. -Carl
-%
-%    2021-11-15: Dino says the JIT will check/upgrade INVOKE_METHOD to INVOKE_FUNCTION
-%    2021-07-23: final classes get INVOKE_FUNCTION, not INVOKE_METHOD
+Cinder adds virtual method tables (vtables) to typed classes.
+These tables help to speed up method dispatch relative to Python's
+dynamic lookup.
+
+Specifically, calls to static methods that appear in statically-typed
+code use the vtable to find an address for the method.
+If the method is part of a final class, then the call is further optimized
+to a direct function-call jump.
+JIT compilation can also upgrade vtable lookups to function calls.
+
+%% related bytecodes: INVOKE_METHOD INVOKE_FUNCTION
+%%  JIT can upgrade I_METH to I_FUN
+%%  final classes get I_FUN
+%% implementation: built on Python vectorcalls, so it's not totally new
+%% beware: with vtables, method resolution is different than Python b/c args. get
+%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
+%%
+%% > This difference in behavior isn’t so much desired as just a consequence of making
+%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
+%% > the callable is placed on the stack, then the arguments, then there is a
+%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
+%% > means that first the callable is resolved, then the arguments. But with
+%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
+%% > itself, since this gives us opportunity to inline-cache the target of the
+%% > call instead of always having to call something dynamic and unknown that’s
+%% > on the stack. So that necessarily implies that first the arguments are
+%% > resolved and placed on the stack, then the callable is resolved as part of
+%% > the `INVOKE_METHOD`. -Carl
 
@@
 come with both a type and an implementation.
-The implementation provides the same interface as the built-in;
-the main difference is that it adds type checks.
+The implementation provides the same interface as the built-in,
+but uses the type to reject some uses.
 
@@
   \item
-    Constructor, basically from dict to CD, checks integrity.
-    All CD valuess have to go through this, whether types or untyped.
-  \item
-    Instances keep a tag, used as exact type in boundary check.
-    (Tag may not be CD[i i] but rather a sentinel generated from it.)
-  \item
-    Writes are protected in every context.
+    A constructor that uses a type and a Python dictionary
+    to initialize a checked dictionary.
+    The constructor ensures that all values of type \code{ChkDict[K, V]}
+    begin with well-typed elements.
+  \item
+    Guarded update functions.
+    Every operation that mutates or extends a checked dictionary must
+    validate untyped arguments.
+  \item
+    A type tag for boundary checks.
+    When a checked dictionary enters typed code from an untyped context,
+    the runtime checks its key and value types for an exact match.
+    For example, a value of type \code{ChkDict[K, V]} matches the type
+    \code{ChkDict[str, dynamic]} only if \code{K} is type \code{str}
+    and \code{V} is dynamic.
 \end{itemize}
-
-%    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
-%    run-time checks inspect this tag for an exact match.\footnote{Immutable
-%    data structures could use a subtype check. \SP{} does not have any of these
-%    because the common Python data structures are all mutable.}
+%
+In general, other checked datatypes have the same three components:
+a constructor, checked update functions, and a tag.
 

### 2021-12-21 pj-7.1/paper.tex
pj: sec 2.2 draft

@@
 
-\subsubsection{Bytecode Optimization}
-
-TODO
-
-optimizations and bytecodes for: length, subscripting, function dispatch, method dispatch, field load/store
-
-table much later in paper
-
-IN ADDITION to optimizations ...
-not sure where this belongs exactly, but we go to a lot of work in the compiler
-to intelligently minimize runtime checks by only inserting them when the source
-value is of dynamic type (this includes also skipping CHECK\_ARGS if the
-function was statically invoked by another static function and thus we know the
-compiler checked argument types)
+\subsubsection{Bytecode Instructions}
+
+Cinder is a bytecode compiler that extends the Python instruction set.
+All the standard Python instructions work similarly to Python 3.8;
+in other words, an untyped Python program has the same semantics in
+Cinder as in the standard interpreter.
+The added instructions help either to express runtime checks or to
+quickly perform a statically-determined action.
+
+The two main instructions that express run-time checks are \code{CAST}
+and \code{CHECK\_ARGS}.
+The former checks a value against a type.
+The latter is a special cast for functions and methods;
+it appears at the top of every function and checks that the
+actual parameters match the types of the formal parameters.
+
+A simple optimizing instruction is \code{FAST\_LEN}, which expects
+a built-in value\footnote{Because of subtyping, it does not suffice to check
+the type. All Python lists support \code{FAST\_LEN} but some subclasses
+of the Python list type may not.}
+and quickly computes its length.
+Another is \code{LOAD\_FIELD}, which scans a fixed offset within an object.
+By contrast, the standard Python instruction for loading a field needs to
+search a dictionary~(\code{LOAD\_ATTR}).
+
+In addition to speeding code up, the \SP{} compiler takes care
+to avoid slowing code down.
+It does so by avoiding casts and run-time checks that are certain to succeed.
+For example, calls from a typed context to a typed function are statically
+checked for compatibility.
+There is no need to run \code{CHECK\_ARGS} on typed inputs, so the bytecode
+compiler inserts a special call instruction to skip past the check.
 

### 2021-12-24 pj-7.1/paper.tex
pj: sec 5 rough draft, production experience

@@
 
+  \newcommand{\bmname}[1]{\textbf{#1}}
 \newcommand{\typefont}[1]{\mathsf{#1}}
@@
   of the \SP{} approach and its soundness.
-  Our performance claims are supported by data from both
-  a public suite a microbenchmarks and the proprietary Instagram codebase.
+  Our performance claims are based on production experience in the Instagram
+  web server.
+  Thanks to \SP{}, global CPU usage has improved by almost 4\% within a year.
 
@@
   % Why does this work matter?
-  \SP{} is the first language that improves performance across the board through
-  the addition of sound types to an untyped language.
+  \SP{} is the first sound gradual language whose piece-meal application
+  to a realistic codebase has consistently improved performance.
   %% Prior work on Nom and Dart 2 showed that concrete types are promising.
@@
     %% Q. why is the python so nominal? style guide? java background?
-    The codebase makes heavy use of first-order classes and objects.
-    First-class classes, objects, and even functions are rare.
+    The codebase depends heavily on first-order functions and objects in
+    performance-critical code.
+    Higher-order code is fairly common, but rarely appears on a critical path.
   \item
@@
 There are no first-class class or object types, as noted in \cref{s:inheritance}.
-There are also no types for first-class functions (callables);
-functions are common at Instagram, but first-class functions are rare.
+There are also no types for first-class functions (callables).
+Pyre can check these types instead, to point out clear logical flaws.
 Finally, \SP{} does not support recursive types due to low demand.
@@
 
-\SP{} does not attempt to type first-class classes:
-partly because they see little use at Instagram (FILL how much?)
-and partly because it is unclear how to incorporate them into the
-nominal type system.
+\SP{} does not attempt to type first-class classes: partly because they have
+yet to appear in performance-critical code at Instagram and partly because it
+is unclear how to incorporate them into the nominal type system.
 The straightforward but restrictive approach is to force code that uses a
@@
 
-\section{Evaluation}
+\section{Production Experience}
 \label{s:eval}
-
-\subsection{Microbenchmarks}
-
-richards etc.
-
-%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
-%%
-%% deltablue
-%% fannkuch
-%% nbody
-%% nqueens
-%% pystone
-%% richards
-%% richards_static
-%%
-%% UHOH except for richards_static, none of these have types!!!
-%% Need to ask if C&D have other typed benchmarks around
-%% ... if not, ask them to port these and measure and tune if needed.
-
-\subsection{Production Experience}
-
-%% 2021-12-15: C&D may have an SP off/on switch soon
-%% - accumulated CPU savings since July 1 = ~2% total IG production CPU
-%%   ... how does that number work?
-%% - in the beginning (?) H1 got 1% from first conversions + 0.7% from all Cython (higher bar)
-%%   - no Cython regressions, SP + JIT is good enough
-%%   - some Cython still exists, imported as library code from another codebase
+%% 2021-12-15: C&D may have an SP off/on switch soon to measure totals
+
+% "The workload for those rough percentage numbers is “Instagram web server
+%  production.” We have pretty extensive production profiling that lets us
+%  identify CPU savings with good granularity when a change lands in prod, and the
+%  observed CPU savings for a variety of staticification changes we’ve landed over
+%  the last half sum up to 2% of total IG prod CPU usage." --Carl
+
+\begin{table}
+  \caption{Overview: \SP{} in production}
+  \label{t:prod-stat}
+  \begin{tabular}{l@{~~}ll@{~~}l}
+    8     & months in production & 3.7\% & global CPU improvement \\
+    530   & modules converted  &  0 & performance regressions
+    %% TODO 530 / ???? how much of codebase?
+  \end{tabular}
+\end{table}
+
+The Instagram web server has been running \SP{} in production since April 2021.
+Overall, the results are very encouraging.
+Instagram's internal profiling tools, which continuously monitor
+performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
+\cref{t:prod-stat} reports a few other production statistics.
+Over 500 modules have been converted thus far.
+None of the converted modules have gotten slower because of \SP{}
+
+\subsection{The Migration Path}
+
+\SP{} first entered the Instagram web server as a replacement for Cython,
+which Instagram had been using to speed up critical modules.
+Cython performed well, but its partial adoption led to an awkward workflow
+because unlike standard Python, Cython is a compiled language.
+Replacing these modules with \SP{} let developers return to a conventional
+Python workflow.
+\SP{} (combined with Cinder) did not slow down any of these
+performance-critical modules, and even sped a few up.
+The net improvement was a 0.7\% reduction in global CPU usage.
+
+Later \SP{} migrations have been led by profiling to find frequently-executed code.
+The goal of these opportunistic migrations was to test whether \SP{} could
+improve typical modules without slowing down their neighbors.
+As a point of comparison, Typed Racket can improve individual Racket
+modules but the type checks that it requires on module boundaries
+can lead to huge slowdowns~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}.
+\SP{} does not appear to suffer in the same way.
+
+During the first half of 2021, the \SP{} team identified critical
+modules and proposed types to the maintainers of these modules.
+Often, the types came with small code changes.
+The accepted types and changes resulted in a 1\% improvement
+in global CPU usage.
+During the second half of 2021, the \SP{} team applied the same
+process at a larger scale by incorporating a tool that drafts
+initial type annotations.
+These changes resulted in a 2\% global CPU improvement.
+
+As of December 2021, the \SP{} team has converted 530 modules.
+Most of these came from the annotation tool (400); the rest
+are from hand conversions.
+In a typical week, the team adds types to four additional modules.
+These modules have seen speedups ranging from 0\% to 25\% depending
+on their nature.
+Regressions have not been an issue.
+The only cause for alarm is when a module gets only a small improvement
+without code changes.
 %% - rough measure: 0 to 25% per module
@@
 
-%% - 530 SP modules
-%%   - 400 from a codegen tool
-%%   - 130 by hand
-%%   - currently +~3/4 per week by hand
-
-%% - pushback one time about removing MI
-%% - otherwise, people happy with low-cost improvements
-
-%% - review of 30 winning diffs "provided the bulk of our wins this half"
-%%   most common changes, not in careful order, are:
-%%   1) adjustments to tests to pass correct types instead of mock objects, as as
-%%      not to fail type checks in SP code. This is an inherent and desired
-%%      incompatibility; next half we plan to work on improving the Python mock
-%%      framework to automatically create mocks of the correct types.
+\subsection{Analyzing Code Changes}
+
+The fact that the \SP{} team has been changing code as well as types
+during its migrations is potentially troubling.
+It could be that the \SP{} approach to gradual typing only pays off
+after significant modifications, or only for a restrictive style of programming.
+
+To understand the code changes, an author affiliated with Instagram reviewed
+approximately 30 patches that significantly improved performance.
+In general, the patches contain relatively minor code changes.
+The biggest changes are to testing code rather than production code.
+Both these points are encouranging.
+The following nine points describe common patterns:
+
+\begin{enumerate}
+  \item
+   adjustments to tests to pass correct types instead of mock objects, as as
+   not to fail type checks in SP code. This is an inherent and desired
+   incompatibility; next half we plan to work on improving the Python mock
+   framework to automatically create mocks of the correct types.
    %% - bg: lookup the dino email about mocks
-%%   2) adjustments to mock assertions in tests about calls to adjust for the
-%%      fact that SP currently converts all call arguments to be positional (we plan
-%%      to fix this incompatibility by falling back to the original call arguments
-%%      when we detect the target function has been patched.)
+
+ \item
+   adjustments to mock assertions in tests about calls to adjust for the
+   fact that SP currently converts all call arguments to be positional (we plan
+   to fix this incompatibility by falling back to the original call arguments
+   when we detect the target function has been patched.)
    %% - bg: ??? bit fuzzy about positional and others
-%%   3) adjustments around class attributes vs instance attributes (Python allows
-%%      this to be a bit fuzzy, where a class attribute can be a fallback default
-%%      for an instance attribute if not set, but SP currently requires an attribute
-%%      to be clearly one or the other, and we have more efficient access for
-%%      instance attributes via LOAD_FIELD that we don’t have for class attributes.)
+
+ \item
+   adjustments around class attributes vs instance attributes (Python allows
+   this to be a bit fuzzy, where a class attribute can be a fallback default
+   for an instance attribute if not set, but SP currently requires an attribute
+   to be clearly one or the other, and we have more efficient access for
+    instance attributes via \code{LOAD\_FIELD} that we don’t have for class attributes.)
    %% - related to github #37 ?
-%%   4) convert @classmethod that don’t actually need the class to @staticmethod,
-%%      allowing for emitting direct INVOKE_FUNCTION instead of INVOKE_METHOD
-%%   5) extract special types that our compiler doesn’t yet support/understand
-%%      (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
-%%      the module we are converting to static. We have work in progress on
-%%      supporting some Enum types, and plan to add a @dataclass intrinsic soon.
-%%   6) Removing the use of super-dynamic Python features like `__setattr__` that
-%%      we don’t support. So far in the cases where we’ve done this it hasn’t
-%%      required a major rewrite.
+
+ \item
+   convert @classmethod that don’t actually need the class to @staticmethod,
+    allowing for emitting direct \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}
+
+ \item
+   extract special types that our compiler doesn’t yet support/understand
+   (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
+   the module we are converting to static. We have work in progress on
+   supporting some Enum types, and plan to add a @dataclass intrinsic soon.
+
+ \item
+   Removing the use of super-dynamic Python features like \code{\_\_setattr\_\_} that
+   we don’t support. So far in the cases where we’ve done this it hasn’t
+   required a major rewrite.
    %% - bg: more details pls ,,, sketch the rewrite
-%%   7) Removing the use of keyword-only arguments (bare `*`) in function
-%%      signatures, since we don’t yet support calling such functions. We plan to
-%%      support this next half.
-%%   8) Convert Python integers and bools used in very hot code paths into SP
-%%      machine primitives.
+
+ \item
+   Removing the use of keyword-only arguments (bare `*`) in function
+   signatures, since we don’t yet support calling such functions. We plan to
+   support this next half.
+
+ \item
+   Convert Python integers and bools used in very hot code paths into SP
+   machine primitives.
    %% - bg: focus on block, convert, no problem?
-%%   9) Occasionally converting a regular Python list or dict to a CheckedList or
-%%      CheckedDict.
+
+ \item
+   Occasionally converting a regular Python list or dict to a CheckedList or
+   CheckedDict.
    %% - bg: how occasional? worried about power of the "gradual soundness" claim
-%%
-%% common theme: gotta change more tests than code
-%%
+\end{enumerate}
+
 %% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
@@
 
+\subsection{Developer Experience}
+
+%% - pushback one time about removing MI
+%% - otherwise, people happy with low-cost improvements
+
+%% - pyre challenges?
+
+\subsection{Microbenchmarks}
+%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
+
+%% TODO jit vs nojit
+
+%% 2021-12-23: TODO decide whether to ask for typed benchmarks to report ratios
+%%  Carl: we could commit to typed 2--3 more, but pystone etc (?!) are notoriously bad
+
+%% Somehow though, we ought to have public code available.
+
+Prior to deploying \SP{} in production, the team used microbenchmarks to study performance.
+These microbenchmarks are admittedly small and quite different than typical application code.
+That said, their use gave the team a coarse idea of whether \SP{} was on track during development.
+We include these microbenchmarks to offer reproducible performance numbers;
+unlike the Instagram web server, the benchmark code is public.
+
+\Cref{t:microbenchmark} reports current microbenchmark results.
+Each row gives the ratio of two running times: fully-typed \SP{} code versus
+unannotated \SP{} code.
+A ratio between 0 and 1 means that performance improved thanks to the types.
+[FILL] numbers.
+
+\begin{table}[t]
+  \caption{Microbenchmark Performance Ratios}
+  \label{t:microbenchmark}
+  %% show untyped Python time?
+  \begin{tabular}{ll}
+    Name               & Typed / Dynamic \\\midrule
+    \bmname{deltablue} &             0\% \\
+    \bmname{fannkuch}  &             0\% \\
+    \bmname{nbody}     &             0\% \\
+    \bmname{nqueens}   &             0\% \\
+    \bmname{pystone}   &             0\% \\
+    \bmname{richards}  &         0.125\% \\
+  \end{tabular}
+\end{table}
+
+%% TODO conclusions? threats?
+
 \section{Related Work}
@@
 Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.
+
+Grace nominal transient aint bad at all
 

### 2021-12-30 pj-7.1/paper.tex
pj: more in-comment notes from carl

@@
 
+%% 2021-12-23: TODO
+%% Carl: The behavior of dynamic type in static code should be:
+%%   a) any type can flow to dynamic
+%%   b) dynamic can flow to any type, but it requires the insertion of a runtime cast for soundness
+%%   c) dynamic type should emit the same bytecode for any given operation that non static code would emit
+%%      ... I think differing behavior around (c) is also unlikely — the structure of our
+%%      code generation is that we call methods on type objects to emit byte code for
+%%      various operations on a value of that type, and our dynamic type simply
+%%      forwards all those calls to the default Python compiler, which should always
+%%      result in dynamic type emitting the same byte code that nonstatci code would.
+%%      This isn’t universal, though (i.e. for some operations, e.g. binary ops, we
+%%      might explicitly handle various cases for operand type pairings, not always by
+%%      calling a polymorphic method on each operand type), which could open the door
+%%      to differing behavior here. But I can’t think of or find any such intentional
+%%      cases, and I suspect if any did crop up we’d consider it a bug.
+
 Like most gradual languages, \SP{} includes a dynamic type
@@
 \Cref{fig:gg-failure} presents two examples, one for each kind of failure:
-%% \footnote{In addition, C types~(\cref{s:ctypes}) are incompatible with type dynamic.}
+%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
 \begin{itemize}
@@
     the typed \code{m1} method with another method that returns the dynamic type.
+
+    %% 2021-12-13: TODO NOTE SWALLOW MORE
+    %% I think the fundamental unusual thing we do here is our per-module
+    %% incremental adoption strategy for our compiler. We don’t apply our compiler
+    %% to the entire codebase, using presence or absence of type annotations as
+    %% the sole indicator of whether to compile code with static or dynamic
+    %% behavior. We apply our compiler only to selected opted-in modules, and the
+    %% rest of the codebase is compiled with Python’s default compiler and no
+    %% type-checking (it may of course be checked by Pyre but that’s unrelated to
+    %% compilation.) So this clearly opens the door to differing behavior between
+    %% code that is compiled by our compiler and code that isn’t, even when the
+    %% former has some or many values/expressions typed as dynamic. It also makes
+    %% such differences somewhat more acceptable in practice — since adoption of
+    %% an individual module into our compiler is opt-in, we can swallow more
+    %% behavior difference implied solely by that change.
+
   \item
@@
 
+%% TODO carl 2021-12-24 ... very important to introduce with ChkDict ... GG not necessary only nice-to-have
+%% > One area where we definitely violate the gradual guarantee and always will
+%% > is primitive machine types. Unlike Python object types these do not flow to
+%% > dynamic, and dynamic can never flow to them! Effectively they form a
+%% > fully-typed subset of the type system that does not participate in gradual
+%% > typing and can never occur in non-static code. So this violates GG because
+%% > there are many examples where removing an annotation of a primitive machine
+%% > type can cause static code to no longer type check.
+%% > 
+%% > We are ok with this violation of the gradual guarantee because the form of
+%% > the gradual guarantee we really care about is not “removing an arbitrary
+%% > annotation” but rather “an untyped module compiles without changes when
+%% > adopted into the SP compiler” — and an untyped module cannot have
+%% > pre-existing uses of the primitive machine types, since those types cannot
+%% > exist outside SP.
+
 \subsubsection{Concrete Types and Shallow Types}
@@
 Additional ``C types'' enable aggressive compilation of primitive
-values~(\cref{s:ctypes}).
+values~(\cref{s:c-types}).
 Union types are well-supported statically and narrow down in accordance
@@
 but hardly supported at run-time.
-Only binary unions of a type and false get enforced with a run-time check
-(\code{Union[T, False]}, or \code{Optional[t0]}).
+Only binary unions of a type and the none value get enforced with a run-time check
+(\code{Union[T, None]}, or \code{Optional[t0]}).
 
@@
 \subsubsection{Unboxed Primitive}
-\label{s:ctypes}
+\label{s:c-types}
+
+%% 2021-12-29: TODO
+%% Using the term “ctypes” for these might be confusing, since the Python stdlib
+%% has a “ctypes” module that isn’t related to our primitive types. We do call the
+%% base type for them CType in our compiler, but when discussing them we usually
+%% call them “primitive types” or “machine types” — “C types” could also work but
+%% I wouldn’t combine it into one word :)
+%%
+%% I just did a very rough grep check and found that we have only 12 modules
+%% currently using primitive types: all int32 and int64 and cbool. So they are
+%% used pretty sparingly, but they are also definitely critical to our ability to
+%% match Cython in a few of the  hottest code paths that used to be Cython.
 
@@
 %%   message, after all its not our goal to reproduce SP exactly
+
+%% TODO 2021-12-28
+%% - model soundness argument = similar to Nom, Redex tests
+%% - while True break another example of modeling helping out
+%%   it's a second implementation, helpful to find bugs
+%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us
 
@@
 
+%% TODO 2021-12-29: Carl
+%% "full soundness" requires strict modules, means you can trust module toplevel
+%% hmph
+
 \begin{theorem}[Type Soundness]
@@
 %  the last half sum up to 2% of total IG prod CPU usage." --Carl
+
+%% TODO 2021-12-29 Carl
+%% We definitely did find some regressions (especially when converting Cython, but
+%% also one or two in non-Cython conversions.) But we were never satisfied with a
+%% regression, and we were always able to apply the JIT or improve SP
+%% optimizations or adjust the code to turn the regression neutral or positive.
+%% (In fact I think all? the non-Cython regressions I can recall boiled down to
+%% “we forgot to enable the JIT for an important function” and just rectifying
+%% that took care of it.)
+
+%% TODO think about it 2021-12-29 Ben
+%% > + Is there any sense in which the 530 converted modules are a big
+%% > sample? Maybe not, since as you say 530 is tiny for the codebase as a
+%% > whole. But do they exercise all sorts of behavior? Do they form a big
+%% > sample of certain parts of the Instagram codebase?
 
@@
 
-%% - pushback one time about removing MI
+%% - pushback one time about removing MI multiple inheritance
 %% - otherwise, people happy with low-cost improvements
@@
 %% - pyre challenges?
+
+FILL
 
@@
 
-Optimize Python integer operations (in addition to the ctypes).
+Optimize Python integer operations (in addition to the c types).
 Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}

### 2021-12-31 conformance_suite/ht_test_checked_dict.py
hand-translated test_checked_dict tests

@@ -1,0 +1,21 @@
+# ht_test_checked_dict.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, str]()
+x["abc"] = "foo"
+x = CheckedDict[str, int]()
+x["abc"] = 42
+x = CheckedDict[int, str]()
+x[42] = "abc"
+
+# def test_checked_dict(self):
+#     x = chkdict[str, str]()
+#     x["abc"] = "foo"
+#     self.assertEqual(repr(x), "{'abc': 'foo'}")
+#     x = chkdict[str, int]()
+#     x["abc"] = 42
+#     x = chkdict[int, str]()
+#     x[42] = "abc"

### 2021-12-31 conformance_suite/ht_test_checked_dict___module__.py

@@ -1,0 +1,16 @@
+# ht_test_checked_dict___module__.py
+# This should pass.
+# This should terminate.
+
+from __static__ import chkdict
+
+class Lol:
+    pass
+x = chkdict[int, Lol]()
+assert type(x).__module__ is "__static__"
+
+# def test_checked_dict___module__(self):
+#     class Lol:
+#         pass
+#     x = chkdict[int, Lol]()
+#     self.assertEqual(type(x).__module__, "__static__")

### 2021-12-31 conformance_suite/ht_test_checked_dict_bad_ctor.py

@@ -1,0 +1,16 @@
+# test_checked_dict_bad_ctor.py
+# This should pass.
+# This should terminate.
+
+from __static__ import chkdict
+
+try:
+    chkdict[str, str](None)
+except Exception:
+    pass
+else:
+    raise Exception("Should fail.")
+
+# def test_checked_dict_bad_ctor(self):
+#     with self.assertRaises(TypeError):
+#         chkdict[str, str](None)

### 2021-12-31 conformance_suite/ht_test_checked_dict_clear.py

@@ -1,0 +1,13 @@
+# ht_test_checked_dict_clear.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+x = CheckedDict[str, str]({"x": "abc"})
+x.clear()
+assert x == {}
+
+# def test_checked_dict_clear(self):
+#     x = chkdict[str, str](x="abc")
+#     x.clear()
+#     self.assertEqual(x, {})

### 2021-12-31 conformance_suite/ht_test_checked_dict_copy.py

@@ -1,0 +1,14 @@
+# ht_test_checked_dict_copy.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, str]({ "x": "abc" })
+assert type(x) == CheckedDict[str, str]
+assert x == {"x": "abc"}
+
+# def test_checked_dict_copy(self):
+#     x = chkdict[str, str](x="abc")
+#     self.assertEqual(type(x), chkdict[str, str])
+#     self.assertEqual(x, {"x": "abc"})

### 2021-12-31 conformance_suite/ht_test_checked_dict_errors.py

@@ -1,0 +1,28 @@
+# ht_test_checked_dict_errors.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+try:
+    x.get(100)
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+try:
+    x.get("x", "abc")
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+# # Reason: Can't be translated by any of the three translator
+# def test_checked_dict_errors(self):
+#     x = chkdict[str, int](x=2)
+#     with self.assertRaises(TypeError):
+#         x.get(100)
+#     with self.assertRaises(TypeError):
+#         x.get("x", "abc")

### 2021-12-31 conformance_suite/ht_test_checked_dict_fromkeys_bad_types.py

@@ -1,0 +1,35 @@
+# ht_test_checked_dict_fromkeys_bad_types.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+try:
+    CheckedDict[str, int].fromkeys([2], 42)
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+try:
+    CheckedDict[str, int].fromkeys("abc", object())
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+try:
+    CheckedDict[str, int].fromkeys("abc")
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+# def test_checked_dict_fromkeys_bad_types(self):
+#     with self.assertRaises(TypeError):
+#         chkdict[str, int].fromkeys([2], 42)
+#     with self.assertRaises(TypeError):
+#         chkdict[str, int].fromkeys("abc", object())
+#     with self.assertRaises(TypeError):
+#         chkdict[str, int].fromkeys("abc")
+

### 2021-12-31 conformance_suite/ht_test_checked_dict_get.py

@@ -1,0 +1,15 @@
+# ht_test_checked_dict_get.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+assert x.get("x") == 2
+assert x.get("y", 100) == 100
+
+
+# def test_checked_dict_get(self):
+#     x = chkdict[str, int](x=2)
+#     self.assertEqual(x.get("x"), 2)
+#     self.assertEqual(x.get("y", 100), 100)

### 2021-12-31 conformance_suite/ht_test_checked_dict_getitem.py

@@ -1,0 +1,12 @@
+# ht_test_checked_dict_getitem.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+assert x.__getitem__("x") == 2
+
+# def test_checked_dict_getitem(self):
+#     x = chkdict[str, int](x=2)
+#     self.assertEqual(x.__getitem__("x"), 2)

### 2021-12-31 conformance_suite/ht_test_checked_dict_items.py

@@ -1,0 +1,21 @@
+# ht_test_checked_dict_items.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+assert list(x.items()) == [("x", 2)]
+x = CheckedDict[str, int]({"x": 2, "y": 3})
+assert list(x.items()) == [("x", 2), ("y", 3)]
+
+# def test_checked_dict_items(self):
+#     x = chkdict[str, int](x=2)
+#     self.assertEqual(
+#         list(x.items()),
+#         [
+#             ("x", 2),
+#         ],
+#     )
+#     x = chkdict[str, int](x=2, y=3)
+#     self.assertEqual(list(x.items()), [("x", 2), ("y", 3)])

### 2021-12-31 conformance_suite/ht_test_checked_dict_keys.py

@@ -1,0 +1,16 @@
+# ht_test_checked_dict_keys.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({ "x": 2 })
+assert list(x.keys()) == ["x"]
+x = CheckedDict[str, int]({ "x": 2, "y": 3})
+assert list(x.keys()) == ["x", "y"]
+
+# def test_checked_dict_keys(self):
+#     x = chkdict[str, int](x=2)
+#     self.assertEqual(list(x.keys()), ["x"])
+#     x = chkdict[str, int](x=2, y=3)
+#     self.assertEqual(list(x.keys()), ["x", "y"])

### 2021-12-31 conformance_suite/ht_test_checked_dict_nonoptional.py

@@ -1,0 +1,30 @@
+# ht_test_checked_dict_nonoptional.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+from typing import Optional
+
+x = CheckedDict[str, Optional[str]]()
+try:
+    x[None] = "abc"
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+x = CheckedDict[Optional[str], str]()
+try:
+    x["abc"] = None
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+# def test_checked_dict_nonoptional(self):
+#     x = chkdict[str, Optional[str]]()
+#     with self.assertRaises(TypeError):
+#         x[None] = "abc"
+#     x = chkdict[Optional[str], str]()
+#     with self.assertRaises(TypeError):
+#         x["abc"] = None

### 2021-12-31 conformance_suite/ht_test_checked_dict_optional.py

@@ -1,0 +1,17 @@
+# ht_test_checked_dict_optional.py
+# This should pass.
+# This should terminate.
+
+from typing import Optional
+from __static__ import CheckedDict
+
+x = CheckedDict[str, Optional[str]]()
+x["abc"] = None
+x = CheckedDict[Optional[str], str]()
+x[None] = "abc"
+
+# def test_checked_dict_optional(self):
+#     x = chkdict[str, Optional[str]]()
+#     x["abc"] = None
+#     x = chkdict[Optional[str], str]()
+#     x[None] = "abc"

### 2021-12-31 conformance_suite/ht_test_checked_dict_pop.py

@@ -1,0 +1,22 @@
+# ht_test_checked_dict_pop.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+y = x.pop("x")
+assert y == 2
+try:
+    x.pop("z")
+except KeyError:
+    pass
+else:
+    raise Exception()
+
+# def test_checked_dict_pop(self):
+#     x = chkdict[str, int](x=2)
+#     y = x.pop("x")
+#     self.assertEqual(y, 2)
+#     with self.assertRaises(KeyError):
+#         x.pop("z")

### 2021-12-31 conformance_suite/ht_test_checked_dict_popitem.py

@@ -1,0 +1,22 @@
+# ht_test_checked_dict_popitem.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2})
+y = x.popitem()
+assert y == ("x", 2)
+try:
+    x.popitem()
+except KeyError:
+    pass
+else:
+    raise Exception()
+
+# def test_checked_dict_popitem(self):
+#     x = chkdict[str, int](x=2)
+#     y = x.popitem()
+#     self.assertEqual(y, ("x", 2))
+#     with self.assertRaises(KeyError):
+#         x.popitem()

### 2021-12-31 conformance_suite/ht_test_checked_dict_setdefault.py

@@ -1,0 +1,14 @@
+# ht_test_checked_dict_setdefault.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, str]()
+x.setdefault("abc", "foo")
+assert x == {"abc": "foo"}
+
+# def test_checked_dict_setdefault(self):
+#     x = chkdict[str, str]()
+#     x.setdefault("abc", "foo")
+#     self.assertEqual(x, {"abc": "foo"})

### 2021-12-31 conformance_suite/ht_test_checked_dict_setdefault_bad_values.py

@@ -1,0 +1,32 @@
+# ht_test_checked_dict_setdefault_bad_values.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]()
+try:
+    x.setdefault("abc", "abc")
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+assert x == {}
+try:
+    x.setdefault(42, 42)
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+assert x == {}
+
+# def test_checked_dict_setdefault_bad_values(self):
+#     x = chkdict[str, int]()
+#     with self.assertRaises(TypeError):
+#         x.setdefault("abc", "abc")
+#     self.assertEqual(x, {})
+#     with self.assertRaises(TypeError):
+#         x.setdefault(42, 42)
+#     self.assertEqual(x, {})

### 2021-12-31 conformance_suite/ht_test_checked_dict_sizeof.py

@@ -1,0 +1,11 @@
+# ht_test_checked_dict_sizeof.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+x = CheckedDict[str, int]({"x": 2}).__sizeof__()
+assert type(x) == int
+
+# def test_checked_dict_sizeof(self):
+#     x = chkdict[str, int](x=2).__sizeof__()
+#     self.assertEqual(type(x), int)

### 2021-12-31 conformance_suite/ht_test_checked_dict_types_enforced.py

@@ -1,0 +1,55 @@
+# ht_test_checked_dict_types_enforced.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, str]()
+try:
+    x[42] = "abc"
+except TypeError:
+    pass
+else:
+    raise Exception()
+assert x == {}
+
+try:
+    x["abc"] = 42
+except TypeError:
+    pass
+else:
+    raise Exception()
+assert x == {}
+
+x = chkdict[str, int]()
+try:
+    x[42] = 42
+except TypeError:
+    pass
+else:
+    raise Exception()
+assert x == {}
+
+try:
+    x["abc"] = "abc"
+except TypeError:
+    pass
+else:
+    raise Exception()
+assert x == {}
+
+# def test_checked_dict_types_enforced(self):
+#     x = chkdict[str, str]()
+#     with self.assertRaises(TypeError):
+#         x[42] = "abc"
+#     self.assertEqual(x, {})
+#     with self.assertRaises(TypeError):
+#         x["abc"] = 42
+#     self.assertEqual(x, {})
+#     x = chkdict[str, int]()
+#     with self.assertRaises(TypeError):
+#         x[42] = 42
+#     self.assertEqual(x, {})
+#     with self.assertRaises(TypeError):
+#         x["abc"] = "abc"
+#     self.assertEqual(x, {})

### 2021-12-31 conformance_suite/ht_test_checked_dict_update.py

@@ -1,0 +1,18 @@
+# ht_test_checked_dict_update.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, str]({"x": "abc"})
+x.update({"y": "foo"})
+assert x == {"x": "abc", "y": "foo"}
+x.update({"z": "bar"})
+assert x == {"x": "abc", "y": "foo", "z": "bar"}
+
+# def test_checked_dict_update(self):
+#     x = chkdict[str, str](x="abc")
+#     x.update(y="foo")
+#     self.assertEqual(x, {"x": "abc", "y": "foo"})
+#     x.update({"z": "bar"})
+#     self.assertEqual(x, {"x": "abc", "y": "foo", "z": "bar"})

### 2021-12-31 conformance_suite/ht_test_checked_dict_update_bad_type.py

@@ -1,0 +1,42 @@
+# ht_test_checked_dict_update_bad_type.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]()
+try:
+    x.update({"x": "abc"})
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+assert x == {}
+
+try:
+    x.update({"x": "abc"})
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+try:
+    x.update({24: 42})
+except TypeError:
+    pass
+else:
+    raise Exception()
+
+assert x == {}
+
+# def test_checked_dict_update_bad_type(self):
+#     x = chkdict[str, int]()
+#     with self.assertRaises(TypeError):
+#         x.update(x="abc")
+#     self.assertEqual(x, {})
+#     with self.assertRaises(TypeError):
+#         x.update({"x": "abc"})
+#     with self.assertRaises(TypeError):
+#         x.update({24: 42})
+#     self.assertEqual(x, {})

### 2021-12-31 conformance_suite/ht_test_checked_dict_values.py

@@ -1,0 +1,12 @@
+# ht_test_checked_dict_values.py
+# This should pass.
+# This should terminate.
+
+from __static__ import CheckedDict
+
+x = CheckedDict[str, int]({"x": 2, "y": 3})
+assert list(x.values()) == [2, 3]
+
+# def test_checked_dict_values(self):
+#     x = chkdict[str, int](x=2, y=3)
+#     self.assertEqual(list(x.values()), [2, 3])

### 2021-12-31 scripts/import_SP_tests.py

@@ -93,6 +93,9 @@ ban_anywhere_in_test = [
     
     # fancy argment spec
     'test_method_prologue_posonly', 'test_check_args_6', 'test_check_args_7',
+
+    # These tests have been hand-translated.
+    'test_checked_dict'
 ]
 
 skip_anywhere_in_test = [

### 2021-12-31 skipped_tests/test_checked_dict.py

@@ -1,9 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict(self):
-    x = chkdict[str, str]()
-    x["abc"] = "foo"
-    self.assertEqual(repr(x), "{'abc': 'foo'}")
-    x = chkdict[str, int]()
-    x["abc"] = 42
-    x = chkdict[int, str]()
-    x[42] = "abc"

### 2021-12-31 skipped_tests/test_checked_dict_bad_ctor.py

@@ -1,4 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_bad_ctor(self):
-    with self.assertRaises(TypeError):
-        chkdict[str, str](None)

### 2021-12-31 skipped_tests/test_checked_dict_clear.py

@@ -1,5 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_clear(self):
-    x = chkdict[str, str](x="abc")
-    x.clear()
-    self.assertEqual(x, {})

### 2021-12-31 skipped_tests/test_checked_dict_copy.py

@@ -1,5 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_copy(self):
-    x = chkdict[str, str](x="abc")
-    self.assertEqual(type(x), chkdict[str, str])
-    self.assertEqual(x, {"x": "abc"})

### 2021-12-31 skipped_tests/test_checked_dict_errors.py

@@ -1,7 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_errors(self):
-    x = chkdict[str, int](x=2)
-    with self.assertRaises(TypeError):
-        x.get(100)
-    with self.assertRaises(TypeError):
-        x.get("x", "abc")

### 2021-12-31 skipped_tests/test_checked_dict_free_list.py

@@ -1,10 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_free_list(self):
-    t1 = chkdict[str, int]
-    t2 = chkdict[str, str]
-    x = t1()
-    x_id1 = id(x)
-    del x
-    x = t2()
-    x_id2 = id(x)
-    self.assertEqual(x_id1, x_id2)

### 2021-12-31 skipped_tests/test_checked_dict_fromkeys_bad_types.py

@@ -1,8 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_fromkeys_bad_types(self):
-    with self.assertRaises(TypeError):
-        chkdict[str, int].fromkeys([2], 42)
-    with self.assertRaises(TypeError):
-        chkdict[str, int].fromkeys("abc", object())
-    with self.assertRaises(TypeError):
-        chkdict[str, int].fromkeys("abc")

### 2021-12-31 skipped_tests/test_checked_dict_get.py

@@ -1,5 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_get(self):
-    x = chkdict[str, int](x=2)
-    self.assertEqual(x.get("x"), 2)
-    self.assertEqual(x.get("y", 100), 100)

### 2021-12-31 skipped_tests/test_checked_dict_getitem.py

@@ -1,4 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_getitem(self):
-    x = chkdict[str, int](x=2)
-    self.assertEqual(x.__getitem__("x"), 2)

### 2021-12-31 skipped_tests/test_checked_dict_items.py

@@ -1,11 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_items(self):
-    x = chkdict[str, int](x=2)
-    self.assertEqual(
-        list(x.items()),
-        [
-            ("x", 2),
-        ],
-    )
-    x = chkdict[str, int](x=2, y=3)
-    self.assertEqual(list(x.items()), [("x", 2), ("y", 3)])

### 2021-12-31 skipped_tests/test_checked_dict_keys.py

@@ -1,6 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_keys(self):
-    x = chkdict[str, int](x=2)
-    self.assertEqual(list(x.keys()), ["x"])
-    x = chkdict[str, int](x=2, y=3)
-    self.assertEqual(list(x.keys()), ["x", "y"])

### 2021-12-31 skipped_tests/test_checked_dict_nonoptional.py

@@ -1,8 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_nonoptional(self):
-    x = chkdict[str, Optional[str]]()
-    with self.assertRaises(TypeError):
-        x[None] = "abc"
-    x = chkdict[Optional[str], str]()
-    with self.assertRaises(TypeError):
-        x["abc"] = None

### 2021-12-31 skipped_tests/test_checked_dict_optional.py

@@ -1,6 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_optional(self):
-    x = chkdict[str, Optional[str]]()
-    x["abc"] = None
-    x = chkdict[Optional[str], str]()
-    x[None] = "abc"

### 2021-12-31 skipped_tests/test_checked_dict_pop.py

@@ -1,7 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_pop(self):
-    x = chkdict[str, int](x=2)
-    y = x.pop("x")
-    self.assertEqual(y, 2)
-    with self.assertRaises(KeyError):
-        x.pop("z")

### 2021-12-31 skipped_tests/test_checked_dict_popitem.py

@@ -1,7 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_popitem(self):
-    x = chkdict[str, int](x=2)
-    y = x.popitem()
-    self.assertEqual(y, ("x", 2))
-    with self.assertRaises(KeyError):
-        x.popitem()

### 2021-12-31 skipped_tests/test_checked_dict_setdefault.py

@@ -1,5 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_setdefault(self):
-    x = chkdict[str, str]()
-    x.setdefault("abc", "foo")
-    self.assertEqual(x, {"abc": "foo"})

### 2021-12-31 skipped_tests/test_checked_dict_setdefault_bad_values.py

@@ -1,9 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_setdefault_bad_values(self):
-    x = chkdict[str, int]()
-    with self.assertRaises(TypeError):
-        x.setdefault("abc", "abc")
-    self.assertEqual(x, {})
-    with self.assertRaises(TypeError):
-        x.setdefault(42, 42)
-    self.assertEqual(x, {})

### 2021-12-31 skipped_tests/test_checked_dict_types_enforced.py

@@ -1,16 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_types_enforced(self):
-    x = chkdict[str, str]()
-    with self.assertRaises(TypeError):
-        x[42] = "abc"
-    self.assertEqual(x, {})
-    with self.assertRaises(TypeError):
-        x["abc"] = 42
-    self.assertEqual(x, {})
-    x = chkdict[str, int]()
-    with self.assertRaises(TypeError):
-        x[42] = 42
-    self.assertEqual(x, {})
-    with self.assertRaises(TypeError):
-        x["abc"] = "abc"
-    self.assertEqual(x, {})

### 2021-12-31 skipped_tests/test_checked_dict_update.py

@@ -1,7 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_update(self):
-    x = chkdict[str, str](x="abc")
-    x.update(y="foo")
-    self.assertEqual(x, {"x": "abc", "y": "foo"})
-    x.update({"z": "bar"})
-    self.assertEqual(x, {"x": "abc", "y": "foo", "z": "bar"})

### 2021-12-31 skipped_tests/test_checked_dict_update_bad_type.py

@@ -1,11 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_update_bad_type(self):
-    x = chkdict[str, int]()
-    with self.assertRaises(TypeError):
-        x.update(x="abc")
-    self.assertEqual(x, {})
-    with self.assertRaises(TypeError):
-        x.update({"x": "abc"})
-    with self.assertRaises(TypeError):
-        x.update({24: 42})
-    self.assertEqual(x, {})

### 2021-12-31 skipped_tests/test_checked_dict_values.py

@@ -1,4 +1,0 @@
-# Reason: Can't be translated by any of the three translator
-def test_checked_dict_values(self):
-    x = chkdict[str, int](x=2, y=3)
-    self.assertEqual(list(x.values()), [2, 3])

### 2021-12-31 pj-7.1/paper.tex

@@
 \newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
+\newcommand{\ipara}[1]{\emph{#1}\/.}
 
@@
 \subsection{Dynamic Python Features}
+\label{s:dynamic-python}
 
@@
 %% Because of the latter concern, type dynamic is a very reasonable default.
+
+\paragraph{Set attribute}
+
+Carl: Removing the use of super-dynamic Python features like
+\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
+we’ve done this it hasn’t required a major rewrite.
 
@@
 \subsection{Analyzing Code Changes}
-
-The fact that the \SP{} team has been changing code as well as types
-during its migrations is potentially troubling.
+%% TODO boring unclear writing
+
+Because the \SP{} team has been changing code as well as types
+during its migrations, the question arises as to whether the code changes
+raise any threats to validity.
 It could be that the \SP{} approach to gradual typing only pays off
@@
 
-To understand the code changes, an author affiliated with Instagram reviewed
-approximately 30 patches that significantly improved performance.
-In general, the patches contain relatively minor code changes.
-The biggest changes are to testing code rather than production code.
-Both these points are encouranging.
-The following nine points describe common patterns:
-
-\begin{enumerate}
-  \item
-   adjustments to tests to pass correct types instead of mock objects, as as
-   not to fail type checks in SP code. This is an inherent and desired
-   incompatibility; next half we plan to work on improving the Python mock
-   framework to automatically create mocks of the correct types.
-   %% - bg: lookup the dino email about mocks
-
- \item
-   adjustments to mock assertions in tests about calls to adjust for the
-   fact that SP currently converts all call arguments to be positional (we plan
-   to fix this incompatibility by falling back to the original call arguments
-   when we detect the target function has been patched.)
-   %% - bg: ??? bit fuzzy about positional and others
-
- \item
-   adjustments around class attributes vs instance attributes (Python allows
-   this to be a bit fuzzy, where a class attribute can be a fallback default
-   for an instance attribute if not set, but SP currently requires an attribute
-   to be clearly one or the other, and we have more efficient access for
-    instance attributes via \code{LOAD\_FIELD} that we don’t have for class attributes.)
-   %% - related to github #37 ?
-
- \item
-   convert @classmethod that don’t actually need the class to @staticmethod,
-    allowing for emitting direct \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}
-
- \item
-   extract special types that our compiler doesn’t yet support/understand
-   (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
-   the module we are converting to static. We have work in progress on
-   supporting some Enum types, and plan to add a @dataclass intrinsic soon.
-
- \item
-   Removing the use of super-dynamic Python features like \code{\_\_setattr\_\_} that
-   we don’t support. So far in the cases where we’ve done this it hasn’t
-   required a major rewrite.
-   %% - bg: more details pls ,,, sketch the rewrite
-
- \item
-   Removing the use of keyword-only arguments (bare `*`) in function
-   signatures, since we don’t yet support calling such functions. We plan to
-   support this next half.
-
- \item
-   Convert Python integers and bools used in very hot code paths into SP
-   machine primitives.
-   %% - bg: focus on block, convert, no problem?
-
- \item
-   Occasionally converting a regular Python list or dict to a CheckedList or
-   CheckedDict.
-   %% - bg: how occasional? worried about power of the "gradual soundness" claim
-\end{enumerate}
-
-%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
+To understand the code changes, we reviewed approximately 30 patches that
+significantly improved performance.
+In general, the patches have minor code changes and most of these changes
+affect tests rather than production code.
+Digging deeper, there are nine common kinds of changes.
+Six of these address type errors; the other three focus on performance.
+
+%% one significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
 %%  removes an extra call layer
 %%  helps with (timing [ context managers / decorators ])
-%%
-%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
-%%  because the pay / payoff isn't yet attractive
-
-\subsection{Developer Experience}
-
-%% - pushback one time about removing MI multiple inheritance
-%% - otherwise, people happy with low-cost improvements
-
-%% - pyre challenges?
-
-FILL
+
+%% TODO section formatting looks good?
+\subsubsection{Code Changes for the Type Checker}
+
+    %% X bg: lookup the dino email about mocks
+    %% And we do go through great lengths to make monkey patching work.  Because
+    %% we need to enforce return types it ends up being not quite as flexible as
+    %% normal Python.  For example one common pattern is to use Python’s mock
+    %% library and we currently end up throwing lots of TypeError’s when a
+    %% function which returns None is mocked and starts returning a MagicMock
+    %% object instead.  But it’s an easy fix in that you can specify the return
+    %% value to be compatible.  There may be more complex cases where you want to
+    %% return a mock for an actual class.  For example in our code base there’s
+    %% frequent mocking of the HttpRequest object and we have a MockReqeust object
+    %% which is a subclass of HttpRequest so that these can still be mocked
+    %% (luckily this actually pre-dated our static conversion work).
+
+\begin{enumerate}
+  \item
+    \ipara{Fix type errors due to mock wrappers}
+    When test code introduces a mock wrapper, it may change the return type of
+    a function.
+    For example, the wrapped version of a function that returns \code{None}
+    will return a \code{MagicMock} object unless the programmer explicitly
+    specifies a return value.
+    %% Carl: This is an inherent and desired
+    %% incompatibility; next half we plan to work on improving the Python mock
+    %% framework to automatically create mocks of the correct types.
+
+  \item
+    \ipara{Change mocked function to expect positional arguments}
+    \SP{} currently rewrites all function calls to use positional arguments.
+    Mock-wrapped functions therefore need to positional arguments instead of keywords.
+    %% Carl: we plan to fix this incompatibility by falling back to the original call
+    %% arguments when we detect the target function has been patched.
+
+  \item
+    \ipara{Change functions to accept positional arguments}
+    Functions that expect only keyword arguments cannot yet be called from \SP{} contexts.
+    These functions need to change.
+    %% Carl: We plan to support this next half.
+
+  \item
+    \ipara{Organize class and instance attributes}
+    Whereas Python allows class attributes to serve as default values for instance attributes,
+    \SP{} does not.
+    Programmers have to decide where an attribute belongs so that \SP{} can optimize
+    reads from instance attributes.
+    %% Carl: (efficient via \code{LOAD\_FIELD})
+    %% - related to github #37 ?
+
+  \item
+    \ipara{Move \code{@dataclass} and \code{Enum} classes}
+    \SP{} does not support classes decorated with \code{@dataclass}
+    and classes that are enums.
+    Programmers have to move these classes to a different module for now.
+    %% Carl: We have work in progress on supporting some Enum types, and plan to add a @dataclass intrinsic soon.
+
+  \item
+    \ipara{Remove unsupported Python features}
+    Occurrences of the dynamic Python features listed in \cref{s:dynamic-python} must be rewritten or moved.
+    According to the \SP{} team, these changes have not required any major rewrites.
+
+\end{enumerate}
+
+%% TODO section formatting looks good?
+\subsubsection{Code Changes for Performance}
+
+\begin{enumerate}
+  \item
+    \ipara{Convert \code{@classmethod} to \code{@staticmethod}}
+    Static methods can get invoked directly as functions, bypassing the class.
+    %% Carl: \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}
+
+  \item
+    \ipara{Use primitive types}
+    Any Python integers and booleans in hot code paths are better as
+    \SP{} primitive values~(\cref{s:c-types}).
+    %% - bg: focus on block, convert, no problem?
+
+  \item
+    \ipara{Use Checked Types}
+    Occasionally, converting a Python dictionary or Python list to a
+    checked type has a big impact on performance.
+    As noted in \cref{s:checked-type}, these conversions affect all downstream
+    clients of the data structure.
+    %% - bg: how occasional? worried about power of the "gradual soundness" claim
+
+\end{enumerate}
+
+\subsubsection{Developer Experience}
+
+For the most part, developers at Instagram have been happy adopting \SP{}.
+The transition from Pyre to \SP{} has thus far delivered performance benefits
+that outweigh the cost of code changes.
+
+One significant pushback dealt with multiple inheritance.
+The \SP{} team proposed a rewrite from multiple inheritance to single inheritance
+to improve performance and static checks.
+But the code maintainers preferred to keep multiple inheritance.
+Because \SP{} defaults to Python behavior, the maintainers were able
+to keep multiple inheritance alongside \SP{} code.
+
+%% any more?
 
@@
   %% show untyped Python time?
+  %% transpose?
   \begin{tabular}{ll}
@@
 %% 2021-12-15: WHAT ARE THESE?
+%% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
+%% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
+%% - functions have to expect positionals (can't do keyword only)
+%% - @dataclass enum.Enum
+%%
+%% significant selection bias
+%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
+%%  because the pay / payoff isn't yet attractive
 

### 2022-01-04 pj-7.1/paper.tex
pj: 3, quick draft of technical figures
idea is, use a very simple model to show where dyn comes in:

- basic surface lang. with function defs, class defs, and chkdict
- evaluation types show that Union and PyDict become Dyn
- grammar disallows nested classes
- typing judgment shows class override restrictions
- typing judgment shows pydict shallowness, chkdict depth
- subtyping / compatibility = upcast, materialization = downcast

the materialize rules need run-time casts,
we need to show that the casts are simple & full,
and argue that the cast locations are a subset of Nom

soundness argume

@@
 
+\newcommand{\spann}[2]{#1\!:\!#2}
 \newcommand{\bmname}[1]{\textbf{#1}}
 \newcommand{\typefont}[1]{\mathsf{#1}}
+\newcommand{\codefont}[1]{\emph{#1}}
 \newcommand{\paramtype}[2]{#1[#2]}
@@
 \newcommand{\sptint}{\typefont{Int}}
+\newcommand{\sptbool}{\typefont{Bool}}
 \newcommand{\sptfloat}{\typefont{Float}}
@@
 \newcommand{\sptobject}{\typefont{Object}}
-\newcommand{\sptnone}{\typefont{NoneType}}
+\newcommand{\sptnone}{\typefont{None}}
 \newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
-\newcommand{\sptfun}[2]{#1 \rightarrow #2}
-\newcommand{\sptunion}[2]{#1 \cup #2}
 \newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
+\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
 \newcommand{\sptrawpydict}{\typefont{PyDict}}
@@
 \newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
-
-\newcommand{\spexpr}{e}
-\newcommand{\spvalue}{v}
+\newcommand{\sptenv}{\Gamma}
+\newcommand{\sptenvnil}{\cdot}
+\newcommand{\sptenvempty}{\sptenvnil}
+\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
+\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
+\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
+
+\newcommand{\spx}{\code{x}}
+\newcommand{\spf}{\code{f}}
+\newcommand{\spc}{\code{c}}
+\newcommand{\spprog}{\codefont{prog}}
+\newcommand{\spstmt}{\codefont{stmt}}
+\newcommand{\spexpr}{\codefont{expr}}
+\newcommand{\spvalue}{\codefont{val}}
+\newcommand{\vdashsub}[1]{\vdash_{#1}}
+\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
+\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
+\newcommand{\spsubt}{\mathrel{\leq:}}
+\newcommand{\spcompat}{\sqsubseteq}
+\newcommand{\spmatr}{\prec}
+\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
+\newcommand{\spfundef}[4]{\mathrm{def}~#1(#2)\code{ -> }#3: #4}
+\newcommand{\spclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
+\newcommand{\spself}{\code{self}}
+\newcommand{\spobject}{\code{object}}
+\newcommand{\spnone}{\code{none}}
+\newcommand{\spint}{\codefont{int}}
+\newcommand{\spbool}{\codefont{bool}}
+\newcommand{\spfloat}{\codefont{float}}
+\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
+\newcommand{\spchkdict}[1]{\code{chkdict}(#1)}
+\newcommand{\spapp}[2]{#1(#2)}
+\newcommand{\spref}[2]{#1[#2]}
+\newcommand{\spset}[3]{#1[#2] = #3}
 
@@
 \usepackage{amsthm}
+\usepackage{mathpartir}
 \newtheorem{theorem}{Theorem}
@@
 Only binary unions of a type and the none value get enforced with a run-time check
-(\code{Union[T, None]}, or \code{Optional[t0]}).
+($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).
 
@@
 
-\section{Model}
+\section{Model of Core \SP{}}
 \label{s:model}
 %% purpose = types, checks, and optimizations can be formalized and analyzed
+
+%% TODO 2021-12-30 outline
+%% NOTE strict modules not modeled
+%% - what is language
+%% - what properties
+%% - what did in model, how to navigate?
+%% - bug reports
 
@@
 %% - model soundness argument = similar to Nom, Redex tests
+%%   - Nom has dispatch modes .. SP does not, simpler
+%%     - are dyn -> stat overrides allowed and do they affect soundness?
 %% - while True break another example of modeling helping out
@@
 
-\subsection{Surface Language}
+%% TODO model omits primitives because their typing is easy, separate hierarchy, clear boundaries
+%% TODO paper-model focus on boundaries ... where can dyn slip in ... not a full semantics ...
+
+\subsection{Surface Syntax and Types}
 
 \begin{figure}[t]
-  %% ported types from redex model
   \begin{langarray}
+    \spprog & \defeq &
+      \spstmt \langmid \spstmt,~\spprog
+  \\
+    \spstmt & \defeq &
+      \spvardef{\spx}{\sptype}{\spexpr} \langmid
+  \\ & &
+      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
+  \\ & &
+      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr} \ldots}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr} \ldots}
+  \\
+    \spexpr & \defeq &
+      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
+  \\ & &
+      \sppydict{\spx: \spexpr, \ldots} \langmid
+      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
+  \\ & &
+      \spapp{\spx}{\spexpr} \langmid
+      \spref{\spexpr}{\spexpr} \langmid
+      \spset{\spexpr}{\spexpr}{\spexpr}
+  \\
     \sptype & \defeq &
@@
       \sptclass \langmid
-      \sptinstanceof{\sptclass} \langmid
-      \sptfun{\sptype}{\sptype} \langmid
-      \sptunion{\sptclass}{\sptclass}
+      \sptobject \langmid
+      \sptint \langmid
+      \sptbool \langmid
+      \sptfloat \langmid
+  \\ & &
+      \sptpydict{\sptype}{\sptype} \langmid
+      \sptchkdict{\sptype}{\sptype} \langmid
+      \sptunion{\sptype, \ldots}
   \\
-    \sptclass & \defeq &
-      \sptint \langmid
-      \sptfloat \langmid
-      \sptobject \langmid
-      \sptpydict{\sptclass}{\sptclass} \langmid
-      \sptchkdict{\sptclass}{\sptclass} \langmid
-      \mbox{(user defined class)}
+    \sptenv & \defeq &
+      \sptenvnil \langmid
+      \sptvardef{\spx}{\sptype},~\sptenv \langmid
+      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
+      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype} \ldots}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype} \ldots},~\sptenv
+  \\
+    \spx, \spf, \spc & \defeq & \mbox{variable names}
   \end{langarray}
 
-  \begin{center}\parbox{0.8\columnwidth}{
-    Abbreviation: $\sptoptional{\sptclass} \defeq \sptunion{\sptclass}{\sptnone}$
-
-    EXCEPT THAT $\sptoptional{\sptclass}$ may be a generic parameter just like any other class $\sptclass$
-
-    Function types cannot expect or return other function types.
-  }\end{center}
-
-  \caption{Surface Types}
+%%  \begin{center}\parbox{0.8\columnwidth}{
+  \bigskip
+    \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}
+%%  }\end{center}
+
+  \caption{Surface Syntax and Types}
   \label{f:surface-types}
@@
 
-\subsection{Evaluation Language, Optimizations}
+%% TODO assume well-formed / normal-form
+%% - no circular field defs (x:t = x)
+%% - unions are tidy ... Opt[Dyn] = Dyn
+%% kk
+
+%% TODO surface typing judgment omitted
+%% - straightforward
+%% - eval typing far more interesting
+%% - for now, main goal is to narrow unions ... that's kinda boring / not worth effort reading & writing
+
+\subsection{Evaluation Types, Materialization, Casts}
+
+\begin{figure}[t]
+  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
+  \begin{mathpar}
+    \inferrule*{
+      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
+      \\\\
+      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
+    }{
+      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
+    }
+  \end{mathpar}
+
+  \fbox{$\sptenv \wtprog \spstmt \dashv \sptenv$}
+  \begin{mathpar}
+    \inferrule*{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+    }{
+      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
+    }
+
+    \inferrule*{
+      \sptenv_1 = \sptfundef{\spf_0}{\sptype_0}{\sptype_1},~\sptenv_0
+      \\\\
+      \sptenv_1 \wtexpr \spexpr_0 : \sptype_1
+    }{
+      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\sptype_0}}{\sptype_1}{\spexpr_0} \dashv \sptenv_1
+    }
+
+    \inferrule*{
+      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0} \ldots}{\sptfundef{\spf_n}{\spc_0, \sptype_{dn}}{\sptype_{cn}}}
+      \\\\
+      %% TODO
+      \emph{lookup parent}
+      \\\\
+      \emph{overrides are subtypes (not merely compatible)}
+      \\\\
+      \emph{fields and methods are well typed}
+    }{
+      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0} \ldots}{\spfundef{\spf_n}{\spself, \spann{\spx_n}{\sptype_{dn}}}{\sptype_{cn}}{\spexpr_n}} \dashv \sptenv_1
+    }
+  \end{mathpar}
+
+  \fbox{$\sptenv \wtexpr \spexpr : \sptype$} (selected rules)
+  \begin{mathpar}
+    \inferrule*{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
+      \\
+      \sptype_1 \spcompat \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+    }
+
+    \inferrule*{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
+      \\
+      \sptype_1 \spmatr \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+    }
+
+    \inferrule*{
+      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
+    }
+
+    \inferrule*{
+      \sptfundef{\spf_0}{\sptype_0}{\sptype_1} \in \sptenv_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptype_1
+    }
+
+    \inferrule*{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+      \\
+      \sptenv_0 \wtexpr \spexpr_1 : \sptype_1
+    }{
+      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
+    }
+
+    \inferrule*{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
+      \\
+      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptype_1
+    }
+
+  \end{mathpar}
+
+  \caption{Evaluation Typing Judgment}
+  \label{f:eval-typing}
+\end{figure}
 
@@
     \\
+      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
+    \\
       \sptype_0 & \mbox{otherwise}
@@
 
-  \caption{Surface Types to Evaluation Types}
+  \fbox{$\sptype \spsubt \sptype$}
+  \begin{mathpar}
+    \inferrule*{
+    }{
+      \sptype_0 \spsubt \sptobject
+    }
+
+    \inferrule*{
+    }{
+      \sptbool \spsubt \sptint
+    }
+
+    \inferrule*{
+      \sptype_0 \spsubt \sptype_1
+    }{
+      \sptype_0 \spsubt \sptoptional{\sptype_1}
+    }
+
+    \inferrule*{
+    }{
+      \sptype_0 \spsubt \sptype_0
+    }
+
+    \inferrule*{
+      \sptype_0 \spsubt \sptype_1
+      \\
+      \sptype_1 \spsubt \sptype_2
+    }{
+      \sptype_0 \spsubt \sptype_2
+    }
+  \end{mathpar}
+
+  \fbox{$\sptype \spcompat \sptype$}
+  \begin{mathpar}
+    \inferrule*{
+      \sptype_0 \spsubt \sptype_1
+    }{
+      \sptype_0 \spcompat \sptype_1
+    }
+
+    \inferrule*{
+    }{
+      \sptype_0 \spcompat \sptdyn
+    }
+  \end{mathpar}
+
+  \fbox{$\sptype \spmatr \sptype$}
+  \begin{mathpar}
+    \inferrule*{
+    }{
+      \sptdyn \spmatr \sptype_0
+    }
+  \end{mathpar}
+
+  \caption{TBA Surface Types to Evaluation Types}
   \label{f:surface-to-eval-types}
@@
 
+\subsubsection{Informal Connection to Nom}
+
+\SP{} types restrict the use of Nom,
+thus model is kind of a restriction of Nom.
+
+So since Nom is sound for any sort of dyn mixing,
+and SP allows fewer mixings,
+and both Nom and SP do similar checks at boundaries,
+can begin to see how SP could be sound.
+
+Or perhaps mention this at the start.
+SP similar to Nom, similar checks, keep that in mind.
+
+\subsection{Redex Development, Navigation, Integration, Conformance Tests}
+
+FILL
+
 \subsection{Bug Reports}
@@
 \label{s:impl}
+
+%% TODO unions tracked statically in SP, erased to dyn unless narrowed
 
@@
 to add soundness?
+
+\subsection{Python Interaction}
+
+FILL
+
+extend model with Python ... two languages ... boundaries ... argue for soundness
 
@@
 %% model: metaclasses
+
+%% 2021-12-30 Whence Nominal GT?
+%%
+%% KC: I think in general adding gradual typing to Java/C# (an also Python) is boring:
+%% - MonNom allows you to cast between an arbitrary type and dyn.
+%% - dyn is effectively Object
+%% - Upcasts, casts from an arbitrary type to Object, are already in the language
+%% - Downcasts, casts from Object to an arbitrary type is also possible, but must be explicit in the fully static language
+%% So MonNom effectively gives programmers another way to name Object, with which downcasts can be automatically inserted by the compiler.
+%% I think this sentence covers all the interesting stuff about MonNom.
+%%
+%% SK: This matches my intuition that gradual nominal is just not very
+%% interesting. KC puts this on a technical footing with his very astute comment,
+%% which to me can be summarized as "every nominal language that has Object is
+%% already gradual" [leaving aside the boring non-Object base type distinction].
+       So the entire remaining game is one of making the programming interface
+%% as painless as possible. Some of that is just a trivial matter of
+%% syntactic defaults; some of it is making the type-checker more forgiving
+%% and replacing static checks with dynamic checks. There doesn't seem to
+%% be much else left.
 

### 2022-01-04 pj-7.1/paper.tex
pj: sec 3, table describing run-time checks
purpose is to show that checks are simple and complete

@@
 \newcommand{\sptype}{\typefont{T}}
+\newcommand{\spteval}{\typefont{S}}
 \newcommand{\sptclass}{\typefont{C}}
@@
 \newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
+\newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
 \newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}
@@
 %% purpose = types, checks, and optimizations can be formalized and analyzed
+
+%% 2022-01-04 argument = nom soundness -> sp model sound -> sp sound via conformance equivalence
+%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)
+
+%% TODO 2022-01-04 (somewhere) explain that a verified compiler is an entirely separate project,
+%%  need bytecode model, full transformations, memory model / machine etc.
+%%  hence we don't model optimizations, little benefit
+%%  opt. don't tell us about the surface language ... the particular SP opts are changing representations but not results
 
@@
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
+      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
     }{
@@
       \\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptype_1
+      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
     }{
@@
 \begin{figure}[t]
+  \begin{langarray}
+    \spvalue & \defeq &
+      \mbox{heap-allocated Python values}
+    \\
+    \spteval & \defeq &
+      \sptdyn \langmid
+      \sptnone \langmid
+      \sptclass \langmid
+      \sptobject \langmid
+      \sptint \langmid
+      \sptbool \langmid
+      \sptfloat \langmid
+    \\ & &
+      \sptrawpydict \langmid
+      \sptchkdict{\sptype}{\sptype} \langmid
+      \sptoptional{\sptype}
+  \end{langarray}
+
+  \bigskip
   \(
@@
 
-  \caption{TBA Surface Types to Evaluation Types}
+  \caption{TBA Type Metafunctions}
   \label{f:surface-to-eval-types}
 \end{figure}
+
+\begin{table}
+  \caption{TBD Run-time checks}
+  \label{t:cast}
+
+  \(\begin{array}{l@{\quad}l}
+    \mbox{Eval. type $\spteval$} & \mbox{Description of run-time check} \\\hline
+    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
+    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
+    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
+    \sptdyn & \mbox{No check needed, accepts any value} \\
+    \sptobject & \mbox{Accepts any non-primitive value} \\  %% except primitives
+    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
+    \sptnone & \mbox{Accepts the Python none value} \\
+    \sptint & \mbox{Accepts integers} \\
+    \sptbool & \mbox{Accepts booleans (\code{0} and \code{1})} \\
+    \sptfloat & \mbox{Accepts floats}
+  \end{array}\)
+\end{table}
 
@@
 %% 2021-12-15: C&D may have an SP off/on switch soon to measure totals
+
+%% 2022-01-04 TODO ask C&D how many teams have adopted SP ... get sense of adoption
 

### 2022-01-05 pj-7.1/paper.tex
pj: sec 3, more model + start describing

@@
 \newcommand{\mfeq}{=}
-\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
+\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
 \newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
@@
 
+\newcommand{\spapp}[2]{#1\,(#2)}
 \newcommand{\spann}[2]{#1\!:\!#2}
@@
 \newcommand{\sptvardef}[2]{\spann{#1}{#2}}
-\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
-\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
+\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
+\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
 
@@
 \newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
-\newcommand{\spsubt}{\mathrel{\leq:}}
+\newcommand{\spsubt}{\mathrel{<:}}
+\newcommand{\spsubteq}{\mathrel{\leq:}}
 \newcommand{\spcompat}{\sqsubseteq}
@@
 \newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
-\newcommand{\spfundef}[4]{\mathrm{def}~#1(#2)\code{ -> }#3: #4}
-\newcommand{\spclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
+\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
+\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
 \newcommand{\spself}{\code{self}}
@@
 \newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
-\newcommand{\spchkdict}[1]{\code{chkdict}(#1)}
-\newcommand{\spapp}[2]{#1(#2)}
-\newcommand{\spref}[2]{#1[#2]}
-\newcommand{\spset}[3]{#1[#2] = #3}
+\newcommand{\spchkdict}[1]{\spapp{\code{chkdict}}{#1}}
+\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
+\newcommand{\spdictref}[2]{#1[#2]}
+\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
+\newcommand{\spobjref}[2]{#1.#2}
+\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
+\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}
 
@@
 %% - shadow frames = JIT feature, lite array to track Python call stack
-%% - strict modules ... avoids toplevel function wrapper
 
@@
 
+%% TODO move prim.types up?
+
 For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
@@
 
-\section{Model of Core \SP{}}
+\subsubsection{Strict Modules}
+\label{s:strict-mod}
+
+Python modules are mutable collections of names.
+Names can be added, deleted, and updated at runtime.
+\SP{} accommodates these behaviors in the same manner as standard Python, which
+necessarily limits the benefits of module types.
+As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
+Strict modules are immutable, and thus enable type-based compiler optimizations.
+
+\section{Formal Model}
 \label{s:model}
@@
 
-%% TODO 2021-12-30 outline
-%% NOTE strict modules not modeled
-%% - what is language
-%% - what properties
-%% - what did in model, how to navigate?
-%% - bug reports
-
 %% NOTE
@@
 %% TODO 2021-12-28
-%% - model soundness argument = similar to Nom, Redex tests
-%%   - Nom has dispatch modes .. SP does not, simpler
-%%     - are dyn -> stat overrides allowed and do they affect soundness?
 %% - while True break another example of modeling helping out
 %%   it's a second implementation, helpful to find bugs
-%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us
-
-\subsection{Goals and Limitations}
-
-Our model of \SP{} includes the key aspects of its types,
+
+To validate the design of \SP{}, we have developed a mechanized model using PLT
+Redex~\cite{RYR}.
+The model encompasses two soundness claims:
+%% addresses two soundness concerns?
+\begin{enumerate}
+  \item
+    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
+    with untyped code do not undermine the validity of types.
+  \item
+    The typed half of \SP{} is a sound type system for Python.
+\end{enumerate}
+This section explores the first claim in depth by way of a tiny formal model
+that illustrates the \SP{} gradual interaction strategy.
+The second claim is justified by our Redex development;
+in particular, by its test suite.
+This section describes the Redex development in broad strokes.
+
+FILL pick names to avoid confusion between the Redex model and the tiny paper model?
+
+\subsection{Goals and Non-Goals}
+
+Our Redex model of \SP{} includes the key aspects of its types,
 statements, and values.
 The model has two main goals:
-illustrate the boundaries where typed and untyped code can mix,
+illustrate the boundaries where typed and dynamic code can mix,
 and show how to soundly protect these boundaries.
 Compared to other gradual languages, these boundaries are quite restrictive---which
-partly explains \SP{} performance~(\cref{s:eval}).
-Indeed, the model is nearly a subset of the model for Nom~\cite{mt-oopsla-2021},
-hence we focus on explaining the gaps and omit formal proofs.
+partly explains how it is able to perform well in practice~(\cref{s:eval}).
+Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021},
+which is a compelling argument for their soundness.
 
@@
 Some aspects of \SP{} are left out because they are straightforward to
-handle soundly (FILL examples ... while loop?).
-Omitting these lets us focus on the difficult corners.
+handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
+Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
+are also omitted.
 Other Python features are left out because their \SP{} semantics is identical
-to Python.
-These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
-\SP{} does not rely on types for these features to guide program transformations;
+to the baseline Python semantics.
+These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
 see \cref{s:impl} for further discussion.
 
-%% TODO model omits primitives because their typing is easy, separate hierarchy, clear boundaries
-%% TODO paper-model focus on boundaries ... where can dyn slip in ... not a full semantics ...
+Lastly, the model does not include type-directed optimizations.
+Validating the correctness of optimizations is a separate task that requires
+significant components; namely, a model of the \SP{} bytecode and a faithful
+rendering of its type-based transformations.
+Our model establishes soundness as a baseline.
+Future work can use soundness to reason informally about optimizations,
+or even take it as a starting point for compiler verification.
 
@@
   \\ & &
-      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr} \ldots}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr} \ldots}
+      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
   \\
@@
       \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
-  \\ & &
       \sppydict{\spx: \spexpr, \ldots} \langmid
@@
       \spapp{\spx}{\spexpr} \langmid
-      \spref{\spexpr}{\spexpr} \langmid
-      \spset{\spexpr}{\spexpr}{\spexpr}
+      \spdictref{\spexpr}{\spexpr} \langmid
+      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
+      \spobjref{\spc}{\spx} \langmid
+      \spobjset{\spc}{\spx}{\spexpr} \langmid
+      \spobjapp{\spc}{\spf}{\spexpr}
   \\
@@
       \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
-      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype} \ldots}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype} \ldots},~\sptenv
+      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
   \\
@@
 %% - for now, main goal is to narrow unions ... that's kinda boring / not worth effort reading & writing
+
+The \SP{} surface syntax is the same as Python 3.8.
+Programmers may write type annotations using the PEP 484 syntax;
+unannotated code gets the dynamic type by default.
+The available static types provide a nominal type system with first-order functions.
+These surface types are meant to be descriptive.
+By design, they are not fully sound.
+
+\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
+using a tiny grammar.
+A program is a sequence of statements;
+a statement defines a variable, function, or class.
+These definitions may only appear on the top level
+and they all require full type annotations.
+Functions must have one positional argument.
+Classes must declare one parent (perhaps $\spobject$), one field, and one method.
+Expressions describe values and simple computations.
+The basic values consist of the following: the none value, the top object, integers,
+booleans (technically the integers \code{0} and \code{1}), and floating point numbers.
+There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
+The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
+and method calls.
+Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
+user-defined class, and union types.
+There is no function type; the only way to define a higher-order function is to use the dynamic type.
+
+Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
+many details of Python including imports, conditionals, and exception handlers.
+These details are crucial in the context of the Redex model, which tests whether \SP{}
+soundly approximates Python.
+They are less important here, where our focus is on type boundaries.
+None of the omitted features give substantially new ways for typed code to interact with dynamic code.
 
@@
     }
-  \end{mathpar}
-
-  \fbox{$\sptenv \wtprog \spstmt \dashv \sptenv$}
-  \begin{mathpar}
+
     \inferrule*{
@@
     \inferrule*{
-      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0} \ldots}{\sptfundef{\spf_n}{\spc_0, \sptype_{dn}}{\sptype_{cn}}}
+      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0}}{\sptfundef{\spf_1}{\spc_0, \sptype_1}{\sptype_2}}, \sptenv_0
       \\\\
-      %% TODO
-      \emph{lookup parent}
+      %% TODO are field overrides exact?
+      \mbox{if $\spann{\spx_0}{\sptype_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_0 = \sptype_0'$}
       \\\\
-      \emph{overrides are subtypes (not merely compatible)}
+      %% TODO can T override dyn?
+      \mbox{if $\sptfundef{\spf_1}{\spc_2, \sptype_1'}{\sptype_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_1' \spsubteq \sptype_1$ and $\sptype_2 \spsubteq \sptype_2'$}
       \\\\
-      \emph{fields and methods are well typed}
+      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \sptype_0
+      \\
+      \spann{\spself}{\spc_0}, \spann{\spx_1}{\sptype_1}, \sptenv_1 \wtexpr \spexpr_1 : \sptype_2
     }{
-      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0} \ldots}{\spfundef{\spf_n}{\spself, \spann{\spx_n}{\sptype_{dn}}}{\sptype_{cn}}{\spexpr_n}} \dashv \sptenv_1
+      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\sptype_1}}{\sptype_2}{\spexpr_1}} \dashv \sptenv_1
     }
@@
 
-  \fbox{$\sptenv \wtexpr \spexpr : \sptype$} (selected rules)
+  \fbox{$\sptenv \wtexpr \spexpr : \sptype$}~selected rules
   \begin{mathpar}
@@
       \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
-      \\
+      \\\\
       \sptenv_0 \wtexpr \spexpr_1 : \sptobject
     }{
-      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
+      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
     }
@@
       \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
-      \\
+      \\\\
       \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
     }{
-      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptype_1
+      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptnone
+    }
+
+    \inferrule*{
+      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
+    }{
+      \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \sptype_0
+    }
+
+    \inferrule*{
+      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
+    }
+
+    \inferrule*{
+      \sptfundef{\spf_0}{\spc_0, \sptype_0}{\sptype_1} \in \spenvapp{\sptenv_0}{\spc_0}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+    }{
+      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \sptype_1
     }
@@
 
-  \fbox{$\sptype \spsubt \sptype$}
+  \fbox{$\sptype \spsubteq \sptype$} reflexive, transitive closure of $\spsubt$, defined below:
   \begin{mathpar}
@@
 
-    \inferrule*{
-    }{
-      \sptype_0 \spsubt \sptype_0
-    }
-
-    \inferrule*{
-      \sptype_0 \spsubt \sptype_1
-      \\
-      \sptype_1 \spsubt \sptype_2
-    }{
-      \sptype_0 \spsubt \sptype_2
-    }
+%    \inferrule*{
+%    }{
+%      \sptype_0 \spsubt \sptype_0
+%    }
+%
+%    \inferrule*{
+%      \sptype_0 \spsubt \sptype_1
+%      \\
+%      \sptype_1 \spsubt \sptype_2
+%    }{
+%      \sptype_0 \spsubt \sptype_2
+%    }
   \end{mathpar}
@@
   \(\begin{array}{l@{\quad}l}
-    \mbox{Eval. type $\spteval$} & \mbox{Description of run-time check} \\\hline
+    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
     \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
@@
   %% most filed by KC, a few by Ben
+  %% + Carl https://github.com/facebookincubator/cinder/issues/64
   \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
@@
 %% 2021-12-15: C&D may have an SP off/on switch soon to measure totals
+
+%% 2021-01-04 TODO have no evidence that "any module" typing works out ... team
+%% tried hard to get network effects ... "We explicitly aimed for converting
+%% related modules like this to improve network effects and get better
+%% performance."
 

### 2022-01-06 pj-7.1/paper.tex
pj: sec 3 soundness argument

@@
 %% - choose words: static vs typed, nonstatic vs dynamic vs untyped
+%% - use ChkDict not ChkList ... avoid confusion
+%% - is \sqle _consistent_ or _compatible_ subtyping?
 
@@
 \newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
-\newcommand{\sptenv}{\Gamma}
+\newcommand{\sptenv}{\typefont{Env}}
 \newcommand{\sptenvnil}{\cdot}
@@
 \newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
+\newcommand{\trule}[1]{\textsc{#1}}
 
@@
 \newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
-\newcommand{\spchkdict}[1]{\spapp{\code{chkdict}}{#1}}
+\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
 \newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
@@
 
+%% 2022-01-05 Carl:
+%% "All the data I’ve been sharing is from a single codebase, the primary IG web
+%% server monolith. The existence of JITted but not SP modules is mostly just due
+%% to the fact that opting a function into the JIT is semantically transparent and
+%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
+%% quite easy to do (the only tradeoff pushing us away from JITting everything is
+%% just how much CPU we spend JIT-compiling and how much memory we spend storing
+%% JIT-compiled code), whereas opting a module into SP requires more work so the
+%% bottleneck is available developer time to do conversions."
+
 % {Language Pipeline}
@@
 To validate the design of \SP{}, we have developed a mechanized model using PLT
-Redex~\cite{RYR}.
+Redex~\cite{kcdeffmrtf-popl-2012}.
 The model encompasses two soundness claims:
@@
 This section describes the Redex development in broad strokes.
-
-FILL pick names to avoid confusion between the Redex model and the tiny paper model?
 
@@
 \subsection{Surface Syntax and Types}
+%% TODO assume well-formed; no circular defs (x:t = x)
 
@@
     \spexpr & \defeq &
-      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
+      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
       \sppydict{\spx: \spexpr, \ldots} \langmid
-      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
+      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
   \\ & &
@@
       \sptnone \langmid
-      \sptclass \langmid
-      \sptobject \langmid
       \sptint \langmid
@@
       \sptfloat \langmid
+      \sptobject \langmid
+      \sptclass \langmid
   \\ & &
@@
 
-%%  \begin{center}\parbox{0.8\columnwidth}{
   \bigskip
   \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}
-%%  }\end{center}
 
@@
 \end{figure}
-
-%% TODO assume well-formed / normal-form
-%% - no circular field defs (x:t = x)
-%% - unions are tidy ... Opt[Dyn] = Dyn
-%% kk
-
-%% TODO surface typing judgment omitted
-%% - straightforward
-%% - eval typing far more interesting
-%% - for now, main goal is to narrow unions ... that's kinda boring / not worth effort reading & writing
 
@@
 By design, they are not fully sound.
+For this reason, we omit the surface typing judgment and merely present their
+declarative syntax.
+%% TODO awkward
 
@@
 Expressions describe values and simple computations.
-The basic values consist of the following: the none value, the top object, integers,
-booleans (technically the integers \code{0} and \code{1}), and floating point numbers.
+The basic values consist of the following: the none value, integers,
+booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
@@
 user-defined class, and union types.
+We assume that all unions are written a concise normal form, e.g., that
+$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
 There is no function type; the only way to define a higher-order function is to use the dynamic type.
 
+%% FILL generally, no way to express higher-order cast ... either trivial C=C or impossible C=D
+
 Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
-many details of Python including imports, conditionals, and exception handlers.
+many details of Python including class variables, imports, conditionals, and exception handlers.
 These details are crucial in the context of the Redex model, which tests whether \SP{}
@@
 
-\subsection{Evaluation Types, Materialization, Casts}
+\subsection{Evaluation Types, Casts, and Typing Judgment}
 
 \begin{figure}[t]
-  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
+  \begin{langarray}
+    \spteval & \defeq &
+      \sptdyn \langmid
+      \sptnone \langmid
+      \sptint \langmid
+      \sptbool \langmid
+      \sptfloat \langmid
+      \sptobject \langmid
+      \sptclass \langmid
+    \\ & &
+      \sptrawpydict \langmid
+      \sptchkdict{\spteval}{\spteval} \langmid
+      \sptoptional{\spteval}
+  \end{langarray}
+
+  \bigskip
+  \(
+    \mftypeF{\sptype_0}
+    \mfeq
+    \left\{\begin{array}{ll}
+      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
+    \\
+      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
+    \\
+      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
+    \\
+      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
+    \\
+      \sptype_0 & \mbox{otherwise}
+    \end{array}\right.
+  \)
+
+  \caption{Evaluation Types, Surface-to-Evaluation Mapping}
+  \label{f:surface-to-eval-types}
+\end{figure}
+
+Evaluation type $\spteval$ are a subset of the surface types.
+These are the types that \SP{} promises to soundly enforce.
+\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
+and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
+As the retraction makes clear, the evaluation types make two main changes:
+\begin{enumerate}
+  \item
+    The parameterized type for Python dictionaries
+    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
+    a raw type $\sptrawpydict$.
+    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
+  \item
+    Unions get replaced with the dynamic type, except for the special case
+    of unions with none ($\sptoptional{\sptype_0}$).
+\end{enumerate}
+
+The \SP{} team chose these particular evaluation types because they can all
+be enforced fully and efficiently with run-time casts.
+
+\subsubsection{Casts}
+
+\Cref{t:cast} describes the casts in detail.
+Most call for a simple tag check, i.e., a Python \code{isinstance} test.
+The sole exception is optional types, which require a tag check and a test for the none value.
+Even checked dictionaries rely on tag checks.
+A checked dict type with keys $\sptype_0$ and values $\sptype_1$ accepts only
+checked dict values that were initialized with exactly the same key and value type.
+As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\sptype_0$
+or $\sptype_1$ is the dynamic type.
+
+There are two key observations to make about casts.
+First, all allowed casts can run in constant time.
+No cast requires traversing a data structure.
+Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
+This latter property means that blame-tracking is trivial;
+casts succeed or fail immediately.
+
+Second, the strict cast for checked dictionaries makes type compatibility simple in \SP{}.
+Values that are partly-dynamic are incompatible with static types.
+For example, suppose that a function expects a list of numbers.
+In a typical gradual language, a list of dynamic values would be a valid input to the function.
+In \SP{}, a list of dynamic values is not a valid input.
+Consequently, \SP{} can use a simple ahead-of-time strategy to enforce soundness.
+Any position where the dynamic type meets a static type needs a cast.
+Other meetings between types can be resolved statically.
+
+\begin{table}[t]
+  \caption{How to Enforce the Evaluation Types}
+  \label{t:cast}
+
+  \(\begin{array}{l@{\quad}l}
+    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
+    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
+    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
+    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
+    %% \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
+    \sptdyn & \mbox{No check needed, accepts any value} \\
+    \sptobject & \mbox{Accepts any value (except primitives,~\cref{s:c-types})} \\
+    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
+    \sptnone & \mbox{Accepts the Python none value} \\
+    \sptint & \mbox{Accepts integers} \\
+    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
+    \sptfloat & \mbox{Accepts floats}
+  \end{array}\)
+\end{table}
+
+\subsubsection{Expression Typing, Cast Insertion}
+
+To better show where \SP{} needs to insert casts, we present a selection of typing rules.
+Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
+These statements come with types and fill an environment ($\sptenv$).
+Relative to the current environment, the expressions within each statement must
+satisfy the judgment in~\cref{f:eval-types}
+
+The first two rules are for function application.
+A typed function may be applied an argument that matches its domain type,
+in which case it computes a value that matches its codomain type.
+A dynamically-typed variable may be applied to any input (that matches the top
+type $\sptobject$) to yield a dynamically-typed result.
+The next rules illustrate reads from Python dictionaries and checked dictionaries.
+A Python dict may be deferenced by any value; the read promises a dynamically-typed result.
+By contrast, a checked dict requires an index value that matches its type
+and promises a typed result.
+Next we have two rules for classes:
+\trule{F-Set} says that writes to a class field must match the declared field
+type; and \trule{M-App} shows that typed methods impose similar constraints as
+typed fuctions.
+
+The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
+for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
+Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
+related by static subtyping ($\spteval_0 \spsubteq \spteval_1$) or if
+$\spteval_1$ is the dynamic type.
+When this relation holds, it is safe to upcast a value from type $\spteval_0$
+to type $\spteval_1$.
+Materialization relates the dynamic type to any non-dynamic type.
+Occurrences of the \trule{Matr} rule are downcasts that require a
+run-time check.
+
+The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
+work~\cite{clps-popl-2019}.
+By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
+Combined with the typing rule, however, materialization is a concise way to find
+where a well-typed program needs to insert casts to ensure soundness.
+For example, suppose that $\spf$ is a function from integers to the dynamic
+type and that $\spx$ is a variable with the dynamic type.
+An application $\spapp{\spf}{\spx}$ can satisfy type $\sptfloat$ using two materializations:
+
+\medskip
 \begin{mathpar}
+  \inferrule*[left=Matr]{
     \inferrule*{
-      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
+      \inferrule*[left=Matr]{
+        \sptenv_0 \wtexpr \spx : \sptdyn
+        \\
+        \sptdyn \spmatr \sptint
+      }{
+        \sptenv_0 \wtexpr \spx : \sptint
+      }
+    }{
+      \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptdyn
+    }
+    \quad
+    \sptdyn \spmatr \sptfloat
+  }{
+    \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptfloat
+  }
+
+  \mbox{Where $\sptenv_0 = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
+\end{mathpar}
+\medskip
+Consequently, this type derivation calls for two casts at run-time.
+
+We end here, with materialization, rather than present a semantics
+for the formalization.
+After all, the main benefit of a full formal semantics is to validate
+the behavior of complex expressions---a job
+better left to the mechanized Redex model.
+
+Furthermore, our casts-via-materialization rule is more aggressive
+than what \SP{} actually does to insert casts because
+\SP{} supports interactions with plain Python code.
+Because it chooses not to recompile Python modules, \SP{} inserts
+casts in an overapproximate way:
+\begin{itemize}
+  \item
+    typed functions and methods begin by checking all their arguments;
+  \item
+    typed classes check all field writes;
+  \item
+    checked dictionaries check all writes; and
+  \item
+    typed code checks the results of function calls, references, etc. whenever there is a materialization from the dynamic type.
+\end{itemize}
+\Cref{s:impl} explains how the type-directed optimizer avoids some
+of these checks.
+
+\begin{figure}[t]
+  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
+  \begin{mathpar}
+    \inferrule*[left=S-App]{
+      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
       \\\\
-      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
+      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
     }{
-      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \spteval_1
     }
 
-    \inferrule*{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
-    }{
-      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
-    }
-
-    \inferrule*{
-      \sptenv_1 = \sptfundef{\spf_0}{\sptype_0}{\sptype_1},~\sptenv_0
-      \\\\
-      \sptenv_1 \wtexpr \spexpr_0 : \sptype_1
-    }{
-      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\sptype_0}}{\sptype_1}{\spexpr_0} \dashv \sptenv_1
-    }
-
-    \inferrule*{
-      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0}}{\sptfundef{\spf_1}{\spc_0, \sptype_1}{\sptype_2}}, \sptenv_0
-      \\\\
-      %% TODO are field overrides exact?
-      \mbox{if $\spann{\spx_0}{\sptype_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_0 = \sptype_0'$}
-      \\\\
-      %% TODO can T override dyn?
-      \mbox{if $\sptfundef{\spf_1}{\spc_2, \sptype_1'}{\sptype_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_1' \spsubteq \sptype_1$ and $\sptype_2 \spsubteq \sptype_2'$}
-      \\\\
-      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \sptype_0
-      \\
-      \spann{\spself}{\spc_0}, \spann{\spx_1}{\sptype_1}, \sptenv_1 \wtexpr \spexpr_1 : \sptype_2
-    }{
-      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\sptype_1}}{\sptype_2}{\spexpr_1}} \dashv \sptenv_1
-    }
-  \end{mathpar}
-
-  \fbox{$\sptenv \wtexpr \spexpr : \sptype$}~selected rules
-  \begin{mathpar}
-    \inferrule*{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
-      \\
-      \sptype_1 \spcompat \sptype_0
-    }{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
-    }
-
-    \inferrule*{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
-      \\
-      \sptype_1 \spmatr \sptype_0
-    }{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
-    }
-
-    \inferrule*{
+    \inferrule*[left=D-App]{
       \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
@@
 
-    \inferrule*{
-      \sptfundef{\spf_0}{\sptype_0}{\sptype_1} \in \sptenv_0
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
-    }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptype_1
-    }
-
-    \inferrule*{
+    \inferrule*[left=PD-Ref]{
       \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
@@
 
-    \inferrule*{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
+    \inferrule*[left=CD-Ref]{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
+      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
     }{
-      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptnone
+      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
     }
 
-    \inferrule*{
-      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
-    }{
-      \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \sptype_0
-    }
-
-    \inferrule*{
-      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
+    %% TODO show dict write rules?
+    %\inferrule*[left=PD-Set]{
+    %  \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+    %  \\\\
+    %  \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+    %  \\\\
+    %  \sptenv_0 \wtexpr \spexpr_2 : \sptobject
+    %}{
+    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
+    %}
+    %
+    %\inferrule*[left=CD-Set]{
+    %  \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+    %  \\\\
+    %  \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+    %  \\\\
+    %  \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
+    %}{
+    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1} : \sptnone
+    %}
+
+    %\inferrule*[left=F-Ref]{
+    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
+    %}{
+    %  \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
+    %}
+
+    \inferrule*[left=F-Set]{
+      \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
       \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
     }{
@@
 
+    \inferrule*[left=M-App]{
+      \sptfundef{\spf_0}{\spc_0, \spteval_0}{\spteval_1} \in \spenvapp{\sptenv_0}{\spc_0}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \spteval_1
+    }
+
+    \inferrule*[left=C-Sub]{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \\
+      \spteval_1 \spcompat \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }
+
+    \inferrule*[left=Matr]{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \\
+      \spteval_1 \spmatr \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }
+
+  \end{mathpar}
+
+  \begin{minipage}[t]{0.5\columnwidth}
+    \fbox{$\spteval \spcompat \spteval$}
+    \begin{mathpar}
       \inferrule*{
-      \sptfundef{\spf_0}{\spc_0, \sptype_0}{\sptype_1} \in \spenvapp{\sptenv_0}{\spc_0}
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
+        \spteval_0 \spsubt \spteval_1
       }{
-      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \sptype_1
+        \spteval_0 \spcompat \spteval_1
       }
 
+      \inferrule*{
+      }{
+        \spteval_0 \spcompat \sptdyn
+      }
   \end{mathpar}
 
-  \caption{Evaluation Typing Judgment}
-  \label{f:eval-typing}
-\end{figure}
-
-\begin{figure}[t]
-  \begin{langarray}
-    \spvalue & \defeq &
-      \mbox{heap-allocated Python values}
-    \\
-    \spteval & \defeq &
-      \sptdyn \langmid
-      \sptnone \langmid
-      \sptclass \langmid
-      \sptobject \langmid
-      \sptint \langmid
-      \sptbool \langmid
-      \sptfloat \langmid
-    \\ & &
-      \sptrawpydict \langmid
-      \sptchkdict{\sptype}{\sptype} \langmid
-      \sptoptional{\sptype}
-  \end{langarray}
-
-  \bigskip
-  \(
-    \mftypeF{\sptype_0}
-    \mfeq
-    \left\{\begin{array}{ll}
-      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
-    \\
-      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
-    \\
-      \sptype_0 & \mbox{otherwise}
-    \end{array}\right.
-  \)
-
-  \fbox{$\sptype \spsubteq \sptype$} reflexive, transitive closure of $\spsubt$, defined below:
+  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
+    \fbox{$\spteval \spmatr \spteval$}
+    \begin{mathpar}
+      \inferrule*[right={$\spteval_0 \neq \sptdyn$}]{
+      }{
+        \sptdyn \spmatr \spteval_0
+      }
+    \end{mathpar}
+  \end{minipage}
+
+  \medskip
+  \fbox{$\spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
   \begin{mathpar}
@@
     }{
-      \sptype_0 \spsubt \sptobject
+      \spteval_0 \spsubt \sptobject
     }
@@
     \inferrule*{
-      \sptype_0 \spsubt \sptype_1
+      \spteval_0 \spsubt \spteval_1
     }{
-      \sptype_0 \spsubt \sptoptional{\sptype_1}
+      \spteval_0 \spsubt \sptoptional{\spteval_1}
     }
@@
 %    }{
-%      \sptype_0 \spsubt \sptype_0
+%      \spteval_0 \spsubt \spteval_0
 %    }
@@
 %    \inferrule*{
-%      \sptype_0 \spsubt \sptype_1
+%      \spteval_0 \spsubt \spteval_1
 %      \\
-%      \sptype_1 \spsubt \sptype_2
+%      \spteval_1 \spsubt \spteval_2
 %    }{
-%      \sptype_0 \spsubt \sptype_2
+%      \spteval_0 \spsubt \spteval_2
 %    }
@@
 
-  \fbox{$\sptype \spcompat \sptype$}
-  \begin{mathpar}
-    \inferrule*{
-      \sptype_0 \spsubt \sptype_1
-    }{
-      \sptype_0 \spcompat \sptype_1
-    }
-
-    \inferrule*{
-    }{
-      \sptype_0 \spcompat \sptdyn
-    }
-  \end{mathpar}
-
-  \fbox{$\sptype \spmatr \sptype$}
-  \begin{mathpar}
-    \inferrule*{
-    }{
-      \sptdyn \spmatr \sptype_0
-    }
-  \end{mathpar}
-
-  \caption{TBA Type Metafunctions}
-  \label{f:surface-to-eval-types}
+  \caption{Expression Typing, Consistent Subtyping, and Materialization}
+  \label{f:eval-types}
 \end{figure}
 
-\begin{table}
-  \caption{TBD Run-time checks}
-  \label{t:cast}
-
-  \(\begin{array}{l@{\quad}l}
-    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
-    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
-    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
-    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
-    \sptdyn & \mbox{No check needed, accepts any value} \\
-    \sptobject & \mbox{Accepts any non-primitive value} \\  %% except primitives
-    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
-    \sptnone & \mbox{Accepts the Python none value} \\
-    \sptint & \mbox{Accepts integers} \\
-    \sptbool & \mbox{Accepts booleans (\code{0} and \code{1})} \\
-    \sptfloat & \mbox{Accepts floats}
-  \end{array}\)
-\end{table}
-
-\subsection{Properties}
-
-%% TODO 2021-12-29: Carl
-%% "full soundness" requires strict modules, means you can trust module toplevel
-%% hmph
+%\subsubsection{Program Typing, Inheritance Restriction}
+%
+%\Cref{f:program-typing} presents the typing judgment for programs,
+%which folds over a sequence of statements to build a type environment.
+%The typing rules for variables and functions are standard.
+%We draw attention to the rule for classes, in particular the override condition.
+%A class may override fields and methods from its parent;
+%if it does, the overriding definition must be a \emph{static subtype} of the original.
+%For example, a method that returns an integer can be overridden by a method that returns
+%a boolean, but it cannot be overriden by a method that returns the dynamic type.
+%This restriction lets the \SP{} compiler avoid some run-time checks; it can assume
+%that static methods return reliable values instead of having to insert checks for subtypes.
+%
+%\begin{figure}[t]
+%  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
+%  \begin{mathpar}
+%    \inferrule*{
+%      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
+%      \\\\
+%      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
+%    }{
+%      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
+%    }
+%
+%    \inferrule*{
+%      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+%    }{
+%      \sptenv_0 \wtprog \spvardef{\spx_0}{\spteval_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\spteval_0},~\sptenv_0
+%    }
+%
+%    \inferrule*{
+%      \sptenv_1 = \sptfundef{\spf_0}{\spteval_0}{\spteval_1},~\sptenv_0
+%      \\\\
+%      \sptenv_1 \wtexpr \spexpr_0 : \spteval_1
+%    }{
+%      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\spteval_0}}{\spteval_1}{\spexpr_0} \dashv \sptenv_1
+%    }
+%
+%    \inferrule*{
+%      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\spteval_0}}{\sptfundef{\spf_1}{\spc_0, \spteval_1}{\spteval_2}}, \sptenv_0
+%      \\\\
+%      %% TODO are field overrides exact?
+%      \mbox{if $\spann{\spx_0}{\spteval_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_0 = \spteval_0'$}
+%      \\\\
+%      %% TODO can T override dyn?
+%      \mbox{if $\sptfundef{\spf_1}{\spc_2, \spteval_1'}{\spteval_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_1' \spsubteq \spteval_1$ and $\spteval_2 \spsubteq \spteval_2'$}
+%      \\\\
+%      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \spteval_0
+%      \\
+%      \spann{\spself}{\spc_0}, \spann{\spx_1}{\spteval_1}, \sptenv_1 \wtexpr \spexpr_1 : \spteval_2
+%    }{
+%      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\spteval_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\spteval_1}}{\spteval_2}{\spexpr_1}} \dashv \sptenv_1
+%    }
+%  \end{mathpar}
+%
+%  \caption{Program Typing}
+%  \label{f:program-typing}
+%\end{figure}
+
+\subsection{Type Soundness}
+
+The central question for the model is whether \SP{}'s approach to gradual typing
+is sound.
+No matter how an expression mixes typed and untyped code, its static
+\emph{evaluation} type should be an accurate prediction about the results
+of evaluation.
+Using the metavariable $\spvalue$ to represent heap-allocated Python values
+and $\sprred$ as the evaluation relation (which would need to maintain a heap),
+an approximate soundness theorem is as follows:
 
 \begin{theorem}[Type Soundness]
-  If\ \(~\vdash \spexpr_0 : \sptype_0\)
+  If\ \(~\vdash \spexpr : \sptype\)
   then one of the following holds:
@@
     \item
-      \(\spexpr_0 \sprred \spvalue_0
+      \(\spexpr \sprred \spvalue
         \mbox{ and }
-        \vdash \spvalue_0 : \mftypeF{\sptype_0}
+        \vdash \spvalue : \mftypeF{\sptype}
       \)
     \item
-      \(\spexpr_0\) diverges
-    \item
-      \(\spexpr_0 \sprred \sperror\)
+      \(\spexpr\) diverges
+    \item
+      \(\spexpr\) reduces to an allowed error
   \end{itemize}
@@
 
-\begin{theorem}[Optimization Soundness]
-  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
-  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
-\end{theorem}
+As mentioned above, our argument for soundness has two parts
+and the question at hand is whether the boundaries to type dynamic
+are adequately protected with run-time checks.
+The Redex model's comprehensive test suite provides the other half of the
+argument; namely, that \SP{} is a sound type system for Python.
+
+The direct path to boundary soundness involves two observations.
+First, every evaluation type can be enforced with a decidable cast.
+This is clearly the case for basic values such as floats.
+It true for parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$
+because of their strict semantics.
+Second, all boundaries relate the dynamic type to an evaluation type.
+There are no significant boundaries that relate a partially static type
+to an evaluation type.
+This is due to the strict semantics for checked types and the fact that
+$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
+
+Alternatively, observe that the boundaries in \SP{} are a subset
+of the boundaries in Nom, a gradual language that comes with a detailed
+proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
+Given an \SP{} program, it is possible to derive a Nom program with
+the same type boundaries by replacing every checked dict instance
+with a fresh class type.
+Because different checked dict instantiations are incompatible with
+one another, the \SP{} boundaries impose the same constraints as the Nom
+boundaries.
+
+%\begin{theorem}[Optimization Soundness]
+%  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
+%  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
+%\end{theorem}
 
@@
 
-\subsubsection{Informal Connection to Nom}
-
-\SP{} types restrict the use of Nom,
-thus model is kind of a restriction of Nom.
-
-So since Nom is sound for any sort of dyn mixing,
-and SP allows fewer mixings,
-and both Nom and SP do similar checks at boundaries,
-can begin to see how SP could be sound.
-
-Or perhaps mention this at the start.
-SP similar to Nom, similar checks, keep that in mind.
-
 \subsection{Redex Development, Navigation, Integration, Conformance Tests}
@@
 FILL
+
+%% - count tests
+%% - summarize expressions
+%% - 
+%% - pbt?
 
@@
 extend model with Python ... two languages ... boundaries ... argue for soundness
+
+\subsection{Bytecode, Optimization}
+\label{s:optimize}
+
+FILL
 
@@
 
-%% 2021-01-04 TODO have no evidence that "any module" typing works out ... team
+%% 2022-01-05 TODO We are currently JITting at least some functions in 873
+%% modules, but we are about to increase this by about an order of magnitude,
+%% independent of Static Python.
+
+%% 2022-01-04 TODO have no evidence that "any module" typing works out ... team
 %% tried hard to get network effects ... "We explicitly aimed for converting

### 2022-01-06 pj-7.1/paper.tex
pj: sec 3, show dict writes instead of reads

@@
 
-%% 2022-01-04 argument = nom soundness -> sp model sound -> sp sound via conformance equivalence
+%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least
 %% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)
@@
 type $\sptobject$) to yield a dynamically-typed result.
-The next rules illustrate reads from Python dictionaries and checked dictionaries.
-A Python dict may be deferenced by any value; the read promises a dynamically-typed result.
-By contrast, a checked dict requires an index value that matches its type
-and promises a typed result.
+The next rules illustrate writes to Python dictionaries and checked dictionaries.
+A Python dict may be updated with any kind of key and value.
+By contrast, a checked dict requires keys and values that match its type parameters.
 Next we have two rules for classes:
@@
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
     }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \spteval_1
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
     }
@@
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+      \sptenv_0 \wtexpr \spexpr_0 : \sptobject
     }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
     }
 
-    \inferrule*[left=PD-Ref]{
+    %% TODO show dict read rules?
+%    \inferrule*[left=PD-Ref]{
+%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+%      \\\\
+%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+%    }{
+%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
+%    }
+%
+%    \inferrule*[left=CD-Ref]{
+%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+%      \\\\
+%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+%    }{
+%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
+%    }
+
+    \inferrule*[lab=PD-Set]{
       \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
@@
       \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
     }{
-      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
+      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
     }
 
-    \inferrule*[left=CD-Ref]{
+    \inferrule*[lab=CD-Set]{
       \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
@@
       \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
     }{
-      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
+      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
     }
-
-    %% TODO show dict write rules?
-    %\inferrule*[left=PD-Set]{
-    %  \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
-    %  \\\\
-    %  \sptenv_0 \wtexpr \spexpr_1 : \sptobject
-    %  \\\\
-    %  \sptenv_0 \wtexpr \spexpr_2 : \sptobject
-    %}{
-    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
-    %}
-    %
-    %\inferrule*[left=CD-Set]{
-    %  \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
-    %  \\\\
-    %  \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
-    %  \\\\
-    %  \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
-    %}{
-    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1} : \sptnone
-    %}
 

### 2022-01-06 pj-7.1/paper.tex
pj: sec 5, use imports analysis to emphasize the mix of T-U code

@@
 %% - use ChkDict not ChkList ... avoid confusion
-%% - is \sqle _consistent_ or _compatible_ subtyping?
+%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
 
@@
 \newcommand{\spcompat}{\sqsubseteq}
+\newcommand{\spconsist}{\spcompat}
 \newcommand{\spmatr}{\prec}
@@
 performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
-\cref{t:prod-stat} reports a few other production statistics.
+\Cref{t:prod-stat} reports a few other production statistics.
 Over 500 modules have been converted thus far.
@@
 
-\subsection{The Migration Path}
+\subsection{Migration Path}
 
@@
 During the second half of 2021, the \SP{} team applied the same
-process at a larger scale by incorporating a tool that drafts
-initial type annotations.
-These changes resulted in a 2\% global CPU improvement.
-
-As of December 2021, the \SP{} team has converted 530 modules.
-Most of these came from the annotation tool (400); the rest
-are from hand conversions.
+process at a larger scale.
+They also changed a code-generating module to output \SP{} code instead
+of untyped Python code;
+this one change brought over four hundred generated modules into the \SP{} fold.
+Overall, these changes resulted in a 2\% global CPU improvement.
+
+As of December 2021, the \SP{} team has converted 541 modules.
+Most of these came from the tool (417); the rest
+are from hand conversions (124).
+These modules are well-mixed with untyped modules in the codebase.
+All together, over thirty thousand identifiers cross a type boundary
+at compile-time by either entering or leaving a \SP{} module.
+Two-thirds of these crossings are exports from \SP{} modules
+to untyped modules;
+in other words, the typed identifiers are widely used throughout the Instagram web server.
+%% data in src/static-python-imports-analysis
+
 In a typical week, the team adds types to four additional modules.
@@
 %%   ... doesn't measure the cooperative effect of static-to-static cross-module calls
+%% 2022-01-06 FILL team did significant work to fight regressions
 
@@
 \label{s:conclusion}
+
+%% 2022-01-06 TODO multiple dimensions of gradual
 

### 2022-01-08 pj-7.1/paper.tex
pj: sec 2, hide ChkList figure, no strong ideas yet for filling it

@@
 
+%% TODO 2022-01-07 List and ChkList are incompatible, need to anticipate somehow ... Chk does not allow any lists!
 Parameterized types (generics) are more difficult to fully enforce without
@@
 for lists is not strong enough to decide whether an incoming value matches the
-type~(\cref{f:list-boundary}).
+type.
 A deeper check is needed to validate the list elements.
@@
 
-\begin{figure}[t]
-  How to enforce type \code{List[int]} simply?
-  \begin{itemize}
-    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
-    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
-    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
-    \item Show venn diagram of the options?
-  \end{itemize}
-  \caption{FILL illustrate list boundary and checks}
-  \label{f:list-boundary}
-\end{figure}
+%\begin{figure}[t]
+%  How to enforce type \code{List[int]} simply?
+%  \begin{itemize}
+%    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
+%    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
+%    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
+%  \end{itemize}
+%  2022-01-07: might be an interesting figure idea, showing the two options for enforcing a data structure
+%    Q How to enforce data structure without traversing
+%     \=> option 1: Shallow
+%     \=> option 2: Deep
+%  \caption{FILL illustrate list boundary and checks}
+%  \label{f:list-boundary}
+%\end{figure}
 
@@
 
-%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least
+%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least ... just add a Gamma, follow chain up one step at a time until refl.
 %% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)
@@
 %% data in src/static-python-imports-analysis
+%% 2022-01-06 TODO gen-code is configuration system / access control ... what are stats omitting these?!
 

### 2022-01-10 pj-7.1/paper.tex

@@
 %% - use ChkDict not ChkList ... avoid confusion
+%%   ... or at least mention both, be clear the focus is Dict b/c older
 %% - do the LESSONS from SP come across clearly? beware making this a "progress report"
@@
 \newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\bcinst}[1]{\code{#1}}
 \newcommand{\defeq}{=}
@@
 %    Q How to enforce data structure without traversing
-%     \=> option 1: Shallow
-%     \=> option 2: Deep
+%     \=> option 1: Shallow; + works for legacy code; - doesn't guarantee full type
+%     \=> option 2: Deep; + guarantees full type; - rejects legacy code
+%     \=> (option 3: traverse/wrrap; + guarantees full type; + works for legacy code; - slow; - complex
 %  \caption{FILL illustrate list boundary and checks}
@@
 
+%% TODO need this para?
+We present these in less detail than the type system highlights because
+we suspect they are less-widely applicable than the type system features.
+Some rely on an optimizing JIT compiler, for example.
+Nevertheless, these highlights are an important part of what makes \SP{} successful.
+
 \subsubsection{Bytecode Instructions}
@@
 The added instructions help either to express runtime checks or to
-quickly perform a statically-determined action.
-
-The two main instructions that express run-time checks are \code{CAST}
-and \code{CHECK\_ARGS}.
-The former checks a value against a type.
-The latter is a special cast for functions and methods;
-it appears at the top of every function and checks that the
-actual parameters match the types of the formal parameters.
-
-A simple optimizing instruction is \code{FAST\_LEN}, which expects
-a built-in value\footnote{Because of subtyping, it does not suffice to check
-the type. All Python lists support \code{FAST\_LEN} but some subclasses
-of the Python list type may not.}
-and quickly computes its length.
-Another is \code{LOAD\_FIELD}, which scans a fixed offset within an object.
-By contrast, the standard Python instruction for loading a field needs to
-search a dictionary~(\code{LOAD\_ATTR}).
-
-In addition to speeding code up, the \SP{} compiler takes care
-to avoid slowing code down.
-It does so by avoiding casts and run-time checks that are certain to succeed.
-For example, calls from a typed context to a typed function are statically
-checked for compatibility.
-There is no need to run \code{CHECK\_ARGS} on typed inputs, so the bytecode
-compiler inserts a special call instruction to skip past the check.
+efficiently perform some action.
+For example, the \code{FAST\_LEN} instruction quickly computes the
+length of a built-in value.
+%% TODO is it really undefined
+\SP{} uses type information to decide where this instruction is appropriate
+and applies it as an optimization.
 
@@
 
+\subsubsection{Classloader}
+\label{s:classloader}
+
+\SP{} keeps track of typed functions and typed classes with a specialized classloader.
+At runtime, the classloader keeps a registry of typed objects and whether or not they
+are mutable.
+The registry helps the bytecode reference objects and types via their module names.
+The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
+patches to mutable data.
+Overall, it is an essential piece of infrastructure to enable soundness and optimizations.
+
 \section{Formal Model}
@@
 %% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)
-
-%% TODO 2022-01-04 (somewhere) explain that a verified compiler is an entirely separate project,
-%%  need bytecode model, full transformations, memory model / machine etc.
-%%  hence we don't model optimizations, little benefit
-%%  opt. don't tell us about the surface language ... the particular SP opts are changing representations but not results
 
@@
 that illustrates the \SP{} gradual interaction strategy.
+%% Compared to other gradual languages, these boundaries are quite restrictive---which
+%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
+%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
 The second claim is justified by our Redex development;
@@
 This section describes the Redex development in broad strokes.
-
-\subsection{Goals and Non-Goals}
-
-Our Redex model of \SP{} includes the key aspects of its types,
-statements, and values.
-The model has two main goals:
-illustrate the boundaries where typed and dynamic code can mix,
-and show how to soundly protect these boundaries.
-Compared to other gradual languages, these boundaries are quite restrictive---which
-partly explains how it is able to perform well in practice~(\cref{s:eval}).
-Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021},
-which is a compelling argument for their soundness.
-
-The model intentionally does not cover all of Python.
-Some aspects of \SP{} are left out because they are straightforward to
-handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
-Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
-are also omitted.
-Other Python features are left out because their \SP{} semantics is identical
-to the baseline Python semantics.
-These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
-see \cref{s:impl} for further discussion.
-
-Lastly, the model does not include type-directed optimizations.
-Validating the correctness of optimizations is a separate task that requires
-significant components; namely, a model of the \SP{} bytecode and a faithful
-rendering of its type-based transformations.
-Our model establishes soundness as a baseline.
-Future work can use soundness to reason informally about optimizations,
-or even take it as a starting point for compiler verification.
 
@@
 
-%\begin{theorem}[Optimization Soundness]
-%  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
-%  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
-%\end{theorem}
-
-%\begin{theorem}[Optimization Effectiveness]
-%  FILL how do we know optimizations are useful rather than trivial?
-%  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
-%  %%   for general heuristic opts.
-%  %% - is [opt(e) != e] true for interesting programs?
-%\end{theorem}
-
-\subsection{Redex Development, Navigation, Integration, Conformance Tests}
-
-FILL
-
-%% - count tests
-%% - summarize expressions
-%% - 
-%% - pbt?
-
-\subsection{Bug Reports}
-
-While formalizing Static Python, we submitted N bug reports to the language developers.
-The developers acknowledged M of these bugs as issues to fix.
-
-\begin{center}
-  %% most filed by KC, a few by Ben
-  %% + Carl https://github.com/facebookincubator/cinder/issues/64
-  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
-\end{center}
+\subsection{Redex Development, \SP{} Conformance}
+%% pbt for soundness?
+
+By contrast to the small model language presented in this section,
+our Redex model covers both the typed--untyped boundaries in \SP{}
+and a large subset of Python.
+Notable Python features include module imports, loops, exception handlers, and
+delete statements (for class members and attributes).
+To ensure that these features are implemented correctly and soundly, the
+model includes hundreds of test programs adapted from the \SP{} test suite.\footnote{
+  Most of the N tests are direct translations of \SP{} conformance tests to
+  the syntax of the model.
+  A program that implements the translation is included with our Redex development.
+  The other N tests are not directly expressible in the model; for these, we
+  translated salient aspects of each test by hand.}
+These tests exercise a variety of Python behaviors, thereby giving us confidence
+that the model is sound for Python.
+Some tests intentionally raise type errors, some raise exceptions, and the rest
+compute a specific result.
+We have additionally verified that the model and \SP{} agree on these end-to-end
+behaviors.
+Thus we believe that \SP{} is sound as well.
+%% ... mention that SP is a large and hard-to-analyze language ... opaque?
+
+The model intentionally does, however, not cover all of Python.
+Some aspects of \SP{} are left out because they are straightforward to
+handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
+Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
+are also omitted.
+Other Python features are left out because their \SP{} semantics is identical
+to the baseline Python semantics.
+These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
+see \cref{s:impl} for further discussion.
+Lastly, the model does not include type-directed optimizations.
+Validating the correctness of optimizations is a separate task that requires
+significant components; namely, a model of the \SP{} bytecode and a faithful
+rendering of its type-based transformations.
+Our model establishes soundness as a baseline.
+Future work can use soundness to reason informally about optimizations,
+or even take it as a starting point for compiler verification.
+
+\subsection{The Payoff: Issues Reported}
+
+While modelling \SP{}, we found several issues with the language
+and submitted them as bug reports to the \SP{} developers.
+Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
+Four of these were critical soundness issues, one of which resulted in
+a segmentation fault.
+All but one of these soundness issue have been fixed.
+Five issues were relatively minor; these dealt with confusing error messages
+and incorrect tests.
+The remaining 16 issues report bugs in language design and implementation.
+
+In addition to these formal issue reports,
+we had several long discussions together (between Brown and Instagram)
+about finer points in the language design.
 
@@
 
-\subsection{Python Interaction}
+\subsection{Interactions with Python}
 
@@
 
-extend model with Python ... two languages ... boundaries ... argue for soundness
-
-\subsection{Bytecode, Optimization}
+model with Python ... two languages ... boundaries ... argue for soundness
+
+%% C-types are among the most complex because of casts? no not really complex, perhaps notable
+
+\subsection{Bytecode and Optimizations}
 \label{s:optimize}
 
-FILL
+\SP{} can generate more-efficent code than Python because it targets
+the Cinder runtime.
+Cinder extends Python with bytecode instructions to check types at run-time,
+to construct \SP{} data structures, and to perform optimized actions.
+\Cref{t:bytecode} lists a few representative bytecode instructions.
+
+\begin{table}
+  \caption{Cinder bytecode instructions}
+  \label{t:bytecode}
+
+  \begin{tabular}{lll}
+    %% TODO remove _
+    %% TODO pick instructions
+    {Instruction} & {Purpose} & {Description} \\\hline
+    \bcinst{CAST} & Soundness & Assert that a value matches a type \\
+    \bcinst{CHECK\_ARGS} & Soundness & Cast all inputs to a function \\[1ex]
+
+    \bcinst{BUILD\_CHECKED\_MAP} & Constructor & Make a checked dictionary \\
+    %\bcinst{BUILD\_CHECKED\_LIST} & Constructor & Make a checked list \\
+    \bcinst{TP\_ALLOC} & Constructor & Make a \SP{} object \\[1ex]
+
+    \bcinst{INVOKE\_FUNCTION} & Optimization & Execute a direct function call \\
+    \bcinst{INVOKE\_METHOD} & Optimization & Execute a method via the object's vtable \\
+    \bcinst{LOAD\_FIELD} & Optimization & Read an object field \\
+    \bcinst{STORE\_FIELD} & Optimization & Write to an object field \\
+    \bcinst{FAST\_LEN} & Optimization & Get length of a built-in value \\
+    \bcinst{REFINE\_TYPE} & Optimization & Type declaration for the JIT \\
+
+%%% Other bytecodes ... either from Carl's talk or the SP repo
+%CONVERT_PRIMITIVE
+%INT_LOAD_CONST_OLD
+%JUMP_IF_NONZERO_OR_POP
+%JUMP_IF_ZERO_OR_POP
+%LIST_DEL
+%LOAD_ITERABLE_ARG
+%LOAD_LOCAL
+%LOAD_MAPPING_ARG
+%POP_JUMP_IF_NONZERO
+%POP_JUMP_IF_ZERO
+%PRIMITIVE_BINARY_OP
+%PRIMITIVE_BOX
+%PRIMITIVE_COMPARE_OP
+%PRIMITIVE_LOAD_CONST
+%PRIMITIVE_UNARY_OP
+%PRIMITIVE_UNBOX
+%RETURN_PRIMITIVE
+%SEQUENCE_GET
+%SEQUENCE_REPEAT
+%SEQUENCE_SET
+%STORE_LOCAL
+
+  \end{tabular}
+\end{table}
+
+The two instructions that express run-time checks are \code{CAST}
+and \code{CHECK\_ARGS}.
+The former checks a value against a type.
+The latter is for functions and methods; it checks all inputs to a function
+against their declared types.
+
+By convention, a \code{CHECK\_ARGS} instruction always appears on the first
+line of a typed function body.
+The optimizer uses this convention to skip argument checks by jumping
+past them when it is safe to do so; namely, whenever a typed function (or method)
+gets called with well-typed arguments.
+%% TODO ask C&D about materialize for fn calls
+
+\Cref{t:bytecode} presents two instructions that allocate \SP{}-specific data structures.
+The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
+Cinder comes with a similar instruction to build checked lists.
+The developers will add more instructions along the same lines in the future.
+
+The remaining instructions are for optimization.
+Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
+to Python's dynamic call dispatch.
+The former jumps to a statically-determined address;
+the latter uses a v-table lookup to quickly find an address at run-time.
+The load and store instructions for fields improve upon Python's generic attribute lookup.
+In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
+The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
+of using the Python \code{\_\_len\_\_} attribute.
+Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
+type is not clear from the context.
 
@@
 \label{s:dynamic-python}
+
+%% TODO 2022-01-09 be careful here ... the model may handle module vars, but it's
+%% just that they're untyped and (maybe) not tested
 
@@
 by compiling source code to C extension modules.
+
+JS++ is a sound gradual type system for JavaScript~\cite{jspp}.
+By default, it separates JS++ from JavaScript values:
+every value with a type must be defined in JS++ code,
+and every value that may have come from Javascript has
+a special type \code{external} to mark it as such.
+Programmers can manually cast external values to a
+more precise type.
+The casts may fail at runtime, and may create a copy of
+the original value if they succeed (depends on the type).
+%% 2022-01-09 how does JS++ protect arrays sent to JS?
 
@@
 \appendix
+
+\section{GitHub Issues}
+\label{a:github-issues}
+
+FILL if we're going to list issues, might as well report the titles, labels, and open/closed status.
+
+The 25 GitHub issues that we filed during the formalization effort may be found at the following three links:
+
+\begin{itemize}
+  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
+  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
+  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
+\end{itemize}
 

### 2022-01-10 pj-7.1/paper.tex
pj: sec 4 complete draft, python + opts

@@
 
-\section{Formal Model}
+\section{Model}
 \label{s:model}
@@
 \end{itemize}
-\Cref{s:impl} explains how the type-directed optimizer avoids some
+\Cref{s:impl} explains
+how the type-directed optimizer avoids some
 of these checks.
@@
       \spteval_1 \spmatr \spteval_0
+      %% side condition: result type is not a primitive
     }{
@@
 
-%% TODO discuss optimizations here, not in sec 2
-
 %% TODO SP side channels:
@@
 
-%% TODO wherever we discuss optimizations, also talk about de-slowdowns
-%% Carl: not sure where this belongs exactly, but we go to a lot of work in the
-%%   compiler to intelligently minimize runtime checks by only inserting them when
-%%   the source value is of dynamic type (this includes also skipping CHECK_ARGS if
-%%   the function was statically invoked by another static function and thus we
-%%   know the compiler checked argument types)
-
-[FILL]
-
-Purpose: explain significant gaps between the model and the implementation.
-E.g. what would the TypeScript team need to know before using our model
-to add soundness?
-
-\subsection{Interactions with Python}
-
-FILL
-
-model with Python ... two languages ... boundaries ... argue for soundness
-
-%% C-types are among the most complex because of casts? no not really complex, perhaps notable
+This section explains significant aspects of \SP{} that lie outside the
+scope of our model.
+These aspects include sound interactions with Python code,
+the bytecode language and optimizations, and dynamic Python features
+that \SP{} does not yet ascribe types to.
+
+\subsection{Open-World Python Interactions}
+
+Although \SP{} is technically a new language, it is designed for gradual adoption.
+Python programmers should be able to add types one module at a time to an
+existing codebase.
+Consequently, \SP{} supports interactions with Python modules in the only viable
+way: by letting the Python code run with minimal constraints.
+After all, if programmers need to fix every module in a codebase before they can
+run tests after typing a first module, then the gradual story is lost.
+
+Interactions with open-world~\cite{vss-popl-2017} Python code pose a minor
+threat to soundness.
+Typed functions (and methods) cannot assume that their arguments are
+well-typed.
+Arguments from typed contexts can be validated by either static checks
+or materialization casts, but arguments from untyped contexts are unchecked.
+For this reason, \SP{} compiles every typed function to check its inputs.
+\Cref{s:optimize} explains how the optimizer avoids some of these checks.
+%% 2022-01-09 worth mentioning primitive types? they get cast like anything else
+%% ... I guess the trick is that primitives don't materialize in typed code
+
+FILL sound generics / checked types also need to guard their writes
+
+There are two cases, however, where new \SP{} types do add constraints to
+existing Python code.
+First, sound generic types (e.g. $\sptchkdict{\sptype_0}{\sptype_1}$)
+force value-suppliers to use a checked constructor.
+Existing code can \emph{use}\/ a checked generic via the same API as on
+un-checked one, but can create new values only using the \SP{} constructors.
+Second, the strict setting for modules prevents updates to module-level variables.
+If existing code happens to mutate a member of a newly-strict module, a run-time
+error will occur.
 
@@
 
-By convention, a \code{CHECK\_ARGS} instruction always appears on the first
-line of a typed function body.
-The optimizer uses this convention to skip argument checks by jumping
-past them when it is safe to do so; namely, whenever a typed function (or method)
-gets called with well-typed arguments.
-%% TODO ask C&D about materialize for fn calls
-
 \Cref{t:bytecode} presents two instructions that allocate \SP{}-specific data structures.
@@
 type is not clear from the context.
+
+\paragraph{Indirect Optimization}
+
+In addition to upgrading bytecode instructions to optimized ones, \SP{} takes care
+to minimize the type checks that it executes at run-time.
+In other words, the goal is to slow code down as little as possible.
+
+Part of this goal is met by inserting casts only in positions where the dynamic
+type flows into a static type.
+The materialze rule in the model illustrates this policy.
+Typed functions, however, are compiled with a \bcinst{CHECK\ARGS} instruction
+that casts all arguments.
+By convention, this instruction always appears on the first line of a typed
+function body.
+The optimizer uses this convention to skip argument checks by jumping past them
+when it is safe to do so; namely, whenever a typed function (or method) gets
+called in a typed context.
+In this way, all untyped calls get the necessary argument check and all typed
+calls benefit from materialize checks at the call-site.
 

### 2022-01-11 pj-7.1/paper.tex
pj: fix slash error (those dang tex _s)
pj: sec 2.1.1, add carl info, cleanup, clarify

@@
 %%   helps the static checker find new errors. ChkDict is one example.
-%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
-%% - what are the idiomatic base SP types ... int or Int ?
 %% - choose words: static vs typed, nonstatic vs dynamic vs untyped
@@
 %% - do the LESSONS from SP come across clearly? beware making this a "progress report"
+%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
 
@@
 \newcommand{\sptclass}{\typefont{C}}
+\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
 \newcommand{\sptint}{\typefont{Int}}
@@
 
-%% 2021-12-14:
+%% 2021-12-14 TODO:
 %% - anything to say about Any being a "meta" type in Python typing module?
-%% - 3 classes of code ... how does dynamic differ from untyped?
-
-%% 2021-12-23: TODO
-%% Carl: The behavior of dynamic type in static code should be:
-%%   a) any type can flow to dynamic
-%%   b) dynamic can flow to any type, but it requires the insertion of a runtime cast for soundness
-%%   c) dynamic type should emit the same bytecode for any given operation that non static code would emit
-%%      ... I think differing behavior around (c) is also unlikely — the structure of our
-%%      code generation is that we call methods on type objects to emit byte code for
-%%      various operations on a value of that type, and our dynamic type simply
-%%      forwards all those calls to the default Python compiler, which should always
-%%      result in dynamic type emitting the same byte code that nonstatci code would.
-%%      This isn’t universal, though (i.e. for some operations, e.g. binary ops, we
-%%      might explicitly handle various cases for operand type pairings, not always by
-%%      calling a polymorphic method on each operand type), which could open the door
-%%      to differing behavior here. But I can’t think of or find any such intentional
-%%      cases, and I suspect if any did crop up we’d consider it a bug.
+%%   2022-01-10 bg: not sure what the consequences of "meta type" are
 
@@
 Whenever an expression or variable lacks a type annotation, \SP{} uses the
-dynamic type as a default and skips most (but not all!) compile-time checks.
+dynamic type as a default and skips most \emph{but not all}\/ compile-time checks.
 Every untyped Python program is thus a well-formed \SP{} program.
-Furthermore, programmers can omit the annotations for specific
-function arguments without losing the benefits of other type checks.
-Freedom to omit annotations helps programmers convert modules to \SP{}
-with little initial refactoring.
+Furthermore, programmers can add annotations to part of a module
+while leaving the rest untyped; the untyped positions receive the dynamic type.
 
 In addition to serving as a default, the dynamic type also replaces any undefined types.
-For example, suppose that the type \code{X} is undefined in \SP{} and that a
-program contains the annotation \code{ChkDict[int, X]}, which describes a
-dictionary with integer keys and \code{X}-typed values.
-\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
+For example, suppose that the type $\sptX{}$ is undefined in \SP{} and that a
+program contains the annotation $\sptchkdict{\sptint}{\sptX{}}$, which describes a
+dictionary with integer keys and $\sptX{}$-typed values.
+\SP{} interprets this annotation as \sptchkdict{\sptint}{\sptdyn}$ and allows any sort
 of values inside the dictionary at runtime.
@@
 Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
-efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions~(\cref{s:idiomatic-types}).
+efficiently, e.g., the type $\sptunion{\sptype_0, \ldots, \sptype_n}$ of arbitrary-width unions~(\cref{s:idiomatic-types}).
 By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.
@@
 The dynamic type is also different than the flexible dynamic type provided
-by true gradual languages~\cite{svcb-snapl-2015}.
-Replacing part of a type with dynamic can lead to both compile-time errors
-and run-time errors.
-In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
-for a standard type precision relation.
+by gradual languages that satisfy the gradual guarantees~\cite{svcb-snapl-2015}.
+Replacing part of a type with dynamic can lead to a compile-time error because
+unlike untyped Python code, the \SP{} compiler does analyze dynamic-typed code.
+Similarly, adding a dynamic type can raise a run-time error because
+\SP{} does not allow interactions among different instantiations of a generic type.
 %% \footnote{We
@@
 
-    %% 2021-12-13: TODO NOTE SWALLOW MORE
-    %% I think the fundamental unusual thing we do here is our per-module
-    %% incremental adoption strategy for our compiler. We don’t apply our compiler
-    %% to the entire codebase, using presence or absence of type annotations as
-    %% the sole indicator of whether to compile code with static or dynamic
-    %% behavior. We apply our compiler only to selected opted-in modules, and the
-    %% rest of the codebase is compiled with Python’s default compiler and no
-    %% type-checking (it may of course be checked by Pyre but that’s unrelated to
-    %% compilation.) So this clearly opens the door to differing behavior between
-    %% code that is compiled by our compiler and code that isn’t, even when the
-    %% former has some or many values/expressions typed as dynamic. It also makes
-    %% such differences somewhat more acceptable in practice — since adoption of
-    %% an individual module into our compiler is opt-in, we can swallow more
-    %% behavior difference implied solely by that change.
-
   \item
@@
 \end{itemize}
-These behaviors are a consequence of \SP{} being driven by the needs of an
-existing codebase.
-The literature on gradual types has served only as a guide, not a mandate.
+These behaviors enable efficient run-time checks and additional type-directed
+optimizations.
+Part of the reason \SP{} weighs these benefits more heavily than the costs
+is that programmers have to opt-in to a new feature to risk the errors.
+The static errors come only after enabling the \SP{} compiler on a module.
+The dynamic errors come only after switching to a sound generic type.
+
+More to the point, the \SP{} team has little interest in guarantees about
+\emph{removing an arbitrary annotation}.
+By contrast, there is great interest in making sure that \emph{an untyped
+module compiles with minimal code changes in \SP{}}.
 
@@
         return 0
-    # Error: type dynamic does not match int
+    # Error: dynamic cannot override int
   \end{verbatim}
-    \caption{Removing a type in \code{B} breaks the static gradual guarantee}
+    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
     \label{f:gg-failure-stat}
@@
   \end{verbatim}
-    \caption{Removing part of the type for \code{x} breaks the dynamic gradual guarantee}
+    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
     \label{f:gg-failure-dyn}
   \end{subfigure}
-  \caption{\SP{} values performance over the gradual guarantees.}
+  \caption{\SP{} maintains neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
   \label{fig:gg-failure}
@@
 
-%% TODO carl 2021-12-24 ... very important to introduce with ChkDict ... GG not necessary only nice-to-have
+%% TODO carl 2021-12-24 ... very important to introduce machine types with ChkDict ... emphasizes that the GG is not necessary, but only nice-to-have
 %% > One area where we definitely violate the gradual guarantee and always will
@@
 %% > type can cause static code to no longer type check.
-%% > 
-%% > We are ok with this violation of the gradual guarantee because the form of
-%% > the gradual guarantee we really care about is not “removing an arbitrary
-%% > annotation” but rather “an untyped module compiles without changes when
-%% > adopted into the SP compiler” — and an untyped module cannot have
-%% > pre-existing uses of the primitive machine types, since those types cannot
-%% > exist outside SP.
 
@@
 with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
-but hardly supported at run-time.
+but are hardly supported at run-time.
 Only binary unions of a type and the none value get enforced with a run-time check
@@
   \begin{tabular}{lll}
-    %% TODO remove _
-    %% TODO pick instructions
     {Instruction} & {Purpose} & {Description} \\\hline
@@
 The materialze rule in the model illustrates this policy.
-Typed functions, however, are compiled with a \bcinst{CHECK\ARGS} instruction
+Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
 that casts all arguments.
@@
 
-%% TODO 2022-01-09 be careful here ... the model may handle module vars, but it's
-%% just that they're untyped and (maybe) not tested
+%% TODO 2022-01-09 be careful here ... the model may indeed handle module vars
+%%  the thing is, they're untyped and (maybe) not tested
 

### 2022-01-11 pj-7.1/paper.tex
pj: add missing $
pj: fb -> 'Meta'
pj: split sec 2 into types & runtime

@@
 %% - do the LESSONS from SP come across clearly? beware making this a "progress report"
+%%   - emphasize the PAYOFF from modeling early on, don't bury in sec 3
 %% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
+%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
 
@@
 \newcommand{\sptfloat}{\typefont{Float}}
-\newcommand{\sptdyn}{\typefont{Dynamic}}
+\newcommand{\sptdyn}{\typefont{Dyn}}
 \newcommand{\sptobject}{\typefont{Object}}
@@
 \newcommand{\sptrawpydict}{\typefont{PyDict}}
+\newcommand{\sptrawchkdict}{\typefont{ChkDict}} %% not a real SP type, but useful for TeX
 \newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
@@
 
-%\usepackage{cleveref}
+\usepackage[framemethod=TikZ]{mdframed}
+\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
 \usepackage{amsthm}
@@
 \authorinfo{(\email{carljm@fb.com}) FILL}
-\affiliation[b]{Facebook, Inc.}
+\affiliation[b]{Meta}
 \author[b]{Dino Viehland}
@@
   %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={Database programming, General-purpose programming, Program
-    verification, Programming education},
+  area={General-purpose programming},
   %% License options include: cc-by (default), cc-by-nc
@@
 
-\section{Typing for Performance}
+\section{Introduction}
+%% alt title:
+%%  {A Staged Gradual Language}
+%%  {Typing for Performance}
 \label{s:intro}
@@
 
-\section{\SP{} and the Cinder JIT}
+\section{A Programmer's View of \SP{}}
 \label{s:tour}
 %% purpose = SP looks like Pyre, adds soundness to certain types
-
-%% 2022-01-05 Carl:
-%% "All the data I’ve been sharing is from a single codebase, the primary IG web
-%% server monolith. The existence of JITted but not SP modules is mostly just due
-%% to the fact that opting a function into the JIT is semantically transparent and
-%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
-%% quite easy to do (the only tradeoff pushing us away from JITting everything is
-%% just how much CPU we spend JIT-compiling and how much memory we spend storing
-%% JIT-compiled code), whereas opting a module into SP requires more work so the
-%% bottleneck is available developer time to do conversions."
 
@@
 
-\SP{} is a type checker and bytecode compiler.
-It validates programs written in ordinary Python syntax
-and generates type-specific instructions for the Cinder runtime.
-The type system can express common Python idioms and is guaranteed
-sound through run-time checks, which in turn enable optimizations.
-The type system also includes a dynamic type as an escape hatch to
-untyped behavior.
-An unannotated variable behaves exactly as it would in Python.
-
-Cinder is an extension of CPython 3.8 that adds a method-based
-JIT compiler, virtual method tables, and several bytecode instructions
-to directly express type checks and type-based optimizations.
-These enhancements take full advantage of \SP{} type information.
-They are built on top of CPython so that untyped code runs on
-the platform.
-
-Using \SP{} and Cinder is as easy as installing a new version
-of CPython.
-The main executable compiles programs to bytecode on the fly,
-same as CPython, allowing standard developer tools to work.
-By contrast, the only effective way to get type-directed optimizations
-without changing CPython is to use C extension modules,
-which introduce an extra compilation step and get in the way
-of tooling~(\cref{s:related}).
+\SP{} is part of a large codebase that includes a type system, a tailored
+bytecode, and a method-based JIT compiler.
+In essence, it is a new language.
+The interface that it offers to programmers, however, is a familiar one.
+\SP{} can handle all untyped\footnote{Python 3.8 code that uses type
+annotations may, of course, contain type errors.} Python 3.8 programs, and it
+compiles code on-the-fly to give the same workflow and enable the same
+developer tools as standard Python.
+
+Indeed, the advanced features of \SP{} are offered strictly on an opt-in basis.
+Programmers whose decide to use its new APIs---in particular, for the type system
+and the JIT compiler---can usuaally benefit without imposing maintenance costs
+on their co-workers.
 
 \begin{figure}
-  %% TODO replace with a real program
-%  \begin{minipage}[t]{0.5\columnwidth}
-  \begin{verbatim}
+  \begin{subfigure}[t]{0.22\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
+def f(x):
+  return x["A"]
+
+f({"A": 1})
+  \end{verbatim}
+  \end{spcodebox}
+  %\caption{TODO}
+  %\label{f:TODO}
+  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
 from __static__ import PyDict
-
-    def f(x: PyDict[str, int]):
+def f(x: PyDict[Str, Int]):
   return x["A"]
 
-    f({"A": 1}) # ==> 1
+f({"A": 1})
   \end{verbatim}
+  \end{spcodebox}
+  %\caption{TODO}
+  %\label{f:TODO}
+  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
+from __static__ import ChkDict
+def f(x: ChkDict[Str, Int]):
+  return x["A"]
+
+f(chkdict[Str, Int]({"A": 1}))
+  \end{verbatim}
+  \end{spcodebox}
+  %\caption{TODO}
+  %\label{f:TODO}
+  \end{subfigure}
+
+%%
+%%%  \begin{minipage}[t]{0.5\columnwidth}
+%%  \begin{verbatim}
+%%    from __static__ import PyDict
+%%    def f(x: PyDict[str, int]):
+%%      return x["A"]
+%%
+%%    f({"A": 1}) # ==> 1
+%%  \end{verbatim}
 %  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
@@
 %  \end{verbatim}
-  \caption{A first \SP{} program}
+  \caption{A first \SP{} program and two migrations}
   \label{fig:sp-example}
@@
 
-\Cref{fig:sp-example} presents a first example program.
-It defines a function \code{f} and calls it from typed code.
-The syntax is normal Python with type annotations.
-Only the import statement makes special use of \SP{};
-it imports a type \code{PyDict} that describes Python dictionaries.
-
-Although this program is typed, a different untyped module is free
-to import the function \code{f} and invoke it with any sort of
-argument.
-For this reason, the compiled version of \cref{fig:sp-example} must
-include a few run-time casts:
+\Cref{fig:sp-example} presents a first example program and two modified versions that
+utilize \SP{} types.
+The basic program defines a function \code{f} and calls it with a dictionary
+value.
+\SP{} can run this program as-is, and even JIT-compile the function.
+The other two versions import dictionary types from the \SP{} standard library:
 \begin{itemize}
   \item
-    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
-    \SP{} inserts a check that accepts any Python dictionary.
-  \item
-    Because the domain check does not validate the elements of an incoming dictionary,
-    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
+    The $\sptrawpydict$ type describes normal Python dictionaries.
+    At compile-time, \SP{} uses this type to check the program for logical errors.
+    At run-time, \SP{} enforces the type by checking that all inputs to \code{f} are
+    dictionary values.
+    These checks enable optimizations within the function body.
+
+  \item
+    The $\sptrawchkdict$ type describes a dictionary data structure that is new to \SP{}.
+    Unlike Python dictionaries, these checked dictionaries are guaranteed to
+    contain well-typed keys and values even if they escape to untyped code.
+    As the body of \code{f} illustrates, the syntax for using a checked dict is standard.
+    Creating a checked dict, however, requires a parameterized constructor call.
 \end{itemize}
-These checks ensure a \emph{shape-level} notion of type soundness~(\cref{s:model}).
-In short: if an expression has type \code{T} and reduces to a value, then the value
-matches the top-level constructor, or shape, of the type.
-The shape for a Python dictionary is merely \code{dict}.
-Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
-type that predicts the keys and values in a special kind of dictionary.
-
-Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
-Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
-there is no need to check that it sends type-correct inputs.
-Thus, the compiled call is optimized to use a Cinder bytecode instruction
-that skips the domain check and immediately enters the function body.
-
-The checks and optimizations in this example illustrate the general balancing act
-of \SP{}.
-Type soundness requires run-time checks, but also enables optimizations.
-The gamble is that the optizations help more that the checks hurt.
-
-\subsection{Type System Highlights}
+These typed versions demonstrate the multi-level nature of the \SP{} interface.
+At one level, there are types that describe standard Python values that can be added
+to a program with little-to-no code changes.
+Beyond that, \SP{} offers special-purpose types with stronger guarantees that impose
+a higher maintenance cost.
+
+\subsection{Type System Design Goals}
 
@@
 because of the engineering context at Instagram.
-The following contextual points are important to bear in mind as we present
-notable aspects of the type system:
+The following contextual points are important to bear in mind:
 \begin{itemize}
@@
 \end{itemize}
-
-\subsubsection{Type Dynamic}
+%
+In general, the types that \SP{} implements all enable significant optimizations
+and are simple to validate with run-time checks.
+Types that do meet these criteria can be deferred to Pyre, even if they are widely-used.\footnote{
+At present, the \SP{} team
+has no plans to support all Pyre types. The top priority is to implement
+core types efficiently.}
+
+The core supported types describe basic data (integers, booleans, strings),
+data structures (lists, dicts, promises),
+and nominal classes.
+Union types are well-supported statically and narrow down in accordance
+with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
+but are, for the most part, not supported at run-time.
+Only binary unions of a type and the none value get enforced with a run-time check
+($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).
+
+Three other unsupported types bear special mention.
+There are no first-class class or object types~(\cref{s:inheritance}).
+There are also no types for first-class functions (callables).
+Pyre can check these types instead.
+Third, \SP{} does not support recursive types.
+Neither Pyre nor mypy have recursive types either.
+%% demand seems low ... baffling
+%% mypy = https://github.com/python/mypy/issues/731
+%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
+
+\subsection{Type Dynamic}
 \label{s:type-dynamic}
@@
 dictionary with integer keys and $\sptX{}$-typed values.
-\SP{} interprets this annotation as \sptchkdict{\sptint}{\sptdyn}$ and allows any sort
+\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
 of values inside the dictionary at runtime.
@@
   \begin{subfigure}[t]{0.5\columnwidth}
+  \begin{spcodebox}
   \begin{verbatim}
     class A:
-      def m1(self)->int:
+      def m1(self) -> Int:
         return 0
@@
         return 0
-    # Error: dynamic cannot override int
+    # Error: Dyn cannot override Int
   \end{verbatim}
+  \end{spcodebox}
     \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
@@
   \begin{subfigure}[t]{0.5\columnwidth}
+  \begin{spcodebox}
   \begin{verbatim}
@@
 
-    def f(x: ChkDict[str, dyn]):
+    def f(x: ChkDict[Str, Dyn]):
       return x["A"]
 
-    d = ChkDict[str, int]({"A": 1})
+    d = ChkDict[Str, Tnt]({"A": 1})
     f(d)
-    # Error: f expected ChkDict[str, dyn]
+    # Error: f expected ChkDict[Str, Dyn]
   \end{verbatim}
+  \end{spcodebox}
     \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
@@
 
-\subsubsection{Concrete Types and Shallow Types}
+\subsection{Concrete Types and Shallow Types}
 \label{s:checked-type}
@@
 
-\subsubsection{Gradual Class Hierarchies}
+\subsection{Gradual Class Hierarchies}
 \label{s:inheritance}
@@
 
-\subsubsection{Tailored Types}
-\label{s:idiomatic-types}
-
-\SP{} is selective about which types to support and which to omit.
-In general, supported types are widely-used and come with an efficient
-implementation.
-Other widely-used types are deferred to Pyre for static checks until
-the day when \SP{} adds support.\footnote{At present, the \SP{} team
-has no plans to support all Pyre types. The top priority is to implement
-core types efficiently.}
-
-The core supported types describe basic data (integers, booleans, strings),
-data structures (lists, dicts, promises),
-and nominal classes.
-Additional ``C types'' enable aggressive compilation of primitive
-values~(\cref{s:c-types}).
-Union types are well-supported statically and narrow down in accordance
-with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
-but are hardly supported at run-time.
-Only binary unions of a type and the none value get enforced with a run-time check
-($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).
-
-Three unsupported types bear special mention.
-There are no first-class class or object types, as noted in \cref{s:inheritance}.
-There are also no types for first-class functions (callables).
-Pyre can check these types instead, to point out clear logical flaws.
-Finally, \SP{} does not support recursive types due to low demand.
-Neither Pyre nor mypy have recursive types either.
-%% demand seems low ... baffling
-%% mypy = https://github.com/python/mypy/issues/731
-%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
-
-\subsection{Runtime System Highlights}
+\subsection{Strict Modules}
+\label{s:strict-mod}
+
+Python modules are mutable collections of names.
+Names can be added, deleted, and updated at runtime.
+\SP{} accommodates these behaviors in the same manner as standard Python, which
+necessarily limits the benefits of module types.
+As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
+Strict modules are immutable, and thus enable type-based compiler optimizations.
+
+\subsection{Unboxed Primitive}
+\label{s:c-types}
+
+%% 2021-12-29: TODO
+%% Using the term “ctypes” for these might be confusing, since the Python stdlib
+%% has a “ctypes” module that isn’t related to our primitive types. We do call the
+%% base type for them CType in our compiler, but when discussing them we usually
+%% call them “primitive types” or “machine types” — “C types” could also work but
+%% I wouldn’t combine it into one word :)
+
+For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
+that describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
+Cinder implements these types with unboxed C values, which are much simpler and cheaper
+than their Python counterparts.
+\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
+These are not themselves primitive.
+
+Primitives are carefully limited by the type system.
+Neither a module-level nor a closure-level variable may have a primitive
+type---because untyped code can mutate such variables.
+Primitive types are incompatible with the dynamic type,
+and there are no implicit upcasts from primitives
+to a matching Python type.
+Even a conjunction disallows mixing; for example, the \code{and} operator
+requires either two Python booleans or two primitive booleans.
+The only way for Python values and primitive values to interact
+is through dedicated conversion functions.
+Cinder handles conversions at the boundaries between \SP{} code
+and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
+Within typed code, programmers must satisfy the type checker with
+appropriate conversions.
+
+% For now, programmers have to write and manage primitive types.
+% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.
+
+% https://github.com/facebookincubator/cinder/issues/52
+
+\section{Runtime System Highlights}
+%% alt title: Runtime System Highlights
 %% Non-Highlights:
@@
 
-Cinder, the \SP{} runtime system, extends the standard Python runtime in
-several ways.
-Some of these ensure the integrity of types.
-Others use type soundness to generate efficient code.
-
-%% TODO need this para?
-We present these in less detail than the type system highlights because
-we suspect they are less-widely applicable than the type system features.
-Some rely on an optimizing JIT compiler, for example.
-Nevertheless, these highlights are an important part of what makes \SP{} successful.
-
-\subsubsection{Bytecode Instructions}
+The \SP{} runtime system, formally known as Cinder, extends
+CPython 3.8 in several ways to take full advantage of static types.
+Cinder includes tailored bytecode instructions, virtual method tables, new
+datatypes, a registry of typed modules, and a JIT compiler.
+
+The key ingredients in Cinder may be of interest to other teams seeking to add
+sound gradual typing to a widely-used language.
+For example, a team of Node developers may wish to fork that runtime system
+and experiment with bytecode instructions that rely on statci types.
+
+\subsection{Bytecode Instructions}
 
@@
 
-\subsubsection{Virtual Method Tables}
+\subsection{Virtual Method Tables}
 \label{s:vtable}
@@
 
-\subsubsection{Checked Data Structures}
+\subsection{Checked Data Structures}
 \label{s:chkdict-impl}
@@
 
-\subsubsection{Unboxed Primitive}
-\label{s:c-types}
-
-%% 2021-12-29: TODO
-%% Using the term “ctypes” for these might be confusing, since the Python stdlib
-%% has a “ctypes” module that isn’t related to our primitive types. We do call the
-%% base type for them CType in our compiler, but when discussing them we usually
-%% call them “primitive types” or “machine types” — “C types” could also work but
-%% I wouldn’t combine it into one word :)
-%%
-%% I just did a very rough grep check and found that we have only 12 modules
-%% currently using primitive types: all int32 and int64 and cbool. So they are
-%% used pretty sparingly, but they are also definitely critical to our ability to
-%% match Cython in a few of the  hottest code paths that used to be Cython.
-
-%% TODO move prim.types up?
-
-For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
-that describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
-Cinder implements these types with unboxed C values, which are much simpler and cheaper
-than their Python counterparts.
-\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
-These are not themselves primitive.
-
-Primitives are carefully limited by the type system.
-Neither a module-level nor a closure-level variable may have a primitive
-type---because untyped code can mutate such variables.
-Primitive types are incompatible with the dynamic type,
-and there are no implicit upcasts from primitives
-to a matching Python type.
-Even a conjunction disallows mixing; for example, the \code{and} operator
-requires either two Python booleans or two primitive booleans.
-The only way for Python values and primitive values to interact
-is through dedicated conversion functions.
-Cinder handles conversions at the boundaries between \SP{} code
-and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
-Within typed code, programmers must satisfy the type checker with
-appropriate conversions.
-
-% For now, programmers have to write and manage primitive types.
-% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.
-
-% https://github.com/facebookincubator/cinder/issues/52
-
-\subsubsection{Strict Modules}
-\label{s:strict-mod}
-
-Python modules are mutable collections of names.
-Names can be added, deleted, and updated at runtime.
-\SP{} accommodates these behaviors in the same manner as standard Python, which
-necessarily limits the benefits of module types.
-As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
-Strict modules are immutable, and thus enable type-based compiler optimizations.
-
-\subsubsection{Classloader}
+\subsection{Classloader}
 \label{s:classloader}
@@
 Overall, it is an essential piece of infrastructure to enable soundness and optimizations.
+
+\subsection{Method-at-a-Time JIT}
+
+FILL Cinder includes a JIT compiler for its bytecode.
+The typical way to configure is to supply a list of functions to be JIT compiled.
+Compilation works on any \SP{} function whether or not it is typed
+(thus there is some incentive to use \SP{} even without adding type annotations).
+For the biggest payoff, though, the JIT needs type information.
+
+%% 2022-01-05 Carl:
+%% "All the data I’ve been sharing is from a single codebase, the primary IG web
+%% server monolith. The existence of JITted but not SP modules is mostly just due
+%% to the fact that opting a function into the JIT is semantically transparent and
+%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
+%% quite easy to do (the only tradeoff pushing us away from JITting everything is
+%% just how much CPU we spend JIT-compiling and how much memory we spend storing
+%% JIT-compiled code), whereas opting a module into SP requires more work so the
+%% bottleneck is available developer time to do conversions."
 
@@
 \subsection{The Payoff: Issues Reported}
+%% 2022-01-11 FILL TODO reword, excitement!
 
@@
 The remaining 16 issues report bugs in language design and implementation.
+The \SP{} team has acknowledged these bugs as such via GitHub a label (\code{sp-correctness}).
 
@@
 
+%% 2022-01-11 TODO how many Checked modules in production?
+%% 2022-01-11 TODO how many strict modules in production?
+%% 2022-01-11 TODO talk about primitive types:
+%% Carl: "I just did a very rough grep check and found that we have only 12 modules
+%% currently using primitive types: all int32 and int64 and cbool. So they are
+%% used pretty sparingly, but they are also definitely critical to our ability to
+%% match Cython in a few of the  hottest code paths that used to be Cython."
+
 %% 2022-01-05 TODO We are currently JITting at least some functions in 873
@@
 by compiling source code to C extension modules.
+\SP{} even used Cython at some point and found the mixed-workflow codebase painful.
 
@@
   \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
+  %% extra, can't remember if we inspired Dino or not: https://github.com/facebookincubator/cinder/issues/32
 \end{itemize}

### 2022-01-12 pj-7.1/paper.tex
abstract cleanup

@@
 
+  \vglue10pt
+
   \paragraph{Context}
@@
   %% TODO acknowlegde the success of new language / tracing JIT approaches
+
+  \vglue10pt
 
@@
 
+  \vglue10pt
+
   \paragraph{Approach}
   % What was done that unveiled new knowledge?
-  In collaboration with the \SP{} team, we have evaluated the language
-  as it exists today and developed a model to capture the essence of its
+  We evaluate the language in its current state
+  to develope a model that capture the essence of its
   approach to gradual types.
-  We drew upon personal communications, bug reports, and the \SP{}
-  regression test suite.
+  We draw upon personal communication, bug reports, and the \SP{}
+  regression test suite to develop this model.
+
+  \vglue10pt
 
@@
   On a more technical level, this paper contributes two artifacts:
-  a model of \SP{} that passes all relevant tests from the \SP{} codebase
-  and a performance evaluation of \SP{}.
+  a model of \SP{} that passes all relevant tests from the \SP{} codebase,
+  and a performance evaluation of \SP{}. The process of formalization
+  has also found several errors in the implementation, inlcuding fatal errors.
+
+  \vglue10pt
 
@@
   Thanks to \SP{}, global CPU usage has improved by almost 4\% within a year.
+
+  \vglue10pt
 

### 2022-01-13 pj-7.1/paper.tex
pj: Meta + Menlo Park
pj: more sk + carl notes
pj: change tobias macro to allow vglue
somehow this was an error for me (texlive 2017) but not sk
pj: intro revised

@@
 %% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
+%% - "we" = all authors, NOT JUST BROWN!!!
 
@@
 \newcommand{\ipara}[1]{\emph{#1}\/.}
+\newcommand{\numbenchmark}{2}
 
@@
 \makeatletter
-\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
+\newcommand*\abstractpart[1]{\leavevmode\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
 \makeatother
@@
 \authorinfo{(\email{carljm@fb.com}) FILL}
-\affiliation[b]{Meta}
+\affiliation[b]{Meta, Menlo Park, CA, USA}
 \author[b]{Dino Viehland}
@@
 \begin{abstract}
-%  Soundness, migratory typing.
-%  Novel run-time check strategy, combines concrete and transient for ergonomics.
-
   \let\paragraph\abstractpart
@@
   Gradually-typed languages allow typed and untyped code to interoperate,
-  but typically come with some kind of drawback.
+  but typically come with significant drawbacks.
   In some languages, the types are unreliable;
@@
   approach to gradual typing.
-
-  %% TODO acknowlegde the success of new language / tracing JIT approaches
 
@@
   the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
-  Concrete types provide soundness in an efficient but difficult-to-use way.
+  Concrete types provide full soundness and low performance overhead, but impose
+  nonlocal constraints.
   Transient types are sound in a shallow sense and easier to use; they help
@@
   We evaluate the language in its current state
-  to develope a model that capture the essence of its
+  and develop a model that captures the essence of its
   approach to gradual types.
@@
   way to lower the maintenance cost of the concrete approach.
+  We also find that method-based JIT technology can eliminate the
+  costs of the transient approach.
   On a more technical level, this paper contributes two artifacts:
-  a model of \SP{} that passes all relevant tests from the \SP{} codebase,
+  a model of \SP{} that passes hundreds of tests from the \SP{} codebase,
   and a performance evaluation of \SP{}. The process of formalization
-  has also found several errors in the implementation, inlcuding fatal errors.
+  found several errors in the implementation, including fatal errors.
 
@@
   Other language designers may wish to replicate its approach,
-  especially those who maintain unsound type checkers for untyped languages
-  such as JavaScript.
+  especially those who currently maintain unsound gradual languages and are
+  seeking a path to soundness.
 
@@
 Gradual typing has attracted significant interest as a solution to
-the impasse between static and dynamic typing.
+the debate between static and dynamic typing.
 The premise is simple: let programmers introduce types in part of a
 codebase and while leaving the rest untyped.
-Run-time checks can in principle enforce the assumptions that
-typed code makes about untyped components, thereby ensuring that
-the types are sound no matter how the untyped code behaves.
+Run-time checks enforce the critical assumptions that typed code makes about
+untyped components, thereby ensuring that the types are sound no matter how
+untyped code behaves.
 
@@
 The most promising attempt is the \emph{concrete} semantics
-for gradual types [CITE kafka nom etc], but its low performance
-overhead stems from severe restrictions on untyped code.
-%% In other words, concrete types impose low performance costs
-%% but high migration costs.
-%% ... you can have any color as long as it's black -h.ford
-%% ... concrete expresses few programs, but runs them all nicely
-Whether developers can accept these restrictions is unclear;
-indeed, they may prefer a whole-sale migration over an awkward
-gradual transition.\footnote{Successful migrations of untyped codebases
+for gradual types [CITE kafka nom etc], but it imposes nonlocal
+constraints on untyped code.
+In particular, concrete-typed client code is incompatible with
+values created by untyped code.
+Rather than search for a coordinated set of edits to make concrete
+types work, developers may prefer a whole-sale migration to
+a typed language.\footnote{Successful migrations of untyped codebases
 to a typed language are rare, but not unheard of.
-Twitter ported its server-side code from Ruby to Scala
-and Dropbox moved its core sync engine from Python mypy to Rust: see
-\shorturl{http://www.}{artima.com/scalazine/articles/twitter\_on\_scala.html} and
-\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}.}
+Twitter ported its server-side code from Ruby to Scala~\cite{twitter-scala}
+and Dropbox moved its core sync engine from Python mypy to Rust~\cite{dropbox-rust}.}
 
@@
 
-This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
+This paper reports an exception to the rule among industry-made gradual type systems.
 The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
 Python along with a runtime system that uses soundness to drive optimizations.
-The language design is a variation of concrete types that eases its restrictions.
-Instead of asking programmers to migrate from untyped code to sound concrete types
-in one leap, \SP{} offers an intermediate step via shallow types that enforce
-only the top-level shape of values.
-Programmers can begin with shallow claims and gradually dial up soundness.
-On a suite of benchmarks and production modules, Static Python out-performs untyped
-Python across the board.
+The language is staged to let programmers choose between easy migrations and
+full-strength optimizations, a design that we call \emph{gradual soundness}.
+To a first approximation, there are three main stages:
+\begin{enumerate}
+  \item
+    \SP{} contains a full language of \emph{shallow types}\/ that describe idiomatic
+    Python code at a coarse granularity.
+    Adapting an untyped module to use shallow types requires at most a few local code
+    changes.
+  \item
+    A small set of \emph{concrete types}\/ describe generic data structures
+    that guarantee the types of their elements.
+    If programmers modify their code to build concrete structures instead of Python
+    ones (a potentially nonlocal change), then \SP{} can speed up reads from these
+    structures.
+  \item
+    Additional types and flags to enable further optimizations.
+    These include \emph{primitive types}\/ that enable unboxed arithmetic
+    and \emph{strict module semantics}\/ that makes modules immutable.
+\end{enumerate}
+
+Over the past year, Instagram has been applying gradual soundness to its
+primary web server monolith.
+The results are very positive.
+Although only a handful of modules use \SP{} types (hundreds among thousands),
+and only a few critical modules rely on concrete types and primitive types (dozens),
+global CPU usage has improved by almost 4\% according to internal profiling tools.
+Type-directed optimizations outweigh the cost of enforcing soundness.
+Consequently, we believe the details of the \SP{} approach are of interest to
+the gradual typing community at large.
 
 \paragraph*{Contributions}
+This paper makes two contributions:
 
@@
   \item
-    Evidence that gradual soundness delivers Python-level performance at a low migration cost
-    and faster performance as programmers gradually incorporate sound types.
-    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
-    After refactoring to sound types, \SP{} performs well on a microbenchmark
-    suite and in production.
-  \item
-    A formal description of the \SP{} core language and an analysis of its soundness
-    and optimizations.
-    Soundness is gradual in the sense that some types are fully sound and others validate
-    only top-level shapes.
-  \item
-    Discussions of the expressiveness limitations in \SP{} relative to
-    idealized gradual type systems, and of the programming discipline that Instagram
-    follows in their gradually-sound codebase.
+    \emph{Evidence}.
+    We present evidence that \SP{} improves performance with few code changes.
+    At Instagram, the web server has become significantly more efficient
+    after the application of \SP{} to high-profile modules.
+    Because the server code is closed-source, we additionally present data for
+    \numbenchmark{} public benchmarks.
+    Using only shallow types, \SP{} runs on-par with Python despite the costs of
+    enforcing soundness.
+    With fine-tuned types, \SP{} consistently out-performs the Python baseline.
+
+  \item
+    \emph{Mechanization}.
+    To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
+    and ran over 200 tests adapted from the \SP{} regression suite.
+    The model brings across the essence of \SP{} and lends support to the claim
+    that core \SP{} is indeed sound.
+    Perhaps more importantly, the \emph{process} of modeling revealed 20 significant
+    issues in \SP{}.
+    Four of these were critical soundness bugs.
+
 \end{itemize}
+
+\paragraph*{Outline}
+
+This paper begins with an informal description of \SP{} in two parts.
+First, we present a user-oriented tour of the language~(\cref{s:tour}).
+Second, we present the key ingredients of the runtime system, Cinder,
+that supports \SP{}~(\cref{s:runtime}).
+\Cref{s:model} uses a small formal model to introduce our Redex mechanization
+and to convince readers that \SP{} is based on a sound core.
+\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
+\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
+production and on public microbenchmarks.
+The paper concludes with related work~(\cref{s:related}),
+future work~(\cref{s:future}), and a final summary~(\cref{s:conclusion}).
 
@@
 In particular, the maintainers of optionally-typed languages
-may find that adding transient and concrete types is a low-overhead
-way to turn types into reliable claims.
-Second, we want to lend focus to researchers.
-Some of the restrictions that \SP{} adopts may be useful to
-ground theoretical work.
-Other restrictions might be lifted by future research.
-
-\section{A Programmer's View of \SP{}}
+may wish to focus on shallow types and JIT compilation as a
+first step toward sound gradual types.
+Second, we want to entice researchers.
+The \SP{} type system has many restrictions that are suprising
+at first glance; for example, functions are supported only by
+shallow types.
+Some of these restrictions might be lifted by future research.
+Others might be useful to adopt in new language designs.
+
+\section{A Tour of \SP{}}
 \label{s:tour}
@@
 
+%% 2022-01-12 Although the Static Python compiler can technically be
+%% used apart from strict modules, and we do that in tests for faster tests, it’s
+%% not sound and we don’t allow it in prod.
+
 Python modules are mutable collections of names.
@@
 \section{Runtime System Highlights}
+\label{s:runtime}
 %% alt title: Runtime System Highlights
@@
 
+%% 2022-01-12
+%% We did have some cases where strict semantics caused a problem, but in the
+%% majority of conversions it didn’t (or if it did initially, the fix was
+%% trivial.) There was at least one case where I would have converted more code to
+%% static but couldn’t easily do so because of strictness issues with a larger
+%% framework reliant on metaclasses, and so I instead extracted the hottest code
+%% paths into a new strict/static module and called them. I have a plan to go back
+%% to this and improve the framework itself to be strict so more code can be
+%% converted. I know of at least one other case where we did modify a framework
+%% that relied on side-effecting decorators so that it could be used without the
+%% side effects.
+
+%% 2022-01-12 All SP mods are strict. Although the Static Python compiler can technically be
+%% used apart from strict modules, and we do that in tests for faster tests, it’s
+%% not sound and we don’t allow it in prod. The way a real module gets compiled as
+%% Static Python is in fact via the strict modules import loader, which detects
+%% both `import __strict__` and `import __static__` markers, treating the former
+%% as strict only and the latter as both strict and static. So we don’t provide
+%% any straightforward way (either in IG codebase or generally with Cinder) to
+%% have a module imported as static without also being strict.
+
+%% 2022-01-12 Looks like 3 use CheckedDict and 6 use CheckedList.
+%% 2022-01-12
+
 %% 2022-01-11 TODO how many Checked modules in production?
@@
     %% Carl: \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}
+
+%% 2022-01-12
+%% I don’t think it was a super common change, but there were some cases where we
+%%     changed a @classmethod that didn’t actually use its cls argument to a
+%%     @staticmethod. It looks like we have 24 static modules using a
+%%     @staticmethod somewhere in the module, but I think in the majority of those
+%%     there just happened to be some pre-existing staticmethods (they aren’t
+%%     uncommon in our codebase); I think a minority were intentional
+%%     optimizations on our part. And the pre-existing ones may or may not be
+%%     meaningfully improving performance. So I’m not sure how valuable it is to
+%%     report this metric.
 
@@
 
-\section{Conclusion / Lessons / Design Principles}
+\section{The Lessons}
 \label{s:conclusion}
+
+In conclusion, we summarize the
 

### 2022-01-13 pj-7.1/paper.tex
pj: some thoughts in the conclusion

@@
 
-In conclusion, we summarize the
+\SP{} is ambitious in many ways:
+\begin{itemize}
+  \item Type checker and compiler for Python syntax
+  \item Fork of Python 3.8
+  \item Add significant runtime components: vtable, classloader
+  \item Add a method-based JIT
+  \item Aiming for tens of thousands of typed modules (surely not all 45k in the server, but otoh the plan must be to hit other parts of IG too)
+\end{itemize}
+
+But \SP{} is also carefully targeted:
+\begin{itemize}
+  \item
+    Not aiming for an exhaustive type checker.
+    The availability of tools like Pyre already do a fine job.
+    (Don't forget that consequently, it's helpful for undefined types go to dynamic.)
+  \item
+    Shallow types come first in the implementation, then concrete / checked
+    (ditto for adoption)
+  \item
+    Not aiming for the full gradual guarantees.
+    Only really matters for the initial migration, the common case.
+    Indeed, it's important to not have the guarantee for primitives for their performance.
+  \item
+    The type checker is module-at-a-time, not fine-grained.
+\end{itemize}
 

### 2022-01-13 pj-7.1/paper.tex
pj: related, progressive types

@@
 %% 2022-01-09 how does JS++ protect arrays sent to JS?
+
+Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
+of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
+Primitive types are progressive types (and a little more) for base values.
+Concrete types are essentially a progressive form of shallow types.
 

### 2022-01-13 pj-7.1/paper.tex
pj: progress on sec 2 edits

@@
 This paper begins with an informal description of \SP{} in two parts.
-First, we present a user-oriented tour of the language~(\cref{s:tour}).
+First, we present a user-oriented summary of the language~(\cref{s:tour}).
 Second, we present the key ingredients of the runtime system, Cinder,
@@
 mix typed and untyped code.
-An untyped class may inherit from a typed one and vice-versa.
+An untyped class can inherit from a typed one and vice-versa.
 The point of this feature is to let programmers add types to a single class
@@
 
-\subsection{Unboxed Primitive}
+\subsection{Progressive Primitive Types}
 \label{s:c-types}
@@
 
-For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
-that describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
-Cinder implements these types with unboxed C values, which are much simpler and cheaper
-than their Python counterparts.
+For performance-critical applications, \SP{} includes a set of primitive types
+that describe booleans and sized numbers.
+Example types include \code{int64}, \code{uint64}, \code{double}.
 \SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
-These are not themselves primitive.
-
-Primitives are carefully limited by the type system.
-Neither a module-level nor a closure-level variable may have a primitive
-type---because untyped code can mutate such variables.
-Primitive types are incompatible with the dynamic type,
-and there are no implicit upcasts from primitives
-to a matching Python type.
-Even a conjunction disallows mixing; for example, the \code{and} operator
-requires either two Python booleans or two primitive booleans.
-The only way for Python values and primitive values to interact
-is through dedicated conversion functions.
-Cinder handles conversions at the boundaries between \SP{} code
-and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
-Within typed code, programmers must satisfy the type checker with
-appropriate conversions.
+
+The purpose of these types is to enable unboxed values and arithmetic at
+runtime.
+Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
+refinement over the semantics of basic Python types.
+Primitive types cannot be implicitly cast to non-primitive types,
+including the dynamic type (another violation of the static gradual guarantee).
+Operations such as \code{+} do not accept a mix of primitive and non-primitive data.
+These restrictions give programmers a static separation between fast-running
+primitives and standard data.
+
+% Neither a module-level nor a closure-level variable may have a primitive
+% type---because untyped code can mutate such variables.
 
@@
 %% - shadow frames = JIT feature, lite array to track Python call stack
+
+%% 2022-01-12
+%% Cinder implements primitives with unboxed C values, which are much simpler and cheaper
+%% than their Python counterparts.
+%%
+%% Cinder handles conversions at the boundaries between \SP{} code
+%% and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
+%% Within typed code, programmers must satisfy the type checker with
+%% appropriate conversions.
 

### 2022-01-13 pj-7.1/paper.tex
pj: sec 2 edits thanks sk

@@
 %% - choose words: static vs typed, nonstatic vs dynamic vs untyped
+%%   trying to avoid nonstatic? staticifying?
 %% - use ChkDict not ChkList ... avoid confusion
@@
 \newcommand{\sptint}{\typefont{Int}}
+\newcommand{\sptstr}{\typefont{Str}}
 \newcommand{\sptbool}{\typefont{Bool}}
 \newcommand{\sptfloat}{\typefont{Float}}
+\newcommand{\sptflout}{\typefont{Flout}}
 \newcommand{\sptdyn}{\typefont{Dyn}}
@@
 \newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
+\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
 \newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
@@
 bytecode, and a method-based JIT compiler.
-In essence, it is a new language.
-The interface that it offers to programmers, however, is a familiar one.
-\SP{} can handle all untyped\footnote{Python 3.8 code that uses type
-annotations may, of course, contain type errors.} Python 3.8 programs, and it
-compiles code on-the-fly to give the same workflow and enable the same
+In essence, \SP{} is a type system for a new language.
+The interface that it offers to programmers, however, seeks to replicate the
+Python experience.
+\SP{} runs untyped Python 3.8 programs with minimal changes to their behavior,
+and it compiles code on-the-fly to give the same workflow and enable the same
 developer tools as standard Python.
-
-Indeed, the advanced features of \SP{} are offered strictly on an opt-in basis.
-Programmers whose decide to use its new APIs---in particular, for the type system
-and the JIT compiler---can usuaally benefit without imposing maintenance costs
-on their co-workers.
-
+The advanced features of the \SP{} type system are offered on an opt-in basis
+and arranged so that programmers can begin adding types one module at a time.
+
+%% TODO clean up arrow between subfigs
 \begin{figure}
-  \begin{subfigure}[t]{0.22\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
+  \begin{subfigure}[t]{0.15\columnwidth}\noindent\begin{lstlisting}
+
 def f(x):
@@
 f({"A": 1})
-  \end{verbatim}
-  \end{spcodebox}
-  %\caption{TODO}
-  %\label{f:TODO}
-  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
+  \end{lstlisting}
+\end{subfigure}\(~~\rightarrow~~\)\begin{subfigure}[t]{0.32\columnwidth}\noindent\begin{lstlisting}
 from __static__ import PyDict
@@
 f({"A": 1})
-  \end{verbatim}
-  \end{spcodebox}
-  %\caption{TODO}
-  %\label{f:TODO}
-  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
+  \end{lstlisting}
+  \end{subfigure}\(~~\rightarrow~~\)\begin{subfigure}[t]{0.34\columnwidth}\noindent\begin{lstlisting}
 from __static__ import ChkDict
@@
 f(chkdict[Str, Int]({"A": 1}))
-  \end{verbatim}
-  \end{spcodebox}
-  %\caption{TODO}
-  %\label{f:TODO}
+  \end{lstlisting}
   \end{subfigure}
 
-%%
-%%%  \begin{minipage}[t]{0.5\columnwidth}
-%%  \begin{verbatim}
-%%    from __static__ import PyDict
-%%    def f(x: PyDict[str, int]):
-%%      return x["A"]
-%%
-%%    f({"A": 1}) # ==> 1
-%%  \end{verbatim}
-%  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
-%  \begin{verbatim}
-%    from __static__ import PyDict
-%
-%    def f(x: PyDict[str, int]):
-%      check_args(x, PyDict)
-%      return cast(x["A"], int)
-%
-%    invoke_function(f, {"A": 1}) # ==> 1
-%  \end{verbatim}
-%  %  ....
-%  %  CHECK_ARGS
-%  %  ....
-%  %  INVOKE_FUNCTION
-%  \end{minipage}
-%  \begin{verbatim}
-%    from __static__ import CheckedDict
-%
-%    def f(x: CheckedDict[str, int]):
-%      return x["A"]
-%
-%    f(CheckedDict({"A": 1}))
-%  \end{verbatim}
   \caption{A first \SP{} program and two migrations}
@@
 value.
-\SP{} can run this program as-is, and even JIT-compile the function.
+\SP{} can run this program as-is and even JIT-compile the function.
 The other two versions import dictionary types from the \SP{} standard library:
@@
   \item
-    The $\sptrawpydict$ type describes normal Python dictionaries.
+    The $\sptrawpydict$ type describes normal Python dictionaries in a \emph{shallow} sense.
     At compile-time, \SP{} uses this type to check the program for logical errors.
@@
   \item
-    The $\sptrawchkdict$ type describes a dictionary data structure that is new to \SP{}.
+    The $\sptrawchkdict$ type describes a \emph{concrete} dictionary data structure provided by \SP{}.
     Unlike Python dictionaries, these checked dictionaries are guaranteed to
@@
 Beyond that, \SP{} offers special-purpose types with stronger guarantees that impose
-a higher maintenance cost.
-
-\subsection{Type System Design Goals}
+nonlocal maintenance costs (i.e., on constructor calls).
+
+\subsection{Type System Context and Design Goals}
 
@@
 Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
-Relative to these, the \SP{} takes some interesting turns
-because of the engineering context at Instagram.
-The following contextual points are important to bear in mind:
+In addition, \SP{} is strongly influenced by the following engineering constraints:
 \begin{itemize}
@@
     Performance is the bottom line.
-    Unless \SP{} helps production code run faster, the team will be disbanded
-    and reallocated.
-    %% 2021-12-14: too extreme? and what about soundness?
-  \item
-    %% Q. why is the python so nominal? style guide? java background?
-    The codebase depends heavily on first-order functions and objects in
-    performance-critical code.
-    Higher-order code is fairly common, but rarely appears on a critical path.
-  \item
-    Prior to investing in \SP{}, Instagram used Cython to compile
-    performance-critical code to C extension modules.
-    Cython performed well, but mixing pure Python with a few compiled modules
-    made for an awkward developer experience.
-    \SP{} aims to match the Python workflow.
-  \item
-    The Instagram codebase is already typed.
-    Developers use Pyre [CITE], an optional type checker for Python,
-    to write type annotations and to drive IDE tools (e.g., autocompletion).
+    At the end of the day, \SP{} needs to make code run faster in production.
+  \item
+    Critical code often relies on first-order functions and objects,
+    at least in the Instagram web server.
+    Precise types for higher-order functions and first-class classes
+    are thus a low priority.
+  \item
+    The codebase in which \SP{} is being applied makes heavy use of Pyre [CITE]
+    type annotations.
+    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
+    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
+    % of types that enable optimizations.
+    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
+  \item
+    A module-level granularity is acceptable.
+    Once \SP{} is enabled for a module, it compiles all code in that module
+    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
+    If this behavior is a problem for certain expressions, it is easy enough to move
+    them to an untyped Python module.
 \end{itemize}
@@
 In general, the types that \SP{} implements all enable significant optimizations
-and are simple to validate with run-time checks.
-Types that do meet these criteria can be deferred to Pyre, even if they are widely-used.\footnote{
-At present, the \SP{} team
-has no plans to support all Pyre types. The top priority is to implement
-core types efficiently.}
+and can be validated with constant-time dynamic checks.
+Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.
+
+At present, \SP{} has no plans to support all Pyre types.
+The goal is to implement a small language of types efficiently.
+After all, programmers can always lint their code with Pyre before running
+\SP{}.
 
@@
 but are, for the most part, not supported at run-time.
-Only binary unions of a type and the none value get enforced with a run-time check
+Only binary unions of a type and the Python \code{None} value get enforced with a run-time check
 ($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).
 
-Three other unsupported types bear special mention.
-There are no first-class class or object types~(\cref{s:inheritance}).
-There are also no types for first-class functions (callables).
-Pyre can check these types instead.
-Third, \SP{} does not support recursive types.
-Neither Pyre nor mypy have recursive types either.
-%% demand seems low ... baffling
+Three other unsupported types bear special mention:
+first-class class and object types~(\cref{s:inheritance});
+first-class function (callable) types; and
+recursive types.
+Pyre has some support for the first two, but not for recursive types.
+In any event, the demand for all three types is low in our applications
+of \SP{}.
 %% mypy = https://github.com/python/mypy/issues/731
@@
 Whenever an expression or variable lacks a type annotation, \SP{} uses the
-dynamic type as a default and skips most \emph{but not all}\/ compile-time checks.
-Every untyped Python program is thus a well-formed \SP{} program.
-Furthermore, programmers can add annotations to part of a module
-while leaving the rest untyped; the untyped positions receive the dynamic type.
+dynamic type as a default and skips most \emph{but not all}\/ static checks.
+Programmers can thus add annotations to part of a module
+while leaving the rest unannotated.
 
 In addition to serving as a default, the dynamic type also replaces any undefined types.
-For example, suppose that the type $\sptX{}$ is undefined in \SP{} and that a
-program contains the annotation $\sptchkdict{\sptint}{\sptX{}}$, which describes a
-dictionary with integer keys and $\sptX{}$-typed values.
+For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
+(i.e. it is a misspelling of $\sptfloat$) and that a
+program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
+dictionary with integer keys and $\sptflout$-typed values.
 \SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
@@
 At first glance, this behavior might seem odd: why not reject undefined types with an error?
-The reason is to reduce friction with the Pyre annotations that are already
-widely-used at Instagram.
+The reason is to reduce friction with existing Pyre annotations.
 Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
-efficiently, e.g., the type $\sptunion{\sptype_0, \ldots, \sptype_n}$ of arbitrary-width unions~(\cref{s:idiomatic-types}).
+efficiently, e.g., the type $\sptset{\sptint}$ for sets of integers.
 By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.
 
-%% intro sentence
-The dynamic type is also different than the flexible dynamic type provided
-by gradual languages that satisfy the gradual guarantees~\cite{svcb-snapl-2015}.
-Replacing part of a type with dynamic can lead to a compile-time error because
-unlike untyped Python code, the \SP{} compiler does analyze dynamic-typed code.
-Similarly, adding a dynamic type can raise a run-time error because
-\SP{} does not allow interactions among different instantiations of a generic type.
+The \SP{} dynamic type is also different than the flexible dynamic type provided
+by gradual languages that satisfy the gradual
+guarantees~\cite{svcb-snapl-2015}.
+In \SP{}, replacing part of a type with dynamic can lead to a static error when
+the compiler analyzes that part of the module.
+Similarly, the addition of a dynamic type can lead to a run-time error because of the
+strict checks that \SP{} applies to generic types.
+\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
+%% \footnote{Technically, \SP{} does not even
+%% have syntax for the dynamic type. The idiomatic way to ask for dynamic is to
+%% omit a type annotation.}
 %% \footnote{We
@@
 %%   %% tailored = describes exactly the migrations that SP allows
-\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
 %% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
@@
     The subclass definition raises a compile-time error because it attempts to override
-    the typed \code{m1} method with another method that returns the dynamic type.
-
-  \item
-    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
-    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
-    It sends a dictionary with integer values to a function that expects
-    dictionaries with dynamic values.
+    the typed \code{m} method to return the dynamic type.
+
+  \item
+    \Cref{f:gg-failure-dyn} sends a checked dictionary with integer values to a
+    function that expects dictionaries with dynamic values.
     At runtime, the function rejects this argument because its type is not an
@@
 Part of the reason \SP{} weighs these benefits more heavily than the costs
-is that programmers have to opt-in to a new feature to risk the errors.
-The static errors come only after enabling the \SP{} compiler on a module.
-The dynamic errors come only after switching to a sound generic type.
-
-More to the point, the \SP{} team has little interest in guarantees about
+is that programmers have to opt-in to a feature in order to risk the errors.
+The static error above comes only after enabling the \SP{} compiler on a module.
+The dynamic error comes only after adopting a concrete type (for checked dictionaries).
+
+More to the point, \SP{} has little interest in guarantees about
 \emph{removing an arbitrary annotation}.
-By contrast, there is great interest in making sure that \emph{an untyped
-module compiles with minimal code changes in \SP{}}.
+It is, however, very interested in making sure that \emph{an untyped
+module compiles with minimal code changes}.
 
 \begin{figure}
-  \begin{subfigure}[t]{0.5\columnwidth}
-  \begin{spcodebox}
-  \begin{verbatim}
+  \begin{subfigure}[t]{0.48\columnwidth}
+    \begin{lstlisting}
 class A:
-      def m1(self) -> Int:
+  def m(self) -> Int:
     return 0
@@
 class B(A):
-      def m1(self):
+  def m(self):
+    # Error: Dyn cannot override Int
     return 0
-    # Error: Dyn cannot override Int
-  \end{verbatim}
-  \end{spcodebox}
+  \end{lstlisting}
     \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
@@
   \end{subfigure}
-  ~
-  \begin{subfigure}[t]{0.5\columnwidth}
-  \begin{spcodebox}
-  \begin{verbatim}
+  \hfill
+  \begin{subfigure}[t]{0.48\columnwidth}
+  \begin{lstlisting}
 from __static__ import ChkDict
@@
 
-    d = ChkDict[Str, Tnt]({"A": 1})
+d = ChkDict[Str, Int]({"A": 1})
 f(d)
 # Error: f expected ChkDict[Str, Dyn]
-  \end{verbatim}
-  \end{spcodebox}
+  \end{lstlisting}
     \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
@@
   \end{subfigure}
-  \caption{\SP{} maintains neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
+  \caption{\SP{} ensures neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
   \label{fig:gg-failure}
@@
 
-Every \SP{} type corresponds to a simple run-time check.
+Every \SP{} type corresponds to a constant-time dynamic check.
 When an unknown value reaches a boundary to typed code at run-time,
 \SP{} uses these checks to quickly decide whether or not the value
-matches the type.\footnote{In other words, \SP{} provides \emph{immediate accountability}~\cite{mt-oopsla-2017}.}
+matches the type (think: \code{isinstance} checks).
 Unlike structural gradual type systems, \SP{} does not traverse
@@
 
-These checks try to faithfully enforce the full type, but some are incomplete.
-We call the incomplete ones \emph{shallow} types; examples follow.
+These dynamic checks try to faithfully enforce a full type, but some are incomplete---especially
+for generic types.
+We call the incomplete ones \emph{shallow} types.
+The \SP{} strategy for fully-sound generics is to package each value with a type
+tag; following the \textsf{StrongScript} authors~\cite{rzv-ecoop-2015}, we call
+these \emph{concrete} types.
 
 First off, for a basic nominally-typed language (think Java 1.4, before
-generics), simple and full checks are easy to implement.
+generics), constant-time checks are easy to implement.
 Every type has a name and every value can carry a tag that corresponds to the
@@
 
-%% TODO 2022-01-07 List and ChkList are incompatible, need to anticipate somehow ... Chk does not allow any lists!
-Parameterized types (generics) are more difficult to fully enforce without
-changing the language.
-Consider a list datatype \code{List[T]} that is parameterized by the
-e type of its elements.
-If a function expects lists of integers (\code{List[int]}), then a simple check
-for lists is not strong enough to decide whether an incoming value matches the
-type.
-A deeper check is needed to validate the list elements.
-But, the default \code{List} type cannot support such a check because untyped
-code is free to create lists.
-
-The \SP{} solution is to provide two versions of every parameterized type,
-both of which support efficient run-time checks.\footnote{Currently, only a handful
-of built-in data structures provide shallow and concrete versions. User-defined
-classes will soon follow this approach.}
-One version is shallow.
-The other is \emph{concrete}: complete, but new.
-Consider \code{List[T]} as an example:
+Parameterized types (generics) are more difficult to enforce.
+Consider a dictionary datatype \code{Dict[K,V]} that is parameterized by types
+for its keys and values.
+If a function expects a dict from integers to integers (\code{Dict[Int,Int]}),
+then merely checking for a dict instance is not enough to decide whether an
+incoming value matches the type.
+For example, the incoming value may have been created by untyped code.
+
+The \SP{} solution is to provide two incompatible versions of every parameterized type,
+both of which support efficient run-time checks.\footnote{Currently, only a handful of built-in data structures provide shallow
+and concrete versions. User-defined classes will soon follow this approach.}
+For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
 \begin{enumerate}
   \item
-    \emph{Shallow type \code{List}}.
-    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
-    the elements.
-    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
-    values at run-time.
-  \item
-    \emph{Concrete type \code{ChkList[T]}}.
-    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
-    for concrete types.
-    The data structure implements the list API and comes with a checked constructor that
-    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
+    \emph{Shallow type $\sptrawpydict$}.
+    The default type $\sptrawpydict$ gets enforced with a constant-time check
+    for dict values that ignores the elements inside a dict.
+    Consequently, a parameterized static type such as $\sptpydict{\sptint}{\sptint}$
+    is merely a promise of dict values at run-time; it does not promise anything about
+    dict keys and values.
+  \item
+    \emph{Concrete type $\sptchkdict{\sptype}{\sptype}$}.
+    \SP{} introduces a parameterized type for concrete, checked dictionaries
+    along with a value constructor that records its type constraints.
     \cref{s:chkdict-impl} discusses the implementation further.
@@
 %
-These options let programmers decide whether to change their code to benefit from
-fully-sound types, or leave it as-is for a lower payoff.
-
-%\begin{figure}[t]
-%  How to enforce type \code{List[int]} simply?
-%  \begin{itemize}
-%    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
-%    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
-%    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
-%  \end{itemize}
-%  2022-01-07: might be an interesting figure idea, showing the two options for enforcing a data structure
-%    Q How to enforce data structure without traversing
-%     \=> option 1: Shallow; + works for legacy code; - doesn't guarantee full type
-%     \=> option 2: Deep; + guarantees full type; - rejects legacy code
-%     \=> (option 3: traverse/wrrap; + guarantees full type; + works for legacy code; - slow; - complex
-%  \caption{FILL illustrate list boundary and checks}
-%  \label{f:list-boundary}
-%\end{figure}
-
+These options let programmers decide whether to edit their code to support concrete types
+or leave it as-is with shallow types.
 Often, the shallow types are more attractive because concrete types impose
-a ripple effect throughout the codebase.
+nonlocal changes.
 If one type changes to concrete, then all data constructors that reach this type need
-to change \emph{and then} all their clients need to use the concrete type as well.
-\Cref{s:concrete-migration} discusses this migration cost of concrete types further.
+to change \emph{and} all their clients need to use the concrete type as well.
+
+The upside of concrete types is that they enforce stronger type constraints.
+These constraints can catch bugs and usually lead to more-efficient code.
+Faster performance is not guaranteed, though, because it depends on how the dict
+gets used.
+\Cref{t:shallow-vs-concrete} presents a detailed comparison.
+Shallow types pay a constant-time cost whenever an untyped value enters typed code
+and whenever a typed context reads from a shallow value.\footnote{The predicted
+costs for reads and writes assume that all \SP{} types can be validated with a
+constant-time check.}
+The net cost of reads may be high.
+Concrete types pay a potentially-high cost for constructors,
+a constant-time cost for casts and for writes,
+and zero cost for dereferences.
+If a value is frequently written to and rarely read from, shallow types may actually
+be faster.
+
+\begin{table}[t]
+  \caption{Worst-case costs for shallow and concrete types}
+  \label{t:shallow-vs-concrete}
+
+  \centering
+  \(\begin{array}{lrrrr}
+     & \mbox{Constructor} & \mbox{Cast} & \mbox{Read} & \mbox{Write} \\\hline
+    \sptrawpydict                &           0 & O(1) & O(1) &     0 \\
+    \sptchkdict{\sptype}{\sptype} &        O(N) & O(1) &    0 &  O(1)
+  \end{array}\)
+\end{table}
 
@@
 mix typed and untyped code.
-An untyped class can inherit from a typed one and vice-versa.
-The point of this feature is to let programmers add types to a single class
-independently of its ancestors and children.
+An untyped class can inherit from a typed one and vice-versa,
+letting programmers add types to a single class independently of its ancestors
+and children.
 
@@
 classes are Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}.}
-The closest related work, for Nom~\cite{mt-oopsla-2017}, is much more flexible.
-\SP{} explores a simpler point in the design space, starting with
-a few restrictions on types and inheritance:
+The closest related work, for Nom~\cite{mt-oopsla-2017}, implements
+a rather flexible point in the design space.
+\SP{} implements a simpler design that places a few restrictions on types and
+inheritance:
 
@@
     For example, a method that returns a number cannot be overridden by a method
-    that returns the dynamic type~(example: \cref{f:gg-failure-stat}).
+    that returns the dynamic type; \cref{f:gg-failure-stat} shows the static error
+    that arises from such a static override.
 
@@
 %
-With this context in mind, every class member is either \emph{reliably} typed or untyped.
-In particular, a typed member cannot be overridden by a less-precise one that uses
-type dynamic.
+With this context in mind, \SP{} keeps track of whether each class is typed of untyped.
+Each typed class can furthermore assume that if a method has a precise (non-dynamic)
+type, then all typed overrides are subtypes of this type.
 \SP{} can therefore optimize dispatch from typed callers to typed methods
@@
 % \SP{}
-The wrappers are handled efficiently by \SP{}'s vtables~(\cref{s:vtable}).
+The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).
 
@@
 Example types include \code{int64}, \code{uint64}, \code{double}.
-\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
+\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives.
 
@@
 Primitive types cannot be implicitly cast to non-primitive types,
-including the dynamic type (another violation of the static gradual guarantee).
-Operations such as \code{+} do not accept a mix of primitive and non-primitive data.
+including the dynamic type---another violation of the static gradual guarantee.
+Operations such as \code{and} (boolean conjunction) do not accept a mix of primitive and non-primitive data.
 These restrictions give programmers a static separation between fast-running
@@
 Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
+%% \footnote{In other words, \SP{} provides \emph{immediate accountability}~\cite{mt-oopsla-2017}.}
 
@@
 \SP{} even used Cython at some point and found the mixed-workflow codebase painful.
+%%  \item
+%%    Prior to investing in \SP{}, Instagram used Cython to compile
+%%    performance-critical code to C extension modules.
+%%    Cython performed well, but mixing pure Python with a few compiled modules
+%%    made for an awkward developer experience.
+%%    \SP{} aims to match the Python workflow.
 

### 2022-01-13 pj-7.1/paper.tex
pj: sec 3 pass

@@
 CPython 3.8 in several ways to take full advantage of static types.
-Cinder includes tailored bytecode instructions, virtual method tables, new
-datatypes, a registry of typed modules, and a JIT compiler.
-
-The key ingredients in Cinder may be of interest to other teams seeking to add
-sound gradual typing to a widely-used language.
-For example, a team of Node developers may wish to fork that runtime system
-and experiment with bytecode instructions that rely on statci types.
+Cinder includes tailored bytecode instructions, virtual method tables,
+concrete datatypes, a registry of typed modules, and a JIT compiler.
+
+These main ingredients of Cinder may be of interest to other teams seeking to
+add sound gradual typing to an existing language.
+For example, Node developers may wish to fork that runtime and experiment with
+bytecode instructions that leverage sound static types.
 
@@
 All the standard Python instructions work similarly to Python 3.8;
-in other words, an untyped Python program has the same semantics in
+in other words, an untyped Python program has the same behavior in
 Cinder as in the standard interpreter.
-The added instructions help either to express runtime checks or to
-efficiently perform some action.
-For example, the \code{FAST\_LEN} instruction quickly computes the
-length of a built-in value.
-%% TODO is it really undefined
+The added instructions help with one of three tasks:
+expressing runtime checks,
+initializing concrete-typed values,
+or efficiently performing a standard action.
+As an example of the third kind, the \code{FAST\_LEN} instruction quickly
+computes the length of a built-in value.
 \SP{} uses type information to decide where this instruction is appropriate
@@
 dynamic lookup.
-
-Specifically, calls to static methods that appear in statically-typed
+Calls to static methods that appear in statically-typed
 code use the vtable to find an address for the method.
@@
 to a direct function-call jump.
-JIT compilation can also upgrade vtable lookups to function calls.
-
-%% related bytecodes: INVOKE_METHOD INVOKE_FUNCTION
-%%  JIT can upgrade I_METH to I_FUN
-%%  final classes get I_FUN
-%% implementation: built on Python vectorcalls, so it's not totally new
+(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
+instructions.)
+Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
+function calls.
+
+The implementation of vtables happens to be built on the Python \code{vectorcall}
+protocol; it is not a from-scratch development.
+That being said, Cinder vtables are specialized to check untyped overrides
+of typed methods against their expected return type (instead of using a layer of
+indirection for the check).
+
 %% beware: with vtables, method resolution is different than Python b/c args. get
 %%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
-%%
+
 %% > This difference in behavior isn’t so much desired as just a consequence of making
@@
 
-As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
-come with both a type and an implementation.
-The implementation provides the same interface as the built-in,
-but uses the type to reject some uses.
-
-For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
-with keys of type \code{K} and values of type \code{V}.
+The concrete versions of built-in data structures come with both a type and an
+implementation.
+The implementation provides the same interface as the built-in but uses a
+type tag to reject certain operations.
+
+For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
+with keys of type $K$ and values of type $V$.
 The implementation has three main components:
@@
   \item
-    A constructor that uses a type and a Python dictionary
+    First, the constructor uses a type and a Python dictionary ($\sptrawpydict$)
     to initialize a checked dictionary.
-    The constructor ensures that all values of type \code{ChkDict[K, V]}
+    This constructor ensures that all values of type $\sptchkdict{K}{V}$
     begin with well-typed elements.
   \item
-    Guarded update functions.
+    Second, all write operations check their input.
     Every operation that mutates or extends a checked dictionary must
-    validate untyped arguments.
-  \item
-    A type tag for boundary checks.
+    validate any untyped arguments that it receives.
+  \item
+    Third, the dictionary value stores a type tag to support casts from the dynamic type.
     When a checked dictionary enters typed code from an untyped context,
-    the runtime checks its key and value types for an exact match.
-    For example, a value of type \code{ChkDict[K, V]} matches the type
-    \code{ChkDict[str, dynamic]} only if \code{K} is type \code{str}
-    and \code{V} is dynamic.
+    the runtime checks its type tag to see whether the key and value types are an exact match.
+    For example, type $\sptchkdict{K}{V}$ matches the type
+    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
+    and $V$ is equal to $\sptdyn$,
 \end{itemize}
@@
 patches to mutable data.
-Overall, it is an essential piece of infrastructure to enable soundness and optimizations.
 
 \subsection{Method-at-a-Time JIT}
-
-FILL Cinder includes a JIT compiler for its bytecode.
-The typical way to configure is to supply a list of functions to be JIT compiled.
-Compilation works on any \SP{} function whether or not it is typed
-(thus there is some incentive to use \SP{} even without adding type annotations).
-For the biggest payoff, though, the JIT needs type information.
+\label{s:jit}
+
+Cinder includes a JIT (just-in-time) compiler for its bytecode.
+At the moment, programmers enable the JIT by supplying a list of method names
+to the compiler.
+Any method can be JIT-compiled whether or not it uses \SP{} types,
+though types are needed to enable the most-effective optimizations.
 

### 2022-01-13 pj-7.1/paper.tex
pj: abstract / sec 1, mention property-based tests

@@
   On a more technical level, this paper contributes two artifacts:
-  a model of \SP{} that passes hundreds of tests from the \SP{} codebase,
+  a model of \SP{} that passes property-based tests and hundreds of tests from the \SP{} codebase,
   and a performance evaluation of \SP{}. The process of formalization
@@
   % or results and evaluation support this work?
-  Our model of \SP{} is implemented in PLT Redex and tested against
-  the \SP{} regression suite.
+  Our model of \SP{} is implemented in PLT Redex and tested using
+  property-based soundness tests and the \SP{} regression suite.
   This paper includes a small core of the model to convey the main ideas
@@
     To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
-    and ran over 200 tests adapted from the \SP{} regression suite.
+    and ran both property-based soundness tests and over 200 tests adapted from the \SP{} regression suite.
     The model brings across the essence of \SP{} and lends support to the claim

### 2022-01-13 pj-7.1/paper.tex
pj: sec 4 pass

@@
 \newcommand{\spf}{\code{f}}
-\newcommand{\spc}{\code{c}}
+\newcommand{\spc}{\sptclass}
 \newcommand{\spprog}{\codefont{prog}}
@@
 \newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
+\newcommand{\wtsub}{\vdash}
 \newcommand{\spsubt}{\mathrel{<:}}
@@
 \newcommand{\mffont}[1]{\mathit{#1}}
-\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
+\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}} % {\mfapply{\mffont{T}_{\downarrow}\!\mffont{S}}{#1}}
 \newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
@@
 \label{s:model}
-%% purpose = types, checks, and optimizations can be formalized and analyzed
-
-%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least ... just add a Gamma, follow chain up one step at a time until refl.
-%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)
-
-%% NOTE
-%% - remark: model gives errors of the same kind in SP, but not always the same
-%%   message, after all its not our goal to reproduce SP exactly
 
@@
 
-To validate the design of \SP{}, we have developed a mechanized model using PLT
+To validate the design of \SP{}, we have developed a model of the language in PLT
 Redex~\cite{kcdeffmrtf-popl-2012}.
@@
   \item
-    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
-    with untyped code do not undermine the validity of types.
+    \SP{}'s overall approach to gradual interactions is sound; in particular,
+    interactions with untyped code do not undermine the validity of types.
   \item
@@
 \end{enumerate}
-This section explores the first claim in depth by way of a tiny formal model
+This section explores the first claim in depth by way of a small formalization
+(as opposed to the Redex \emph{mechanization})
 that illustrates the \SP{} gradual interaction strategy.
@@
 %% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
-The second claim is justified by our Redex development;
-in particular, by its test suite.
-This section describes the Redex development in broad strokes.
+The second claim is justified by our Redex development and its test suite~(\cref{s:redex}).
 
 \subsection{Surface Syntax and Types}
-%% TODO assume well-formed; no circular defs (x:t = x)
 
@@
   \\
-    \spx, \spf, \spc & \defeq & \mbox{variable names}
+    \spx, \spf & \defeq & \mbox{variable names}
   \end{langarray}
@@
 
-The \SP{} surface syntax is the same as Python 3.8.
-Programmers may write type annotations using the PEP 484 syntax;
-unannotated code gets the dynamic type by default.
-The available static types provide a nominal type system with first-order functions.
-These surface types are meant to be descriptive.
-By design, they are not fully sound.
-For this reason, we omit the surface typing judgment and merely present their
-declarative syntax.
-%% TODO awkward
-
-\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
-using a tiny grammar.
+\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
 A program is a sequence of statements;
@@
 The basic values consist of the following: the none value, integers,
-booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
+booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.\footnote{By contrast to
+Python, in which \code{int} is a type and \code{None} is a value, our formalization uses lowercase names for
+values and capitalized names for types.}
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
@@
 user-defined class, and union types.
-We assume that all unions are written a concise normal form, e.g., that
+We assume that all unions are written in a tidy normal form, e.g., that
 $\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
-There is no function type; the only way to define a higher-order function is to use the dynamic type.
-
-%% FILL generally, no way to express higher-order cast ... either trivial C=C or impossible C=D
-
-Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
+There is no function type in the model, or in \SP{} for that matter.
+The only way to define a higher-order function is to use the dynamic type.
+
+The surface types are meant to describe programs, but they are not designed
+to be fully sound.
+For this reason, we omit the surface typing judgment, which scans for
+logical errors, and merely present the declarative syntax of surface types.
+We also omit a well-formedness judgment that checks for circular definitions.
+
+Relative to \SP{} and our Redex mechanization, the formalization in \cref{f:surface-types} omits
 many details of Python including class variables, imports, conditionals, and exception handlers.
-These details are crucial in the context of the Redex model, which tests whether \SP{}
+These details are crucial in the mechanization, which tests whether \SP{}
 soundly approximates Python.
@@
     \\
-      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
-    \\
       \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
@@
     \\
+      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
+    \\
       \sptype_0 & \mbox{otherwise}
@@
 
-Evaluation type $\spteval$ are a subset of the surface types.
+Evaluation types $\spteval$ are a subset of the surface types.
 These are the types that \SP{} promises to soundly enforce.
@@
 and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
-As the retraction makes clear, the evaluation types make two main changes:
+As the retraction shows, the evaluation types make two main changes:
 \begin{enumerate}
@@
 
-The \SP{} team chose these particular evaluation types because they can all
-be enforced fully and efficiently with run-time casts.
-
 \subsubsection{Casts}
 
+\SP{} evaluation types can all be enforced fully and in constant-time
+with casts.
 \Cref{t:cast} describes the casts in detail.
-Most call for a simple tag check, i.e., a Python \code{isinstance} test.
+Most call for a tag check, i.e., a Python \code{isinstance} test.
 The sole exception is optional types, which require a tag check and a test for the none value.
 Even checked dictionaries rely on tag checks.
-A checked dict type with keys $\sptype_0$ and values $\sptype_1$ accepts only
+A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
 checked dict values that were initialized with exactly the same key and value type.
-As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\sptype_0$
-or $\sptype_1$ is the dynamic type.
-
-There are two key observations to make about casts.
-First, all allowed casts can run in constant time.
+As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\spteval_0$
+or $\spteval_1$ is the dynamic type.
+
+As promised, all casts run in constant time.
 No cast requires traversing a data structure.
 Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
-This latter property means that blame-tracking is trivial;
-casts succeed or fail immediately.
-
-Second, the strict cast for checked dictionaries makes type compatibility simple in \SP{}.
-Values that are partly-dynamic are incompatible with static types.
+This latter property means that blame-tracking is trivial because
+casts either succeed or fail immediately.
+%% In other words, \SP{} provides \emph{immediate
+%% accountability}~\cite{mt-oopsla-2017} for cast errors.
+
+The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
+Values that are partly-dynamic are simply incompatible with static types.
 For example, suppose that a function expects a list of numbers.
@@
 In \SP{}, a list of dynamic values is not a valid input.
-Consequently, \SP{} can use a simple ahead-of-time strategy to enforce soundness.
+Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
 Any position where the dynamic type meets a static type needs a cast.
 Other meetings between types can be resolved statically.
+\SP{} does not need to worry that a list of dynamic values might be masquerading
+under a fully-static type.
 
@@
 Relative to the current environment, the expressions within each statement must
-satisfy the judgment in~\cref{f:eval-types}
+satisfy the typing judgment in~\cref{f:eval-types}
 
 The first two rules are for function application.
-A typed function may be applied an argument that matches its domain type,
+A typed function may be applied to an argument that matches its domain type,
 in which case it computes a value that matches its codomain type.
@@
 The next rules illustrate writes to Python dictionaries and checked dictionaries.
-A Python dict may be updated with any kind of key and value.
-By contrast, a checked dict requires keys and values that match its type parameters.
+A shallow Python dict may be updated with any kind of key and value.
+By contrast, a concrete checked dict requires keys and values that match its type parameters.
 Next we have two rules for classes:
@@
 Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
-related by static subtyping ($\spteval_0 \spsubteq \spteval_1$) or if
+related by static subtyping ($\spsubteq$) or if
 $\spteval_1$ is the dynamic type.
@@
 Combined with the typing rule, however, materialization is a concise way to find
-where a well-typed program needs to insert casts to ensure soundness.
+where a well-typed program needs casts to ensure soundness.
 For example, suppose that $\spf$ is a function from integers to the dynamic
 type and that $\spx$ is a variable with the dynamic type.
-An application $\spapp{\spf}{\spx}$ can satisfy type $\sptfloat$ using two materializations:
+An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptfloat$ using two materializations:
 
@@
 After all, the main benefit of a full formal semantics is to validate
-the behavior of complex expressions---a job
-better left to the mechanized Redex model.
+the behavior of complex expressions---and this job is
+better left to the Redex mechanization.
 
 Furthermore, our casts-via-materialization rule is more aggressive
-than what \SP{} actually does to insert casts because
-\SP{} supports interactions with plain Python code.
-Because it chooses not to recompile Python modules, \SP{} inserts
+than what \SP{} actually does to insert casts.
+Because \SP{} supports interactions with unanalyzed Python modules, it inserts
 casts in an overapproximate way:
@@
   \item
-    typed code checks the results of function calls, references, etc. whenever there is a materialization from the dynamic type.
+    typed code checks the results of function calls, references, and other elimination forms
+    whenever there is a materialization from the dynamic type.
 \end{itemize}
 \Cref{s:impl} explains
-how the type-directed optimizer avoids some
+how the type-directed optimizer helps avoid some
 of these checks.
 
-\begin{figure}[t]
+\begin{figure}[p]
   \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
@@
       \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
-      \\
-      \spteval_1 \spcompat \spteval_0
+      \\\\
+      \sptenv_0 \wtsub \spteval_1 \spcompat \spteval_0
     }{
@@
       \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
-      \\
+      \\\\
       \spteval_1 \spmatr \spteval_0
@@
 
-  \begin{minipage}[t]{0.5\columnwidth}
-    \fbox{$\spteval \spcompat \spteval$}
+  \begin{minipage}[t]{0.6\columnwidth}
+    \fbox{$\sptenv \wtsub \spteval \spcompat \spteval$}
     \begin{mathpar}
       \inferrule*{
-        \spteval_0 \spsubt \spteval_1
+        \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
       }{
-        \spteval_0 \spcompat \spteval_1
+        \sptenv_0 \wtsub \spteval_0 \spcompat \spteval_1
       }
@@
       }{
-        \spteval_0 \spcompat \sptdyn
+        \sptenv_0 \wtsub \spteval_0 \spcompat \sptdyn
       }
@@
 
-  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
+  \end{minipage}\begin{minipage}[t]{0.4\columnwidth}
     \fbox{$\spteval \spmatr \spteval$}
@@
   \medskip
-  \fbox{$\spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
+  \fbox{$\sptenv \wtsub \spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
   \begin{mathpar}
@@
     }{
-      \spteval_0 \spsubt \sptobject
+      \sptenv_0 \wtsub \spteval_0 \spsubt \sptobject
     }
@@
     }{
-      \sptbool \spsubt \sptint
+      \sptenv_0 \wtsub \sptbool \spsubt \sptint
     }
@@
     \inferrule*{
-      \spteval_0 \spsubt \spteval_1
+      \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
     }{
-      \spteval_0 \spsubt \sptoptional{\spteval_1}
+      \sptenv_0 \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
+    }
+
+    \inferrule*{
+      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
+      \\
+      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
+    }{
+      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
     }
@@
 No matter how an expression mixes typed and untyped code, its static
-\emph{evaluation} type should be an accurate prediction about the results
+\emph{evaluation type} should be an accurate prediction about the results
 of evaluation.
@@
     \item
+      \(\spexpr\) reduces to an allowed error
+    \item
       \(\spexpr\) diverges
-    \item
-      \(\spexpr\) reduces to an allowed error
   \end{itemize}
@@
 As mentioned above, our argument for soundness has two parts
-and the question at hand is whether the boundaries to type dynamic
+The question at hand is whether the boundaries to type dynamic
 are adequately protected with run-time checks.
-The Redex model's comprehensive test suite provides the other half of the
+The Redex mechanization's comprehensive test suite provides the other half of the
 argument; namely, that \SP{} is a sound type system for Python.
@@
 Alternatively, observe that the boundaries in \SP{} are a subset
-of the boundaries in Nom, a gradual language that comes with a detailed
+of the possible boundaries in Nom, a gradual language that comes with a detailed
 proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
-Given an \SP{} program, it is possible to derive a Nom program with
+Given an \SP{} program, we can derive a Nom program with
 the same type boundaries by replacing every checked dict instance
 with a fresh class type.
-Because different checked dict instantiations are incompatible with
-one another, the \SP{} boundaries impose the same constraints as the Nom
+Because distinct checked dict instantiations are incompatible,
+the \SP{} boundaries impose the same constraints as the translated Nom
 boundaries.
@@
 \subsection{Redex Development, \SP{} Conformance}
+\label{s:redex}
 %% pbt for soundness?
 
-By contrast to the small model language presented in this section,
-our Redex model covers both the typed--untyped boundaries in \SP{}
-and a large subset of Python.
+By contrast to the small formalization presented in this section,
+our Redex mechanization covers both type boundaries and a large subset of Python.
 Notable Python features include module imports, loops, exception handlers, and
@@
 model includes hundreds of test programs adapted from the \SP{} test suite.\footnote{
-  Most of the N tests are direct translations of \SP{} conformance tests to
+  Most of the 214 tests are direct translations of \SP{} conformance tests to
   the syntax of the model.
   A program that implements the translation is included with our Redex development.
-  The other N tests are not directly expressible in the model; for these, we
+  The other 52 tests are not directly expressible in the model; for these, we
   translated salient aspects of each test by hand.}
@@
 Some tests intentionally raise type errors, some raise exceptions, and the rest
-compute a specific result.
-We have additionally verified that the model and \SP{} agree on these end-to-end
-behaviors.
-Thus we believe that \SP{} is sound as well.
-%% ... mention that SP is a large and hard-to-analyze language ... opaque?
-
-The model intentionally does, however, not cover all of Python.
+compute a specific result; \SP{} agrees on these end-to-end behaviors.
+We have additional applied Redex's random testing tools~\cite{kf-sfp-2009}
+to validate soundness.
+%% FILL how many random tests?
+%% FILL what bugs did random catch?
+
+The model intentionally does not cover all of Python.
 Some aspects of \SP{} are left out because they are straightforward to
-handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
+handle soundly.
+These include primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}.
 Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
@@
 Validating the correctness of optimizations is a separate task that requires
-significant components; namely, a model of the \SP{} bytecode and a faithful
+significant components: a model of the \SP{} bytecode and a faithful
 rendering of its type-based transformations.
 Our model establishes soundness as a baseline.
-Future work can use soundness to reason informally about optimizations,
-or even take it as a starting point for compiler verification.
+Future work can use soundness as a starting point for compiler verification,
+or informally leverage soundness to reason about possible optimizations.
 
@@
 The remaining 16 issues report bugs in language design and implementation.
-The \SP{} team has acknowledged these bugs as such via GitHub a label (\code{sp-correctness}).
-
-In addition to these formal issue reports,
-we had several long discussions together (between Brown and Instagram)
-about finer points in the language design.
+The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label (\code{sp-correctness}).
+
+% In addition to these formal issue reports,
+% we had several long discussions together (between Brown and Instagram)
+% about finer points in the language design.
 
@@
 \begin{table}
-  \caption{Cinder bytecode instructions}
+  \caption{Representative bytecode instructions}
   \label{t:bytecode}
@@
 %%    \SP{} aims to match the Python workflow.
+
+optimized transient, sp suggests method-at-a-time equally effective as tracing
 

### 2022-01-13 pj-7.1/paper.tex
pj: sec 4, reword para on not-modeled features

@@
 are also omitted.
-Other Python features are left out because their \SP{} semantics is identical
-to the baseline Python semantics.
-These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
-see \cref{s:impl} for further discussion.
-Lastly, the model does not include type-directed optimizations.
-Validating the correctness of optimizations is a separate task that requires
-significant components: a model of the \SP{} bytecode and a faithful
-rendering of its type-based transformations.
+Other Python features are left out either because their \SP{} semantics is identical
+to the baseline Python semantics or because they would entail a separate project~(\cref{s:impl}).
+In particular, modelling type-directed optimization would require substantial
+components: a model of the \SP{} bytecode and a faithful rendering of its
+transformations.
 Our model establishes soundness as a baseline.

### 2022-01-14 pj-7.1/paper.tex
pj: sec 5 edits

@@
 
-%% TODO unions tracked statically in SP, erased to dyn unless narrowed
-
 %% TODO SP side channels:
@@
 
-%% TODO built-in vs. user defined ... 
-%% Carl: I don't think this is quite true the way it's phrased here? in general
-%%   for SP classes we wrap non-static subclass override methods at runtime (in our
-%%   vtable implementation) to enforce correct return type, so that we don't need
-%%   exact types, we can trust that Liskov is not broken.
-%%     The cases where we tend to be limited to exact types is in optimizing
-%%   operations on builtin types, where they aren't really static types and
-%%   methods called on them aren't going through SP vtables, so we don't have
-%%   the vtable wrappers to enforce LSP, we just know the behavior of the exact
-%%   builtin type
-
 This section explains significant aspects of \SP{} that lie outside the
 scope of our model.
-These aspects include sound interactions with Python code,
+These include sound interactions with Python code,
 the bytecode language and optimizations, and dynamic Python features
-that \SP{} does not yet ascribe types to.
-
-\subsection{Open-World Python Interactions}
+that \SP{} does not type.
+
+\subsection{Interactions with Open-World Python Code}
 
 Although \SP{} is technically a new language, it is designed for gradual adoption.
-Python programmers should be able to add types one module at a time to an
+Python programmers should be able to add types module-by-module to an
 existing codebase.
 Consequently, \SP{} supports interactions with Python modules in the only viable
-way: by letting the Python code run with minimal constraints.
-After all, if programmers need to fix every module in a codebase before they can
-run tests after typing a first module, then the gradual story is lost.
-
-Interactions with open-world~\cite{vss-popl-2017} Python code pose a minor
-threat to soundness.
-Typed functions (and methods) cannot assume that their arguments are
+way: by letting the Python code run with zero static constraints and minimal
+dynamic constraints.
+
+There are two case in which \SP{} types add dynamic constraints to Python code.
+First, concrete generic types (e.g. $\sptchkdict{\sptype}{\sptype}$)
+reject inputs that were not created via a checked contructor.
+If a client of a Python module decides to impose concrete types, the Python module
+may need to change.
+Second, the strict setting for modules prevents updates to module-level variables.
+If existing code mutates a binding in a strict module, a run-time error will
+occur.
+
+Interactions with \emph{open-world}~\cite{vss-popl-2017} Python code pose a minor
+threat to soundness because
+typed functions (and methods) cannot assume that their arguments are
 well-typed.
-Arguments from typed contexts can be validated by either static checks
-or materialization casts, but arguments from untyped contexts are unchecked.
+All arguments sent from typed contexts get validated either statically or
+via materialization casts, but arguments from untyped contexts are unchecked.
 For this reason, \SP{} compiles every typed function to check its inputs.
-\Cref{s:optimize} explains how the optimizer avoids some of these checks.
-%% 2022-01-09 worth mentioning primitive types? they get cast like anything else
-%% ... I guess the trick is that primitives don't materialize in typed code
-
-FILL sound generics / checked types also need to guard their writes
-
-There are two cases, however, where new \SP{} types do add constraints to
-existing Python code.
-First, sound generic types (e.g. $\sptchkdict{\sptype_0}{\sptype_1}$)
-force value-suppliers to use a checked constructor.
-Existing code can \emph{use}\/ a checked generic via the same API as on
-un-checked one, but can create new values only using the \SP{} constructors.
-Second, the strict setting for modules prevents updates to module-level variables.
-If existing code happens to mutate a member of a newly-strict module, a run-time
-error will occur.
+Similarly, concrete types such as $\sptchkdict{\sptype}{\sptype}$ also
+check their writes.
+%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
+%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness
 
@@
 
-\SP{} can generate more-efficent code than Python because it targets
+\SP{} can generate efficient code because it targets
 the Cinder runtime.
-Cinder extends Python with bytecode instructions to check types at run-time,
-to construct \SP{} data structures, and to perform optimized actions.
-\Cref{t:bytecode} lists a few representative bytecode instructions.
+Cinder extends the Python 3.8 bytecode with instructions to check types at run-time,
+to construct \SP{} data structures, and to perform actions in a way that benefits from
+static information..
 
 \begin{table}
-  \caption{Representative bytecode instructions}
+  \caption{Representative Cinder bytecode instructions}
   \label{t:bytecode}
@@
 
+\Cref{t:bytecode} lists a few representative bytecode instructions.
 The two instructions that express run-time checks are \code{CAST}
@@
 
-\Cref{t:bytecode} presents two instructions that allocate \SP{}-specific data structures.
+Two of the listed instructions allocate \SP{}-specific data structures.
 The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
 Cinder comes with a similar instruction to build checked lists.
-The developers will add more instructions along the same lines in the future.
+%% The developers will add more instructions along the same lines in the future.
 
@@
 The former jumps to a statically-determined address;
-the latter uses a v-table lookup to quickly find an address at run-time.
+the latter uses a vtable lookup to quickly find an address at run-time.
 The load and store instructions for fields improve upon Python's generic attribute lookup.
@@
 type flows into a static type.
-The materialze rule in the model illustrates this policy.
+The materialize rule in the model illustrates this policy.
 Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
-that casts all arguments.
+that defensively casts all arguments.
 By convention, this instruction always appears on the first line of a typed
@@
 The optimizer uses this convention to skip argument checks by jumping past them
-when it is safe to do so; namely, whenever a typed function (or method) gets
+when it is safe to do so; namely, whenever a typed function or method gets
 called in a typed context.
-In this way, all untyped calls get the necessary argument check and all typed
-calls benefit from materialize checks at the call-site.
 
@@
 Python code.
-For each, we claim that the dynamic type is a reasonable choice;
-accurate static types would be difficult to maintain.
+For each, the dynamic type is a very reasonable choice;
+accurate static types would be burdensome to maintain.
 
@@
 \SP{} does not attempt to type first-class classes: partly because they have
-yet to appear in performance-critical code at Instagram and partly because it
+yet to appear in performance-critical code and partly because it
 is unclear how to incorporate them into the nominal type system.
@@
 structural objects.
-Adding an interface layer to \SP{} is an open question.
+How best to add an interface layer in \SP{} is an open question.
 
@@
 Python allows classes to inherit from a list of parents.
-When resolving a method call to such a class, Python
+To resolve method calls to such a class, Python
 dynamically traverses the parent list in a fixed order
@@
 Method calls to these classes execute the same way as
-in Python, with no type-directed optimizations to speed up dispatch.
+in Python, with no type-directed optimizations to improve dispatch.
 
 \paragraph{Dynamic Execution}
-%% TODO is eval a side channel, or safe?
-%% TODO check that eval/exec args really are unoptimized / sound etc.
+
+%% 2022-01-13: __setattr__ is along the same lines
+%%Carl: Removing the use of super-dynamic Python features like
+%%\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
+%%we’ve done this it hasn’t required a major rewrite.
+
 Results computed by calls to \code{eval} and \code{exec}
 have the dynamic type.
-Their inputs run without being rewritten
-by the optimizer.
+Their inputs run without being rewritten by the optimizer.
 
@@
 can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
-Assuming these findings carry over to Python, we recommend either similar
-adjustments or the introduction of semantically-descriptive replacements
-over attempting to type eval.
-
-\paragraph{Module-Level Variables}
-
-Any Python module can read and write to the module-level variables of another module.
-When compiling a module, \SP{} therefore assumes that its module-level variables
-may be modified by untyped code and assigns the dynamic type.
-
-Cinder offers \emph{strict modules} as an alternative to the Python
-semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
-If a programmer chooses to declare a module as strict, then its module-level varibles
-are immutable and thus typeable.
-
-Another potential solution is for the Cinder runtime to check that writes to module-level variables
-preserve their types.
-There are two downsides to this idea:
-it will add some performance overhead,
-and it will force developers to rewrite untyped code in order to fix any errors that arise.
-%% Because of the latter concern, type dynamic is a very reasonable default.
-
-\paragraph{Set attribute}
-
-Carl: Removing the use of super-dynamic Python features like
-\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
-we’ve done this it hasn’t required a major rewrite.
+Assuming these findings carry over to Python, similar adjustments would
+preferable over attempting to type eval.
+
+%% 2022-01-13 covered pretty well earlier
+%%\paragraph{Module-Level Variables}
+%%
+%%Any Python module can read and write to the module-level variables of another module.
+%%When compiling a module, \SP{} therefore assumes that its module-level variables
+%%may be modified by untyped code and assigns the dynamic type.
+%%
+%%Strict modules as an alternative to the Python
+%%semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
+%%If a programmer chooses to declare a module as strict, then its module-level varibles
+%%are immutable and thus typeable.
+%%
+%%Another potential solution is for the Cinder runtime to check that writes to module-level variables
+%%preserve their types.
+%%There are two downsides to this idea:
+%%it will add some performance overhead,
+%%and it will force developers to rewrite untyped code in order to fix any errors that arise.
+%%%% Because of the latter concern, type dynamic is a very reasonable default.
 
@@
 
-\begin{table}
-  \caption{Overview: \SP{} in production}
-  \label{t:prod-stat}
-  \begin{tabular}{l@{~~}ll@{~~}l}
-    8     & months in production & 3.7\% & global CPU improvement \\
-    530   & modules converted  &  0 & performance regressions
-    %% TODO 530 / ???? how much of codebase?
-  \end{tabular}
-\end{table}
-
 The Instagram web server has been running \SP{} in production since April 2021.
@@
 %% 2022-01-06 FILL team did significant work to fight regressions
+
+\begin{table}
+  \caption{Overview: \SP{} in production}
+  \label{t:prod-stat}
+  \begin{tabular}{l@{~~}ll@{~~}l}
+    8     & months in production & 3.7\% & global CPU improvement \\
+    530   & modules converted  &  0 & performance regressions
+    %% TODO 530 / ???? how much of codebase?
+  \end{tabular}
+\end{table}
 

### 2022-01-14 pj-7.1/paper.tex
pj: sec 6 edit

@@
 \label{s:eval}
-%% 2021-12-15: C&D may have an SP off/on switch soon to measure totals
-
-%% 2022-01-12
-%% We did have some cases where strict semantics caused a problem, but in the
-%% majority of conversions it didn’t (or if it did initially, the fix was
-%% trivial.) There was at least one case where I would have converted more code to
-%% static but couldn’t easily do so because of strictness issues with a larger
-%% framework reliant on metaclasses, and so I instead extracted the hottest code
-%% paths into a new strict/static module and called them. I have a plan to go back
-%% to this and improve the framework itself to be strict so more code can be
-%% converted. I know of at least one other case where we did modify a framework
-%% that relied on side-effecting decorators so that it could be used without the
-%% side effects.
-
-%% 2022-01-12 All SP mods are strict. Although the Static Python compiler can technically be
-%% used apart from strict modules, and we do that in tests for faster tests, it’s
-%% not sound and we don’t allow it in prod. The way a real module gets compiled as
-%% Static Python is in fact via the strict modules import loader, which detects
-%% both `import __strict__` and `import __static__` markers, treating the former
-%% as strict only and the latter as both strict and static. So we don’t provide
-%% any straightforward way (either in IG codebase or generally with Cinder) to
-%% have a module imported as static without also being strict.
-
-%% 2022-01-12 Looks like 3 use CheckedDict and 6 use CheckedList.
-%% 2022-01-12
-
-%% 2022-01-11 TODO how many Checked modules in production?
-%% 2022-01-11 TODO how many strict modules in production?
-%% 2022-01-11 TODO talk about primitive types:
-%% Carl: "I just did a very rough grep check and found that we have only 12 modules
-%% currently using primitive types: all int32 and int64 and cbool. So they are
-%% used pretty sparingly, but they are also definitely critical to our ability to
-%% match Cython in a few of the  hottest code paths that used to be Cython."
-
-%% 2022-01-05 TODO We are currently JITting at least some functions in 873
-%% modules, but we are about to increase this by about an order of magnitude,
-%% independent of Static Python.
 
@@
 %% performance."
-
-%% 2022-01-04 TODO ask C&D how many teams have adopted SP ... get sense of adoption
-
-% "The workload for those rough percentage numbers is “Instagram web server
-%  production.” We have pretty extensive production profiling that lets us
-%  identify CPU savings with good granularity when a change lands in prod, and the
-%  observed CPU savings for a variety of staticification changes we’ve landed over
-%  the last half sum up to 2% of total IG prod CPU usage." --Carl
 
@@
 
-%% TODO think about it 2021-12-29 Ben
-%% > + Is there any sense in which the 530 converted modules are a big
-%% > sample? Maybe not, since as you say 530 is tiny for the codebase as a
-%% > whole. But do they exercise all sorts of behavior? Do they form a big
-%% > sample of certain parts of the Instagram codebase?
-
 The Instagram web server has been running \SP{} in production since April 2021.
@@
 Over 500 modules have been converted thus far.
-None of the converted modules have gotten slower because of \SP{}
-
-\subsection{Migration Path}
-
-\SP{} first entered the Instagram web server as a replacement for Cython,
-which Instagram had been using to speed up critical modules.
-Cython performed well, but its partial adoption led to an awkward workflow
-because unlike standard Python, Cython is a compiled language.
-Replacing these modules with \SP{} let developers return to a conventional
-Python workflow.
-\SP{} (combined with Cinder) did not slow down any of these
-performance-critical modules, and even sped a few up.
-The net improvement was a 0.7\% reduction in global CPU usage.
-
-Later \SP{} migrations have been led by profiling to find frequently-executed code.
-The goal of these opportunistic migrations was to test whether \SP{} could
-improve typical modules without slowing down their neighbors.
-As a point of comparison, Typed Racket can improve individual Racket
-modules but the type checks that it requires on module boundaries
-can lead to huge slowdowns~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}.
-\SP{} does not appear to suffer in the same way.
-
-During the first half of 2021, the \SP{} team identified critical
-modules and proposed types to the maintainers of these modules.
-Often, the types came with small code changes.
-The accepted types and changes resulted in a 1\% improvement
-in global CPU usage.
-During the second half of 2021, the \SP{} team applied the same
-process at a larger scale.
-They also changed a code-generating module to output \SP{} code instead
-of untyped Python code;
-this one change brought over four hundred generated modules into the \SP{} fold.
-Overall, these changes resulted in a 2\% global CPU improvement.
-
-As of December 2021, the \SP{} team has converted 541 modules.
-Most of these came from the tool (417); the rest
-are from hand conversions (124).
-These modules are well-mixed with untyped modules in the codebase.
-All together, over thirty thousand identifiers cross a type boundary
-at compile-time by either entering or leaving a \SP{} module.
-Two-thirds of these crossings are exports from \SP{} modules
-to untyped modules;
-in other words, the typed identifiers are widely used throughout the Instagram web server.
-%% data in src/static-python-imports-analysis
-%% 2022-01-06 TODO gen-code is configuration system / access control ... what are stats omitting these?!
-
-In a typical week, the team adds types to four additional modules.
-These modules have seen speedups ranging from 0\% to 25\% depending
-on their nature.
-Regressions have not been an issue.
-The only cause for alarm is when a module gets only a small improvement
-without code changes.
-%% - rough measure: 0 to 25% per module
-%%   ... but depends on nature of module
-%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls
-%% 2022-01-06 FILL team did significant work to fight regressions
+Despite some initial regressions, none of the converted modules
+ran slower after small rewrites and enabling the JIT on certain functions.
 
@@
   \label{t:prod-stat}
-  \begin{tabular}{l@{~~}ll@{~~}l}
-    8     & months in production & 3.7\% & global CPU improvement \\
-    530   & modules converted  &  0 & performance regressions
-    %% TODO 530 / ???? how much of codebase?
+  \begin{tabular}{l@{~~}l@{\qquad}l@{~~}l}
+    %% 541 modules out of ~45000 total
+    8     & months in production   & 541   & typed, strict modules \\
+    3.7\% & global CPU improvement &   9   & using concrete types \\
+    0     & net regressions        &  12   & using primitive types \\
+    %% all prims are int32 and int64 and cbool
+    %% 873 modules use JIT for some functions (about to increase by OOM)
   \end{tabular}
@@
 
+\subsection{Migration Path}
+
+\SP{} first entered the web server codebase as a replacement for a
+few critical Cython modules.
+Cython had improved performance by compiling these modules to C,
+but its partial adoption had led to an awkward workflow (with a
+few compiled modules spread across an interpreted codebase).
+Replacing these modules with \SP{} let programmers return to a conventional
+Python workflow.
+Furthermore, \SP{} types combined with Cinder resulted in a
+0.7\% reduction in global CPU usage.
+Primitive types~(\cref{s:c-types}) were essential for matching Cython.
+
+Later \SP{} migrations have been directed by profiling to find frequently-executed code.
+During the first half of 2021, the \SP{} team identified critical
+modules and proposed types plus small code changes to the maintainers of these modules.
+The accepted proposals resulted in a 1\% improvement in global CPU usage.
+During the second half of 2021, the \SP{} team applied the same
+process at a larger scale.
+They also modified a code-generating module to output \SP{} code;
+this one change brought over four hundred generated modules into the typed fold.
+Overall, the second-half changes added a 2\% global CPU improvement.
+
+As of December 2021, the \SP{} team has converted 541 modules.
+Most of these came from the code-generating tool (417); the rest
+are from hand conversions (124).
+These modules are well-mixed with untyped modules in the codebase.
+According to an analysis of identifiers that cross module-dependence
+boundaries, over 30,000 identifiers go between typed and untyped code.
+Two-thirds of these crossings are exports from \SP{} modules
+to untyped modules; in other words, typed identifiers are widely used
+throughout the web server.
+%% data in src/static-python-imports-analysis
+%% TODO gotta compare to TR etc.
+
+% In a typical week, the team adds types to four additional modules.
+% These modules have seen speedups ranging from 0\% to 25\% depending
+% on their nature.
+% Regressions have not been an issue.
+% The only cause for alarm is when a module gets only a small improvement
+% without code changes.
+
 \subsection{Analyzing Code Changes}
-%% TODO boring unclear writing
-
-Because the \SP{} team has been changing code as well as types
-during its migrations, the question arises as to whether the code changes
-raise any threats to validity.
+
+Because the \SP{} team has been changing code as well as types during its
+migrations, the question arises as to whether the code changes raise any
+threats to validity.
 It could be that the \SP{} approach to gradual typing only pays off
-after significant modifications, or only for a restrictive style of programming.
-
-To understand the code changes, we reviewed approximately 30 patches that
-significantly improved performance.
+after significant modifications.
+%% Carl: "I reviewed the ~30 conversion diffs"
+We therefore reviewed the $\sim 30$ conversion diffs that significantly improved
+performance.
+
 In general, the patches have minor code changes and most of these changes
 affect tests rather than production code.
-Digging deeper, there are nine common kinds of changes.
-Six of these address type errors; the other three focus on performance.
+Digging deeper, there are nine common kinds of changes: six address
+type errors and the other three are for performance.
 
@@
 
-%% TODO section formatting looks good?
 \subsubsection{Code Changes for the Type Checker}
@@
     \ipara{Fix type errors due to mock wrappers}
-    When test code introduces a mock wrapper, it may change the return type of
-    a function.
+    When test code introduces a mock wrapper, it may inadvertantly change the
+    return type of a function.
     For example, the wrapped version of a function that returns \code{None}
     will return a \code{MagicMock} object unless the programmer explicitly
-    specifies a return value.
+    specifies \code{None} as the desired return value.
     %% Carl: This is an inherent and desired
@@
     \SP{} currently rewrites all function calls to use positional arguments.
-    Mock-wrapped functions therefore need to positional arguments instead of keywords.
+    Mock-wrapped functions therefore need to use positional arguments instead of keywords.
     %% Carl: we plan to fix this incompatibility by falling back to the original call
@@
     Functions that expect only keyword arguments cannot yet be called from \SP{} contexts.
-    These functions need to change.
     %% Carl: We plan to support this next half.
@@
     and classes that are enums.
-    Programmers have to move these classes to a different module for now.
+    For now, programmers have to move these classes to an untyped module.
     %% Carl: We have work in progress on supporting some Enum types, and plan to add a @dataclass intrinsic soon.
@@
     Occurrences of the dynamic Python features listed in \cref{s:dynamic-python} must be rewritten or moved.
-    According to the \SP{} team, these changes have not required any major rewrites.
+    %% According to the \SP{} team, these changes have not required any major rewrites.
 
@@
   \item
+    \ipara{Use primitive types}
+    Any Python integers and booleans in hot code paths are better as
+    \SP{} primitive values~(\cref{s:c-types}).
+    As noted above, primitives were required for \SP{} to match the performance of
+    Cython.
+    Twelve modules currently use primitives.
+    %% - bg: focus on block, convert, no problem?
+
+  \item
+    \ipara{Use concrete types}
+    Three modules currently use a concrete dictionary type
+    (of the form $\sptchkdict{\sptype}{\sptype}$)
+    and six modules use a checked list type.
+    %% - bg: how occasional? worried about power of the "gradual soundness" claim
+
+  \item
     \ipara{Convert \code{@classmethod} to \code{@staticmethod}}
-    Static methods can get invoked directly as functions, bypassing the class.
-    %% Carl: \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}
-
-%% 2022-01-12
+    Static methods can get invoked directly as functions, bypassing the class vtable.
+    Twenty-four modules use static methods, though many of these uses pre-date \SP{}.
+    %% 2022-01-12 Carl:
     %% I don’t think it was a super common change, but there were some cases where we
@@
 
-  \item
-    \ipara{Use primitive types}
-    Any Python integers and booleans in hot code paths are better as
-    \SP{} primitive values~(\cref{s:c-types}).
-    %% - bg: focus on block, convert, no problem?
-
-  \item
-    \ipara{Use Checked Types}
-    Occasionally, converting a Python dictionary or Python list to a
-    checked type has a big impact on performance.
-    As noted in \cref{s:checked-type}, these conversions affect all downstream
-    clients of the data structure.
-    %% - bg: how occasional? worried about power of the "gradual soundness" claim
-
 \end{enumerate}
 
+%% More Migration Notes
+
+%% 2022-01-12 Carl:
+%% We did have some cases where strict semantics caused a problem, but in the
+%% majority of conversions it didn’t (or if it did initially, the fix was
+%% trivial.) There was at least one case where I would have converted more code to
+%% static but couldn’t easily do so because of strictness issues with a larger
+%% framework reliant on metaclasses, and so I instead extracted the hottest code
+%% paths into a new strict/static module and called them. I have a plan to go back
+%% to this and improve the framework itself to be strict so more code can be
+%% converted. I know of at least one other case where we did modify a framework
+%% that relied on side-effecting decorators so that it could be used without the
+%% side effects.
+
 \subsubsection{Developer Experience}
 
-For the most part, developers at Instagram have been happy adopting \SP{}.
-The transition from Pyre to \SP{} has thus far delivered performance benefits
+For the most part, programmers have been happy adopting \SP{}.
+The transition from Pyre to \SP{} has delivered performance benefits
 that outweigh the cost of code changes.
+Even the use of strict modules has been acceptable; only two modules
+required non-trivial rewrites to avoid mutation.
 
 One significant pushback dealt with multiple inheritance.
-The \SP{} team proposed a rewrite from multiple inheritance to single inheritance
-to improve performance and static checks.
-But the code maintainers preferred to keep multiple inheritance.
-Because \SP{} defaults to Python behavior, the maintainers were able
-to keep multiple inheritance alongside \SP{} code.
-
-%% any more?
+The \SP{} team proposed a rewrite from multiple inheritance to single inheritance,
+but the code maintainers preferred to keep the design as-is.
 
@@
 %% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
-
-%% TODO jit vs nojit
-
-%% 2021-12-23: TODO decide whether to ask for typed benchmarks to report ratios
-%%  Carl: we could commit to typed 2--3 more, but pystone etc (?!) are notoriously bad
-
-%% Somehow though, we ought to have public code available.
-
-Prior to deploying \SP{} in production, the team used microbenchmarks to study performance.
-These microbenchmarks are admittedly small and quite different than typical application code.
-That said, their use gave the team a coarse idea of whether \SP{} was on track during development.
-We include these microbenchmarks to offer reproducible performance numbers;
-unlike the Instagram web server, the benchmark code is public.
-
-\Cref{t:microbenchmark} reports current microbenchmark results.
-Each row gives the ratio of two running times: fully-typed \SP{} code versus
-unannotated \SP{} code.
-A ratio between 0 and 1 means that performance improved thanks to the types.
-[FILL] numbers.
+%% https://docs.google.com/spreadsheets/d/1IO25bL_8LwDqx3ciQTj9dMetsFnZFdiQIyjFvgYWIpI/edit?usp=sharing
+
+\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
+These microbenchmarks are admittedly small and quite different than typical application code,
+but they provide an auditable and reproducible way to measure performance.
+Each row presents two ratios.
+The first compares a simply-typed version of the benchmark to untyped Python code.
+The second compares a hand-refined benchmark to untyped Python.
+The refinements are modest, mostly adding primitive types to hot loops;
+they do not require siginificant expertise to reproduce.
+
+All runtimes for this table came from a 64-bit CentOS Linux system using a build
+of \SP{} and Cinder compiled with default settings.
+Each benchmark version ran 11 times total: once to warm up the JIT and 10 times to
+compute an average runtime (using the \code{time} utility).
+Every benchmark version invoked the JIT on the same functions---even the untyped versions.
 
 \begin{table}[t]
-  \caption{Microbenchmark Performance Ratios}
+  \caption{Microbenchmark performance ratios. NB: data for deltablue and nbody is pending.}
   \label{t:microbenchmark}
-  %% show untyped Python time?
-  %% transpose?
-  \begin{tabular}{ll}
-    Name               & Typed / Dynamic \\\midrule
-    \bmname{deltablue} &             0\% \\
-    \bmname{fannkuch}  &             0\% \\
-    \bmname{nbody}     &             0\% \\
-    \bmname{nqueens}   &             0\% \\
-    \bmname{pystone}   &             0\% \\
-    \bmname{richards}  &         0.125\% \\
+  \begin{tabular}{lrr}
+    Name               & Typed / Python & Refined / Python \\\midrule
+    \bmname{deltablue} &            TBD &              TBD \\
+    \bmname{nbody}     &            TBD &              TBD \\
+    \bmname{richards}  &         0.53\% &           0.22\% \\
+    \bmname{fannkuch}  &         1.03\% &           0.46\%
+    %% fannkuch typedbasic2 = 0.90
   \end{tabular}
@@
 
-%% TODO conclusions? threats?
+Almost all ratios are less than one, showing that \SP{} achieves a speedup over
+Python despite the cost of enforcing soundness.
+The one exception is the basic typed version of \bmname{fannkuch}, which sees a minor
+slowdown.
+This benchmark is primarily about number-crunching; and indeed, changing a few simple
+types to primitive types without modifying the code leads to a minor speedup (ratio = $0.90\%$).
 
@@
 
+%% TODO conclusions about SP? threats to validity?
+
 \SP{} is ambitious in many ways:
@@
 
-But \SP{} is also carefully targeted:
+\noindent{}But \SP{} is also carefully targeted:
 \begin{itemize}
@@
 \end{itemize}
+
+\noindent{}Lesson: the migration works, with a dedicated support team.
 

### 2022-01-14 pj-7.1/paper.tex
pj: macro for net CPU change

@@
 \newcommand{\ipara}[1]{\emph{#1}\/.}
+
 \newcommand{\numbenchmark}{2}
+\newcommand{\CPUchange}{3.7\%}
 
@@
   web server.
-  Thanks to \SP{}, global CPU usage has improved by almost 4\% within a year.
+  Thanks to \SP{}, global CPU usage has improved by \CPUchange{} within a year.
 
@@
 and only a few critical modules rely on concrete types and primitive types (dozens),
-global CPU usage has improved by almost 4\% according to internal profiling tools.
+global CPU usage has improved by \CPUchange{} according to internal profiling tools.
 Type-directed optimizations outweigh the cost of enforcing soundness.
@@
 Instagram's internal profiling tools, which continuously monitor
-performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
+performance, attribute a \CPUchange{} improvement in global CPU to \SP{} conversions.
 \Cref{t:prod-stat} reports a few other production statistics.
@@
     8     & months in production   & 541   & typed, strict modules \\
-    3.7\% & global CPU improvement &   9   & using concrete types \\
+    \CPUchange{} & global CPU improvement &   9   & using concrete types \\
     0     & net regressions        &  12   & using primitive types \\

### 2022-01-14 pj-7.1/paper.tex
pj: sec 6, pct -> ratio

@@
     \bmname{nbody}     &          TBD &            TBD \\
-    \bmname{richards}  &         0.53\% &           0.22\% \\
-    \bmname{fannkuch}  &         1.03\% &           0.46\%
+    \bmname{richards}  &         0.53 &           0.22 \\
+    \bmname{fannkuch}  &         1.03 &           0.46
     %% fannkuch typedbasic2 = 0.90
@@
 This benchmark is primarily about number-crunching; and indeed, changing a few simple
-types to primitive types without modifying the code leads to a minor speedup (ratio = $0.90\%$).
+types to primitive types without modifying the code leads to a minor speedup (ratio = $0.90$).
 

### 2022-01-14 pj-7.1/paper.tex
pj: sec 4, example spec error caught

@@
 \subsection{The Payoff: Issues Reported}
-%% 2022-01-11 FILL TODO reword, excitement!
 
@@
 The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label (\code{sp-correctness}).
+
+For example, we discovered that \SP{} handled the following function incorrectly when
+we tried to model it:
+%% \url{https://github.com/facebookincubator/cinder/issues/62}
+
+\medskip
+\begin{minipage}[t]{0.25\columnwidth}~
+\end{minipage}\begin{minipage}[t]{0.4\columnwidth}
+\begin{lstlisting}
+  from typing import Optional
+
+  def f(x: Optional[str]) -> str:
+    while True:
+      if x is None:
+        break
+      return x
+
+  print(f(None))
+  # None
+\end{lstlisting}
+\end{minipage}
+
+\noindent{}The function expects either a string or the \code{None} value and promises to return a string.
+When called with \code{None}, however, the function breaks out of the \code{while} loop
+and implicitly returns \code{None}---disobeying the return type.
+\SP{} had failed to account for the \code{break} and implicit return.
+More concerningly, the associated test case \emph{expected} the program to type check.
+Creating a formal model helped us catch this specification error.
 
@@
 \label{s:related}
+
+Concrete types
 

### 2022-01-14 pj-7.1/paper.tex
pj: related work, no cites

@@
     % of types that enable optimizations.
+    % depth-first types.
     % (\SP{} defers to Pyre the task of statically checking the full Python language.)
@@
 
-Concrete types
-
-Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
-%% \footnote{In other words, \SP{} provides \emph{immediate accountability}~\cite{mt-oopsla-2017}.}
-
-Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.
-
-Grace nominal transient aint bad at all
-
-Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.
-
-PyPy is another runtime for Python
-Pycket is built on PyPy and improves both Typed Racket~\cite{bbst-oopsla-2017}
-and Reticulated Python~\cite{vsc-dls-2019}.
-
-Type systems for Python ...
-mypy pyre pytype;
-Reticulated;
-
-Optimizing type systems for Python ...
-Reticulated;
-mypyc;
-Developer experience is one of the main advantages of \SP{} over mypyc,
-which implements type-directed optimizations for CPython
-by compiling source code to C extension modules.
-\SP{} even used Cython at some point and found the mixed-workflow codebase painful.
-%%  \item
-%%    Prior to investing in \SP{}, Instagram used Cython to compile
-%%    performance-critical code to C extension modules.
-%%    Cython performed well, but mixing pure Python with a few compiled modules
-%%    made for an awkward developer experience.
-%%    \SP{} aims to match the Python workflow.
-
-optimized transient, sp suggests method-at-a-time equally effective as tracing
-
-JS++ is a sound gradual type system for JavaScript~\cite{jspp}.
-By default, it separates JS++ from JavaScript values:
-every value with a type must be defined in JS++ code,
-and every value that may have come from Javascript has
-a special type \code{external} to mark it as such.
-Programmers can manually cast external values to a
-more precise type.
-The casts may fail at runtime, and may create a copy of
-the original value if they succeed (depends on the type).
-%% 2022-01-09 how does JS++ protect arrays sent to JS?
+Concrete types originated in Thorn~\cite{bloom,wnlov-popl-2010}
+and StrongScript~\cite{rzv-ecoop-2015},
+and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
+It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
+in the original Nom paper inspired the \SP{} team to pursue an implementation.
+
+In addition to concrete types, Thorn and StrongScript have optional like types.
+The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
+types, though like types make no claims about soundness.
+
+Many aspects of our Redex model for \SP{} are adapted from the Full Monty core calculus
+for Python~\cite{pmmwplck-oopsla-2013}.
+In particular, the Full Monty paper reminded us that booleans \emph{are} integers in Python;
+\SP{} had overlooked this detail.\footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
+%% 2022-01-14 TODO check full monty with KC
+
+Our production experience with \SP{} suggests that a method-based JIT compiler can
+usually eliminate the costs of shallow types.
+This finding complements prior positive experience with shallow types and JIT
+technology; e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
+and in Grace with the Graal partial-evaluation-based JIT~\cite{FILL}.
+Similarly, the Pycket team used a tracing JIT to reduce the costs of higher-order
+checks~\cite{bbst-oopsla-2017}.
+It remains to be seen whether a non-tracing JIT is effective for higher-order
+casts.
+
+\SP{} is one of many type systems for Python.
+Other optional type checkers include Pyre, mypy, and PyType;
+these systems do not ensure safe interactions with untyped Python.
+Another sound type system is Reticulated,
+which enforces structural types (rather than nominal) using its transient
+semantics~\cite{vss-popl-2017,FILL}.
+Other \emph{optimizing} type systems for Python
+include Reticulated and mypyc.
+The latter compiles Python source to C extension modules; unlike \SP{},
+it does not offer a pure-Python developer experience.
+
+Although sound gradual type systems are rare in industry, there are at
+least three others besides \SP{}.
+Dart 2 is a nominally-typed language with a dynamic type that is similar
+to the top object type but allows implicit downcasts to every other type~\cite{FILL}.
+C\# includes a dynamic type that defers type checking until runtime~\cite{bmt-ecoop-2010}.
+JS++ is a sound type system for JavaScript that allows untyped values
+using a catch-all \code{external} type~\cite{jspp}.
+%% Programmers can manually cast external values to a more precise type.
+%% The non-immediate casts make copies (how does that work for functions?!)
+%% also, how does JS++ protect arrays sent to JS?
 
@@
 of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
-Primitive types are progressive types (and a little more) for base values.
+Primitive types are progressive types for base values.
 Concrete types are essentially a progressive form of shallow types.
 
-\section{Future Work}
-\label{s:future}
-
-%% Immediate future: fix adoption friction points
-%% 2021-12-15: WHAT ARE THESE?
-%% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
-%% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
-%% - functions have to expect positionals (can't do keyword only)
-%% - @dataclass enum.Enum
+%% \section{Future Work}
+%% \label{s:future}
 %% 
-%% significant selection bias
-%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
-%%  because the pay / payoff isn't yet attractive
-
-%% engineering: JIT profiling, instead of hand-requested JIT list
-%% model: metaclasses
+%% %% Immediate future: fix adoption friction points
+%% %% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
+%% %% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
+%% %% - functions have to expect positionals (can't do keyword only)
+%% %% - @dataclass enum.Enum
+%% %%
+%% %% significant selection bias toward low-hanging fruit
+%% %% model: metaclasses
+%% %% engineering: JIT profiling, instead of hand-requested JIT list
+%% 
+%% Adapt SP ideas to a new setting, test performance takeaways.
+%% 
+%% Improve SP with generic types, structural types (lambda), \ldots.
+%% 
+%% Use confined GT to relax ChkDict ... let programmers decide whether
+%% the type should reject or convert untyped data.
+%% Maybe a ChkDict function could compile to an "overloaded" version with
+%% fast and slow paths for ChkDict vs normal dict.
+%% 
+%% Build an automatic migration tool for Checked data.
+%% 
+%% Build a static analysis that hoists transient annotations to an early, shared point.
+%% Take care to give quality error messages.
+%% 
+%% Optimize Python integer operations (in addition to the c types).
+%% Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
+%% or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
+
+\section{The Lessons}
+\label{s:conclusion}
 
@@
 %% already gradual" [leaving aside the boring non-Object base type distinction].
-       So the entire remaining game is one of making the programming interface
+%%     So the entire remaining game is one of making the programming interface
 %% as painless as possible. Some of that is just a trivial matter of
@@
 %% be much else left.
-
-Adapt SP ideas to a new setting, test performance takeaways.
-
-Improve SP with generic types, structural types (lambda), \ldots.
-
-Use confined GT to relax ChkDict ... let programmers decide whether
-the type should reject or convert untyped data.
-Maybe a ChkDict function could compile to an "overloaded" version with
-fast and slow paths for ChkDict vs normal dict.
-
-Build an automatic migration tool for Checked data.
-
-Build a static analysis that hoists transient annotations to an early, shared point.
-Take care to give quality error messages.
-
-Optimize Python integer operations (in addition to the c types).
-Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
-or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
-
-\section{The Lessons}
-\label{s:conclusion}
 
@@
 
-\subsection{Nominal, Checked Types}
-\label{s:concrete-migration}
-
-CheckedDict enables strong type checks and optimizations, but is painful to use.
-It is painful even though ChkDict supports the full dict API
-(any context that uses a Python dict can use a ChkDict without code changes).
-
-The problem is that the ChkDict type rejects all Python values.
-It accepts only ChkDict values, built through a special constructor
-that installs a tag for type tests and guards writes.
-Suppose that \code{f} is an untyped dict function.
-As is, it can process ChkDict but gets no benefit from type checks nor from optimizations.
-Adding a ChkDict domain type enables optimizations within \code{f} but raises a non-local
-problem: all callers of \code{f} must be sure to create a ChkDict.
-These ChkDicts must also be monomorphic to match whatever type \code{f} chose.
-It can take many edits to get a program running again after adding a ChkDict type.
-
-Lessons:
-\begin{itemize}
-  \item
-    Nominal types are not compatible with structual values, such as Python values.
-    Programmers are forced to edit old code to use such types ... unless the recent Nom work has a better idea~\cite{mt-oopsla-2021}
-  \item
-    Monomorphic nominal types are even worse.
-    Editing old code to fit their rigid constraints may not be feasible.
-
-\end{itemize}
-
-Enabling checked types has a huge ripple effect.
-The worst case is changing an annotation to a expect a checked data structure:
-\begin{enumerate}
-  \item all callers must create a checked value by invoking the right constructor, and
-  \item all clients of those callers can no longer expect an unchecked type.
-\end{enumerate}
-Programmers have to trace the annotation back to value-creation points, and then
-ensure that all uses of those values are compatible.
-
-\subsection{Pyre Compatibility}
-
-Many decisions in \SP{} are influenced by the widespread use of Pyre at Instagram.
-\begin{itemize}
-  \item
-    To minimize changes needed to adapt Pyre-annotated programs
-    to \SP{}, the syntax of \SP{} is compatible with Pyre.
-    Pyre can run useful checks on every \SP{} program.
-  \item
-    Because Pyre is a mature type checker, the \SP{} team has taken a
-    depth-first approach to development.
-    The language implements a small set of sound types efficiently and ignores
-    other types.
-    Of course, programmers who use advanced Pyre types can still use Pyre to
-    get compile-time feedback.
-  \item
-    The fact that \SP{} interprets unknown Pyre types as type dynamic changes
-    users' expectations about what the dynamic type should mean.
-    It is not necessarily a label for untyped Python code; dynamic code
-    can be analyzed to catch static bugs.
-    %% FILL what are the checks?
-\end{itemize}
-
-\subsection{Conclusion}
-
 Sound types in Static Python give programmers a way to improve performance.
@@
 Too soon to say.
-
-\subsection{Communication}
-
-How did we communicate effectively?
-Starting point was GT survey [CITE].
-Moved forward with example programs, see emails and github issues.
-Our formalism had to cover their unit tests.
-FILL
 
@@
 \label{a:github-issues}
-
-FILL if we're going to list issues, might as well report the titles, labels, and open/closed status.
 
@@
 
+\section{Fine-Grained Benchmark Data}
+
+FILL
+
 \end{document}

### 2022-01-14 pj-7.1/paper.tex
pj: discussion draft

@@
 
+%% 2022-01-14 TODO re 3.7%
+%% That percentage represents the aggregate CPU improvement observed from all of
+%% our individual diffs that converted modules to Static Python over the past
+%% year. In a sizable number of cases a conversion to Static Python did mean that
+%% functions in that module started being JITted that previously were not JITted,
+%% because for most of the half we had a policy in place that all Static Python
+%% modules are always JITted. (We’ve now decoupled these things for future
+%% conversions, so opting functions into the JIT is now orthogonal to opting them
+%% into SP.)
+%% 
+%% So the 3.7% does not include any CPU wins that Cinder or the JIT provide in any
+%% non-static modules, it only includes improvements observed from specifically
+%% converting a module from nonstatic to static. But it does in some cases include
+%% the effects of going from nonstatic + no-JIT to SP+JIT for a given module. In
+%% other cases a conversion may have gone from nonstatic+JIT to SP+JIT.
+%% Unfortunately we don’t have any way to go back and tease apart these cases for
+%% each individual module conversion.
+%% 
+%% What we are doing now is an experiment where we turn off SP in a set of hosts
+%% and observe them relative to the rest of prod, without changing the JIT list.
+%% This will show some % regression, which we expect to be some number <3.7%, and
+%% that will give a solid figure for the benefit we are getting specifically from
+%% SP across all static modules, totally independent of the JIT. Unfortunately
+%% there have been some hiccups in this experiment and we don’t have that number
+%% ready yet, but we hope to have it in the next week or so.
+
+%% 2022-01-14 FILL TODO 
+%% > I still have trouble understanding the CPU number. There must be many
+%% > servers running many workloads and some statistics to merge them. Do
+%% > you have a blog post or something that explains the profiling in more
+%% > detail?
+%% 
+%% I don’t know of anything reasonably current in terms of blog posts. For IG web
+%% prod, there are not “many workloads” — there is no specialization in the IG web
+%% tier. There are many thousands of hosts, all running the exact same workload —
+%% all those hosts are handling an undifferentiated random selection from a single
+%% giant pool of incoming IG requests. So for all intents and purposes we can
+%% consider them identical. (There are in practice some differences in the hosts
+%% themselves — e.g. different generations of CPU hardware — that we do have to
+%% take into account when running the experiments below.)
+%% 
+%% The details of how we measure a CPU regression or improvement are interesting,
+%% though. What we do is we take a set of hosts that we want to measure (typically
+%% an experiment group and a control group, where the control group is identical
+%% to the rest of production and the experiment group has some change made to its
+%% code or configuration whose effect we want to measure), and we slowly increase
+%% the traffic to those hosts until they are at max CPU utilization — they cannot
+%% handle any more traffic without falling over. Then we measure the
+%% requests-per-second handled by each set of hosts; if the experiment group is
+%% able to handle 2% more requests per second than the control group, while both
+%% are at max CPU utilization, this effect is considered a 2% global CPU win.
+%% 
+%% This metric has the advantage that it is highly holistic and results-oriented;
+%% it is a measure of our actual capacity to serve more requests with fewer hosts,
+%% and thus it can subsume and account for, and give apples-to-apples comparison
+%% of, many different possible reasons for an increase or decrease in a host’s
+%% efficiency in serving requests.
+%% 
+%% The disadvantage of the metric is that the raw number it measures (requests per
+%% second) is also highly dependent on the product code and client behavior at
+%% that moment in time, and is not stable over time, because the average cost in
+%% CPU cycles of “one request” is not stable over time. If the behavior of the IG
+%% iOS or Android app changes so that it is sending more heavy requests and fewer
+%% lightweight requests, RPS across the fleet will naturally decrease even though
+%% the efficiency of the servers has not changed. Therefore, this metric is really
+%% good for saying “at X point in time, Y change has a Z% overall effect on our
+%% CPU efficiency” but it cannot be used to compare raw numbers from two different
+%% points in time; we cannot use it to say “between X1 point in time and X2 point
+%% in time, RPS increased Z%, therefore our efficiency has increased by Z%.”
+%% 
+%% So the 3.7% number we gave you is an aggregate win from doing many such
+%% individual experiments over the course of the past year, where each experiment
+%% was measuring the effect of converting X module (or small set of modules) from
+%% nonstatic to static (where in some cases it was nonstatic+JIT to SP+JIT, and in
+%% other cases it was nonstatic noJIT to SP+JIT.) And the updated number we hope
+%% to get you soon is an overall measure from a single experiment, today, of the
+%% effect of disabling SP everywhere, while leaving the JIT list untouched. We
+%% expect the latter number to be smaller than 3.7% because it will remove the JIT
+%% factor entirely, but there’s also an unknown factor (which could net positive
+%% or negative) of how the behavior of long-ago-converted SP modules has changed,
+%% or the capabilities of SP has changed, since their conversion, which changes
+%% the net win of SP on those modules since the time when we converted them.
+
 %% 2022-01-04 TODO have no evidence that "any module" typing works out ... team
@@
 
-\section{The Lessons}
+\section{Lessons}
 \label{s:conclusion}
@@
 
-%% TODO conclusions about SP? threats to validity?
-
-\SP{} is ambitious in many ways:
+\SP{} is an ambitious undertaking.
+Its developers are maintaining a fork of the Python 3.8 runtime,
+a type checker,
+an optimizing ahead-of-time compiler,
+and a method-based JIT compiler.
+The team has converted hundreds of untyped modules to \SP{}
+and plans to convert thousands more in the future.
+
+And yet, \SP{} is carefully restricted to set attainable goals
+for its implementors and for adopters:
 \begin{itemize}
-  \item Type checker and compiler for Python syntax
-  \item Fork of Python 3.8
-  \item Add significant runtime components: vtable, classloader
-  \item Add a method-based JIT
-  \item Aiming for tens of thousands of typed modules (surely not all 45k in the server, but otoh the plan must be to hit other parts of IG too)
+  \item
+    The syntax of sound static types is relatively small compared
+    to Pyre~\cite{pyre} and other types-for-Python proposals~\cite{FILL}.
+    Higher-order types are absent.
+    The focus is instead on first-order class types, base types, and standard data structures.
+  \item
+    The type checker operates at a module-level granularity rather
+    than at the level of statements or expressions.
+  \item
+    Gradual class hierarchies are permitted, but restricted to single inheritance.
+    Furthermore, static-to-static method overrides cannot reduce the precision of types---a restriction
+    that enables optimizations but breaks the static gradual guarantee.
+  \item
+    Shallow types such as $\sptrawpydict$ accommodate Python code but offer coarse
+    soundness guarantees.
+    These types have been the top priority for \SP{}.
+    Concrete types support full soundness, but impose nonlocal maintenace costs.
+    \SP{} supports few concrete types at the moment and plans to add more.
+  \item
+    Progressive primitive types are available to help developers maximize performance.
+    These types are intentionally incompatible with the dynamic type to help programmers
+    avoid casts at run-time.
 \end{itemize}
-
-\noindent{}But \SP{} is also carefully targeted:
-\begin{itemize}
-  \item
-    Not aiming for an exhaustive type checker.
-    The availability of tools like Pyre already do a fine job.
-    (Don't forget that consequently, it's helpful for undefined types go to dynamic.)
-  \item
-    Shallow types come first in the implementation, then concrete / checked
-    (ditto for adoption)
-  \item
-    Not aiming for the full gradual guarantees.
-    Only really matters for the initial migration, the common case.
-    Indeed, it's important to not have the guarantee for primitives for their performance.
-  \item
-    The type checker is module-at-a-time, not fine-grained.
-\end{itemize}
-
-\noindent{}Lesson: the migration works, with a dedicated support team.
-
-%% 2022-01-06 TODO multiple dimensions of gradual
-
-Sound types in Static Python give programmers a way to improve performance.
-The change is not forced, not intrusive;
-programmers can keep using unsound types when working toward a deadline.
-
-Conjecture that soundness will find bugs too, and lead to more reliable products.
-Too soon to say.
+%
+Many of these restrictions go against the grain of the mainstream
+research community, but they have been effective in practice.
+The current \CPUchange{} increase in CPU usage is a huge improvement
+at the scale of the Instagram web server.
+
+In conclusion, we conjecture that the \SP{} approach to gradual typing
+is a viable way to incrementally add types to a production codebase.
+Determining whether this conjecture holds more broadly is a critical
+question that calls for research on two fronts.
+First, \SP{} must be applied to additional projects in a module-at-a-time manner.
+A exhaustive performance evaluation~\cite{FILL} of a few small projects would be
+especially useful.
+Second, the \SP{} approach should be adapted to other languages.
+An optionally-typed language such as TypeScript would be an ideal starting
+point for such an experiment.
 

### 2022-01-15 pj-7.1/paper.tex
italic correction must go inside the italics, otherwise it doesn't correct anything

@@
 Whenever an expression or variable lacks a type annotation, \SP{} uses the
-dynamic type as a default and skips most \emph{but not all}\/ static checks.
+dynamic type as a default and skips most \emph{but not all\/} static checks.
 Programmers can thus add annotations to part of a module

### 2022-01-15 pj-7.1/paper.tex
typo
pj: appendix, import table

@@
 
+\usepackage{lscape}
 \usepackage[framemethod=TikZ]{mdframed}
@@
 %
-With this context in mind, \SP{} keeps track of whether each class is typed of untyped.
+With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
 Each typed class can furthermore assume that if a method has a precise (non-dynamic)
@@
 All runtimes for this table came from a 64-bit CentOS Linux system using a build
-of \SP{} and Cinder compiled with default settings.
+of \SP{} and Cinder compiled with default settings
+(commit \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{\code{d0d071a}}).
 Each benchmark version ran 11 times total: once to warm up the JIT and 10 times to
@@
 
-The 25 GitHub issues that we filed during the formalization effort may be found at the following three links:
+The 25 GitHub issues that arose from our work may be found at the following three links:
 
@@
 
-\section{Pyre Limitations}
-
-FILL condense the email thread in comment (see source)
-
-% Pyre limitations
-% 4 messages
-% Ben Greenman <benjaminlgreenman@gmail.com>	Fri, Sep 24, 2021 at 11:18 AM
-% To: Carl Meyer <carljm@fb.com>
-% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
-% >>> When we “add types” to a function (by adopting its module into Static
-% >>> Python, meaning Static Python now cares about the type annotations on that
-% >>> function), there may be other code calling that function with types that
-% >>> don’t match what it is annotated to receive. This then causes new runtime
-% >>> type errors for those calls once the module is made static. So sometimes we
-% >>> do have to fix those – we consider that to be fixing pre-existing type
-% >>> errors that Pyre missed. We probably see a lot less of that because the
-% >>> codebase is already type annotated and statically checked by Pyre, than we
-% >>> would if we were adopting Static Python into a previously-not-type-checked
-% >>> codebase.
-% >>
-% >> I'm surprised that you see any errors like this. Is Pyre unsound at
-% >> these pre-existing spots?
-% >
-% > Yes, Pyre is unsound in various ways (as a purely static checker of a
-% > gradually typed language must inherently be):
-% >
-% > 1. Some code just isn’t covered or checked by Pyre, because Pyre’s gradual
-% > adoption strategy is to only check functions that have annotations in their
-% > signature, or in extreme cases because the “code” that’s calling into Python
-% > isn’t Python code at all, but rather Python C extension code or Cython code.
-% > So Pyre has no visibility into calls from code it isn’t even checking.
-% > 2. Any code with annotations involving the `Any` type (or missing
-% > annotations) will have values that Pyre types as `Any`, and Pyre will
-% > silently allow you to do anything with those values. (It has no other
-% > choice, this is the only way that static-only gradual typing can work and
-% > actually be gradual.) So anything typed as `Any` by Pyre is a potential
-% > unsoundness. And although I say our codebase is largely type annotated,
-% > given that lots of it is pre-typing code that can’t be accurately statically
-% > annotated without rewriting it, there are plenty of Any annotations
-% > (especially `Dict[str, Any]` annotations) floating around.
-% > 3. People can (and do) just silence Pyre by adding a `# pyre-fixme` comment
-% > that silences a type error on the following line, if they are doing some
-% > mass adoption of types or otherwise just don’t feel up to fixing a type
-% > error for whatever reason. This tends to happen most often in cases where an
-% > annotation is just overly restrictive (e.g. claims to require `str` but in
-% > fact the method handles `None` just fine, and the argument really should be
-% > annotated as `str | None`.) So we have to fix the annotation to actually
-% > match the runtime behavior, and then we also are able to remove the
-% > pyre-fixme comments; everyone wins :)
-% > 4. If the entire codebase and all its type annotations and type stubs agree
-% > to lie about some dynamic wrinkle, Pyre will be satisfied that everything is
-% > good (and will have no way to know otherwise). A particular case of this we
-% > hit recently is that lots of values annotated as `str` in our codebase can
-% > actually at runtime be `Promise` objects that dynamically behave like a
-% > `str` if you treat them as one, and implement lazy translation of that
-% > string to the current user’s language. So as we adopt such code into Static
-% > Python, we have to correct those annotations from `str` to `str | Promise`
-% > and handle the fallout. A similar case is the `weakref` module in the Python
-% > standard library, which allows you to create “weak references” to objects
-% > that will not keep them alive in the garbage collector, but will just
-% > resolve to None if the object dies. There’s a `weakref.proxy` that creates a
-% > transparent weak-reference proxy to any object, and sometimes methods will
-% > return one of those but be annotated as returning the proxied type. This is
-% > fine for Pyre, as long as the proxy is sufficiently transparent at the level
-% > of Python semantics, but it doesn’t work for Static Python.
-% 
-% Points 1, 2, and 3 make sense.
-% 
-% I'm wondering if there are ways to write a fully-typed Pyre program (that
-% doesn't use those escape hatches) but is still unsound.* It sounds like
-% point 4 might be on this track ... but also that unchecked code (points 1,2)
-% might be involved.
-% 
-% For the weakrefs case, would Pyre approve a function like this one:
-% 
-% ```
-%   def f()->C:
-%     return weakref.proxy(C())
-% ```
-% 
-% If so, is that because Pyre comes with a builtin annotation for
-% `weakrefs.proxy` that lets it return the Any type?
-% 
-% 
-% * For comparison, TypeScript is unsound by design in a few ways. Any computed
-% string can index any object. And all object properties are covariant, despite
-% being mutable. I wonder if Pyre has similar "static unsoundness" by design.
-% Lu, Kuang-Chen <kuang-chen_lu@brown.edu>	Fri, Sep 24, 2021 at 11:50 AM
-% To: Ben Greenman <benjaminlgreenman@gmail.com>
-% Cc: Shriram Krishnamurthi <shriram@gmail.com>
-% (Internal discussion)
-% 
-% I think point 4, especially the `str` example, comes from the lack of structural typing.
-Interfaces can partially solve this problem.
-But with Nom-style interfaces, they have to define an interface and annotate the classes involved.
-% [Quoted text hidden]
-% Carl Meyer <carljm@fb.com>	Fri, Sep 24, 2021 at 12:37 PM
-% To: Ben Greenman <benjaminlgreenman@gmail.com>
-% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
-% [Quoted text hidden]
-% I think it could be reasonable to define “fully-typed” as to inherently exclude all four points above!
-This would also exclude most real-world Python programs, but might still be an interesting definition, because even then there are some unsoundnesses remaining (that I didn’t cover above because they are not usually a problem for us in practice; in most cases I think because they involve parts of the type system we don’t even pretend to support yet in Static Python.) I’ll list some of these known unsoundnesses within the type system below.
-% 
-% In practice the Python language and standard library include features that are considered too painful to use if fully typed, generally because sound typing would require typing things as `object` (the top type), requiring `isinstance` checks everywhere to keep the static checker happy, so an unsound `Any` is preferred for usability reasons.
-Weakref proxies are just one example of this; there are many others.
-% 
-% https://github.com/python/typeshed is the shared repository of type stubs for the Python standard library and popular third-party packages (used by all the static checkers: Pyre, mypy, pyright, pytype…).
-You can search it for `Any` to find many many more examples.
-% 
-% 
-% > For the weakrefs case, would Pyre approve a function like this one:
-% >
-% > ```
-% >  def f()->C:
-% >    return weakref.proxy(C())
-% > ```
-% >
-% > If so, is that because Pyre comes with a builtin annotation for
-% > `weakrefs.proxy` that lets it return the Any type?
-% 
-% Yes, exactly, this is in the shared “typeshed” repository: https://github.com/python/typeshed/blob/master/stdlib/_weakref.pyi#L33
-% 
-% >
-% > * For comparison, TypeScript is unsound by design in a few ways. Any computed
-% > string can index any object. And all object properties are covariant, despite
-% > being mutable. I wonder if Pyre has similar "static unsoundness" by design.
-% 
-% Right, and the Python type system has some similar unsound-by-design choices. A few that I’m aware of:
-% 
-% - Mutable attributes ought to be invariant (because they are in effect both getters and setters, thus appearing in both covariant and contravariant position), thus subclasses should not be allowed to change the type of a base class attribute.
-But in practice all Python type checkers permit a subclass to narrow the type of a base class attribute.
-Now that `Final[]` annotations are available on attributes (which would make them soundly covariant), it might be more reasonable to close this hole, here’s discussion on the mypy bug tracker: https://github.com/python/mypy/issues/3208.
-We are currently sound here in Static Python, but in an odd way; we should error on the subclass attribute type override itself, but instead we ignore it and continue to use the base class attribute type for the subclass also.
-% 
-% - There are some issues with `typing.Type[]` types (these are the types of class objects themselves) and the fact that they are covariant (which is widely relied on.) This is unsound because incompatible constructor signature overriding is allowed in subclasses (which is also widely used).
-It’s also unsound because the type of the `self` argument to any overridden method will be Parent on the parent method and Child on the child method, which is the opposite of the normal requirement that method arguments (appearing in a contravariant position) must only be widened on subclass overrides.
-This means that if some function accepts a type and calls an unbound method on it, providing the `self` arg explicitly, the covariance of `Type` is unsound.
-On these we are currently OK only because we don’t yet support `Type[]` annotations.
-% 
-% - Type checking argument splatting in calls is unsound.
-If we have a dynamic list `l` or dict `d` and we have a call like `foo(*l)` or `foo(**d)`, unless foo accepts `(*args, **kwds)` it is impossible to statically type check the call.
-Making this sound would have to either require the callee to accept (*args, **kwds) or require that the list or dict is a statically known literal; the latter totally defeats the purpose of argument splatting, the former is more conceivable but still makes it a lot less useful, so in practice type checkers just allow these calls.
-This one is OK in Static Python because we simply fallback to dynamic calls (and thus runtime argument type checks in the callee) for these calls (I mean it’s OK in that it doesn’t result in cascading unsoundness — it still means runtime error rather than static one for the call itself.)
-% 
-% There are more cases than just these, but I think the others are more esoteric.
-% 
-% Carl
+\section{On Migrations from Pyre to \SP{}}
+
+Although the Instagram web server has extensive Pyre annotations,
+these annotations are not always enough to satisfy \SP{}.
+In fact, code may have latent bugs that Pyre did not catch;
+we list common reasons for such bugs below:
+
+\begin{itemize}
+  \item
+    Pyre misses some bugs because it does not monitor run-time interactions
+    with untyped code.
+
+  \item
+    Similarly, some uses of the Pyre dynamic type (\code{Any}) end up
+    raising errors when \SP{} monitors their interactions with precisely-typed code.
+
+  \item
+    Occurrences of the special comment \lstinline$# pyre-fixme$ disable Pyre checks.
+
+  \item
+    Some annotations intentionally lie to reduce the work of maintaining types.
+    For example:
+
+    \begin{itemize}
+      \item
+        The standard \code{typeshed} repo of Python type annotations declares
+        that a weak reference has the dynamic type~\cite{FILL}.
+        This declaration is easy to use because it does not force programmers
+        to handle the case where their weakly-held object has been collected,
+        but it is also a potential source of bugs.
+      \item
+        Mutable class attributes are covariant in Pyre and other
+        Python type checkers, including mypy~\cite{FILL}.
+      \item
+        The meta-type \code{Type} is unsoundly covariant as well in Pyre
+        and others.
+      \item
+        Argument splatting, i.e. applying a function that expects positional
+        arguments to a list (\code{f(*lst)}), is not statically checked
+        either.
+    \end{itemize}
+
+\end{itemize}
+
+None of these issues are problems with Pyre.
+Quite naturally, it cannot find bugs in code that it does not analyze!
+Our point is simply that optional type systems can miss a variety of issues.
 
@@
 
-FILL
+\newcommand{\colname}[1]{\textbf{#1}}
+
+The tables in this section present fine-grained data for the microbenchmark programs.
+Each benchmark has several versions of its code that accomplish the same work:
+the original untyped version (\colname{Orig}) and a typed version (\colname{T-Max}) converted for maximum
+performance under Static Python and the JIT.
+Some benchmarks have a basic typed version (\colname{T-Min}) that adds only
+type annotations and necessary casts to the original code,
+without using any more advanced Static Python
+features such as primitive types or otherwise optimizing the code for Static
+Python.
+Fannkuch has a second basic version (\colname{T-Min2}) that uses just a few primitive integers
+where it can be done without significant changes to the code.
+
+The other axes of our test matrix are \colname{SP} (whether the Static Python compiler is
+used), \colname{JIT} (whether the Cinder JIT is enabled), and \colname{SF} (whether the JIT
+shadow-frame mode is enabled, which reduces Python frame allocation costs.) The
+matrix is not complete, since using the Static Python compiler on untyped code
+has no noticeable effect, and shadow-frame mode is only relevant under the JIT.
+
+Benchmarks are run on a 64-bit CentOS Linux system, using a build of Cinder
+compiled with default configure settings. Overall process user time is measured
+via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
+initial warmup run to ensure there is an up-to-date bytecode cache file and
+compilation time will not be included), and all individual data points are
+recorded. All data points are in seconds.
+
+JIT runs use a JIT list that includes only the benchmark code itself, to
+minimize overhead of compiling unrelated standard library code not used in the
+benchmark.
+
+Benchmarks run on the following commit:
+
+\smallskip
+{\centering\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}
+
+Fannkuch in particular is very numeric-heavy, so using unboxed primitive
+integers in the typed version can be a big win. For this to work maximally
+efficiently, we also need to avoid operations that don't yet support primitives
+and thus would require expensive boxing. In typed-and-optimized Fannkuch (\colname{T-Max}), these changes move
+a lot of the work out of heavily-optimized C builtin methods (for list
+slice/insert/pop) and into simple Python code doing the equivalent operations
+manually with simple iteration and indexing. In the short run this is a massive
+pessimization for non-static Python (or even SP without the JIT, since
+currently we always use boxed integers in the interpreter loop.) In the long
+run it suggests that Static Python + JIT could make it feasible to implement
+many more of these core datastructure operations in pure Python rather than in
+C.
+
+Because of this subtlety, we include benchmark results for both typed and
+untyped versions of each benchmark. In the short term, perhaps the most
+relevant comparison is the performance of Static Python on the typed benchmark
+vs the performance of nonstatic Python on the untyped benchmark; this gives a
+fair picture of expected perf gains converting code to Static Python with some
+willingness to optimize.
+
+We also observe significant untapped opportunity to improve Static Python's
+performance; for example, $20\%$ of Typed Fannkuch SP+JIT time is spent on
+bounds-checking array accesses; most of this runtime bounds-checking cost could
+be eliminated if the Static Python compiler tracks known array sizes and
+integer values more thoroughly. And it would also be possible to support list
+slicing, list pop, and list insert with primitive arguments, thus avoiding the
+need to rewrite these operations in Python for maximum efficiency.
+
+\colname{T-Min} Fannkuch without the JIT is slow because of a known issue where we
+don't cache function lookups for \code{INVOKE\_FUNCTION} in the interpreter loop,
+making the calls to \code{list.insert} and \code{list.pop} slower; this is fixable.
+
+\begin{landscape}
+
+\begin{table}[tph]
+  \caption{Richards microbenchmark data}
+  \label{t:mb:richards}
+  \footnotesize
+
+  \begin{tabular}{rrrrrrrrrrrrrrr}
+  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ & $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ & $2.6$ & $3.88$ & $10.17$ \\
+  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ & $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ & $2.6$ & $4.01$ & $10.18$ \\
+  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ & $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ & $2.61$ & $4.17$ & $10.73$ \\
+  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ & $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ & $2.9$ & $3.92$ & $9.96$ \\
+  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ & $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ & $2.63$ & $3.79$ & $10.04$ \\
+  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ & $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ & $2.64$ & $3.95$ & $9.99$ \\
+  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ & $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ & $2.62$ & $3.9$ & $10.16$ \\
+  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ & $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ & $2.6$ & $3.84$ & $9.69$ \\
+  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ & $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ & $2.6$ & $4.04$ & $9.82$ \\
+  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ & $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ & $2.62$ & $3.85$ & $9.85$ \\
+\end{tabular}
+
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}[tph]
+  \caption{Fannkuch microbenchmark data}
+  \label{t:mb:fannkuch}
+  \footnotesize
+
+  \begin{tabular}{rrrrrrrrrrrrrrr}
+  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) \\\hline
+  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ & $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ & $2.48$ & $2.52$ & $7.28$ \\
+  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ & $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ & $2.52$ & $2.48$ & $7.28$ \\
+  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ & $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ & $2.47$ & $2.53$ & $7.24$ \\
+  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ & $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ & $2.48$ & $2.57$ & $7.35$ \\
+  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ & $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ & $2.5$ & $2.56$ & $7.34$ \\
+  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ & $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ & $2.43$ & $2.47$ & $7.17$ \\
+  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ & $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ & $2.47$ & $2.69$ & $7.18$ \\
+  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ & $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ & $2.47$ & $2.45$ & $7.14$ \\
+  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ & $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ & $2.43$ & $2.55$ & $7.23$ \\
+  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ & $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ & $2.52$ & $2.49$ & $7.2$ \\
+\end{tabular}
+
+\begin{tabular}{rrrrrr}
+  \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $3.16$ & $3.08$ & $4.59$ & $2.81$ & $2.94$ & $4.02$ \\
+  $3.26$ & $3.08$ & $4.51$ & $2.66$ & $2.85$ & $4.22$ \\
+  $3.11$ & $3.18$ & $4.59$ & $2.69$ & $2.78$ & $4.01$ \\
+  $3.18$ & $3.14$ & $4.73$ & $2.75$ & $2.8$ & $4.01$ \\
+  $3.14$ & $3.12$ & $4.63$ & $2.75$ & $2.92$ & $4.07$ \\
+  $3.23$ & $3.09$ & $4.53$ & $2.78$ & $2.83$ & $4.36$ \\
+  $3.41$ & $3.21$ & $4.48$ & $2.84$ & $2.81$ & $4.19$ \\
+  $3.43$ & $3.14$ & $4.53$ & $2.7$ & $2.86$ & $4.25$ \\
+  $3.08$ & $3.24$ & $4.53$ & $2.76$ & $2.78$ & $4.08$ \\
+  $3.16$ & $3.18$ & $4.45$ & $2.89$ & $2.76$ & $4.16$ \\
+\end{tabular}
+\end{table}
+
+\end{landscape}
+
+\begin{table}[tph]
+  \caption{Exact commands to invoke each benchmark}
+  \label{t:mb:commands}
+  \tiny
+
+  \begin{tabular}{ll}
+  \colname{Benchmark} & \colname{Command} \\\hline
+  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
+  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
+  Richards \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
+  (\colname{SP}) &  \\
+  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/richards\_static.py 100} \\
+  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{JIT}) & \texttt{ Tools/benchmarks/richards\_static.py 100} \\
+  Richards \colname{T-Max} & \texttt{time ./python Tools/benchmarks/richards\_static.py 100} \\
+  () &  \\
+  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
+  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
+  Richards \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
+  (\colname{SP}) &  \\
+  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/richards\_static\_basic.py 100} \\
+  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{JIT}) & \texttt{dcards Tools/benchmarks/richards\_static\_basic.py 100} \\
+  Richards \colname{T-Min} & \texttt{time ./python Tools/benchmarks/richards\_static\_basic.py 100} \\
+  () &  \\
+  Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
+  (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/richards.py 100} \\
+  Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
+  (\colname{JIT}) & \texttt{hmarks/richards.py 100} \\
+  Richards \colname{Orig} & \texttt{time ./python Tools/benchmarks/richards.py 100} \\
+  () &  \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
+  (\colname{SP}) &  \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/fannkuch\_static.py 5} \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
+  (\colname{JIT}) & \texttt{ Tools/benchmarks/fannkuch\_static.py 5} \\
+  Fannkuch \colname{T-Max} & \texttt{time ./python Tools/benchmarks/fannkuch\_static.py 5} \\
+  () &  \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  (\colname{SP}) &  \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
+  (\colname{JIT}) & \texttt{dcards Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
+  () &  \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  (\colname{SP}) &  \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min2} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  () &  \\
+  Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
+  (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/fannkuch.py 5} \\
+  Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
+  (\colname{JIT}) & \texttt{hmarks/fannkuch.py 5} \\
+  Fannkuch \colname{Orig} & \texttt{time ./python Tools/benchmarks/fannkuch.py 5} \\
+  () &  \\
+\end{tabular}
+
+\end{table}
 

### 2022-01-15 pj-7.1/paper.tex
sp: quick fixes
- rewording in sec 5 (incl. the 'bizarre' edit)
- mention class overrides in sec 1

@@
 The \SP{} type system has many restrictions that are suprising
-at first glance; for example, functions are supported only by
-shallow types.
+at first glance.
+For example, functions are supported only by shallow types
+and method overrides that use the dynamic type are more constrained
+than overrides in untyped code.
 Some of these restrictions might be lifted by future research.
@@
 
-There are two case in which \SP{} types add dynamic constraints to Python code.
+There are two case in which \SP{} types impose constraints on untyped Python code.
 First, concrete generic types (e.g. $\sptchkdict{\sptype}{\sptype}$)
@@
 \SP{} does not attempt to type first-class classes: partly because they have
-yet to appear in performance-critical code and partly because it
-is unclear how to incorporate them into the nominal type system.
-The straightforward but restrictive approach is to force code that uses a
-first-class class to expect subtypes of a particular named static class.
+yet to appear in performance-critical code in the Instagram web server
+and partly because they do not fit well with a nominal type system.
+The straightforward but restrictive approach for first-class nominal class types
+is to force code that uses a first-class class to expect subtypes of a
+particular named static class.
 Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)

### 2022-01-15 pj-7.1/paper.tex
pj: fill cites

@@
 %% TODO before submission
-%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
-%%   helps the static checker find new errors. ChkDict is one example.
 %% - choose words: static vs typed, nonstatic vs dynamic vs untyped
 %%   trying to avoid nonstatic? staticifying?
-%% - use ChkDict not ChkList ... avoid confusion
-%%   ... or at least mention both, be clear the focus is Dict b/c older
-%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
-%%   - emphasize the PAYOFF from modeling early on, don't bury in sec 3
 %% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
 %% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
-%% - "we" = all authors, NOT JUST BROWN!!!
 
@@
 Industry teams have developed innovative type systems that accommodate
-untyped designs, but are unsound [CITE].
+untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017}.
 These systems intentionally check nothing at run-time when untyped values enter
@@
 Academic teams have primarily focused on the theory of sound
-gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
+gradual types, formulating correctness properties and studying ever-more-descriptive
+types~(e.g.~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
 A few academics have studied the cost of run-time checks
-in detail [CITE] and proposed implementation methods [CITE],
-compiler technology [CITE],
-and even weakened semantics [CITE], but these efforts have not yet
-decisively closed the performance gap.
+in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
+and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
+compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
+and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
+but these efforts have not yet decisively closed the performance gap.
 The most promising attempt is the \emph{concrete} semantics
-for gradual types [CITE kafka nom etc], but it imposes nonlocal
+for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
 constraints on untyped code.
@@
 production and on public microbenchmarks.
-The paper concludes with related work~(\cref{s:related}),
-future work~(\cref{s:future}), and a final summary~(\cref{s:conclusion}).
+The paper concludes with related work~(\cref{s:related})
+and a final summary~(\cref{s:conclusion}).
 
@@
 and prior work on types for Python.
-Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
+Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~\cite{pep484} are two notable sources.
 In addition, \SP{} is strongly influenced by the following engineering constraints:
@@
   \item
-    The codebase in which \SP{} is being applied makes heavy use of Pyre [CITE]
+    The codebase in which \SP{} is being applied makes heavy use of Pyre~\cite{pyre}
     type annotations.
@@
 
-%% TODO 2021-12-28
-%% - while True break another example of modeling helping out
-%%   it's a second implementation, helpful to find bugs
+%% test suite is for counterex, not proof
+%% 
+%% (as opposed mech CONFUSSING let each sec explain purpose)
+%% 
+%% keep none None & explain why, ambiguity in stx
+%% 
+%% fully-sound weird, kill phrase ... say whats happening .... surface t is linter
+%% ignore WF checker
+%% 
+%% check para may be red.
+%% "The strict cast for checked dictionaries ...."
+%% 
+%% soundness, 1st para garbled don't need such a long reminder
+%% 
+%% Notable py **that we handle**
+%% soundly => correctly
+%% 
+%% addl -> additionally (4.4)
+%% 
+%% "is triue" missing word
+%% 
+%% move 4.5 up to 4.1, start with the payoff --- found soundness bugs in a tool that's running in prod at instagram
+%% 
+%% rename 4.3 to "boundry type soundness" satate a lemma instead of full TS keep the focus, don't jump up
+%% - make the proof case via Nom keep it short they have "detailed" proof
+%% - "interested reader" can follow another direct path (edit current para2)
+%% 
+%% 4.0 put less emphasis on part2 of the soundness. ... it's not a threat its a boring swept-under-rug part ... maybe 4.5 is the new 4.0???
+%% 
+%% 4.4 redex = since no formalization of py + sp .. follow well-established path of using test suite to check conformance with **reality**
+%% - cite Py \Js Jan paper, Politz getter-setter, Gardner js
+%% 
+%% 4.4 redex += everything in 4.1--4.3 is captured in redex, done 3 steps
+%%  1 human verification, carefully spot-check python faithful, boundaries vs 4.1--4.3
+%%  2 run againt SP tests .. indeed part of the reason for getting so much python is to run against an actual test suite
+%%  3 b/c in redex, use pbt to verify conventional soundness
+%% 
+%% cut "soundness as baseline" final 2 sentences of 4.4(?) they're too weak, end with the compiler verification out-of-scope part
 
@@
 technology; e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
-and in Grace with the Graal partial-evaluation-based JIT~\cite{FILL}.
+and in Grace with the Graal partial-evaluation-based JIT~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
 Similarly, the Pycket team used a tracing JIT to reduce the costs of higher-order
@@
 which enforces structural types (rather than nominal) using its transient
-semantics~\cite{vss-popl-2017,FILL}.
+semantics~\cite{vss-popl-2017,v-thesis-2019}.
 Other \emph{optimizing} type systems for Python
@@
 Dart 2 is a nominally-typed language with a dynamic type that is similar
-to the top object type but allows implicit downcasts to every other type~\cite{FILL}.
+to the top object type but allows implicit downcasts to every other type~\cite{dart-types}.
 C\# includes a dynamic type that defers type checking until runtime~\cite{bmt-ecoop-2010}.
@@
     The syntax of sound static types is relatively small compared
-    to Pyre~\cite{pyre} and other types-for-Python proposals~\cite{FILL}.
+    to Pyre~\cite{pyre} and other types-for-Python proposals~\cite{mypy,pytype}.
     Higher-order types are absent.
@@
 First, \SP{} must be applied to additional projects in a module-at-a-time manner.
-A exhaustive performance evaluation~\cite{FILL} of a few small projects would be
+A exhaustive performance evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
 especially useful.
@@
         The standard \code{typeshed} repo of Python type annotations declares
-        that a weak reference has the dynamic type~\cite{FILL}.
+        that a weak reference has the dynamic type~\cite{typeshed}.
         This declaration is easy to use because it does not force programmers
@@
         Mutable class attributes are covariant in Pyre and other
-        Python type checkers, including mypy~\cite{FILL}.
+        Python type checkers, including mypy~\cite{mypy}.
       \item
@@
 \smallskip
-{\centering\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}
+{\centering\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}
 

### 2022-01-15 pj-7.1/paper.tex
pj: cleanup tables, no landscape

@@
 
-%\overfullrule=1mm
+\overfullrule=1mm
 
@@
 
-\usepackage{lscape}
 \usepackage[framemethod=TikZ]{mdframed}
@@
 Python.
-Fannkuch has a second basic version (\colname{T-Min2}) that uses just a few primitive integers
+Fannkuch has a second basic version (\colname{T-Min-2}) that uses just a few primitive integers
 where it can be done without significant changes to the code.
@@
 
-\begin{landscape}
-
 \begin{table}[tph]
@@
 
-  \begin{tabular}{rrrrrrrrrrrrrrr}
-  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ & $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ & $2.6$ & $3.88$ & $10.17$ \\
-  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ & $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ & $2.6$ & $4.01$ & $10.18$ \\
-  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ & $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ & $2.61$ & $4.17$ & $10.73$ \\
-  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ & $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ & $2.9$ & $3.92$ & $9.96$ \\
-  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ & $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ & $2.63$ & $3.79$ & $10.04$ \\
-  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ & $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ & $2.64$ & $3.95$ & $9.99$ \\
-  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ & $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ & $2.62$ & $3.9$ & $10.16$ \\
-  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ & $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ & $2.6$ & $3.84$ & $9.69$ \\
-  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ & $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ & $2.6$ & $4.04$ & $9.82$ \\
-  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ & $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ & $2.62$ & $3.85$ & $9.85$ \\
+\begin{tabular}{rrrrrr}
+  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ \\
+  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ \\
+  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ \\
+  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ \\
+  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ \\
+  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ \\
+  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ \\
+  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ \\
+  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ \\
+  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ \\
 \end{tabular}
 
+\begin{tabular}{rrrrrr}
+  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ \\
+  $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ \\
+  $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ \\
+  $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ \\
+  $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ \\
+  $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ \\
+  $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ \\
+  $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ \\
+  $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ \\
+  $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ \\
+\end{tabular}
+
+\begin{tabular}{rrr}
+  \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $2.6$ & $3.88$ & $10.17$ \\
+  $2.6$ & $4.01$ & $10.18$ \\
+  $2.61$ & $4.17$ & $10.73$ \\
+  $2.9$ & $3.92$ & $9.96$ \\
+  $2.63$ & $3.79$ & $10.04$ \\
+  $2.64$ & $3.95$ & $9.99$ \\
+  $2.62$ & $3.9$ & $10.16$ \\
+  $2.6$ & $3.84$ & $9.69$ \\
+  $2.6$ & $4.04$ & $9.82$ \\
+  $2.62$ & $3.85$ & $9.85$ \\
+\end{tabular}
+
 \end{table}
-\end{landscape}
-
-\begin{landscape}
+
 \begin{table}[tph]
@@
 
-  \begin{tabular}{rrrrrrrrrrrrrrr}
-  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) \\\hline
-  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ & $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ & $2.48$ & $2.52$ & $7.28$ \\
-  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ & $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ & $2.52$ & $2.48$ & $7.28$ \\
-  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ & $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ & $2.47$ & $2.53$ & $7.24$ \\
-  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ & $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ & $2.48$ & $2.57$ & $7.35$ \\
-  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ & $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ & $2.5$ & $2.56$ & $7.34$ \\
-  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ & $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ & $2.43$ & $2.47$ & $7.17$ \\
-  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ & $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ & $2.47$ & $2.69$ & $7.18$ \\
-  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ & $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ & $2.47$ & $2.45$ & $7.14$ \\
-  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ & $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ & $2.43$ & $2.55$ & $7.23$ \\
-  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ & $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ & $2.52$ & $2.49$ & $7.2$ \\
+  \begin{tabular}{rrrrrr}
+  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ \\
+  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ \\
+  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ \\
+  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ \\
+  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ \\
+  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ \\
+  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ \\
+  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ \\
+  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ \\
+  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ \\
 \end{tabular}
@@
 \begin{tabular}{rrrrrr}
-  \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $3.16$ & $3.08$ & $4.59$ & $2.81$ & $2.94$ & $4.02$ \\
-  $3.26$ & $3.08$ & $4.51$ & $2.66$ & $2.85$ & $4.22$ \\
-  $3.11$ & $3.18$ & $4.59$ & $2.69$ & $2.78$ & $4.01$ \\
-  $3.18$ & $3.14$ & $4.73$ & $2.75$ & $2.8$ & $4.01$ \\
-  $3.14$ & $3.12$ & $4.63$ & $2.75$ & $2.92$ & $4.07$ \\
-  $3.23$ & $3.09$ & $4.53$ & $2.78$ & $2.83$ & $4.36$ \\
-  $3.41$ & $3.21$ & $4.48$ & $2.84$ & $2.81$ & $4.19$ \\
-  $3.43$ & $3.14$ & $4.53$ & $2.7$ & $2.86$ & $4.25$ \\
-  $3.08$ & $3.24$ & $4.53$ & $2.76$ & $2.78$ & $4.08$ \\
-  $3.16$ & $3.18$ & $4.45$ & $2.89$ & $2.76$ & $4.16$ \\
+  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ \\
+  $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ \\
+  $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ \\
+  $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ \\
+  $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ \\
+  $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ \\
+  $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ \\
+  $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ \\
+  $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ \\
+  $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ \\
 \end{tabular}
+
+\begin{tabular}{rrrrrr}
+  \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $2.48$ & $2.52$ & $7.28$ & $3.16$ & $3.08$ & $4.59$ \\
+  $2.52$ & $2.48$ & $7.28$ & $3.26$ & $3.08$ & $4.51$ \\
+  $2.47$ & $2.53$ & $7.24$ & $3.11$ & $3.18$ & $4.59$ \\
+  $2.48$ & $2.57$ & $7.35$ & $3.18$ & $3.14$ & $4.73$ \\
+  $2.5$ & $2.56$ & $7.34$ & $3.14$ & $3.12$ & $4.63$ \\
+  $2.43$ & $2.47$ & $7.17$ & $3.23$ & $3.09$ & $4.53$ \\
+  $2.47$ & $2.69$ & $7.18$ & $3.41$ & $3.21$ & $4.48$ \\
+  $2.47$ & $2.45$ & $7.14$ & $3.43$ & $3.14$ & $4.53$ \\
+  $2.43$ & $2.55$ & $7.23$ & $3.08$ & $3.24$ & $4.53$ \\
+  $2.52$ & $2.49$ & $7.2$ & $3.16$ & $3.18$ & $4.45$ \\
+\end{tabular}
+
+\begin{tabular}{rrr}
+  \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $2.81$ & $2.94$ & $4.02$ \\
+  $2.66$ & $2.85$ & $4.22$ \\
+  $2.69$ & $2.78$ & $4.01$ \\
+  $2.75$ & $2.8$ & $4.01$ \\
+  $2.75$ & $2.92$ & $4.07$ \\
+  $2.78$ & $2.83$ & $4.36$ \\
+  $2.84$ & $2.81$ & $4.19$ \\
+  $2.7$ & $2.86$ & $4.25$ \\
+  $2.76$ & $2.78$ & $4.08$ \\
+  $2.89$ & $2.76$ & $4.16$ \\
+\end{tabular}
+
 \end{table}
-
-\end{landscape}
 
@@
   () &  \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
   (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
   (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
   (\colname{SP}) &  \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
   (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
   (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-  Fannkuch \colname{T-Min2} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+  Fannkuch \colname{T-Min-2} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
   () &  \\

### 2022-01-15 pj-7.1/paper.tex
pj: sec 4 overhaul

@@
 \newcommand{\mfeq}{=}
-\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
+\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
 \newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
 \newcommand{\ipara}[1]{\emph{#1}\/.}
+\newcommand{\nnum}[1]{$N=#1$}
+\newcommand{\totalnum}[1]{$#1$ total}
 
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 \newtheorem{definition}{Definition}
@@
 
-%% test suite is for counterex, not proof
-%% 
-%% (as opposed mech CONFUSSING let each sec explain purpose)
-%% 
-%% keep none None & explain why, ambiguity in stx
-%% 
-%% fully-sound weird, kill phrase ... say whats happening .... surface t is linter
-%% ignore WF checker
-%% 
-%% check para may be red.
-%% "The strict cast for checked dictionaries ...."
-%% 
-%% soundness, 1st para garbled don't need such a long reminder
-%% 
-%% Notable py **that we handle**
-%% soundly => correctly
-%% 
-%% addl -> additionally (4.4)
-%% 
-%% "is triue" missing word
-%% 
-%% move 4.5 up to 4.1, start with the payoff --- found soundness bugs in a tool that's running in prod at instagram
-%% 
-%% rename 4.3 to "boundry type soundness" satate a lemma instead of full TS keep the focus, don't jump up
-%% - make the proof case via Nom keep it short they have "detailed" proof
-%% - "interested reader" can follow another direct path (edit current para2)
-%% 
-%% 4.0 put less emphasis on part2 of the soundness. ... it's not a threat its a boring swept-under-rug part ... maybe 4.5 is the new 4.0???
-%% 
-%% 4.4 redex = since no formalization of py + sp .. follow well-established path of using test suite to check conformance with **reality**
-%% - cite Py \Js Jan paper, Politz getter-setter, Gardner js
-%% 
-%% 4.4 redex += everything in 4.1--4.3 is captured in redex, done 3 steps
-%%  1 human verification, carefully spot-check python faithful, boundaries vs 4.1--4.3
-%%  2 run againt SP tests .. indeed part of the reason for getting so much python is to run against an actual test suite
-%%  3 b/c in redex, use pbt to verify conventional soundness
-%% 
-%% cut "soundness as baseline" final 2 sentences of 4.4(?) they're too weak, end with the compiler verification out-of-scope part
-
-To validate the design of \SP{}, we have developed a model of the language in PLT
+To validate the design of \SP{}, we developed a model of the language in PLT
 Redex~\cite{kcdeffmrtf-popl-2012}.
-The model encompasses two soundness claims:
-%% addresses two soundness concerns?
-\begin{enumerate}
-  \item
-    \SP{}'s overall approach to gradual interactions is sound; in particular,
-    interactions with untyped code do not undermine the validity of types.
-  \item
-    The typed half of \SP{} is a sound type system for Python.
-\end{enumerate}
-This section explores the first claim in depth by way of a small formalization
-(as opposed to the Redex \emph{mechanization})
-that illustrates the \SP{} gradual interaction strategy.
-%% Compared to other gradual languages, these boundaries are quite restrictive---which
-%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
-%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
-The second claim is justified by our Redex development and its test suite~(\cref{s:redex}).
+The model covers a substantial part of the Python language, including
+module imports, loops, exception handlers, and delete statements.
+The model follows \SP{}'s approach to typing these features.
+
+Because there is no prior formalization of Python and \SP{}, we cannot
+\emph{verify} that the model matches reality.
+Instead, we have applied the well-established method of testing end-to-end
+conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
+with a test suite.
+In particular, we translated 214 tests from the \SP{} regression suite to the
+syntax of the model and confirmed that the results
+do match,\footnote{For most of the 214 tests, the translation is automatic.
+  A few tests required hand-pruning to remove features that the model
+  does not handle (\totalnum{52}).}
+which suggests that the model conforms to actual \SP{}.
+
+Beyond the conformance suite, we have applied two further tests for correctness.
+First, we used Redex's random testing tools~\cite{kf-sfp-2009} to check
+type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
+%% FILL how many programs exactly?
+%% FILL what bugs did testing catch?
+Second, we have manually inspected the model to convince ourselves
+of its soundness.
+Though not a proof, the inspection gives us confidence that the model,
+and by extension \SP{}, is essentially sound.
+
+\paragraph{The Payoff: Issues Reported}
+
+The modelling process helped uncover several critical issues with the
+implementation of \SP{}.
+This is a very important payoff given that \SP{} is running in production at
+Instagram.
+Overall, we reported 25 bug reports to the \SP{} team via GitHub~(listed in
+\cref{a:github-issues}).
+Four of these were soundness issues, one of which could be easily exploited to raise
+a segmentation fault.
+All but one of these soundness bugs have been fixed.
+Five issues were relatively minor; these dealt with confusing error messages
+and incorrect tests.
+The remaining 16 issues report bugs in language design and implementation.
+The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label: \lstinline{sp-correctness}.
+
+To give one example issue, \SP{} accepted the following program whereas our model
+reported a type error:
+%% \url{https://github.com/facebookincubator/cinder/issues/62}
+
+\medskip
+\begin{minipage}[t]{0.25\columnwidth}~
+\end{minipage}\begin{minipage}[t]{0.4\columnwidth}
+\begin{lstlisting}
+  from typing import Optional
+
+  def f(x: Optional[str]) -> str:
+    while True:
+      if x is None:
+        break
+      return x
+
+  print(f(None))
+  # None
+\end{lstlisting}
+\end{minipage}
+
+\noindent{}The function \code{f} expects either a string or the \code{None} value and promises to return a string.
+When called with \code{None}, however, the function breaks out of the \code{while} loop
+and implicitly returns \code{None}---disobeying the return type.
+\SP{} had failed to account for the \code{break} and implicit return.
+More concerningly, the associated test case \emph{expected} the program to type check.
+Creating a formal model helped us catch this specification error.
+
+% In addition to these formal issue reports,
+% we had several long discussions together (between Brown and Instagram)
+% about finer points in the language design.
+
+\paragraph{Section Outline}
+
+The rest of this section illustrates core aspects of our Redex model by way of
+a small formalization.
+\Cref{s:surface} presents a surface syntax to highlight the boundaries between
+typed and untyped code.
+\Cref{s:eval-types} explains how types get enforced at run-time.
+\Cref{s:ts} argues that the overall approach toward boundaries is sound.
 
 \subsection{Surface Syntax and Types}
+\label{s:surface}
 
@@
       \spvardef{\spx}{\sptype}{\spexpr} \langmid
-  \\ & &
       \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
@@
       \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
+  \\ & &
       \sppydict{\spx: \spexpr, \ldots} \langmid
@@
   \\ & &
-      \spapp{\spx}{\spexpr} \langmid
+      \spapp{\spf}{\spexpr} \langmid
       \spdictref{\spexpr}{\spexpr} \langmid
@@
 Functions must have one positional argument.
-Classes must declare one parent (perhaps $\spobject$), one field, and one method.
+Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
 Expressions describe values and simple computations.
 The basic values consist of the following: the none value, integers,
-booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.\footnote{By contrast to
-Python, in which \code{int} is a type and \code{None} is a value, our formalization uses lowercase names for
-values and capitalized names for types.}
+booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
@@
 and method calls.
+
+In Python, the syntax for expressions overlaps with the syntax of types.
+The Python name \code{None}, for example, is both an expression and a type.
+By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
+Expressions use lowercase names (e.g. \code{none}) and types uses capitalized
+names (\code{None}).
+
 Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
@@
 
-The surface types are meant to describe programs, but they are not designed
-to be fully sound.
-For this reason, we omit the surface typing judgment, which scans for
-logical errors, and merely present the declarative syntax of surface types.
-We also omit a well-formedness judgment that checks for circular definitions.
+\SP{} does not enforce all surface types against untyped code, only the evaluation
+types presented in the next subsection~(\cref{s:eval-types}).
+For this reason, we omit the surface typing judgment, which is a kind of linter
+that merely scans typed code for logical errors.
 
@@
 \subsection{Evaluation Types, Casts, and Typing Judgment}
+\label{s:eval-types}
 
@@
 
-The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
-Values that are partly-dynamic are simply incompatible with static types.
-For example, suppose that a function expects a list of numbers.
-In a typical gradual language, a list of dynamic values would be a valid input to the function.
-In \SP{}, a list of dynamic values is not a valid input.
-Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
-Any position where the dynamic type meets a static type needs a cast.
-Other meetings between types can be resolved statically.
-\SP{} does not need to worry that a list of dynamic values might be masquerading
-under a fully-static type.
+%% 2022-01-15: unclear if this is helpful or redundant, hiding for now
+%The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
+%Values that are partly-dynamic are simply incompatible with static types.
+%For example, suppose that a function expects a list of numbers.
+%In a typical gradual language, a list of dynamic values would be a valid input to the function.
+%In \SP{}, a list of dynamic values is not a valid input.
+%Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
+%Any position where the dynamic type meets a static type needs a cast.
+%Other meetings between types can be resolved statically.
+%\SP{} does not need to worry that a list of dynamic values might be masquerading
+%under a fully-static type.
 
@@
 
-\subsection{Type Soundness}
-
-The central question for the model is whether \SP{}'s approach to gradual typing
-is sound.
-No matter how an expression mixes typed and untyped code, its static
-\emph{evaluation type} should be an accurate prediction about the results
-of evaluation.
-Using the metavariable $\spvalue$ to represent heap-allocated Python values
-and $\sprred$ as the evaluation relation (which would need to maintain a heap),
-an approximate soundness theorem is as follows:
-
-\begin{theorem}[Type Soundness]
-  If\ \(~\vdash \spexpr : \sptype\)
-  then one of the following holds:
-  \begin{itemize}
-    \item
-      \(\spexpr \sprred \spvalue
-        \mbox{ and }
-        \vdash \spvalue : \mftypeF{\sptype}
-      \)
-    \item
-      \(\spexpr\) reduces to an allowed error
-    \item
-      \(\spexpr\) diverges
-  \end{itemize}
-\end{theorem}
-
-As mentioned above, our argument for soundness has two parts
-The question at hand is whether the boundaries to type dynamic
-are adequately protected with run-time checks.
-The Redex mechanization's comprehensive test suite provides the other half of the
-argument; namely, that \SP{} is a sound type system for Python.
-
-The direct path to boundary soundness involves two observations.
-First, every evaluation type can be enforced with a decidable cast.
-This is clearly the case for basic values such as floats.
-It true for parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$
-because of their strict semantics.
-Second, all boundaries relate the dynamic type to an evaluation type.
-There are no significant boundaries that relate a partially static type
-to an evaluation type.
+\subsection{Type Boundary Soundness}
+\label{s:ts}
+
+The soundness of our model is easily seen by reduction to Nom, a gradual
+language that allows fine-grained interactions with untyped code and comes with
+a detailed proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
+Given a program written in the syntax of \cref{f:surface-types}, one can derive
+a Nom program with the same type boundaries by replacing every checked
+dict instance with a fresh class type.
+Because distinct checked dict types are incompatible, the
+original boundaries impose the same constraints as the translated Nom ones.
+
+For interested readers, we offer the following direct argument as well.
+Boundaries to less-typed code would be \emph{unsound} if an untyped
+(or partially-untyped) value could enter a typed context without
+a validating check.
+The question is thus whether all boundaries are properly guarded.
+Our answer has two parts.
+First, observe that every evaluation type comes with a decidable
+cast~(\cref{t:cast}).
+For basic types such as floats, a tag check clearly suffices.
+For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
+the exact-match semantics also provides a decidable, though strict, test.
+Second, observe that boundaries can only relate the dynamic type to an
+evaluation type; there are no significant boundaries that relate a partially
+static type to an evaluation type.
 This is due to the strict semantics for checked types and the fact that
 $\sptoptional{\sptdyn}$ normalizes to the dynamic type.
-
-Alternatively, observe that the boundaries in \SP{} are a subset
-of the possible boundaries in Nom, a gradual language that comes with a detailed
-proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
-Given an \SP{} program, we can derive a Nom program with
-the same type boundaries by replacing every checked dict instance
-with a fresh class type.
-Because distinct checked dict instantiations are incompatible,
-the \SP{} boundaries impose the same constraints as the translated Nom
-boundaries.
-
-\subsection{Redex Development, \SP{} Conformance}
-\label{s:redex}
-%% pbt for soundness?
-
-By contrast to the small formalization presented in this section,
-our Redex mechanization covers both type boundaries and a large subset of Python.
-Notable Python features include module imports, loops, exception handlers, and
-delete statements (for class members and attributes).
-To ensure that these features are implemented correctly and soundly, the
-model includes hundreds of test programs adapted from the \SP{} test suite.\footnote{
-  Most of the 214 tests are direct translations of \SP{} conformance tests to
-  the syntax of the model.
-  A program that implements the translation is included with our Redex development.
-  The other 52 tests are not directly expressible in the model; for these, we
-  translated salient aspects of each test by hand.}
-These tests exercise a variety of Python behaviors, thereby giving us confidence
-that the model is sound for Python.
-Some tests intentionally raise type errors, some raise exceptions, and the rest
-compute a specific result; \SP{} agrees on these end-to-end behaviors.
-We have additional applied Redex's random testing tools~\cite{kf-sfp-2009}
-to validate soundness.
-%% FILL how many random tests?
-%% FILL what bugs did random catch?
+Therefore casts at static occurrences of the materialization rule are
+enough to protect all boundaries.
+
+\section{Scaling to Python}
+\label{s:impl}
 
@@
 Some aspects of \SP{} are left out because they are straightforward to
-handle soundly.
-These include primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}.
-Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
-are also omitted.
-Other Python features are left out either because their \SP{} semantics is identical
-to the baseline Python semantics or because they would entail a separate project~(\cref{s:impl}).
-In particular, modelling type-directed optimization would require substantial
+handle.
+These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
+and the boundary to untyped Python~(\cref{s:py-boundary}).
+%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
+%% are also omitted.
+Other Python features are left out because their \SP{} semantics is identical
+to standard Python~(\cref{s:py-boundary}).
+Our model also does not cover the type-directed optimizer~(\cref{s:optimize}).
+Modelling type-directed optimization would require substantial additional
 components: a model of the \SP{} bytecode and a faithful rendering of its
 transformations.
-Our model establishes soundness as a baseline.
-Future work can use soundness as a starting point for compiler verification,
-or informally leverage soundness to reason about possible optimizations.
-
-\subsection{The Payoff: Issues Reported}
-
-While modelling \SP{}, we found several issues with the language
-and submitted them as bug reports to the \SP{} developers.
-Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
-Four of these were critical soundness issues, one of which resulted in
-a segmentation fault.
-All but one of these soundness issue have been fixed.
-Five issues were relatively minor; these dealt with confusing error messages
-and incorrect tests.
-The remaining 16 issues report bugs in language design and implementation.
-The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label (\code{sp-correctness}).
-
-For example, we discovered that \SP{} handled the following function incorrectly when
-we tried to model it:
-%% \url{https://github.com/facebookincubator/cinder/issues/62}
-
-\medskip
-\begin{minipage}[t]{0.25\columnwidth}~
-\end{minipage}\begin{minipage}[t]{0.4\columnwidth}
-\begin{lstlisting}
-  from typing import Optional
-
-  def f(x: Optional[str]) -> str:
-    while True:
-      if x is None:
-        break
-      return x
-
-  print(f(None))
-  # None
-\end{lstlisting}
-\end{minipage}
-
-\noindent{}The function expects either a string or the \code{None} value and promises to return a string.
-When called with \code{None}, however, the function breaks out of the \code{while} loop
-and implicitly returns \code{None}---disobeying the return type.
-\SP{} had failed to account for the \code{break} and implicit return.
-More concerningly, the associated test case \emph{expected} the program to type check.
-Creating a formal model helped us catch this specification error.
-
-% In addition to these formal issue reports,
-% we had several long discussions together (between Brown and Instagram)
-% about finer points in the language design.
-
-\section{Scaling to Python}
-\label{s:impl}
-
-%% TODO SP side channels:
-%% ?? eval,
-%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
-%% and mutable closure cells.
-
-This section explains significant aspects of \SP{} that lie outside the
-scope of our model.
-These include sound interactions with Python code,
-the bytecode language and optimizations, and dynamic Python features
-that \SP{} does not type.
 
 \subsection{Interactions with Open-World Python Code}
+\label{s:py-boundary}
 
@@
 
-Interactions with \emph{open-world}~\cite{vss-popl-2017} Python code pose a minor
+Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
 threat to soundness because
@@
 %% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness
+
+\subsection{Dynamic Python Features}
+\label{s:dynamic-python}
+
+%% TODO 2022-01-09 be careful here ... the model may indeed handle module vars
+%%  the thing is, they're untyped and (maybe) not tested
+
+\SP{} does not ascribe types to the following Python features.
+These are not covered in our model because the implementation simply
+assigns the dynamic type and lets the runtime treat them as untyped
+Python code.
+For each, the dynamic type is a very reasonable choice;
+accurate static types would be burdensome to maintain.
+
+\paragraph{First-Class Classes}
+
+\SP{} does not attempt to type first-class classes: partly because they have
+yet to appear in performance-critical code in the Instagram web server
+and partly because they do not fit well with a nominal type system.
+The straightforward but restrictive approach for first-class nominal class types
+is to force code that uses a first-class class to expect subtypes of a
+particular named static class.
+Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
+but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
+MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
+structural objects.
+How best to add an interface layer in \SP{} is an open question.
+
+% in particular , Python's multiple inheritance would complicate the mixin story.
+% Classic mixins rely on structural types~\cite{bc-oopsla-1990}.
+
+%Notes on first-class classes being untyped:
+%- mixin-based code tends to structrural types
+%- nominality overly restrictive, get forced into <: hierarchy
+%- but don't have NO IDEA for wat to do
+%  + classes and mixins paper shows one idea,
+%    ... need complicated mixin study??? (many sub objects)
+%    built new type system to be mixin-aware, layered atop java
+%    unsure if such extension works for SP
+%    furthermore, class-and-mixins depends on Java single inheritance
+%  + bracha cook mixin pattern
+%  + tate muehlboeck structural object + interfaces
+
+\paragraph{Multiple Inheritance}
+
+%% https://docs.python.org/3/reference/datamodel.html
+%% TODO check Dino email, confirm with C&D that layout conflict is unrelated to vtables
+
+Python allows classes to inherit from a list of parents.
+To resolve method calls to such a class, Python
+dynamically traverses the parent list in a fixed order
+seeking a first match.
+
+Due to the dynamic method resolution order (MRO),
+\SP{} does not track types for classes with multiple parents.
+Method calls to these classes execute the same way as
+in Python, with no type-directed optimizations to improve dispatch.
+
+\paragraph{Dynamic Execution}
+
+%% 2022-01-13: __setattr__ is along the same lines
+%%Carl: Removing the use of super-dynamic Python features like
+%%\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
+%%we’ve done this it hasn’t required a major rewrite.
+
+Results computed by calls to \code{eval} and \code{exec}
+have the dynamic type.
+Their inputs run without being rewritten by the optimizer.
+
+Studies of JavaScript and R have shown that many uses of dynamic execution
+can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
+Assuming these findings carry over to Python, similar adjustments would
+preferable over attempting to type eval.
+
+%% 2022-01-13 covered pretty well earlier
+%%\paragraph{Module-Level Variables}
+%%
+%%Any Python module can read and write to the module-level variables of another module.
+%%When compiling a module, \SP{} therefore assumes that its module-level variables
+%%may be modified by untyped code and assigns the dynamic type.
+%%
+%%Strict modules as an alternative to the Python
+%%semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
+%%If a programmer chooses to declare a module as strict, then its module-level varibles
+%%are immutable and thus typeable.
+%%
+%%Another potential solution is for the Cinder runtime to check that writes to module-level variables
+%%preserve their types.
+%%There are two downsides to this idea:
+%%it will add some performance overhead,
+%%and it will force developers to rewrite untyped code in order to fix any errors that arise.
+%%%% Because of the latter concern, type dynamic is a very reasonable default.
 
@@
 called in a typed context.
-
-\subsection{Dynamic Python Features}
-\label{s:dynamic-python}
-
-%% TODO 2022-01-09 be careful here ... the model may indeed handle module vars
-%%  the thing is, they're untyped and (maybe) not tested
-
-\SP{} does not ascribe types to the following Python features.
-These are not covered in our model because the implementation simply
-assigns the dynamic type and lets the runtime treat them as untyped
-Python code.
-For each, the dynamic type is a very reasonable choice;
-accurate static types would be burdensome to maintain.
-
-\paragraph{First-Class Classes}
-
-\SP{} does not attempt to type first-class classes: partly because they have
-yet to appear in performance-critical code in the Instagram web server
-and partly because they do not fit well with a nominal type system.
-The straightforward but restrictive approach for first-class nominal class types
-is to force code that uses a first-class class to expect subtypes of a
-particular named static class.
-Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
-but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
-MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
-structural objects.
-How best to add an interface layer in \SP{} is an open question.
-
-% in particular , Python's multiple inheritance would complicate the mixin story.
-% Classic mixins rely on structural types~\cite{bc-oopsla-1990}.
-
-%Notes on first-class classes being untyped:
-%- mixin-based code tends to structrural types
-%- nominality overly restrictive, get forced into <: hierarchy
-%- but don't have NO IDEA for wat to do
-%  + classes and mixins paper shows one idea,
-%    ... need complicated mixin study??? (many sub objects)
-%    built new type system to be mixin-aware, layered atop java
-%    unsure if such extension works for SP
-%    furthermore, class-and-mixins depends on Java single inheritance
-%  + bracha cook mixin pattern
-%  + tate muehlboeck structural object + interfaces
-
-\paragraph{Multiple Inheritance}
-
-%% https://docs.python.org/3/reference/datamodel.html
-%% TODO check Dino email, confirm with C&D that layout conflict is unrelated to vtables
-
-Python allows classes to inherit from a list of parents.
-To resolve method calls to such a class, Python
-dynamically traverses the parent list in a fixed order
-seeking a first match.
-
-Due to the dynamic method resolution order (MRO),
-\SP{} does not track types for classes with multiple parents.
-Method calls to these classes execute the same way as
-in Python, with no type-directed optimizations to improve dispatch.
-
-\paragraph{Dynamic Execution}
-
-%% 2022-01-13: __setattr__ is along the same lines
-%%Carl: Removing the use of super-dynamic Python features like
-%%\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
-%%we’ve done this it hasn’t required a major rewrite.
-
-Results computed by calls to \code{eval} and \code{exec}
-have the dynamic type.
-Their inputs run without being rewritten by the optimizer.
-
-Studies of JavaScript and R have shown that many uses of dynamic execution
-can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
-Assuming these findings carry over to Python, similar adjustments would
-preferable over attempting to type eval.
-
-%% 2022-01-13 covered pretty well earlier
-%%\paragraph{Module-Level Variables}
-%%
-%%Any Python module can read and write to the module-level variables of another module.
-%%When compiling a module, \SP{} therefore assumes that its module-level variables
-%%may be modified by untyped code and assigns the dynamic type.
-%%
-%%Strict modules as an alternative to the Python
-%%semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
-%%If a programmer chooses to declare a module as strict, then its module-level varibles
-%%are immutable and thus typeable.
-%%
-%%Another potential solution is for the Cinder runtime to check that writes to module-level variables
-%%preserve their types.
-%%There are two downsides to this idea:
-%%it will add some performance overhead,
-%%and it will force developers to rewrite untyped code in order to fix any errors that arise.
-%%%% Because of the latter concern, type dynamic is a very reasonable default.
 

### 2022-01-16 pj-7.1/paper.tex
pj: more sec 3, 4 fixes thanks sk

@@
     Precise types for higher-order functions and first-class classes
-    are thus a low priority.
+    are thus a low priority; instead, the dynamic type is available as a coarse
+    approximation.
   \item
@@
   \item
-    Methods and fields may be overridden in arbitrary ways by untyped code.
+    Non-final methods and fields may be overridden in arbitrary ways by untyped code.
     In typed code, however, overrides must be compatible subtypes.
@@
     that returns the dynamic type; \cref{f:gg-failure-stat} shows the static error
-    that arises from such a static override.
+    that arises from such an override.
 
@@
 protocol; it is not a from-scratch development.
-That being said, Cinder vtables are specialized to check untyped overrides
+Additionally, however, Cinder vtables are specialized to check untyped overrides
 of typed methods against their expected return type (instead of using a layer of
-indirection for the check).
+indirection for the check, as found languages that enforce types with higher-order
+wrappers~\cite{clzv-ecoop-2018,gf-icfp-2018}).
 
@@
 
-\subsection{Checked Data Structures}
+\subsection{Concrete Data Structures}
 \label{s:chkdict-impl}
@@
   \item
-    First, the constructor uses a type and a Python dictionary ($\sptrawpydict$)
-    to initialize a checked dictionary.
-    This constructor ensures that all values of type $\sptchkdict{K}{V}$
-    begin with well-typed elements.
-  \item
-    Second, all write operations check their input.
+    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
+    It checks that the elements of the dictionary match the key and value types.
+  \item
+    All write operations check their input.
     Every operation that mutates or extends a checked dictionary must
@@
   \item
-    Third, the dictionary value stores a type tag to support casts from the dynamic type.
-    When a checked dictionary enters typed code from an untyped context,
-    the runtime checks its type tag to see whether the key and value types are an exact match.
+    The dictionary value stores a compound type tag 
+    (e.g. one token representing a three-part type $\sptchkdict{K}{V}$)
+    to support casts from the dynamic type.
+    When a concrete dictionary enters typed code from an untyped context,
+    its tag must match the expected type exactly.
     For example, type $\sptchkdict{K}{V}$ matches the type
     $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
-    and $V$ is equal to $\sptdyn$,
+    and $V$ is equal to $\sptdyn$.
 \end{itemize}
@@
 module imports, loops, exception handlers, and delete statements.
-The model follows \SP{}'s approach to typing these features.
+It follows \SP{}'s approach to typing these features.
+The model is available at the following URL:
+
+\smallskip
+{\centering\url{https://github.com/brownplt/insta-model}}
 
@@
 
-\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
+\Cref{f:surface-types} presents a core surface syntax for model programs.
 A program is a sequence of statements;
@@
 $\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
-There is no function type in the model, or in \SP{} for that matter.
-The only way to define a higher-order function is to use the dynamic type.
 
@@
 Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
-This latter property means that blame-tracking is trivial because
+This latter property means that blame-tracking~\cite{ff-icfp-2002} is trivial because
 casts either succeed or fail immediately.
@@
     \sptdyn & \mbox{No check needed, accepts any value} \\
-    \sptobject & \mbox{Accepts any value (except primitives,~\cref{s:c-types})} \\
+    \sptobject & \mbox{Accepts any value, except primitives~(\cref{s:c-types})} \\
     \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
@@
 Relative to the current environment, the expressions within each statement must
-satisfy the typing judgment in~\cref{f:eval-types}
+satisfy the typing judgment in~\cref{f:eval-types}.
 
@@
 \medskip
-Consequently, this type derivation calls for two casts at run-time.
+Consequently, this derivation calls for two casts to match the \trule{Matr} applications.
 
@@
 
-Furthermore, our casts-via-materialization rule is more aggressive
-than what \SP{} actually does to insert casts.
-Because \SP{} supports interactions with unanalyzed Python modules, it inserts
-casts in an overapproximate way:
+Furthermore, our casts-via-materialization rule assumes that the
+type checker has access to the whole program.
+In practice, \SP{} approximates our ideal casts in two steps
+to support interactions with unanalyzed Python modules.
+First, \SP{} inserts casts in an overapproximate way:
 \begin{itemize}
@@
   \item
-    checked dictionaries check all writes; and
+    concrete-typed dictionaries check all writes; and
   \item
@@
 \end{itemize}
-\Cref{s:impl} explains
-how the type-directed optimizer helps avoid some
-of these checks.
+Second, the type-directed optimizer skips casts wherever a well-typed context
+interacts with a typed value.
+\Cref{s:impl} explains the optimizer in more detail.
 
@@
 For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
-the exact-match semantics also provides a decidable, though strict, test.
+the exact-match semantics also provides a decidable test.
 Second, observe that boundaries can only relate the dynamic type to an
@@
 static type to an evaluation type.
-This is due to the strict semantics for checked types and the fact that
+This is due to the exact-match semantics for checked types and the fact that
 $\sptoptional{\sptdyn}$ normalizes to the dynamic type.
@@
 to standard Python~(\cref{s:py-boundary}).
-Our model also does not cover the type-directed optimizer~(\cref{s:optimize}).
-Modelling type-directed optimization would require substantial additional
+Our model also does not cover the optimizer~(\cref{s:optimize}) because
+a formal account of type-directed optimization would require substantial additional
 components: a model of the \SP{} bytecode and a faithful rendering of its

### 2022-01-16 pj-7.1/paper.tex
pj: abstract, sec 8 fixes

@@
 \newcommand{\numbenchmark}{2}
+\newcommand{\numSPtest}{214}
 \newcommand{\CPUchange}{3.7\%}
@@
 \author[a]{Kuang-Chen Lu}
-\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
+\authorinfo{(\email{LuKuangchen1024@gmail.com}) } %% { is a PhD student at Brown University.}
 \affiliation[a]{Brown University, Providence, RI, USA}
 \author[a]{Ben Greenman}
-\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
+\authorinfo{(\email{benjamin.l.greenman@gmail.com}) } %% {is a PLT member and a postdoc at Brown University.}
 \author[b]{Carl Meyer}
-\authorinfo{(\email{carljm@fb.com}) FILL}
+\authorinfo{(\email{carljm@fb.com}) } %% FILL
 \affiliation[b]{Meta, Menlo Park, CA, USA}
 \author[b]{Dino Viehland}
-\authorinfo{(\email{dinoviehland@fb.com}) FILL}
+\authorinfo{(\email{dinoviehland@fb.com}) } %% FILL
 \author[a]{Shriram Krishnamurthi}
-\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}
+\authorinfo{(\email{shriram@brown.edu}) } %% {is the Vice President of Programming Languages (no, not really) at Brown University.}
 
@@
   costs of the transient approach.
-  On a more technical level, this paper contributes two artifacts:
-  a model of \SP{} that passes property-based tests and hundreds of tests from the \SP{} codebase,
-  and a performance evaluation of \SP{}. The process of formalization
+  On a more technical level, this paper describes two contributions: a model of
+  \SP{} and a performance evaluation of \SP{}. The process of formalization
   found several errors in the implementation, including fatal errors.
@@
   Our model of \SP{} is implemented in PLT Redex and tested using
-  property-based soundness tests and the \SP{} regression suite.
+  property-based soundness tests and \numSPtest{} tests from the \SP{} regression suite.
   This paper includes a small core of the model to convey the main ideas
@@
   web server.
-  Thanks to \SP{}, global CPU usage has improved by \CPUchange{} within a year.
+  Converting modules to \SP{} over the past year has led to a \CPUchange{}
+  improvement in CPU usage in the web server.
 
@@
 values created by untyped code.
-Rather than search for a coordinated set of edits to make concrete
-types work, developers may prefer a whole-sale migration to
-a typed language.\footnote{Successful migrations of untyped codebases
-to a typed language are rare, but not unheard of.
-Twitter ported its server-side code from Ruby to Scala~\cite{twitter-scala}
-and Dropbox moved its core sync engine from Python mypy to Rust~\cite{dropbox-rust}.}
 
@@
 We call the incomplete ones \emph{shallow} types.
-The \SP{} strategy for fully-sound generics is to package each value with a type
+The \SP{} strategy for sound generics is to package each value with a type
 tag; following the \textsf{StrongScript} authors~\cite{rzv-ecoop-2015}, we call
@@
 Parameterized types (generics) are more difficult to enforce.
-Consider a dictionary datatype \code{Dict[K,V]} that is parameterized by types
-for its keys and values.
-If a function expects a dict from integers to integers (\code{Dict[Int,Int]}),
-then merely checking for a dict instance is not enough to decide whether an
-incoming value matches the type.
-For example, the incoming value may have been created by untyped code.
-
+For example, suppose that a typed function expects inputs that match
+the type \code{Dict[Int,Int]}; that is, dictionaries with integer keys and
+integer values.
+A constant-time check for dictionaries is not enough to protect the typed
+function because it fails to validate the elements inside a dictionary.
 The \SP{} solution is to provide two incompatible versions of every parameterized type,
@@
   \item
-    \emph{Shallow type $\sptrawpydict$}.
-    The default type $\sptrawpydict$ gets enforced with a constant-time check
+    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
     for dict values that ignores the elements inside a dict.
-    Consequently, a parameterized static type such as $\sptpydict{\sptint}{\sptint}$
+    Consequently, a parameterized surface type such as $\sptpydict{\sptint}{\sptint}$
     is merely a promise of dict values at run-time; it does not promise anything about
@@
   \item
-    \emph{Concrete type $\sptchkdict{\sptype}{\sptype}$}.
-    \SP{} introduces a parameterized type for concrete, checked dictionaries
-    along with a value constructor that records its type constraints.
-    \cref{s:chkdict-impl} discusses the implementation further.
+    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
+    keys and values, but is incompatible with built-in Python dictionaries.
+    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
 \end{enumerate}
@@
 with a test suite.
-In particular, we translated 214 tests from the \SP{} regression suite to the
+In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
 syntax of the model and confirmed that the results
-do match,\footnote{For most of the 214 tests, the translation is automatic.
+do match,\footnote{For most of the \numSPtest{} tests, the translation is automatic.
   A few tests required hand-pruning to remove features that the model
@@
 point for such an experiment.
+Gradual, progressive soundness may prove itself to be an effective general strategy for
+gradual types in practice.
 

### 2022-01-16 pj-7.1/paper.tex
pj: global -> 'web server', remove author bios

@@
 and only a few critical modules rely on concrete types and primitive types (dozens),
-global CPU usage has improved by \CPUchange{} according to internal profiling tools.
+web server CPU usage has improved by \CPUchange{} according to internal profiling tools.
 Type-directed optimizations outweigh the cost of enforcing soundness.
@@
 Instagram's internal profiling tools, which continuously monitor
-performance, attribute a \CPUchange{} improvement in global CPU to \SP{} conversions.
+performance, attribute a \CPUchange{} improvement in web server CPU usage to \SP{} conversions.
 \Cref{t:prod-stat} reports a few other production statistics.
@@
     8     & months in production   & 541   & typed, strict modules \\
-    \CPUchange{} & global CPU improvement &   9   & using concrete types \\
+    \CPUchange{} & web server CPU improvement &   9   & using concrete types \\
     0     & net regressions        &  12   & using primitive types \\
@@
 Furthermore, \SP{} types combined with Cinder resulted in a
-0.7\% reduction in global CPU usage.
+0.7\% reduction in web server CPU usage.
 Primitive types~(\cref{s:c-types}) were essential for matching Cython.
@@
 modules and proposed types plus small code changes to the maintainers of these modules.
-The accepted proposals resulted in a 1\% improvement in global CPU usage.
+The accepted proposals resulted in a 1\% improvement in web server CPU usage.
 During the second half of 2021, the \SP{} team applied the same
@@
 this one change brought over four hundred generated modules into the typed fold.
-Overall, the second-half changes added a 2\% global CPU improvement.
+Overall, the second-half changes added a 2\% web server CPU improvement.
 
@@
 Quite naturally, it cannot find bugs in code that it does not analyze!
-Our point is simply that optional type systems can miss a variety of issues.
+The takeaway is simply that optional type systems can miss a variety of issues.
 

### 2022-01-16 pj-7.1/paper.tex
pj: reword CPU utilization

@@
   web server.
-  Converting modules to \SP{} over the past year has led to a \CPUchange{}
-  improvement in CPU usage in the web server.
+  Migrations to \SP{} in the server have caused a \CPUchange{} increase in
+  requests handled per second at max CPU utilization.
 
@@
 and only a few critical modules rely on concrete types and primitive types (dozens),
-web server CPU usage has improved by \CPUchange{} according to internal profiling tools.
+these migrations have resulted in a \CPUchange{} increase in production requests handled
+per second for servers at max CPU utilization.
 Type-directed optimizations outweigh the cost of enforcing soundness.
@@
 though types are needed to enable the most-effective optimizations.
-
-%% 2022-01-05 Carl:
-%% "All the data I’ve been sharing is from a single codebase, the primary IG web
-%% server monolith. The existence of JITted but not SP modules is mostly just due
-%% to the fact that opting a function into the JIT is semantically transparent and
-%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
-%% quite easy to do (the only tradeoff pushing us away from JITting everything is
-%% just how much CPU we spend JIT-compiling and how much memory we spend storing
-%% JIT-compiled code), whereas opting a module into SP requires more work so the
-%% bottleneck is available developer time to do conversions."
 
@@
 
-%% 2022-01-14 TODO re 3.7%
-%% That percentage represents the aggregate CPU improvement observed from all of
-%% our individual diffs that converted modules to Static Python over the past
-%% year. In a sizable number of cases a conversion to Static Python did mean that
-%% functions in that module started being JITted that previously were not JITted,
-%% because for most of the half we had a policy in place that all Static Python
-%% modules are always JITted. (We’ve now decoupled these things for future
-%% conversions, so opting functions into the JIT is now orthogonal to opting them
-%% into SP.)
-%% 
-%% So the 3.7% does not include any CPU wins that Cinder or the JIT provide in any
-%% non-static modules, it only includes improvements observed from specifically
-%% converting a module from nonstatic to static. But it does in some cases include
-%% the effects of going from nonstatic + no-JIT to SP+JIT for a given module. In
-%% other cases a conversion may have gone from nonstatic+JIT to SP+JIT.
-%% Unfortunately we don’t have any way to go back and tease apart these cases for
-%% each individual module conversion.
-%% 
-%% What we are doing now is an experiment where we turn off SP in a set of hosts
-%% and observe them relative to the rest of prod, without changing the JIT list.
-%% This will show some % regression, which we expect to be some number <3.7%, and
-%% that will give a solid figure for the benefit we are getting specifically from
-%% SP across all static modules, totally independent of the JIT. Unfortunately
-%% there have been some hiccups in this experiment and we don’t have that number
-%% ready yet, but we hope to have it in the next week or so.
-
-%% 2022-01-14 FILL TODO 
-%% > I still have trouble understanding the CPU number. There must be many
-%% > servers running many workloads and some statistics to merge them. Do
-%% > you have a blog post or something that explains the profiling in more
-%% > detail?
-%% 
-%% I don’t know of anything reasonably current in terms of blog posts. For IG web
-%% prod, there are not “many workloads” — there is no specialization in the IG web
-%% tier. There are many thousands of hosts, all running the exact same workload —
-%% all those hosts are handling an undifferentiated random selection from a single
-%% giant pool of incoming IG requests. So for all intents and purposes we can
-%% consider them identical. (There are in practice some differences in the hosts
-%% themselves — e.g. different generations of CPU hardware — that we do have to
-%% take into account when running the experiments below.)
-%% 
-%% The details of how we measure a CPU regression or improvement are interesting,
-%% though. What we do is we take a set of hosts that we want to measure (typically
-%% an experiment group and a control group, where the control group is identical
-%% to the rest of production and the experiment group has some change made to its
-%% code or configuration whose effect we want to measure), and we slowly increase
-%% the traffic to those hosts until they are at max CPU utilization — they cannot
-%% handle any more traffic without falling over. Then we measure the
-%% requests-per-second handled by each set of hosts; if the experiment group is
-%% able to handle 2% more requests per second than the control group, while both
-%% are at max CPU utilization, this effect is considered a 2% global CPU win.
-%% 
-%% This metric has the advantage that it is highly holistic and results-oriented;
-%% it is a measure of our actual capacity to serve more requests with fewer hosts,
-%% and thus it can subsume and account for, and give apples-to-apples comparison
-%% of, many different possible reasons for an increase or decrease in a host’s
-%% efficiency in serving requests.
-%% 
-%% The disadvantage of the metric is that the raw number it measures (requests per
-%% second) is also highly dependent on the product code and client behavior at
-%% that moment in time, and is not stable over time, because the average cost in
-%% CPU cycles of “one request” is not stable over time. If the behavior of the IG
-%% iOS or Android app changes so that it is sending more heavy requests and fewer
-%% lightweight requests, RPS across the fleet will naturally decrease even though
-%% the efficiency of the servers has not changed. Therefore, this metric is really
-%% good for saying “at X point in time, Y change has a Z% overall effect on our
-%% CPU efficiency” but it cannot be used to compare raw numbers from two different
-%% points in time; we cannot use it to say “between X1 point in time and X2 point
-%% in time, RPS increased Z%, therefore our efficiency has increased by Z%.”
-%% 
-%% So the 3.7% number we gave you is an aggregate win from doing many such
-%% individual experiments over the course of the past year, where each experiment
-%% was measuring the effect of converting X module (or small set of modules) from
-%% nonstatic to static (where in some cases it was nonstatic+JIT to SP+JIT, and in
-%% other cases it was nonstatic noJIT to SP+JIT.) And the updated number we hope
-%% to get you soon is an overall measure from a single experiment, today, of the
-%% effect of disabling SP everywhere, while leaving the JIT list untouched. We
-%% expect the latter number to be smaller than 3.7% because it will remove the JIT
-%% factor entirely, but there’s also an unknown factor (which could net positive
-%% or negative) of how the behavior of long-ago-converted SP modules has changed,
-%% or the capabilities of SP has changed, since their conversion, which changes
-%% the net win of SP on those modules since the time when we converted them.
-
 %% 2022-01-04 TODO have no evidence that "any module" typing works out ... team
@@
 Overall, the results are very encouraging.
-Instagram's internal profiling tools, which continuously monitor
-performance, attribute a \CPUchange{} improvement in web server CPU usage to \SP{} conversions.
-\Cref{t:prod-stat} reports a few other production statistics.
-Over 500 modules have been converted thus far.
+Instagram's internal profiling tools attribute a \CPUchange{} improvement
+in CPU utilization~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
+This is a big improvement at Instagram scale.
+
+Developers have converted 541 modules to \SP{} thus far.
 Despite some initial regressions, none of the converted modules
-ran slower after small rewrites and enabling the JIT on certain functions.
-
-\begin{table}
-  \caption{Overview: \SP{} in production}
-  \label{t:prod-stat}
-  \begin{tabular}{l@{~~}l@{\qquad}l@{~~}l}
+ran slower after small rewrites~(\cref{s:code-changes})
+and/or enabling the JIT on certain functions.
+Only 9 modules use concrete types and only 12 modules use primitive types,
+but these features delivered critical performance in key modules.
+
+%% all prims are int32 and int64 and cbool
 %% 541 modules out of ~45000 total
-    8     & months in production   & 541   & typed, strict modules \\
-    \CPUchange{} & web server CPU improvement &   9   & using concrete types \\
-    0     & net regressions        &  12   & using primitive types \\
-    %% all prims are int32 and int64 and cbool
 %% 873 modules use JIT for some functions (about to increase by OOM)
-  \end{tabular}
-\end{table}
+
+\subsection{How to Interpret the CPU Utilization Result}
+\label{s:what-is-cpu}
+
+The Instagram web server code runs on a large number of host machines, each of
+which continuously handles requests from a common pool.
+Improvements to the server codebase should make these machines more efficient
+as they handle arbitrary requests.
+
+To measure the effect of a code change on these hosts, profiling tools select
+two representative sets of machines to run as experimental and control groups.
+The experimental groups gets the latest version of the server code;
+the control group gets the previous version.
+Next, the profiling tools slowly increase the number of requests that these
+machines receive until each is running at max CPU utilization.
+Once the machines are fully allocated, the tools measure how many requests
+per second each group is able to handle.
+If the experimental group can serve more requests per second, then the change
+is recorded as a success.
+Assuming that the experimental and control groups have equivalent hardware
+capabilities, and assuming that both groups receive a representative sample
+of requests, then the results of an experiment (increase/decrease in CPU utilization)
+should predict actual performance.
+
+The main threat to validity is that this CPU utilization number depends on
+product code and on client behavior.
+Over time, the size of the average request may increase and thereby increase
+the CPU cost of serving one request.
+Nevertheless, CPU utilization is the primary metric that Instagram
+uses for production changes.
 
 \subsection{Migration Path}
+\label{s:migration-path}
 
@@
 few critical Cython modules.
-Cython had improved performance by compiling these modules to C,
+Cython improved performance by compiling these modules to C,
 but its partial adoption had led to an awkward workflow (with a
@@
 \subsection{Analyzing Code Changes}
+\label{s:code-changes}
 
@@
 
-%% TODO section formatting looks good?
 \subsubsection{Code Changes for Performance}
@@
 \subsection{Microbenchmarks}
+\label{s:microbenchmarks}
 %% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks

### 2022-01-16 pj-7.1/paper.tex
pj: cleanup everywhere, esp. 3 and 6

@@
 
-\usepackage[framemethod=TikZ]{mdframed}
-\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
+\usepackage{tikz}
 \usepackage{amsthm}
@@
 codebase and while leaving the rest untyped.
-Run-time checks enforce the critical assumptions that typed code makes about
+Run-time casts and checks enforce the critical assumptions that typed code makes about
 untyped components, thereby ensuring that the types are sound no matter how
@@
 gradual types, formulating correctness properties and studying ever-more-descriptive
-types~(e.g.~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
+types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
 A few academics have studied the cost of run-time checks
@@
 bytecode, and a method-based JIT compiler.
-In essence, \SP{} is a type system for a new language.
-The interface that it offers to programmers, however, seeks to replicate the
+In essence, \SP{} is the type system for an entirely new language.
+The interface that it offers to programmers, however, replicates the
 Python experience.
-\SP{} runs untyped Python 3.8 programs with minimal changes to their behavior,
+\SP{} runs untyped Python 3.8 programs with minimal changes to their
+behavior,\footnote{\Cref{s:behavior-changes} lists differences with Python 3.8 that
+have caused friction in practice.}
 and it compiles code on-the-fly to give the same workflow and enable the same
@@
 
-%% TODO clean up arrow between subfigs
 \begin{figure}
-  \begin{subfigure}[t]{0.15\columnwidth}\noindent\begin{lstlisting}
+  \begin{minipage}{0.15\columnwidth}\noindent\begin{lstlisting}
 
@@
   \end{lstlisting}
-\end{subfigure}\(~~\rightarrow~~\)\begin{subfigure}[t]{0.32\columnwidth}\noindent\begin{lstlisting}
+\end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.32\columnwidth}\noindent\begin{lstlisting}
 from __static__ import PyDict
@@
   \end{lstlisting}
-  \end{subfigure}\(~~\rightarrow~~\)\begin{subfigure}[t]{0.34\columnwidth}\noindent\begin{lstlisting}
+  \end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.34\columnwidth}\noindent\begin{lstlisting}
 from __static__ import ChkDict
@@
   \end{lstlisting}
-  \end{subfigure}
+  \end{minipage}
 
@@
     As the body of \code{f} illustrates, the syntax for using a checked dict is standard.
-    Creating a checked dict, however, requires a parameterized constructor call.
+    Creating a checked dict, however, requires a type-parameterized constructor call.
 \end{itemize}
@@
 and prior work on types for Python.
-Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~\cite{pep484} are two notable sources.
+Pyre~\cite{pyre}, Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, and PEP
+484~\cite{pep484} are three notable sources.
 In addition, \SP{} is strongly influenced by the following engineering constraints:
@@
   \item
-    The codebase in which \SP{} is being applied makes heavy use of Pyre~\cite{pyre}
+    The codebase in which \SP{} is being applied makes heavy use of Pyre
     type annotations.
@@
 In general, the types that \SP{} implements all enable significant optimizations
-and can be validated with constant-time dynamic checks.
+and can be validated with constant-time casts.
 Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.
@@
 with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
-but are, for the most part, not supported at run-time.
-Only binary unions of a type and the Python \code{None} value get enforced with a run-time check
-($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).
+but are, for the most part, not supported at run-time~(\cref{s:eval-types}).
+The only exception is that binary unions of a type and the Python \code{None}
+value get enforced with a run-time check ($\sptunion{\sptype, \sptnone}$, or
+$\sptoptional{\sptype}$).
 
@@
 For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
-(i.e. it is a misspelling of $\sptfloat$) and that a
+(i.e., it is a misspelling of $\sptfloat$) and that a
 program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
@@
 Unlike structural gradual type systems, \SP{} does not traverse
-incoming values nor does it create wrappers to enforce behavior.
+incoming values, nor does it create wrappers to enforce behavior.
 
@@
 The \SP{} solution is to provide two incompatible versions of every parameterized type,
-both of which support efficient run-time checks.\footnote{Currently, only a handful of built-in data structures provide shallow
-and concrete versions. User-defined classes will soon follow this approach.}
+both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
+and concrete versions at present. In particular, concrete user-defined classes have yet to gain concrete support.}
 For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
@@
 nonlocal changes.
-If one type changes to concrete, then all data constructors that reach this type need
-to change \emph{and} all their clients need to use the concrete type as well.
+If one type changes to concrete, then several changes may need to happen:
+all values that reach this type must be initialized as concrete,
+all typed clients of these values must expect concrete values,
+all values that reach those clients must be concrete,
+and so on and so forth.
 
@@
 Shallow types pay a constant-time cost whenever an untyped value enters typed code
-and whenever a typed context reads from a shallow value.\footnote{The predicted
-costs for reads and writes assume that all \SP{} types can be validated with a
-constant-time check.}
+and whenever a typed context reads from a shallow value.
 The net cost of reads may be high.
@@
     \sptrawpydict                &           0 & O(1) & O(1) &     0 \\
-    \sptchkdict{\sptype}{\sptype} &        O(N) & O(1) &    0 &  O(1)
+    \sptchkdict{K}{V} &        O(N) & O(1) &    0 &  O(1)
   \end{array}\)
@@
   \item
-    To benefit from types, classes must be declared on the module top level
+    To benefit from types, classes must be declared at the module top level
     and may have at most one parent.
@@
 Example types include \code{int64}, \code{uint64}, \code{double}.
-\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives.
+There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
 
@@
     The dictionary value stores a compound type tag 
-    (e.g. one token representing a three-part type $\sptchkdict{K}{V}$)
+    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
     to support casts from the dynamic type.
@@
 By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
-Expressions use lowercase names (e.g. \code{none}) and types uses capitalized
+Expressions use lowercase names (e.g., \code{none}) and types uses capitalized
 names (\code{None}).
@@
 
+  \centering
   \(\begin{array}{l@{\quad}l}
@@
     \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
-    %% \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
     \sptdyn & \mbox{No check needed, accepts any value} \\
@@
 There are two case in which \SP{} types impose constraints on untyped Python code.
-First, concrete generic types (e.g. $\sptchkdict{\sptype}{\sptype}$)
+First, concrete generic types (e.g., $\sptrawchkdict$)
 reject inputs that were not created via a checked contructor.
@@
 For this reason, \SP{} compiles every typed function to check its inputs.
-Similarly, concrete types such as $\sptchkdict{\sptype}{\sptype}$ also
+Similarly, concrete types such as $\sptrawchkdict$ also
 check their writes.
@@
 
-Results computed by calls to \code{eval} and \code{exec}
-have the dynamic type.
-Their inputs run without being rewritten by the optimizer.
+Results computed by calls to \code{eval} and \code{exec} have the dynamic type.
+The type system makes no finer prediction about what kind of value such calls will
+return.
 
@@
 Assuming these findings carry over to Python, similar adjustments would
-preferable over attempting to type eval.
+preferable over attempting to type \code{eval}.
 
@@
 to construct \SP{} data structures, and to perform actions in a way that benefits from
-static information..
+static information.
 
@@
 
+  \centering
   \begin{tabular}{lll}
@@
 
-The Instagram web server has been running \SP{} in production since April 2021.
+The Instagram web server has been using \SP{} code to serve requests in
+production since April 2021.
 Overall, the results are very encouraging.
@@
 
-Developers have converted 541 modules to \SP{} thus far.
+Developers have converted over 500 modules to \SP{} thus far.
 Despite some initial regressions, none of the converted modules
@@
 During the first half of 2021, the \SP{} team identified critical
-modules and proposed types plus small code changes to the maintainers of these modules.
+modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
 The accepted proposals resulted in a 1\% improvement in web server CPU usage.
@@
 are from hand conversions (124).
-These modules are well-mixed with untyped modules in the codebase.
+These modules frequently interact with untyped modules in the codebase.
 According to an analysis of identifiers that cross module-dependence
@@
 Because the \SP{} team has been changing code as well as types during its
-migrations, the question arises as to whether the code changes raise any
-threats to validity.
-It could be that the \SP{} approach to gradual typing only pays off
-after significant modifications.
+migrations, the question arises as to whether the changes themselves
+are responsible for more speedups than the types.
 %% Carl: "I reviewed the ~30 conversion diffs"
 We therefore reviewed the $\sim 30$ conversion diffs that significantly improved
-performance.
+performance to get a sense of the code changes at hand.
 
@@
 \subsubsection{Code Changes for the Type Checker}
+\label{s:behavior-changes}
 
@@
     Three modules currently use a concrete dictionary type
-    (of the form $\sptchkdict{\sptype}{\sptype}$)
+    (of the form $\sptchkdict{K}{V}$)
     and six modules use a checked list type.
@@
 The \SP{} team proposed a rewrite from multiple inheritance to single inheritance,
-but the code maintainers preferred to keep the design as-is.
+but the code maintainers preferred to keep the design as-is instead of enabling
+type-driven method dispatch optimizations.
 
@@
 
-\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
+\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.\footnote{\url{https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks}}
 These microbenchmarks are admittedly small and quite different than typical application code,
@@
   \label{t:microbenchmark}
+
+  \centering
   \begin{tabular}{lrr}
@@
 
-Concrete types originated in Thorn~\cite{bloom,wnlov-popl-2010}
+Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
 and StrongScript~\cite{rzv-ecoop-2015},
@@
       \item
-        Argument splatting, i.e. applying a function that expects positional
+        Argument splatting, i.e., applying a function that expects positional
         arguments to a list (\code{f(*lst)}), is not statically checked
@@
   \label{t:mb:richards}
-  \footnotesize
+  \footnotesize \centering
 
@@
   \label{t:mb:fannkuch}
-  \footnotesize
+  \footnotesize\centering
 
@@
   \label{t:mb:commands}
-  \tiny
+  \tiny\centering
 

### 2022-01-18 pj-7.1/paper.tex
pj: draft running title + authors
rewrote soundness text

@@
 
-%\titlerunning{short title}
+\titlerunning{DRAFT do not distribute}
 
@@
 
-%\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
+\authorrunning{DRAFT do not distribute}
 
@@
 It follows \SP{}'s approach to typing these features.
-The model is available at the following URL:
-
-\smallskip
-{\centering\url{https://github.com/brownplt/insta-model}}
-
-Because there is no prior formalization of Python and \SP{}, we cannot
-\emph{verify} that the model matches reality.
-Instead, we have applied the well-established method of testing end-to-end
+The model is available at:
+\begin{center}
+  \url{https://github.com/brownplt/insta-model}
+\end{center}
+Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
+\emph{verify} that the model matches them. Instead, we have applied
+three methods to give confidence that our model matches reality.
+First, we manually reviewed the model in depth---using our
+own expertise---to look for non-conformance.  Second, 
+we used Redex's random testing tools~\cite{kf-sfp-2009} to check
+type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
+%% FILL how many programs exactly?
+%% FILL what bugs did testing catch?
+Finally, we employed the well-established method of testing end-to-end
 conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
@@
 which suggests that the model conforms to actual \SP{}.
-
-Beyond the conformance suite, we have applied two further tests for correctness.
-First, we used Redex's random testing tools~\cite{kf-sfp-2009} to check
-type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
-%% FILL how many programs exactly?
-%% FILL what bugs did testing catch?
-Second, we have manually inspected the model to convince ourselves
-of its soundness.
-Though not a proof, the inspection gives us confidence that the model,
-and by extension \SP{}, is essentially sound.
 

### 2022-01-19 pj-7.1/paper.tex
pj: comment from carl about fastlen

@@
 \newcommand{\numSPtest}{214}
+\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
+\newcommand{\numSPdiff}{588} %% (\numSPtotal - \numSPtest)
 \newcommand{\CPUchange}{3.7\%}
@@
 and prior work on types for Python.
-Pyre~\cite{pyre}, Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, and PEP
-484~\cite{pep484} are three notable sources.
-In addition, \SP{} is strongly influenced by the following engineering constraints:
+Pyre~\cite{pyre}, an optional type checker developed by Facebook,
+greatly influenced the syntax and types of \SP{}.
+Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, a research language with compelling
+performance results, set the basic strategy for run-time checks.
+\SP{}'s novelty comes from the following engineering constraints:
 \begin{itemize}
@@
     type annotations.
-    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
-    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
-    % of types that enable optimizations.
-    % depth-first types.
-    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
+    \SP{} is therefore compatible with Pyre syntax to reduce the
+    adoption burden on its target users.
+    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
   \item
@@
 and applies it as an optimization.
+
+%% TODO 2022-01-16 clarify why fast_len is interesting at all:
+%% Carl: Two parts to the answer here. In the early days of Static Python we were
+%% focused on matching Cython in some very hot code paths, and a couple of those
+%% did take length of containers. So we implemented some things like FAST_LEN that
+%% may not be broadly applicable but we needed to be as fast as possible in those
+%% few Cython conversions. But since then we also implemented a general
+%% optimization for iteration of loops where the container is known to be an exact
+%% list (or array or tuple IIRC): when we see e.g. `for item in mylist` and we
+%% know statically that `mylist` is an exact list/tuple/array, instead of
+%% generating the usual bytecode that goes through the dynamic iterator protocol,
+%% we will generate more efficient bytecode that uses `FAST_LEN` to get the length
+%% and then uses an incremented (primitive) index counter and our optimized
+%% `SEQUENCE_GET` opcode (without bounds checking since in this case we can
+%% guarantee the index counter can never be out of bounds). The visible semantics
+%% remain the same but under the hood we use numeric indexing instead of the
+%% iterator protocol. So this optimization is much more generally applicable, and
+%% relies on FAST_LEN.
+%%
+%% Also I think FAST_LEN is a convenient example optimization for explication
+%% purposes, because it is simple and easy to understand but it’s reasonably
+%% representative of the general approach.
 
@@
   A few tests required hand-pruning to remove features that the model
-  does not handle (\totalnum{52}).}
+  does not handle (\totalnum{52}).
+  \SP{} has \numSPdiff{} other tests (\numSPtotal{} total) that we did not use
+  because they fall outside the scope of the model.
+  \Cref{s:impl} summarizes the out-of-scope features;
+  \cref{a:banned-test} gives a detailed categorization.
+  }
 which suggests that the model conforms to actual \SP{}.
@@
 
+\section{Skipped \SP{} Regression Tests}
+\label{a:banned-test}
+
+As of January 9 (commit \href{https://github.com/facebookincubator/cinder/commit/6d61575e28b2d1f31a197f883486c5f012de98c1}{6d61575}),
+the \SP{} test suite contains \numSPtotal{} tests.
+We use \numSPtest{} of these tests for the model.
+The remaining \numSPdiff{} tests are skipped for the following reasons:
+
+\begin{itemize}
+  \item TBD need categories
+\end{itemize}
+
 \section{Fine-Grained Benchmark Data}
+\label{a:microbench-detail}
 
@@
 
-\smallskip
-{\centering\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}
+\begin{center}
+\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}
+\end{center}
 

### 2022-01-21 pj-7.1/paper.tex
pj: touch up redex numbers

@@
 
+%% TODO confirm with C&D 
+%%  KC points out that variables and types are not too different so issue 50 may apply https://github.com/facebookincubator/cinder/issues/50
+%%  perhaps every type should have an import; that's not a burden for Pyre users
 In addition to serving as a default, the dynamic type also replaces any undefined types.
@@
 we used Redex's random testing tools~\cite{kf-sfp-2009} to check
-type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
-%% FILL how many programs exactly?
-%% FILL what bugs did testing catch?
+type soundness on thousands of examples (1,600 expressions and 11,000
+programs).
+% \footnote{Random testing helped us find and fix several issues in the
+% model.  None of issues had implications for \SP{}'s soundness.}
 Finally, we employed the well-established method of testing end-to-end
-conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
-with a test suite.
+conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
 In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
@@
 which suggests that the model conforms to actual \SP{}.
+
+%% 2022-01-20 redex-check details
+%% 1 no counterexamples in 10000 attempts
+%%   found 1636 well-typed expressions.
+%%   1278 of them reduce to a value.
+%%   358 of them don't reduce to a value within the step limit.
+%%
+%% 2 no counterexamples in 100000 attempts
+%%   found 11963 well-typed programs.
+%%   11924 of them terminate.
+%%   39 of them don't terminate within the step limit.
 
@@
 \begin{itemize}
-  \item TBD need categories
+  \item TBD need categories, expect about 40 of these.
 \end{itemize}

### 2022-01-21 pj-7.1/paper.tex
pj: kc edits
Big ones are:
- Replace 'compatible subtyping' with direct phrase about type precision
- Replace model's claim about module imports with assertions, because imports
  are limited.
- Use String instead of Float b/c Float is not in the model
- Remind in 4.4 that generics get a tag check too
- Mention choice of type precision relation near SGG / DGG

Big, undone ones are:
- What's the plan for Unions? "Well-supported" or less?
- Is 'Flout' an OK example, or should we stick to defined types?
- In 6.2 what dynamic features are actually restricted? (Can't be all those
  from sec 5)
- Sh

@@
 \newcommand{\spbool}{\codefont{bool}}
+\newcommand{\spstr}{\typefont{str}}
 \newcommand{\spfloat}{\codefont{float}}
@@
     at least in the Instagram web server.
-    Precise types for higher-order functions and first-class classes
-    are thus a low priority; instead, the dynamic type is available as a coarse
-    approximation.
+    Precise types for higher-order functions, for first-class classes, and even for
+    locally-defined functions and classes are thus a low priority; instead, the
+    dynamic type is available as a coarse approximation.
   \item
@@
   \end{subfigure}
-  \caption{\SP{} ensures neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
+  \caption{\SP{} provides neither the static (SGG) nor the dynamic (DGG) gradual guarantees (assuming a standard type-precision relation)}
   \label{fig:gg-failure}
@@
     Non-final methods and fields may be overridden in arbitrary ways by untyped code.
-    In typed code, however, overrides must be compatible subtypes.
+    In typed code, however, overrides cannot use less-precise types.
     For example, a method that returns a number cannot be overridden by a method
@@
 The model covers a substantial part of the Python language, including
-module imports, loops, exception handlers, and delete statements.
+assertions, loops, exception handlers, and delete statements.
 It follows \SP{}'s approach to typing these features.
@@
 
-The rest of this section illustrates core aspects of our Redex model by way of
+The rest of this section illustrates our Redex model using
 a small formalization.
@@
     \spexpr & \defeq &
-      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
+      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid \spobject \langmid
   \\ & &
@@
       \sptbool \langmid
-      \sptfloat \langmid
+      \sptstr \langmid
       \sptobject \langmid
@@
 The basic values consist of the following: the none value, integers,
-booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
+booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
@@
 user-defined class, and union types.
-We assume that all unions are written in a tidy normal form, e.g., that
-$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
+We assume that all unions are written in a flat and simplified form, e.g., that
+$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.
 
@@
     \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
-    \sptfloat & \mbox{Accepts floats}
+    \sptstr & \mbox{Accepts strings}
   \end{array}\)
@@
 type and that $\spx$ is a variable with the dynamic type.
-An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptfloat$ using two materializations:
+An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptstr$ using two materializations:
 
@@
     \quad
-    \sptdyn \spmatr \sptfloat
+    \sptdyn \spmatr \sptstr
   }{
-    \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptfloat
+    \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptstr
   }
@@
 cast~(\cref{t:cast}).
-For basic types such as floats, a tag check clearly suffices.
+For basic types such as strings, a tag check clearly suffices.
 For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
-the exact-match semantics also provides a decidable test.
+the exact-match semantics is also decidable (and also implemented with
+tag checks).
 Second, observe that boundaries can only relate the dynamic type to an
@@
 \label{s:behavior-changes}
-
-    %% X bg: lookup the dino email about mocks
-    %% And we do go through great lengths to make monkey patching work.  Because
-    %% we need to enforce return types it ends up being not quite as flexible as
-    %% normal Python.  For example one common pattern is to use Python’s mock
-    %% library and we currently end up throwing lots of TypeError’s when a
-    %% function which returns None is mocked and starts returning a MagicMock
-    %% object instead.  But it’s an easy fix in that you can specify the return
-    %% value to be compatible.  There may be more complex cases where you want to
-    %% return a mock for an actual class.  For example in our code base there’s
-    %% frequent mocking of the HttpRequest object and we have a MockReqeust object
-    %% which is a subclass of HttpRequest so that these can still be mocked
-    %% (luckily this actually pre-dated our static conversion work).
 
@@
     These types have been the top priority for \SP{}.
-    Concrete types support full soundness, but impose nonlocal maintenace costs.
+    Concrete types support full soundness, but impose nonlocal maintenance costs.
     \SP{} supports few concrete types at the moment and plans to add more.
@@
     These types are intentionally incompatible with the dynamic type to help programmers
-    avoid casts at run-time.
+    avoid casts.
 \end{itemize}

### 2022-01-24 pj-7.1/paper.tex

@@
 %%   trying to avoid nonstatic? staticifying?
-%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
 %% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
+%% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
+%% -
 
@@
 \newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
-\newcommand{\sptrawpydict}{\typefont{PyDict}}
-\newcommand{\sptrawchkdict}{\typefont{ChkDict}} %% not a real SP type, but useful for TeX
+\newcommand{\sptrawpydict}{\typefont{Dict}}
+\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
 \newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
-\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
+\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
 \newcommand{\sptenv}{\typefont{Env}}
@@
   \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
-  the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
+  the \emph{concrete\/} and \emph{transient\/} approaches from the literature.
   Concrete types provide full soundness and low performance overhead, but impose
@@
   % capabilities are enabled by the work?
-  Our main finding is that the \emph{gradual soundness}\/ that
+  Our main finding is that the \emph{gradual soundness\/} that
   arises from a mix of concrete and transient types is an effective
@@
 but these efforts have not yet decisively closed the performance gap.
-The most promising attempt is the \emph{concrete} semantics
+The most promising attempt is the \emph{concrete\/} semantics
 for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
@@
   \item
-    \SP{} contains a full language of \emph{shallow types}\/ that describe idiomatic
+    A full language of \emph{shallow types\/} that describe idiomatic
     Python code at a coarse granularity.
@@
   \item
-    A small set of \emph{concrete types}\/ describe generic data structures
+    A second level of \emph{concrete types\/} describes generic data structures
     that guarantee the types of their elements.
     If programmers modify their code to build concrete structures instead of Python
-    ones (a potentially nonlocal change), then \SP{} can speed up reads from these
-    structures.
-  \item
-    Additional types and flags to enable further optimizations.
-    These include \emph{primitive types}\/ that enable unboxed arithmetic
-    and \emph{strict module semantics}\/ that makes modules immutable.
+    ones (a potentially nonlocal change), then \SP{} can perform deeper optimizations.
+  \item
+    Additional progressive types that enable further optimizations.
+    In particular, \emph{primitive types\/} enable unboxed arithmetic.
 \end{enumerate}
@@
   \end{lstlisting}
-\end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.32\columnwidth}\noindent\begin{lstlisting}
-from __static__ import PyDict
-def f(x: PyDict[Str, Int]):
+\end{minipage}\begin{minipage}{0.07\columnwidth}\(~~\rightarrow\!\!\!\!\)\end{minipage}\begin{minipage}{0.32\columnwidth}\noindent\begin{lstlisting}
+from typing import Dict
+def f(x: Dict[str, int]):
   return x["A"]
@@
   \end{lstlisting}
-  \end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.34\columnwidth}\noindent\begin{lstlisting}
-from __static__ import ChkDict
-def f(x: ChkDict[Str, Int]):
+  \end{minipage}\begin{minipage}{0.07\columnwidth}\(~~\rightarrow\!\!\!\!\)\end{minipage}\begin{minipage}{0.38\columnwidth}\noindent\begin{lstlisting}
+from __static__ import CheckedDict
+def f(x: CheckedDict[str, int]):
   return x["A"]
 
-f(chkdict[Str, Int]({"A": 1}))
+f(CheckedDict[str, int]({"A": 1}))
   \end{lstlisting}
@@
   \item
-    The $\sptrawchkdict$ type describes a \emph{concrete} dictionary data structure provided by \SP{}.
+    The $\sptrawchkdict$ type
+    describes a \emph{concrete} dictionary data structure provided by \SP{}.
     Unlike Python dictionaries, these checked dictionaries are guaranteed to
@@
 
-%% TODO confirm with C&D 
-%%  KC points out that variables and types are not too different so issue 50 may apply https://github.com/facebookincubator/cinder/issues/50
-%%  perhaps every type should have an import; that's not a burden for Pyre users
+%% TODO avoid typo ... shorten?
 In addition to serving as a default, the dynamic type also replaces any undefined types.
@@
 class A:
-  def m(self) -> Int:
+  def m(self) -> int:
     return 0
@@
   def m(self):
-    # Error: Dyn cannot override Int
+    # Error: dynamic cannot override int
     return 0
@@
   \begin{lstlisting}
-from __static__ import ChkDict
-
-def f(x: ChkDict[Str, Dyn]):
+from __static__ import CheckedDict
+
+def f(x: CheckedDict[str, dynamic]):
   return x["A"]
 
-d = ChkDict[Str, Int]({"A": 1})
+d = CheckedDict[str, int]({"A": 1})
 f(d)
-# Error: f expected ChkDict[Str, Dyn]
+# Error: f expected CheckedDict[str, dynamic]
   \end{lstlisting}
@@
 For example, suppose that a typed function expects inputs that match
-the type \code{Dict[Int,Int]}; that is, dictionaries with integer keys and
+the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
 integer values.
@@
 both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
-and concrete versions at present. In particular, concrete user-defined classes have yet to gain concrete support.}
+and concrete versions at present.
+In particular, user-defined classes do not have concrete support.
+The shallow check for a user-defined class, however, guarantees the types of all
+non-generic fields and methods.}
 For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
@@
     for dict values that ignores the elements inside a dict.
-    Consequently, a parameterized surface type such as $\sptpydict{\sptint}{\sptint}$
+    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
     is merely a promise of dict values at run-time; it does not promise anything about
@@
 
-\subsection{Strict Modules}
-\label{s:strict-mod}
-
-%% 2022-01-12 Although the Static Python compiler can technically be
-%% used apart from strict modules, and we do that in tests for faster tests, it’s
-%% not sound and we don’t allow it in prod.
-
-Python modules are mutable collections of names.
-Names can be added, deleted, and updated at runtime.
-\SP{} accommodates these behaviors in the same manner as standard Python, which
-necessarily limits the benefits of module types.
-As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
-Strict modules are immutable, and thus enable type-based compiler optimizations.
-
 \subsection{Progressive Primitive Types}
@@
 % https://github.com/facebookincubator/cinder/issues/52
+
+\subsection{Behavioral Changes to Python}
+\label{s:not-python}
+
+FILL
+
+%% 2022-01-12 Although the Static Python compiler can technically be
+%% used apart from strict modules, and we do that in tests for faster tests, it’s
+%% not sound and we don’t allow it in prod.
+
+%% Python modules are mutable collections of names.
+%% Names can be added, deleted, and updated at runtime.
+%% \SP{} accommodates these behaviors in the same manner as standard Python, which
+%% necessarily limits the benefits of module types.
+%% As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
+%% Strict modules are immutable, and thus enable type-based compiler optimizations.
+
+%% 1 strict modules
+%% 2 .....
 
@@
     For example, type $\sptchkdict{K}{V}$ matches the type
-    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
-    and $V$ is equal to $\sptdyn$.
+    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
+    and $V$ is equal to $V'$.
 \end{itemize}
@@
 In Python, the syntax for expressions overlaps with the syntax of types.
-The Python name \code{None}, for example, is both an expression and a type.
+The Python name \code{None} is both an expression and a type,
+and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
 By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
-Expressions use lowercase names (e.g., \code{none}) and types uses capitalized
-names (\code{None}).
+Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
+names (\code{None}, \code{CheckedDict}).
 
@@
     \\
-      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
+      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}\!\!\!\!$}
     \\
@@
     %% TODO show dict read rules?
-%    \inferrule*[left=PD-Ref]{
+%    \inferrule*[left=D-Ref]{
 %      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
@@
 
-    \inferrule*[lab=PD-Set]{
+    \inferrule*[lab=D-Set]{
       \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
@@
 handle.
-These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
+These include primitive types~\cref{s:c-types}, module-level variables,
 and the boundary to untyped Python~(\cref{s:py-boundary}).
@@
 
-There are two case in which \SP{} types impose constraints on untyped Python code.
+%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
+There are two cases in which \SP{} types impose constraints on untyped Python code.
 First, concrete generic types (e.g., $\sptrawchkdict$)
@@
 may need to change.
-Second, the strict setting for modules prevents updates to module-level variables.
-If existing code mutates a binding in a strict module, a run-time error will
-occur.
+Second, typed modules prevent updates to module-level variables.
 
@@
 type flows into a static type.
-The materialize rule in the model illustrates this policy.
+The materialize rule in the model illustrates this policy, which ensures that
+well-typed writes to classes and to concrete dictionaries have no cost.
 Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
@@
     Name               & Typed / Python & Refined / Python \\\midrule
-    \bmname{deltablue} &          TBD &            TBD \\
-    \bmname{nbody}     &          TBD &            TBD \\
+    \bmname{deltablue} &         FILL &           FILL \\
+    \bmname{nbody}     &         FILL &           FILL \\
     \bmname{richards}  &         0.53 &           0.22 \\
@@
 \begin{itemize}
-  \item TBD need categories, expect about 40 of these.
+  \item FILL need categories, expect about 40 of these.
 \end{itemize}

### 2022-01-25 pj-7.1/paper.tex
pj: a few more sec2 fixes

@@
 \newcommand{\SP}{Static Python}
+\newcommand{\PEP}{PEP~484}
 \newcommand{\code}[1]{\texttt{#1}}
@@
 \label{s:tour}
-%% purpose = SP looks like Pyre, adds soundness to certain types
-
-% {Language Pipeline}
+
+% {Language Pipeline for IG}
 % \begin{enumerate}
-%   \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
+%   \item Pyre as an optional, but recommended pre-check. (Specific to IG. Not necessary for the model.)
 %   \item SP type checker
@@
 and prior work on types for Python.
-Pyre~\cite{pyre}, an optional type checker developed by Facebook,
-greatly influenced the syntax and types of \SP{}.
-Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, a research language with compelling
-performance results, set the basic strategy for run-time checks.
+Its syntax for types is based on the definitions in \PEP{} and related proposals~\cite{pep484}.
+Its static semantics for types is inspired by the Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright}
+optional type checkers.
+And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
+a research language with compelling performance results.
 \SP{}'s novelty comes from the following engineering constraints:
@@
   \item
-    The codebase in which \SP{} is being applied makes heavy use of Pyre
-    type annotations.
-    \SP{} is therefore compatible with Pyre syntax to reduce the
+    The codebase in which \SP{} is being applied makes heavy use of
+    \PEP{} type annotations and is regularly analyzed by Pyre.
+    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
     adoption burden on its target users.
@@
 In general, the types that \SP{} implements all enable significant optimizations
-and can be validated with constant-time casts.
-Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.
-
-At present, \SP{} has no plans to support all Pyre types.
+and can be validated quickly with casts.
+Types that do not meet these criteria are deferred to the mature optional
+type checkers that already exist for Python.
+In fact, \SP{} has no plans at present to support all \PEP{} types.
 The goal is to implement a small language of types efficiently.
-After all, programmers can always lint their code with Pyre before running
-\SP{}.
 
@@
 recursive types.
-Pyre has some support for the first two, but not for recursive types.
-In any event, the demand for all three types is low in our applications
-of \SP{}.
-%% mypy = https://github.com/python/mypy/issues/731
+Pyre and mypy some support for the first two, but not for recursive types.
+%% mypy rec.types = https://github.com/python/mypy/issues/731
 %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
@@
 
+TODO reword.
 At first glance, this behavior might seem odd: why not reject undefined types with an error?
-The reason is to reduce friction with existing Pyre annotations.
+The reason is to reduce friction with existing \pep{} annotations.
 Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
@@
 
-Every \SP{} type corresponds to a constant-time dynamic check.
+Every \SP{} type corresponds to a constant-time cast.
 When an unknown value reaches a boundary to typed code at run-time,
@@
 generics), constant-time checks are easy to implement.
-Every type has a name and every value can carry a tag that corresponds to the
+Every type has a name and every value has a tag that corresponds to the
 name of its type.
+A cast must examine the parents of a class in the worst case,
+but these traversals are bounded by a (typically small) constant.
 
@@
 
-\SP{} evaluation types can all be enforced fully and in constant-time
+\SP{} evaluation types can all be enforced fully and efficiently
 with casts.
@@
 
-As promised, all casts run in constant time.
 No cast requires traversing a data structure.
@@
 
-\subsubsection{Developer Experience}
-
-For the most part, programmers have been happy adopting \SP{}.
-The transition from Pyre to \SP{} has delivered performance benefits
-that outweigh the cost of code changes.
-Even the use of strict modules has been acceptable; only two modules
-required non-trivial rewrites to avoid mutation.
-
-One significant pushback dealt with multiple inheritance.
-The \SP{} team proposed a rewrite from multiple inheritance to single inheritance,
-but the code maintainers preferred to keep the design as-is instead of enabling
-type-driven method dispatch optimizations.
+% 2022-01-24 bg: hidden from submission to save space
+%\subsubsection{Developer Experience}
+%
+%For the most part, programmers have been happy adopting \SP{}.
+%The transition from Pyre to \SP{} has delivered performance benefits
+%that outweigh the cost of code changes.
+%Even the use of strict modules has been acceptable; only two modules
+%required non-trivial rewrites to avoid mutation.
+%
+%One significant pushback dealt with multiple inheritance.
+%The \SP{} team proposed a rewrite from multiple inheritance to single inheritance,
+%but the code maintainers preferred to keep the design as-is instead of enabling
+%type-driven method dispatch optimizations.
 
@@
 \SP{} is one of many type systems for Python.
-Other optional type checkers include Pyre, mypy, and PyType;
+Other optional type checkers include Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright};
 these systems do not ensure safe interactions with untyped Python.
@@
     The syntax of sound static types is relatively small compared
-    to Pyre~\cite{pyre} and other types-for-Python proposals~\cite{mypy,pytype}.
+    to the \PEP{} definition.
     Higher-order types are absent.
@@
 
-Although the Instagram web server has extensive Pyre annotations,
+Although the Instagram web server has extensive type annotations that are
+checked by Pyre,
 these annotations are not always enough to satisfy \SP{}.

### 2022-01-25 pj-7.1/paper.tex
pj: categorize left out tests

@@
 
+%% TODO 2022-01-25 recount tests
 \newcommand{\numbenchmark}{2}
-\newcommand{\numSPtest}{214}
-\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
-\newcommand{\numSPdiff}{588} %% (\numSPtotal - \numSPtest)
+\newcommand{\numSPtest}{265}
+\newcommand{\numSPtotal}{779} %% SP commit 6d61575 from Jan 9
+\newcommand{\numSPdiff}{514} %% (\numSPtotal - \numSPtest)
 \newcommand{\CPUchange}{3.7\%}
+% 588 514
 
@@
 At first glance, this behavior might seem odd: why not reject undefined types with an error?
-The reason is to reduce friction with existing \pep{} annotations.
+The reason is to reduce friction with existing \PEP{} annotations.
 Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
@@
   does not handle (\totalnum{52}).
-  \SP{} has \numSPdiff{} other tests (\numSPtotal{} total) that we did not use
+  \SP{} has \numSPdiff{} other tests (\totalnum{\numSPtotal{}}) that we did not use
   because they fall outside the scope of the model.
@@
 \begin{itemize}
-  \item FILL need categories, expect about 40 of these.
+  \item 269 test primitive types.
+  \item 62 test optional and keyword functions.
+  \item 26 test the classloader.
+  \item 24 test coroutines (\code{async}).
+  \item 23 test list and dictionary comprehensions.
+  \item 15 use format strings.
+  \item 12 test class and object slots.
+  \item The others (\nnum{83}) test a wide range of other topics
+    including byte strings, memory management, specific decorators
+    (\code{@staticmethod}), and optimizations.
+
+%% actual list
+%  \item 269 use C_types
+%  \item 62 use fancy_args
+%  \item 26 use xxclassloader
+%  \item 24 use async_methods
+%  \item 23 use comprehension_syntax
+%  \item 15 use format_strings
+%  \item 12 use __slots__
+%
+%  \item 10 use advanced_occurrance_typing
+%  \item 8 use @staticmethod
+%  \item 7 use prod_assert
+%  \item 7 use python_scope
+%  \item 6 use __call__
+%  \item 6 use reveal_type
+%  \item 4 use @_donotcompile
+%  \item 4 use memory_management
+%  \item 3 use ...
+%  \item 3 use __setattr__
+%  \item 3 use immutable_variables
+%  \item 3 use shadow_frame
+%  \item 2 use TypeVar
+%  \item 2 use decorator
+%  \item 2 use float
+%  \item 2 use nested_class
+%  \item 2 use nonexpressible_tests
+%  \item 2 use optimize=1
+%  \item 2 use sys.modules
+%  \item 1 use @property
+%  \item 1 use NamedTuple
+%  \item 1 use Protocol
+%  \item 1 use bad_tests
+%  \item 1 use byte_strings
+%  \item 1 use code_flag
+%  \item 1 use fancy_function_type
+%  \item 1 use final_classes
+%  \item 1 use slicing_syntax
+%  \item 1 use sorted
+%  \item 1 use try_except_redeclare
+%  \item 1 use with_traceback
 \end{itemize}
+
+%%\begin{table}
+%%  %% 2022-01-25: seems to raise more questions than it answers ... next question is what each filter really means.
+%%  \caption{List of filter names and a count of skipped test that failed each filter}
+%%  \label{t:left-out}
+%%
+%%  \begin{tabular}{lr}
+%%    Filter & Tests Caught \\\hline
+%%    C\_types & 269 \\
+%%    fancy\_args & 77 \\
+%%    format\_strings & 60 \\
+%%    xxclassloader & 37 \\
+%%    async\_methods & 25 \\
+%%    comprehension\_syntax & 23 \\
+%%    \_\_slots\_\_ & 15 \\
+%%    immutable\_variables & 14 \\
+%%    python\_scope & 14 \\
+%%    float & 11 \\
+%%    advanced\_occurrance\_typing & 10 \\
+%%    @staticmethod & 8 \\
+%%    reveal\_type & 8 \\
+%%    \_\_call\_\_ & 7 \\
+%%    prod\_assert & 7 \\
+%%    ... & 5 \\
+%%    nested\_class & 5 \\
+%%    @\_donotcompile & 4 \\
+%%    decorator & 4 \\
+%%    memory\_management & 4 \\
+%%    sys.modules & 4 \\
+%%    TypeVar & 3 \\
+%%    \_\_setattr\_\_ & 3 \\
+%%    byte\_strings & 3 \\
+%%    shadow\_frame & 3 \\
+%%    sorted & 3 \\
+%%    nonexpressible\_tests & 2 \\
+%%    optimize=1 & 2 \\
+%%    weakref & 2 \\
+%%    @property & 1 \\
+%%    NamedTuple & 1 \\
+%%    Protocol & 1 \\
+%%    bad\_tests & 1 \\
+%%    code\_flag & 1 \\
+%%    fancy\_function\_type & 1 \\
+%%    final\_classes & 1 \\
+%%    slicing\_syntax & 1 \\
+%%    try\_except\_redeclare & 1 \\
+%%    with\_traceback & 1
+%%  \end{tabular}
+%%\end{table}
 

### 2022-01-26 pj-7.1/paper.tex
pj: many sp edits sec 2 3 5

@@
 Its syntax for types is based on the definitions in \PEP{} and related proposals~\cite{pep484}.
-Its static semantics for types is inspired by the Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright}
+Its static semantics for types is inspired by the Pyre~\cite{pyre} and mypy~\cite{mypy}
 optional type checkers.
@@
 and nominal classes.
-Union types are well-supported statically and narrow down in accordance
-with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
-but are, for the most part, not supported at run-time~(\cref{s:eval-types}).
+Union types are tracked statically and narrow down in accordance
+with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),\footnote{
+  For example, if \code{x} is an integer and \code{y} is a string, then the
+  expression \code{x or y} has a union type.}
+but are not generally supported at run-time~(\cref{s:eval-types}).
 The only exception is that binary unions of a type and the Python \code{None}
@@
 
-%% 2021-12-14 TODO:
-%% - anything to say about Any being a "meta" type in Python typing module?
-%%   2022-01-10 bg: not sure what the consequences of "meta type" are
-
-Like most gradual languages, \SP{} includes a dynamic type
-that allows untyped expressions within a statically-typed context.
+\SP{} includes a dynamic type that allows untyped expressions within a
+statically-typed context.
 Whenever an expression or variable lacks a type annotation, \SP{} uses the
-dynamic type as a default and skips most \emph{but not all\/} static checks.
-Programmers can thus add annotations to part of a module
-while leaving the rest unannotated.
-
-%% TODO avoid typo ... shorten?
-In addition to serving as a default, the dynamic type also replaces any undefined types.
-For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
-(i.e., it is a misspelling of $\sptfloat$) and that a
-program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
-dictionary with integer keys and $\sptflout$-typed values.
-\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
-of values inside the dictionary at runtime.
-
-TODO reword.
-At first glance, this behavior might seem odd: why not reject undefined types with an error?
-The reason is to reduce friction with existing \PEP{} annotations.
-Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
-efficiently, e.g., the type $\sptset{\sptint}$ for sets of integers.
-By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.
-
-The \SP{} dynamic type is also different than the flexible dynamic type provided
-by gradual languages that satisfy the gradual
-guarantees~\cite{svcb-snapl-2015}.
-In \SP{}, replacing part of a type with dynamic can lead to a static error when
-the compiler analyzes that part of the module.
-Similarly, the addition of a dynamic type can lead to a run-time error because of the
-strict checks that \SP{} applies to generic types.
+dynamic type as a default and skips most static checks.
+The overall goal of this design is to let programmers add simple annotations to
+part of a Python module while leaving the rest untyped.
+
+In addition to unannotated positions, the dynamic type is also the default for
+\PEP{} types that \SP{} does not yet understand.
+If existing code uses the type $\sptset{\sptint}$ for sets of integers,
+then \SP{} replaces it with dynamic.
+Thus, \SP{} focuses on types that it can soundly optimize without forcing
+programmers to remove types that other systems (e.g., mypy) can check.
+
+The behavior of the \SP{} dynamic type is subtle, however, because dynamic code
+faces more restrictions than untyped Python code.
+Replacing part of a type with dynamic can lead to either a static error or
+a run-time error.
 \Cref{fig:gg-failure} presents two examples, one for each kind of failure:
@@
 \end{itemize}
-These behaviors enable efficient run-time checks and additional type-directed
-optimizations.
-Part of the reason \SP{} weighs these benefits more heavily than the costs
+These extra restrictions make the \SP{} dynamic type different than the flexible
+dynamic type provided by languages that satisfy the gradual guarantees~\cite{svcb-snapl-2015}.
+But they also enable efficient run-time checks for generics and type-directed optimizations
+for method calls.
+Part of the reason \SP{} weighs these benefits more heavily than the gradual guarantees
 is that programmers have to opt-in to a feature in order to risk the errors.
-The static error above comes only after enabling the \SP{} compiler on a module.
-The dynamic error comes only after adopting a concrete type (for checked dictionaries).
-
-More to the point, \SP{} has little interest in guarantees about
-\emph{removing an arbitrary annotation}.
-It is, however, very interested in making sure that \emph{an untyped
-module compiles with minimal code changes}.
+The static error above comes only after enabling the \SP{} compiler on both a parent
+class and its subclass.
+The dynamic error comes only after adopting a concrete type for checked dictionaries.
+More to the point, \SP{} is less interested in guarantees about \emph{removing
+an arbitrary annotation} and more interested in making sure that \emph{an
+untyped module compiles with minimal code changes}.
 
@@
     is merely a promise of dict values at run-time; it does not promise anything about
-    dict keys and values.
+    keys and values.
   \item
@@
 The upside of concrete types is that they enforce stronger type constraints.
-These constraints can catch bugs and usually lead to more-efficient code.
-Faster performance is not guaranteed, though, because it depends on how the dict
-gets used.
-\Cref{t:shallow-vs-concrete} presents a detailed comparison.
+These constraints can catch bugs and always lead to more-efficient typed code.
+Faster performance is not guaranteed, though, because concrete types must
+check all writes from untyped contexts.
+\Cref{t:shallow-vs-concrete} presents in more detail the worst-case costs of
+shallow and concrete dict types.
 Shallow types pay a constant-time cost whenever an untyped value enters typed code
 and whenever a typed context reads from a shallow value.
-The net cost of reads may be high.
 Concrete types pay a potentially-high cost for constructors,
-a constant-time cost for casts and for writes,
+a constant-time cost for casts and for writes in untyped code,
 and zero cost for dereferences.
-If a value is frequently written to and rarely read from, shallow types may actually
-be faster.
+If a value travels to untyped code, gets written to many times, and gets read few times,
+then the shallow types may run faster overall.
 
 \begin{table}[t]
-  \caption{Worst-case costs for shallow and concrete types}
+  \caption{Worst-case costs for shallow and concrete types (T = typed, U = untyped)}
   \label{t:shallow-vs-concrete}
@@
   \centering
-  \(\begin{array}{lrrrr}
-     & \mbox{Constructor} & \mbox{Cast} & \mbox{Read} & \mbox{Write} \\\hline
-    \sptrawpydict                &           0 & O(1) & O(1) &     0 \\
-    \sptchkdict{K}{V} &        O(N) & O(1) &    0 &  O(1)
+  \(\begin{array}{l@{}rr@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
+                   & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
+    \sptrawpydict         &           0 &        O(1) &          O(1) &              0 &             0 &              0 \\
+    \sptchkdict{K}{V}     &        O(N) &        O(1) &             0 &              0 &             0 &           O(1)
   \end{array}\)
@@
 and use checked dispatch for other combinations.
-One extra step arises when an untyped class overrides a typed method.
+One extra step arises when an untyped method overrides a typed method.
 In this case, \SP{} creates a wrapper around the overriding method to check
@@
 concrete datatypes, a registry of typed modules, and a JIT compiler.
-
 These main ingredients of Cinder may be of interest to other teams seeking to
 add sound gradual typing to an existing language.
-For example, Node developers may wish to fork that runtime and experiment with
+For example, Node developers may wish to fork V8 and experiment with
 bytecode instructions that leverage sound static types.
 
-\subsection{Bytecode Instructions}
+\paragraph{Bytecode Instructions}
 
@@
 
-\subsection{Virtual Method Tables}
+\paragraph{Virtual Method Tables}
 \label{s:vtable}
@@
 protocol; it is not a from-scratch development.
-Additionally, however, Cinder vtables are specialized to check untyped overrides
-of typed methods against their expected return type (instead of using a layer of
-indirection for the check, as found languages that enforce types with higher-order
-wrappers~\cite{clzv-ecoop-2018,gf-icfp-2018}).
+Cinder vtables are specialized to check untyped overrides of typed methods
+using a wrapper implemented in C rather than in Python, to reduce performance costs.
 
@@
 %%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
-
+%%
 %% > This difference in behavior isn’t so much desired as just a consequence of making
@@
 
-\subsection{Concrete Data Structures}
+\paragraph{Concrete Data Structures}
 \label{s:chkdict-impl}
@@
   \item
-    All write operations check their input.
+    All untyped writes must be protected by casts.
     Every operation that mutates or extends a checked dictionary must
@@
 
-\subsection{Classloader}
+\paragraph{Classloader}
 \label{s:classloader}
@@
 \SP{} keeps track of typed functions and typed classes with a specialized classloader.
-At runtime, the classloader keeps a registry of typed objects and whether or not they
-are mutable.
+At runtime, the classloader keeps a registry of typed objects.
 The registry helps the bytecode reference objects and types via their module names.
-The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
-patches to mutable data.
-
-\subsection{Method-at-a-Time JIT}
+
+\paragraph{Method-at-a-Time JIT}
 \label{s:jit}
@@
 handle.
-These include primitive types~\cref{s:c-types}, module-level variables,
+These include primitive types~(\cref{s:c-types}), module-level variables,
 and the boundary to untyped Python~(\cref{s:py-boundary}).
@@
 First, concrete generic types (e.g., $\sptrawchkdict$)
-reject inputs that were not created via a checked contructor.
+reject inputs that were not created via a checked constructor.
 If a client of a Python module decides to impose concrete types, the Python module
@@
 via materialization casts, but arguments from untyped contexts are unchecked.
-For this reason, \SP{} compiles every typed function to check its inputs.
-Similarly, concrete types such as $\sptrawchkdict$ also
-check their writes.
+For this reason, \SP{} compiles every typed function to check its inputs;
+the optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
+Similarly, concrete types such as $\sptrawchkdict$ must check writes that
+come from untyped contexts.
 %% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
@@
 \SP{} does not ascribe types to the following Python features.
-These are not covered in our model because the implementation simply
+These are not covered in the model because the implementation simply
 assigns the dynamic type and lets the runtime treat them as untyped
 Python code.
-For each, the dynamic type is a very reasonable choice;
+For each, the dynamic type is a reasonable choice;
 accurate static types would be burdensome to maintain.
@@
 \SP{} is one of many type systems for Python.
-Other optional type checkers include Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright};
-these systems do not ensure safe interactions with untyped Python.
+Other \emph{optional\/} type checkers include Pyre~\cite{pyre},
+mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
+systems do not ensure safe interactions with untyped Python.
 Another sound type system is Reticulated,

### 2022-01-26 pj-7.1/paper.tex
pj: finish sp team comments

@@
 \newcommand{\numSPtest}{265}
-\newcommand{\numSPtotal}{779} %% SP commit 6d61575 from Jan 9
-\newcommand{\numSPdiff}{514} %% (\numSPtotal - \numSPtest)
+\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
+\newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
 \newcommand{\CPUchange}{3.7\%}
-% 588 514
 
@@
   Migrations to \SP{} in the server have caused a \CPUchange{} increase in
-  requests handled per second at max CPU utilization.
+  requests handled per second at maximum CPU load.
 
@@
 these migrations have resulted in a \CPUchange{} increase in production requests handled
-per second for servers at max CPU utilization.
+per second for servers running at maximum CPU load.
 Type-directed optimizations outweigh the cost of enforcing soundness.
@@
 \SP{} runs untyped Python 3.8 programs with minimal changes to their
-behavior,\footnote{\Cref{s:behavior-changes} lists differences with Python 3.8 that
-have caused friction in practice.}
+behavior~(\cref{s:not-python}),
 and it compiles code on-the-fly to give the same workflow and enable the same
@@
     \SP{} is therefore compatible with the \PEP{} syntax to reduce the
-    adoption burden on its target users.
+    adoption burden.
     %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
@@
 type checkers that already exist for Python.
-In fact, \SP{} has no plans at present to support all \PEP{} types.
+In fact, \SP{} has no plans to support all \PEP{} types.
 The goal is to implement a small language of types efficiently.
@@
 and nominal classes.
-Union types are tracked statically and narrow down in accordance
-with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),\footnote{
-  For example, if \code{x} is an integer and \code{y} is a string, then the
-  expression \code{x or y} has a union type.}
-but are not generally supported at run-time~(\cref{s:eval-types}).
+Union types are tracked statically; for example, if \code{x} is an integer and
+\code{y} is a string, then the expression \code{x or y} has a union type.
+They also narrow down via type tests as a kind of occurrence
+typing~\cite{tf-icfp-2010,gsk-esop-2011}.
+Union types are not, however, generally supported at run-time~(\cref{s:eval-types}).
 The only exception is that binary unions of a type and the Python \code{None}
@@
 A cast must examine the parents of a class in the worst case,
-but these traversals are bounded by a (typically small) constant.
+but these traversals are bounded and typically short.
 
@@
 
-FILL
-
-%% 2022-01-12 Although the Static Python compiler can technically be
-%% used apart from strict modules, and we do that in tests for faster tests, it’s
-%% not sound and we don’t allow it in prod.
-
-%% Python modules are mutable collections of names.
-%% Names can be added, deleted, and updated at runtime.
-%% \SP{} accommodates these behaviors in the same manner as standard Python, which
-%% necessarily limits the benefits of module types.
-%% As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
-%% Strict modules are immutable, and thus enable type-based compiler optimizations.
-
-%% 1 strict modules
-%% 2 .....
+Relative to Python, \SP{} makes a few behavioral changes to enable
+sound types and strong optimizations:
+class attributes must be set in an \code{\_\_init\_\_} method to
+ensure predictable layouts (all \PEP{} checkers agree on this);
+module-level variables cannot be mutated; and
+typed classes can inherit from at most one other typed class.
+
+Several other Python features are unsupported at present:
+enums; the \code{@dataclass} decorator;
+overrides of the \code{\_\_setattr\_\_} and \code{\_\_getattr\_\_} methods;
+keyword arguments with default expressions (rather than values); and
+calls to keyword functions that use \code{**kwargs} with either
+a custom object or a dict with extra keys.
+The \SP{} team plans to lift these restrictions in the future.
 
@@
 syntax of the model and confirmed that the results
-do match,\footnote{For most of the \numSPtest{} tests, the translation is automatic.
-  A few tests required hand-pruning to remove features that the model
-  does not handle (\totalnum{52}).
-  \SP{} has \numSPdiff{} other tests (\totalnum{\numSPtotal{}}) that we did not use
-  because they fall outside the scope of the model.
-  \Cref{s:impl} summarizes the out-of-scope features;
-  \cref{a:banned-test} gives a detailed categorization.
-  }
+do match,
 which suggests that the model conforms to actual \SP{}.
+
+For most of the \numSPtest{} tests, the translation is automatic.
+A few tests required hand-pruning to remove features that the model does not
+handle (\totalnum{52}).
+\SP{} has \numSPdiff{} other tests (\totalnum{\numSPtotal{}}) that we did not
+use because they fall outside the scope of the model.
+\Cref{s:impl} summarizes the out-of-scope features; \cref{a:banned-test} gives
+a detailed categorization.
 
@@
 \label{s:surface}
+
+\Cref{f:surface-types} presents a core surface syntax for model programs.
+A program is a sequence of statements;
+a statement defines a variable, function, or class.
+These definitions may only appear on the top level
+and they all require full type annotations.
+Functions must have one positional argument.
+Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
+Expressions describe values and simple computations.
+The basic values consist of the following: the none value, integers,
+booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
+There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
+The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
+and method calls.
+
+In Python, the syntax for expressions overlaps with the syntax of types.
+The Python name \code{None} is both an expression and a type,
+and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
+By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
+Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
+names (\code{None}, \code{CheckedDict}).
+
+Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
+user-defined class, and union types.
+We assume that all unions are written in a flat and simplified form, e.g., that
+$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.
+
+\SP{} does not enforce all surface types against untyped code, only the evaluation
+types presented in the next subsection~(\cref{s:eval-types}).
+For this reason, we omit the surface typing judgment, which is a kind of linter
+that merely scans typed code for logical errors.
+
+Relative to \SP{} and our Redex mechanization, the formalization in \cref{f:surface-types} omits
+many details of Python including class variables, imports, conditionals, and exception handlers.
+These details are crucial in the mechanization, which tests whether \SP{}
+soundly approximates Python.
+They are less important here, where our focus is on type boundaries.
+None of the omitted features give substantially new ways for typed code to interact with dynamic code.
 
@@
 
-\Cref{f:surface-types} presents a core surface syntax for model programs.
-A program is a sequence of statements;
-a statement defines a variable, function, or class.
-These definitions may only appear on the top level
-and they all require full type annotations.
-Functions must have one positional argument.
-Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
-Expressions describe values and simple computations.
-The basic values consist of the following: the none value, integers,
-booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
-There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
-The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
-and method calls.
-
-In Python, the syntax for expressions overlaps with the syntax of types.
-The Python name \code{None} is both an expression and a type,
-and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
-By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
-Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
-names (\code{None}, \code{CheckedDict}).
-
-Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
-user-defined class, and union types.
-We assume that all unions are written in a flat and simplified form, e.g., that
-$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.
-
-\SP{} does not enforce all surface types against untyped code, only the evaluation
-types presented in the next subsection~(\cref{s:eval-types}).
-For this reason, we omit the surface typing judgment, which is a kind of linter
-that merely scans typed code for logical errors.
-
-Relative to \SP{} and our Redex mechanization, the formalization in \cref{f:surface-types} omits
-many details of Python including class variables, imports, conditionals, and exception handlers.
-These details are crucial in the mechanization, which tests whether \SP{}
-soundly approximates Python.
-They are less important here, where our focus is on type boundaries.
-None of the omitted features give substantially new ways for typed code to interact with dynamic code.
-
 \subsection{Evaluation Types, Casts, and Typing Judgment}
@@
 \subsubsection{Expression Typing, Cast Insertion}
+
+\begin{figure}[p]
+  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
+  \begin{mathpar}
+    \inferrule*[left=S-App]{
+      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
+    }
+
+    \inferrule*[left=D-App]{
+      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \sptobject
+    }{
+      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
+    }
+
+    %% TODO show dict read rules?
+%    \inferrule*[left=D-Ref]{
+%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+%      \\\\
+%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+%    }{
+%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
+%    }
+%
+%    \inferrule*[left=CD-Ref]{
+%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+%      \\\\
+%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+%    }{
+%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
+%    }
+
+    \inferrule*[lab=D-Set]{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
+    }{
+      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
+    }
+
+    \inferrule*[lab=CD-Set]{
+      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
+    }{
+      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
+    }
+
+    %\inferrule*[left=F-Ref]{
+    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
+    %}{
+    %  \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
+    %}
+
+    \inferrule*[left=F-Set]{
+      \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
+    }
+
+    \inferrule*[left=M-App]{
+      \sptfundef{\spf_0}{\spc_0, \spteval_0}{\spteval_1} \in \spenvapp{\sptenv_0}{\spc_0}
+      \\\\
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \spteval_1
+    }
+
+    \inferrule*[left=C-Sub]{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \\\\
+      \sptenv_0 \wtsub \spteval_1 \spcompat \spteval_0
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }
+
+    \inferrule*[left=Matr]{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \\\\
+      \spteval_1 \spmatr \spteval_0
+      %% side condition: result type is not a primitive
+    }{
+      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+    }
+
+  \end{mathpar}
+
+  \begin{minipage}[t]{0.6\columnwidth}
+    \fbox{$\sptenv \wtsub \spteval \spcompat \spteval$}
+    \begin{mathpar}
+      \inferrule*{
+        \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
+      }{
+        \sptenv_0 \wtsub \spteval_0 \spcompat \spteval_1
+      }
+
+      \inferrule*{
+      }{
+        \sptenv_0 \wtsub \spteval_0 \spcompat \sptdyn
+      }
+  \end{mathpar}
+
+  \end{minipage}\begin{minipage}[t]{0.4\columnwidth}
+    \fbox{$\spteval \spmatr \spteval$}
+    \begin{mathpar}
+      \inferrule*[right={$\spteval_0 \neq \sptdyn$}]{
+      }{
+        \sptdyn \spmatr \spteval_0
+      }
+    \end{mathpar}
+  \end{minipage}
+
+  \medskip
+  \fbox{$\sptenv \wtsub \spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
+  \begin{mathpar}
+    \inferrule*{
+    }{
+      \sptenv_0 \wtsub \spteval_0 \spsubt \sptobject
+    }
+
+    \inferrule*{
+    }{
+      \sptenv_0 \wtsub \sptbool \spsubt \sptint
+    }
+
+    \inferrule*{
+      \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
+    }{
+      \sptenv_0 \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
+    }
+
+    \inferrule*{
+      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
+      \\
+      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
+    }{
+      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
+    }
+
+%    \inferrule*{
+%    }{
+%      \spteval_0 \spsubt \spteval_0
+%    }
+%
+%    \inferrule*{
+%      \spteval_0 \spsubt \spteval_1
+%      \\
+%      \spteval_1 \spsubt \spteval_2
+%    }{
+%      \spteval_0 \spsubt \spteval_2
+%    }
+  \end{mathpar}
+
+  \caption{Expression Typing, Consistent Subtyping, and Materialization}
+  \label{f:eval-types}
+\end{figure}
 
@@
 \Cref{s:impl} explains the optimizer in more detail.
-
-\begin{figure}[p]
-  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
-  \begin{mathpar}
-    \inferrule*[left=S-App]{
-      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
-    }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
-    }
-
-    \inferrule*[left=D-App]{
-      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \sptobject
-    }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
-    }
-
-    %% TODO show dict read rules?
-%    \inferrule*[left=D-Ref]{
-%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
-%      \\\\
-%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
-%    }{
-%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
-%    }
-%
-%    \inferrule*[left=CD-Ref]{
-%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
-%      \\\\
-%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
-%    }{
-%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
-%    }
-
-    \inferrule*[lab=D-Set]{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
-    }{
-      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
-    }
-
-    \inferrule*[lab=CD-Set]{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
-    }{
-      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
-    }
-
-    %\inferrule*[left=F-Ref]{
-    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
-    %}{
-    %  \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
-    %}
-
-    \inferrule*[left=F-Set]{
-      \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
-    }{
-      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
-    }
-
-    \inferrule*[left=M-App]{
-      \sptfundef{\spf_0}{\spc_0, \spteval_0}{\spteval_1} \in \spenvapp{\sptenv_0}{\spc_0}
-      \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
-    }{
-      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \spteval_1
-    }
-
-    \inferrule*[left=C-Sub]{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
-      \\\\
-      \sptenv_0 \wtsub \spteval_1 \spcompat \spteval_0
-    }{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
-    }
-
-    \inferrule*[left=Matr]{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
-      \\\\
-      \spteval_1 \spmatr \spteval_0
-      %% side condition: result type is not a primitive
-    }{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
-    }
-
-  \end{mathpar}
-
-  \begin{minipage}[t]{0.6\columnwidth}
-    \fbox{$\sptenv \wtsub \spteval \spcompat \spteval$}
-    \begin{mathpar}
-      \inferrule*{
-        \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
-      }{
-        \sptenv_0 \wtsub \spteval_0 \spcompat \spteval_1
-      }
-
-      \inferrule*{
-      }{
-        \sptenv_0 \wtsub \spteval_0 \spcompat \sptdyn
-      }
-  \end{mathpar}
-
-  \end{minipage}\begin{minipage}[t]{0.4\columnwidth}
-    \fbox{$\spteval \spmatr \spteval$}
-    \begin{mathpar}
-      \inferrule*[right={$\spteval_0 \neq \sptdyn$}]{
-      }{
-        \sptdyn \spmatr \spteval_0
-      }
-    \end{mathpar}
-  \end{minipage}
-
-  \medskip
-  \fbox{$\sptenv \wtsub \spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
-  \begin{mathpar}
-    \inferrule*{
-    }{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \sptobject
-    }
-
-    \inferrule*{
-    }{
-      \sptenv_0 \wtsub \sptbool \spsubt \sptint
-    }
-
-    \inferrule*{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
-    }{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
-    }
-
-    \inferrule*{
-      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
-      \\
-      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
-    }{
-      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
-    }
-
-%    \inferrule*{
-%    }{
-%      \spteval_0 \spsubt \spteval_0
-%    }
-%
-%    \inferrule*{
-%      \spteval_0 \spsubt \spteval_1
-%      \\
-%      \spteval_1 \spsubt \spteval_2
-%    }{
-%      \spteval_0 \spsubt \spteval_2
-%    }
-  \end{mathpar}
-
-  \caption{Expression Typing, Consistent Subtyping, and Materialization}
-  \label{f:eval-types}
-\end{figure}
 
@@
 structural objects.
-How best to add an interface layer in \SP{} is an open question.
+
+On a related note, \SP{} has no support for first-class functions or for
+structural types as defined by PEP~544~\cite{pep544}.
+MonNom interfaces may be a promising way to support these types.
 
@@
 Python allows classes to inherit from a list of parents.
-To resolve method calls to such a class, Python
-dynamically traverses the parent list in a fixed order
-seeking a first match.
-
-Due to the dynamic method resolution order (MRO),
-\SP{} does not track types for classes with multiple parents.
-Method calls to these classes execute the same way as
-in Python, with no type-directed optimizations to improve dispatch.
+\SP{} accommodates this behavior provided that one parent is defined in typed
+code and the untyped parents do not specify their members (via
+\code{\_\_slots\_\_}).
+Other uses of multiple inheritance must stay in untyped code.
+
+\SP{} does not leverage types for classes with multiple parents
+because of vtable layout issues.
+Efficient vtable dispatch requires that each method name
+resides in a fixed slot across all subclasses.
+With multiple inheritance, however, parents may disagree about which method
+lives in each slot.
+%% possible improvement: invoke methods on the first parent statically and on
+%% other parents dynamically and forbid those other paarents from having instance
+%% attributes.
 
@@
 
-Results computed by calls to \code{eval} and \code{exec} have the dynamic type.
+Calls to \code{eval} and \code{exec} have the dynamic type.
 The type system makes no finer prediction about what kind of value such calls will
@@
 can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
-Assuming these findings carry over to Python, similar adjustments would
+Assuming these findings carry over to Python, similar adjustments would be
 preferable over attempting to type \code{eval}.
@@
 Two of the listed instructions allocate \SP{}-specific data structures.
-The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
+The first creates a checked dictionary; the second allocates an uninitialized instance
+of an object~(\cref{s:vtable}).
 Cinder comes with a similar instruction to build checked lists.
@@
 to Python's dynamic call dispatch.
-The former jumps to a statically-determined address;
-the latter uses a vtable lookup to quickly find an address at run-time.
+The former uses the classloader to quickly find the address of a function;
+the latter uses a vtable lookup to find a method.
 The load and store instructions for fields improve upon Python's generic attribute lookup.
@@
 Instagram's internal profiling tools attribute a \CPUchange{} improvement
-in CPU utilization~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
+in CPU efficiency~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
 This is a big improvement at Instagram scale.
@@
 
-\subsection{How to Interpret the CPU Utilization Result}
+\subsection{How to Interpret the CPU Efficiency Result}
 \label{s:what-is-cpu}
@@
 Next, the profiling tools slowly increase the number of requests that these
-machines receive until each is running at max CPU utilization.
+machines receive until each is running at maximum load.
 Once the machines are fully allocated, the tools measure how many requests
@@
 capabilities, and assuming that both groups receive a representative sample
-of requests, then the results of an experiment (increase/decrease in CPU utilization)
+of requests, then the results of an experiment (increase/decrease in CPU efficiency)
 should predict actual performance.
 
-The main threat to validity is that this CPU utilization number depends on
-product code and on client behavior.
-Over time, the size of the average request may increase and thereby increase
-the CPU cost of serving one request.
-Nevertheless, CPU utilization is the primary metric that Instagram
-uses for production changes.
+The one caveat with this CPU efficiency measure is that it is pinned
+to a specific point in time.
+Changes in product code and in client behavior can change the size
+of a typical web server request, which changes the number
+of requests that a server can handle per second.
+Taking the sum of several CPU efficiency changes over a long
+timespan is therefore only a rough measure of their net effect.
+%% In particular, although we report a \CPUchange{} net improvement
+%% in web server CPU efficiency due to \SP{}, removing \SP{} and
+%% taking a new measurement may yield a slightly different degredation
+%% than \CPUchange{}.
 
@@
 Furthermore, \SP{} types combined with Cinder resulted in a
-0.7\% reduction in web server CPU usage.
+0.7\% improvement in web server CPU efficiency.
 Primitive types~(\cref{s:c-types}) were essential for matching Cython.
@@
 Later \SP{} migrations have been directed by profiling to find frequently-executed code.
-During the first half of 2021, the \SP{} team identified critical
+During the first half of 2021, the \SP{} team identified hot
 modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
-The accepted proposals resulted in a 1\% improvement in web server CPU usage.
+The accepted proposals resulted in a 1\% improvement in CPU efficiency.
 During the second half of 2021, the \SP{} team applied the same
@@
 this one change brought over four hundred generated modules into the typed fold.
-Overall, the second-half changes added a 2\% web server CPU improvement.
+Overall, the second-half changes added 2\% to the number of requests
+that servers could handle per second at maximum CPU load.
 
@@
     \ipara{Change mocked function to expect positional arguments}
-    \SP{} currently rewrites all function calls to use positional arguments.
+    \SP{} currently rewrites the bytecode for all typed-to-typed function calls to use positional arguments.
     Mock-wrapped functions therefore need to use positional arguments instead of keywords.
@@
     %% arguments when we detect the target function has been patched.
-
-  \item
-    \ipara{Change functions to accept positional arguments}
-    Functions that expect only keyword arguments cannot yet be called from \SP{} contexts.
-    %% Carl: We plan to support this next half.
 
@@
   \item
-    \ipara{Move \code{@dataclass} and \code{Enum} classes}
-    \SP{} does not support classes decorated with \code{@dataclass}
-    and classes that are enums.
-    For now, programmers have to move these classes to an untyped module.
-    %% Carl: We have work in progress on supporting some Enum types, and plan to add a @dataclass intrinsic soon.
-
-  \item
-    \ipara{Remove unsupported Python features}
-    Occurrences of the dynamic Python features listed in \cref{s:dynamic-python} must be rewritten or moved.
-    %% According to the \SP{} team, these changes have not required any major rewrites.
+    \ipara{Move unsupported Python features}
+    Code that uses the unsupported features listed in~\cref{s:not-python}
+    must be changed or moved to an untyped module.
+    So far, these changes have not required major rewrites.
+
+  \item
+    \ipara{Refine some Pyre annotations}
+    Well-typed Pyre code is not always well-typed \SP{} code.
+    \Cref{a:pyre-limits} lists the main friction points.
 
@@
     %% - bg: how occasional? worried about power of the "gradual soundness" claim
+
+  \item
+    \ipara{Change functions to accept positional arguments}
+    Functions calls that use only keyword arguments are not yet optimized by \SP{}.
+    Rewriting these functions and call-sites to use positional arguments improves performance.
+
+    %% Carl: We plan to support this next half.
 
@@
 but they provide an auditable and reproducible way to measure performance.
+
 Each row presents two ratios.
@@
 The refinements are modest, mostly adding primitive types to hot loops;
-they do not require siginificant expertise to reproduce.
+they do not require significant expertise to reproduce.
 
@@
 (commit \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{\code{d0d071a}}).
-Each benchmark version ran 11 times total: once to warm up the JIT and 10 times to
-compute an average runtime (using the \code{time} utility).
-Every benchmark version invoked the JIT on the same functions---even the untyped versions.
+Each benchmark version ran 11 times total: once to warm up the bytecode caches
+(\code{.pyc} files) and 10 times to compute an average runtime (using the
+\code{time} utility).
+Every benchmark version invoked the JIT on the same functions---even the
+untyped versions.
 
@@
     Name               & Typed / Python & Refined / Python \\\midrule
-    \bmname{deltablue} &         FILL &           FILL \\
+    \bmname{deltablue} &         0.59 &           0.30 \\
     \bmname{nbody}     &         FILL &           FILL \\
@@
 slowdown.
-This benchmark is primarily about number-crunching; and indeed, changing a few simple
-types to primitive types without modifying the code leads to a minor speedup (ratio = $0.90$).
+This benchmark is primarily about number-crunching and the basic version uses standard
+numeric types that \SP{} does not optimize.
+Switching to primitive types without otherwise changing the code
+gives an immediate speedup (ratio = $0.90$).
 
@@
 research community, but they have been effective in practice.
-The current \CPUchange{} increase in CPU usage is a huge improvement
+The current \CPUchange{} increase in CPU efficiency is a huge improvement
 at the scale of the Instagram web server.
@@
 \section{On Migrations from Pyre to \SP{}}
+\label{a:pyre-limits}
 
@@
   \item 12 test class and object slots.
-  \item The others (\nnum{83}) test a wide range of other topics
+  \item The others (\nnum{106}) test a long tail of other topics
     including byte strings, memory management, specific decorators
@@
 \begin{table}[tph]
-  \caption{Exact commands to invoke each benchmark}
+  \caption{DeltaBlue microbenchmark data}
+  \label{t:mb:deltablue}
+  \footnotesize\centering
+
+\begin{tabular}{rrrrrr}
+  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.3$ & $0.4$ & $1.32$ & $0.51$ & $0.64$ & $1.02$ \\
+  $0.32$ & $0.39$ & $1.23$ & $0.54$ & $0.64$ & $1$ \\
+  $0.31$ & $0.4$ & $1.24$ & $0.54$ & $0.64$ & $0.99$ \\
+  $0.3$ & $0.42$ & $1.31$ & $0.56$ & $0.65$ & $1.02$ \\
+  $0.28$ & $0.4$ & $1.24$ & $0.52$ & $0.65$ & $0.96$ \\
+  $0.28$ & $0.39$ & $1.24$ & $0.55$ & $0.65$ & $0.98$ \\
+  $0.29$ & $0.4$ & $1.27$ & $0.55$ & $0.68$ & $0.97$ \\
+  $0.3$ & $0.42$ & $1.29$ & $0.53$ & $0.67$ & $1$ \\
+  $0.29$ & $0.39$ & $1.28$ & $0.53$ & $0.69$ & $0.99$ \\
+  $0.28$ & $0.42$ & $1.32$ & $0.51$ & $0.64$ & $0.99$ \\
+\end{tabular}
+
+\begin{tabular}{rrrrrr}
+  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.6$ & $0.74$ & $1.32$ & $0.69$ & $0.82$ & $1.27$ \\
+  $0.59$ & $0.74$ & $1.32$ & $0.67$ & $0.84$ & $1.27$ \\
+  $0.57$ & $0.71$ & $1.31$ & $0.7$ & $0.89$ & $1.23$ \\
+  $0.58$ & $0.74$ & $1.23$ & $0.73$ & $0.82$ & $1.26$ \\
+  $0.6$ & $0.71$ & $1.32$ & $0.71$ & $0.84$ & $1.26$ \\
+  $0.59$ & $0.73$ & $1.4$ & $0.71$ & $0.84$ & $1.23$ \\
+  $0.56$ & $0.7$ & $1.37$ & $0.71$ & $0.85$ & $1.35$ \\
+  $0.58$ & $0.69$ & $1.3$ & $0.71$ & $0.82$ & $1.31$ \\
+  $0.58$ & $0.69$ & $1.27$ & $0.69$ & $0.81$ & $1.34$ \\
+  $0.6$ & $0.73$ & $1.39$ & $0.68$ & $0.8$ & $1.33$ \\
+\end{tabular}
+
+\begin{tabular}{rrrrrr}
+  \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} \\
+  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.54$ & $0.66$ & $1.38$ & $0.61$ & $0.76$ & $1.17$ \\
+  $0.51$ & $0.62$ & $1.44$ & $0.61$ & $0.75$ & $1.25$ \\
+  $0.55$ & $0.69$ & $1.41$ & $0.6$ & $0.82$ & $1.17$ \\
+  $0.54$ & $0.69$ & $1.41$ & $0.63$ & $0.75$ & $1.16$ \\
+  $0.54$ & $0.7$ & $1.41$ & $0.65$ & $0.75$ & $1.17$ \\
+  $0.55$ & $0.67$ & $1.4$ & $0.66$ & $0.75$ & $1.15$ \\
+  $0.52$ & $0.67$ & $1.43$ & $0.61$ & $0.75$ & $1.15$ \\
+  $0.52$ & $0.66$ & $1.4$ & $0.62$ & $0.75$ & $1.18$ \\
+  $0.52$ & $0.64$ & $1.51$ & $0.65$ & $0.76$ & $1.18$ \\
+  $0.5$ & $0.65$ & $1.41$ & $0.62$ & $0.79$ & $1.15$ \\
+\end{tabular}
+
+\begin{tabular}{rrr}
+  \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+  $0.64$ & $0.78$ & $1.22$ \\
+  $0.64$ & $0.81$ & $1.25$ \\
+  $0.63$ & $0.81$ & $1.23$ \\
+  $0.62$ & $0.79$ & $1.2$ \\
+  $0.67$ & $0.81$ & $1.23$ \\
+  $0.7$ & $0.8$ & $1.23$ \\
+  $0.68$ & $0.78$ & $1.34$ \\
+  $0.66$ & $0.79$ & $1.22$ \\
+  $0.69$ & $0.85$ & $1.23$ \\
+  $0.68$ & $0.79$ & $1.24$ \\
+\end{tabular}
+
+\end{table}
+
+\begin{table}[tph]
+  \caption{Exact commands to invoke each benchmark (1/2)}
   \label{t:mb:commands}
@@
     Fannkuch \colname{Orig} & \texttt{time ./python Tools/benchmarks/fannkuch.py 5} \\
+    () &
+  \end{tabular}
+\end{table}
+
+\begin{table}[tph]
+  \caption{Exact commands to invoke each benchmark (2/2)}
+  \label{t:mb:commands}
+  \tiny\centering
+
+  \begin{tabular}{ll}
+    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
+    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
+    (\colname{SP} \colname{JIT}) & \texttt{s -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
+    DeltaBlue \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
+    (\colname{SP}) &  \\
+    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
+    (\colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame Tools/benchmarks/deltablue\_static.py 100} \\
+    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
+    (\colname{JIT}) & \texttt{s Tools/benchmarks/deltablue\_static.py 100} \\
+    DeltaBlue \colname{T-Max} & \texttt{time ./python Tools/benchmarks/deltablue\_static.py 100} \\
+    () &  \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
+    (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    (\colname{SP}) &  \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
+    (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
+    (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic.py 100} \\
+    () &  \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    (\colname{SP} \colname{JIT}) & \texttt{ildcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    (\colname{SP}) &  \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    (\colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    (\colname{JIT}) & \texttt{ildcards Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min-2} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    () &  \\
+    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
+    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/deltablue.py 100} \\
+    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
+    (\colname{JIT}) & \texttt{hmarks/deltablue.py 100} \\
+    DeltaBlue \colname{Orig} & \texttt{time ./python Tools/benchmarks/deltablue.py 100} \\
     () &  \\
   \end{tabular}
-
 \end{table}

### 2022-01-26 pj-7.1/paper.tex
pj: acks, try and fail to align appendix tables

@@
 
-%\acks{
-%  Thanks to
-%  Guido van Rossum for stimulating tweets.
-%  This work was partly supported by the US National Science Foundation.
-%  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
-%  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
-%  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-%}
+\acks{
+  \SP{} was developed by Sinan Cepel, William Meehan, Carl Meyer, Aniket Panse,
+  and Dino Viehland.
+  The Brown authors contributed to the language by modeling it independently; they
+  are not part of the \SP{} team.
+
+  This work was partly supported by the US National Science Foundation.
+  %% Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
+  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
+
+  Special thanks to Guido van Rossum for helping the authors meet via Twitter
+  and for his ongoing contributions to Python.
+}
 
@@
 
-\begin{table}[tph]
+\begin{table}[tp]
   \caption{Richards microbenchmark data}
@@
 
-\begin{table}[tph]
+\begin{table}[tp]
   \caption{Fannkuch microbenchmark data}
@@
 
-\begin{table}[tph]
+\begin{table}[tp]
   \caption{DeltaBlue microbenchmark data}
@@
 \end{tabular}
-
 \end{table}
 
-\begin{table}[tph]
+\begin{table}[tp]
   \caption{Exact commands to invoke each benchmark (1/2)}
@@
 
-\begin{table}[tph]
+\begin{table}[tp]
   \caption{Exact commands to invoke each benchmark (2/2)}

### 2022-01-26 pj-7.1/paper.tex
pj: down to 23 pages

@@
 \newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
-\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
+\newenvironment{langarray}{\(\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
 \newcommand{\ipara}[1]{\emph{#1}\/.}
@@
 codebase and while leaving the rest untyped.
-Run-time casts and checks enforce the critical assumptions that typed code makes about
+Run-time casts and checks enforce the assumptions that typed code makes about
 untyped components, thereby ensuring that the types are sound no matter how
@@
   \item
-    Additional progressive types that enable further optimizations.
-    In particular, \emph{primitive types\/} enable unboxed arithmetic.
+    A third set of progressive types enable further optimizations.
+    In particular, \emph{primitive types\/} allow for unboxed arithmetic,
+    which has been critical in practice.
 \end{enumerate}
@@
 primary web server monolith.
-The results are very positive.
 Although only a handful of modules use \SP{} types (hundreds among thousands),
@@
 per second for servers running at maximum CPU load.
+These results are very positive:
 Type-directed optimizations outweigh the cost of enforcing soundness.
@@
     issues in \SP{}.
-    Four of these were critical soundness bugs.
+    Four of these were soundness bugs.
 
@@
 behavior~(\cref{s:not-python}),
-and it compiles code on-the-fly to give the same workflow and enable the same
+and it compiles code on-the-fly to enable the same
 developer tools as standard Python.
@@
 and prior work on types for Python.
-Its syntax for types is based on the definitions in \PEP{} and related proposals~\cite{pep484}.
-Its static semantics for types is inspired by the Pyre~\cite{pyre} and mypy~\cite{mypy}
-optional type checkers.
+Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
+Its static semantics for types is inspired by optional type checkers; in
+particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
 And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
@@
 recursive types.
-Pyre and mypy some support for the first two, but not for recursive types.
+Pyre and mypy some support for the first two kinds, but not for recursive types.
 %% mypy rec.types = https://github.com/python/mypy/issues/731
@@
 
-The behavior of the \SP{} dynamic type is subtle, however, because dynamic code
+The behavior of the \SP{} dynamic type is subtle because dynamic code
 faces more restrictions than untyped Python code.
@@
 \begin{table}[t]
-  \caption{Worst-case costs for shallow and concrete types (T = typed, U = untyped)}
+  \caption{Worst-case costs for shallow and concrete types (T=typed, U=untyped)}
   \label{t:shallow-vs-concrete}
@@
 a rather flexible point in the design space.
-\SP{} implements a simpler design that places a few restrictions on types and
+\SP{} implements a simpler design that restricts types and
 inheritance:
@@
 The modelling process helped uncover several critical issues with the
-implementation of \SP{}.
-This is a very important payoff given that \SP{} is running in production at
-Instagram.
+implementation of \SP{}---an important payoff given that \SP{} is running in
+production at Instagram.
 Overall, we reported 25 bug reports to the \SP{} team via GitHub~(listed in
@@
       return x
-
-  print(f(None))
-  # None
 \end{lstlisting}
+%
+%  print(f(None))
+%  # None
 \end{minipage}
@@
 When called with \code{None}, however, the function breaks out of the \code{while} loop
-and implicitly returns \code{None}---disobeying the return type.
+and implicitly returns \code{None} contrary to the return type.
 \SP{} had failed to account for the \code{break} and implicit return.
@@
       \spstmt \langmid \spstmt,~\spprog
-  \\
+    \\[1.5ex]
     \spstmt & \defeq &
@@
       \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
-  \\
+    \\[1.5ex]
     \spexpr & \defeq &
@@
       \spobjapp{\spc}{\spf}{\spexpr}
-  \\
+    \\[1.5ex]
     \sptype & \defeq &
@@
       \sptunion{\sptype, \ldots}
-  \\
+    \\[1.5ex]
     \sptenv & \defeq &
@@
       \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
-  \\
+    \\[1.5ex]
     \spx, \spf & \defeq & \mbox{variable names}
@@
 
-  \medskip
+  \bigskip
   \fbox{$\sptenv \wtsub \spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
@@
 Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
-These statements come with types and fill an environment ($\sptenv$).
+These declarations fill a type environment ($\sptenv$).
 Relative to the current environment, the expressions within each statement must
@@
 
-\paragraph{Indirect Optimization}
+\paragraph{Removing Casts}
 
 In addition to upgrading bytecode instructions to optimized ones, \SP{} takes care
-to minimize the type checks that it executes at run-time.
-In other words, the goal is to slow code down as little as possible.
+to minimize the type casts that it executes at run-time.
+In other words, it takes care to slow code down as little as possible.
 
@@
     \SP{} currently rewrites the bytecode for all typed-to-typed function calls to use positional arguments.
-    Mock-wrapped functions therefore need to use positional arguments instead of keywords.
+    Mock-wrapped functions therefore need to use positional arguments.
     %% Carl: we plan to fix this incompatibility by falling back to the original call
@@
 gives an immediate speedup (ratio = $0.90$).
+\Cref{a:microbench-detail} presents further benchmark details.
 
@@
 
-  This work was partly supported by the US National Science Foundation.
-  %% Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
-  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
-
   Special thanks to Guido van Rossum for helping the authors meet via Twitter
   and for his ongoing contributions to Python.
+  This work was partly supported by the US National Science Foundation.
+  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
 }

### 2022-01-26 conformance_suite/edited_test_invoke_strict_module_deep_unjitable_many_args.py
fix a test

@@ -4,7 +4,8 @@
 
 def f0(): return 42
 def f1(a, b, c, d, e, f, g, h):
-    class C: pass
+    # We deleted the next line because nested classes are not supported.
+    # class C: pass
     return f0() - a + b - c + d - e + f - g + h - 4
 def f2(): return f1(1,2,3,4,5,6,7,8)
 def f3(): return f2()

### 2022-01-27 pj-7.1/paper.tex
pj: carl r2 fixes

@@
 %% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
-%% -
+%% - threats to validity: only 1 migration path, no systematic evaluation (lattice)
 
@@
 \newcommand{\totalnum}[1]{$#1$ total}
+\newcommand{\pctnum}[1]{$#1\%$}
 
 %% TODO 2022-01-25 recount tests
-\newcommand{\numbenchmark}{2}
+\newcommand{\numbenchmark}{3}
 \newcommand{\numSPtest}{265}
@@
 the gradual typing community.
-Industry teams have developed innovative type systems that accommodate
-untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017}.
+Industry teams have developed innovative \emph{optional} type systems that accommodate
+untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017,pep484}.
 These systems intentionally check nothing at run-time when untyped values enter
@@
     Because the server code is closed-source, we additionally present data for
-    \numbenchmark{} public benchmarks.
-    Using only shallow types, \SP{} runs on-par with Python despite the costs of
-    enforcing soundness.
-    With fine-tuned types, \SP{} consistently out-performs the Python baseline.
+    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
+    Using only shallow types, \SP{} runs slightly faster than Python
+    despite the costs of enforcing soundness~(avg. \pctnum{28}).
+    With fine-tuned types, \SP{} consistently out-performs the Python baseline~(avg. \pctnum{67}).
 
@@
 type checkers that already exist for Python.
-In fact, \SP{} has no plans to support all \PEP{} types.
-The goal is to implement a small language of types efficiently.
+In fact, \SP{} has no short-term plans to support all \PEP{} types.
+The main priority is to implement a small language of types efficiently.
 
@@
 recursive types.
-Pyre and mypy some support for the first two kinds, but not for recursive types.
+Pyre and mypy offer some support for the first two kinds, but not for recursive types.
 %% mypy rec.types = https://github.com/python/mypy/issues/731
@@
 
-Every \SP{} type corresponds to a constant-time cast.
-When an unknown value reaches a boundary to typed code at run-time,
-\SP{} uses these checks to quickly decide whether or not the value
-matches the type (think: \code{isinstance} checks).
-Unlike structural gradual type systems, \SP{} does not traverse
-incoming values, nor does it create wrappers to enforce behavior.
-
-These dynamic checks try to faithfully enforce a full type, but some are incomplete---especially
-for generic types.
-We call the incomplete ones \emph{shallow} types.
-The \SP{} strategy for sound generics is to package each value with a type
-tag; following the \textsf{StrongScript} authors~\cite{rzv-ecoop-2015}, we call
-these \emph{concrete} types.
-
-First off, for a basic nominally-typed language (think Java 1.4, before
-generics), constant-time checks are easy to implement.
-Every type has a name and every value has a tag that corresponds to the
-name of its type.
-A cast must examine the parents of a class in the worst case,
-but these traversals are bounded and typically short.
-
-Parameterized types (generics) are more difficult to enforce.
-For example, suppose that a typed function expects inputs that match
+Every \SP{} type is paired with an efficient cast operation that runs in
+constant time regardless of how large an incoming value is.
+Unlike structural gradual type systems~\cite{gtnffvf-jfp-2019},
+no cast traverses an incoming value and no cast creates a wrapper
+to monitor future behaviors.
+Some casts for generic types are, however, incomplete.
+We call the types with incomplete casts \emph{shallow} types
+and the types with full casts \emph{concrete} types, following
+the \code{StrongScript} authors~\cite{rzv-ecoop-2015}.
+
+First off, the casts for non-generic nominal types are straightforward.
+Every type has a name and every value has a tag that corresponds to the name.
+A cast must examine the parents of a class to find a match in the worst case,
+but these traversals are bounded in length and typically short.
+
+Generic types are more difficult to enforce because a tag by itself does not
+describe their contents.
+For example, suppose that a function expects inputs that match
 the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
 integer values.
-A constant-time check for dictionaries is not enough to protect the typed
-function because it fails to validate the elements inside a dictionary.
-The \SP{} solution is to provide two incompatible versions of every parameterized type,
-both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
-and concrete versions at present.
-In particular, user-defined classes do not have concrete support.
-The shallow check for a user-defined class, however, guarantees the types of all
-non-generic fields and methods.}
-For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
+A constant-time check for Python dictionaries is not enough to protect the
+function against dicts with ill-typed elements.
+The \SP{} solution is provide a second kind of dictionary that ensures
+the validity of its elements.
+These two dicts have distinct types:
 \begin{enumerate}
@@
     The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
-    for dict values that ignores the elements inside a dict.
+    for dict values that ignores elements.
     Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
-    is merely a promise of dict values at run-time; it does not promise anything about
-    keys and values.
+    does not promise anything about keys and values.
   \item
@@
 These options let programmers decide whether to edit their code to support concrete types
-or leave it as-is with shallow types.
+or leave it as-is with shallow types.\footnote{Not all generic \SP{} types have shallow
+  and concrete versions at present.
+  In particular, user-defined classes do not have concrete support.
+  The shallow check for a user-defined class does, however, guarantee the types of all
+  non-generic fields and methods.}
 Often, the shallow types are more attractive because concrete types impose
@@
 The upside of concrete types is that they enforce stronger type constraints.
-These constraints can catch bugs and always lead to more-efficient typed code.
-Faster performance is not guaranteed, though, because concrete types must
+These constraints can catch bugs and always lead to faster typed code.
+Faster performance is not guaranteed in general, though, because concrete types must
 check all writes from untyped contexts.
@@
 a constant-time cost for casts and for writes in untyped code,
-and zero cost for dereferences.
+and zero cost for reads.
 If a value travels to untyped code, gets written to many times, and gets read few times,
-then the shallow types may run faster overall.
+then the shallow types may run faster in the end.
 
@@
 
-Cinder adds virtual method tables (vtables) to typed classes.
+%% vtables go in:
+%% - typed classes
+%% - builtin classes that SP treats as typed
+%% - typed and untyped subtypes of classes with vtables
+
+Cinder adds virtual method tables (vtables) to classes.
 These tables help to speed up method dispatch relative to Python's
@@
 Python allows classes to inherit from a list of parents.
-\SP{} accommodates this behavior provided that one parent is defined in typed
-code and the untyped parents do not specify their members (via
-\code{\_\_slots\_\_}).
-Other uses of multiple inheritance must stay in untyped code.
+\SP{} accommodates this behavior provided that at most one parent
+is typed (or uses \code{\_\_slots\_\_} to specify its layout; plain Python has the same restriction).
 
@@
 lives in each slot.
+Similarly, \SP{} cannot optimize instance variable reads because their slots
+may have a conflict among parents.
 %% possible improvement: invoke methods on the first parent statically and on
@@
 \begin{table}[t]
-  \caption{Microbenchmark performance ratios. NB: data for deltablue and nbody is pending.}
+  \caption{Microbenchmark performance ratios. NB: data for nbody is pending.}
+  %% FILL don't forget to update \numbenchmark{}
   \label{t:microbenchmark}
@@
 \SP{} is one of many type systems for Python.
-Other \emph{optional\/} type checkers include Pyre~\cite{pyre},
+Other optional type checkers include Pyre~\cite{pyre},
 mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these

### 2022-01-27 pj-7.1/paper.tex
pj: carl MI clarifications

@@
 module-level variables cannot be mutated; and
-typed classes can inherit from at most one other typed class.
+multiple inheritance among typed classes (or untyped classes with
+a typed ancestor) is forbidden.
 
@@
 %% attributes.
+
+%% So in normal Python you will get a layout error anytime you try to inherit from
+%% more than one class defining _slots_. Any SP class with instance attributes
+%% implicitly defines _slots_. So it depends how you interpret "where Python
+%% wouldn't." If you consider SP classes to be slots-defining (because we emit
+%% bytecode such that they are, even though they don't explicitly in the source
+%% code), then our behavior is just the normal Python behavior.
+%% 
+%% There is one edge case where we are more strict, which is a SP class with no
+%% instance attributes, only methods. In this case it will not have _slots_ and
+%% Python would allow it to multiply-inherit with something else that does have
+%% _slots_. But because we also have the vtables problem (which normal Python does
+%% not), we can't ever allow multiple inheritance from two SP classes, even if one
+%% or both of them has no instance slots, and even if that multiple inheritance
+%% occurs in a nonstatic module (the nonstatic subclass will still have a vtable,
+%% because we may statically invoke methods on it, since it is substitutable for
+%% its static parents, and that vtable will still be impossible to construct
+%% correctly.) We implement this restriction in a kind of lazy way by just
+%% piggybacking on Python's existing layout incompatibility check, and adding in a
+%% clause that forces every SP class to always be considered as slots-having, even
+%% if it doesn't actually have any instance variable slots. So in this edge case
+%% only, we will error where Python wouldn't.
+%% 
+%% I summarized all of this in as few words as I could with "at most one parent is
+%% defined in typed code or uses _slots_ to specify its members," which is what
+%% the above boils down to.
+%% 
+%% suppose technically we are slightly more restrictive than we would have to be:
+%% we currently forbid, but technically could support a case of multiple
+%% inheritance where one parent is an SP class with no instance attributes, and
+%% the other parent is a nonstatic class. But this seems not very useful to
+%% support, and just makes the rules even more complicated.
 

### 2022-01-28 pj-7.1/paper.tex
pj: 22 pages finally no thanks to baby cub

@@
 \newcommand{\SP}{Static Python}
+\newcommand{\JSPP}{JS\code{++}}
 \newcommand{\PEP}{PEP~484}
@@
 \subsection{Type System Context and Design Goals}
+\label{s:ts-context}
 
@@
 
-Cinder is a bytecode compiler that extends the Python instruction set.
-All the standard Python instructions work similarly to Python 3.8;
-in other words, an untyped Python program has the same behavior in
-Cinder as in the standard interpreter.
-The added instructions help with one of three tasks:
+All CPython 3.8 instructions have the same behavior in Cinder.
+Cinder adds instructions to help with one of three tasks:
 expressing runtime checks,
@@
 As an example of the third kind, the \code{FAST\_LEN} instruction quickly
-computes the length of a built-in value.
+computes the length of a built-in value, which helps to optimize for loops.
 \SP{} uses type information to decide where this instruction is appropriate
@@
 to a direct function-call jump.
+%% Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
+%% function calls.
 (Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
 instructions.)
-Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
-function calls.
 
@@
 protocol; it is not a from-scratch development.
-Cinder vtables are specialized to check untyped overrides of typed methods
+Cinder vtables are further specialized to check untyped overrides of typed methods
 using a wrapper implemented in C rather than in Python, to reduce performance costs.
@@
 In general, other checked datatypes have the same three components:
-a constructor, checked update functions, and a tag.
+a constructor, checked update functions, and a compound tag.
 
@@
 Any method can be JIT-compiled whether or not it uses \SP{} types,
-though types are needed to enable the most-effective optimizations.
+although types are needed to enable the most-effective optimizations.
 
@@
 
-To give one example issue, \SP{} accepted the following program whereas our model
+To give one example issue, \SP{} incorrectly accepted the following program whereas our model
 reported a type error:
@@
 More concerningly, the associated test case \emph{expected} the program to type check.
-Creating a formal model helped us catch this specification error.
+Our model caught this specification error.
 
@@
 
-\Cref{f:surface-types} presents a core surface syntax for model programs.
+\Cref{f:surface-types} presents a core syntax for programs.
 A program is a sequence of statements;
@@
 These definitions may only appear on the top level
-and they all require full type annotations.
+and they all require type annotations.
 Functions must have one positional argument.
@@
 Expressions describe values and simple computations.
-The basic values consist of the following: the none value, integers,
+The basic values are the none value, integers,
 booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
-The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
+Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
 and method calls.
@@
 In Python, the syntax for expressions overlaps with the syntax of types.
-The Python name \code{None} is both an expression and a type,
+For example, the Python name \code{None} is both an expression and a type,
 and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
-By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
+\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
 Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
@@
 checked dict values that were initialized with exactly the same key and value type.
-As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\spteval_0$
+As noted in~\cref{s:type-dynamic}, this exact-match rules applies even when $\spteval_0$
 or $\spteval_1$ is the dynamic type.
@@
 casts either succeed or fail immediately.
-%% In other words, \SP{} provides \emph{immediate
-%% accountability}~\cite{mt-oopsla-2017} for cast errors.
+In other words, \SP{} provides \emph{immediate accountability} same as
+Nom~\cite{mt-oopsla-2017}.
 
@@
 
-To better show where \SP{} needs to insert casts, we present a selection of typing rules.
+To show where \SP{} needs to insert casts, we present a selection of typing rules.
 Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
@@
 a formal account of type-directed optimization would require substantial additional
-components: a model of the \SP{} bytecode and a faithful rendering of its
+components; namely, a model of the \SP{} bytecode and a faithful rendering of its
 transformations.
@@
 %% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
-There are two cases in which \SP{} types impose constraints on untyped Python code.
+There are two cases in which \SP{} types impose dynamic constraints on untyped Python code.
 First, concrete generic types (e.g., $\sptrawchkdict$)
@@
 via materialization casts, but arguments from untyped contexts are unchecked.
-For this reason, \SP{} compiles every typed function to check its inputs;
-the optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
+For this reason, \SP{} compiles every typed function to check its inputs.
+The optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
 Similarly, concrete types such as $\sptrawchkdict$ must check writes that
@@
 
-\SP{} can generate efficient code because it targets
-the Cinder runtime.
-Cinder extends the Python 3.8 bytecode with instructions to check types at run-time,
-to construct \SP{} data structures, and to perform actions in a way that benefits from
-static information.
-
 \begin{table}
@@
 
+\SP{} can generate efficient code because it targets
+the Cinder runtime, which has instructions to check types at run-time, to
+construct \SP{} data structures, and to perform actions in a way that benefits
+from static information.
 \Cref{t:bytecode} lists a few representative bytecode instructions.
-The two instructions that express run-time checks are \code{CAST}
-and \code{CHECK\_ARGS}.
+The two instructions that express run-time checks are \bcinst{CAST}
+and \bcinst{CHECK\_ARGS}.
 The former checks a value against a type.
@@
 
-Two of the listed instructions allocate \SP{}-specific data structures.
+The instructions \bcinst{BUILD\_CHECKED\_MAP} and \bcinst{TP\_ALLOC}
+allocate \SP{}-specific data structures.
 The first creates a checked dictionary; the second allocates an uninitialized instance
@@
 In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
-The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
-of using the Python \code{\_\_len\_\_} attribute.
+%% The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
+%% of using the Python \code{\_\_len\_\_} attribute.
 Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
@@
 
-\paragraph{Removing Casts}
+\paragraph{Reducing Casts}
 
@@
 The optimizer uses this convention to skip argument checks by jumping past them
-when it is safe to do so; namely, whenever a typed function or method gets
+when it is safe to do so; that is, whenever a typed function or method gets
 called in a typed context.
@@
 If the experimental group can serve more requests per second, then the change
-is recorded as a success.
+is a success.
 Assuming that the experimental and control groups have equivalent hardware
@@
 Cython improved performance by compiling these modules to C,
-but its partial adoption had led to an awkward workflow (with a
-few compiled modules spread across an interpreted codebase).
+but its partial adoption had led to an awkward workflow with a
+few compiled modules spread across an interpreted codebase.
 Replacing these modules with \SP{} let programmers return to a conventional
@@
 affect tests rather than production code.
-Digging deeper, there are nine common kinds of changes: six address
-type errors and the other three are for performance.
+Digging deeper, there are nine common kinds of changes: five address
+type errors and the other four are for performance.
 
@@
     \ipara{Use primitive types}
-    Any Python integers and booleans in hot code paths are better as
-    \SP{} primitive values~(\cref{s:c-types}).
-    As noted above, primitives were required for \SP{} to match the performance of
-    Cython.
+    Integers and booleans in hot code paths run fastest with
+    primitive type~(\cref{s:c-types}).
+    %%As noted above, primitives were required for \SP{} to match the performance of
+    %%Cython.
     Twelve modules currently use primitives.
@@
     Functions calls that use only keyword arguments are not yet optimized by \SP{}.
-    Rewriting these functions and call-sites to use positional arguments improves performance.
+    Rewriting these functions and their call-sites to use positional arguments improves performance.
 
@@
 
-\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.\footnote{\url{https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks}}
+\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
 These microbenchmarks are admittedly small and quite different than typical application code,
@@
 All runtimes for this table came from a 64-bit CentOS Linux system using a build
-of \SP{} and Cinder compiled with default settings
-(commit \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{\code{d0d071a}}).
+of \SP{} and Cinder compiled with default settings.
 Each benchmark version ran 11 times total: once to warm up the bytecode caches
@@
 \code{time} utility).
-Every benchmark version invoked the JIT on the same functions---even the
-untyped versions.
+Every benchmark version invoked the JIT on the same set of functions.
+\Cref{a:microbench-detail} presents further details.
 
@@
 gives an immediate speedup (ratio = $0.90$).
-\Cref{a:microbench-detail} presents further benchmark details.
+
+%% FILL threats to validity
 
@@
 and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
-It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
-in the original Nom paper inspired the \SP{} team to pursue an implementation.
-
-In addition to concrete types, Thorn and StrongScript have optional like types.
+%% It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
+%% in the original Nom paper inspired the \SP{} team to pursue an implementation.
+In addition to concrete types, Thorn and StrongScript have optional (unsound) \emph{like} types.
 The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
-types, though like types make no claims about soundness.
+types.
 
@@
 for Python~\cite{pmmwplck-oopsla-2013}.
-In particular, the Full Monty paper reminded us that booleans \emph{are} integers in Python;
-\SP{} had overlooked this detail.\footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
+In particular, the Full Monty paper reminded us that booleans are integers in Python;
+\SP{} had overlooked this detail.
+%% \footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
 %% 2022-01-14 TODO check full monty with KC
 
-Our production experience with \SP{} suggests that a method-based JIT compiler can
-usually eliminate the costs of shallow types.
-This finding complements prior positive experience with shallow types and JIT
-technology; e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
+Production experience with \SP{} suggests that a method-based JIT can
+eliminate the costs of shallow types.
+This finding complements prior work on shallow types;
+e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
 and in Grace with the Graal partial-evaluation-based JIT~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
-Similarly, the Pycket team used a tracing JIT to reduce the costs of higher-order
+Similarly, the Pycket team successfully used a tracing JIT to reduce the costs of higher-order
 checks~\cite{bbst-oopsla-2017}.
-It remains to be seen whether a non-tracing JIT is effective for higher-order
-casts.
+%% It remains to be seen whether a non-tracing JIT is effective for higher-order
+%% casts.
 
@@
 Other optional type checkers include Pyre~\cite{pyre},
-mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
-systems do not ensure safe interactions with untyped Python.
-Another sound type system is Reticulated,
-which enforces structural types (rather than nominal) using its transient
-semantics~\cite{vss-popl-2017,v-thesis-2019}.
-Other \emph{optimizing} type systems for Python
-include Reticulated and mypyc.
-The latter compiles Python source to C extension modules; unlike \SP{},
-it does not offer a pure-Python developer experience.
-
-Although sound gradual type systems are rare in industry, there are at
-least three others besides \SP{}.
+mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}.
+Another sound type system is Reticulated, which enforces structural types and
+pioneered the transient semantics~\cite{vss-popl-2017,v-thesis-2019}.
+Optimizing type systems include Reticulated and mypyc~\cite{mypyc}.
+%% The latter compiles Python source to C extension modules; unlike \SP{},
+%% it does not offer a pure-Python developer experience.
+
+Although sound gradual types are rare in industry, there are at
+least three other languages that provide them.
 Dart 2 is a nominally-typed language with a dynamic type that is similar
-to the top object type but allows implicit downcasts to every other type~\cite{dart-types}.
-C\# includes a dynamic type that defers type checking until runtime~\cite{bmt-ecoop-2010}.
-JS++ is a sound type system for JavaScript that allows untyped values
+to a top (object) type but statically allows all method calls~\cite{dart-types}.
+C\# has a dynamic type that delays type checking until runtime~\cite{bmt-ecoop-2010}.
+\JSPP{} is typed JavaScript that allows untyped values
 using a catch-all \code{external} type~\cite{jspp}.
@@
 
-Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
-of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
-Primitive types are progressive types for base values.
-Concrete types are essentially a progressive form of shallow types.
+%% Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
+%% of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
+%% Primitive types are progressive types for base values.
+%% Concrete types are essentially a progressive form of shallow types.
 
@@
 and plans to convert thousands more in the future.
-
-And yet, \SP{} is carefully restricted to set attainable goals
-for its implementors and for adopters:
+And yet, \SP{} is carefully restricted to set attainable goals:
+%% for its implementors and for adopters:
 \begin{itemize}
@@
     The syntax of sound static types is relatively small compared
-    to the \PEP{} definition.
+    to \PEP{}.
     Higher-order types are absent.
-    The focus is instead on first-order class types, base types, and standard data structures.
-  \item
-    The type checker operates at a module-level granularity rather
-    than at the level of statements or expressions.
+    The focus is instead on first-order objects and standard data structures~(\cref{s:ts-context}).
+  \item
+    The type checker analyzes entire modules rather than individual statements,
+    and although it supports a dynamic type, dynamic type-checked code
+    does not have the same freedom as untyped code~(\cref{s:type-dynamic}).
+  \item
+    Shallow types accommodate Python code but offer coarse
+    soundness guarantees.
+    %% These types have been the top priority for \SP{}.
+    Concrete types are deeply sound, but add nonlocal constraints~(\cref{s:checked-type}).
+    %% \SP{} supports few concrete types at the moment and plans to add more.
   \item
     Gradual class hierarchies are permitted, but restricted to single inheritance.
-    Furthermore, static-to-static method overrides cannot reduce the precision of types---a restriction
-    that enables optimizations but breaks the static gradual guarantee.
-  \item
-    Shallow types such as $\sptrawpydict$ accommodate Python code but offer coarse
-    soundness guarantees.
-    These types have been the top priority for \SP{}.
-    Concrete types support full soundness, but impose nonlocal maintenance costs.
-    \SP{} supports few concrete types at the moment and plans to add more.
-  \item
-    Progressive primitive types are available to help developers maximize performance.
-    These types are intentionally incompatible with the dynamic type to help programmers
-    avoid casts.
+    Static-to-static method overrides cannot reduce the precision of types~(\cref{s:inheritance}).
+    %% ---a restriction
+    %% that enables optimizations but breaks the static gradual guarantee.
+  \item
+    Progressive primitive types are available to maximize performance~(\cref{{s:c-types}}).
+    %%These types are intentionally incompatible with the dynamic type to help programmers
+    %%avoid casts.
 \end{itemize}
@@
 research community, but they have been effective in practice.
-The current \CPUchange{} increase in CPU efficiency is a huge improvement
-at the scale of the Instagram web server.
-
-In conclusion, we conjecture that the \SP{} approach to gradual typing
-is a viable way to incrementally add types to a production codebase.
-Determining whether this conjecture holds more broadly is a critical
-question that calls for research on two fronts.
+The current \CPUchange{} increase in Instagram web server CPU efficiency is a huge improvement
+and was obtained incrementally without any careful planning to avoid performance bottlenecks.
+By contrast, bottlenecks are a major concern for expressive, structurally-typed
+gradual languages~\cite{gtnffvf-jfp-2019,tfgnvf-popl-2016}.
+
+In conclusion, the \SP{} approach to gradual typing seems to be a promising
+way to realize the vision of gradual typing.
+Determining whether this conjecture holds more broadly
+calls for research on two fronts.
 First, \SP{} must be applied to additional projects in a module-at-a-time manner.
-A exhaustive performance evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
-especially useful.
+A careful evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
+especially useful to find performance bottlenecks.
 Second, the \SP{} approach should be adapted to other languages.
-An optionally-typed language such as TypeScript would be an ideal starting
-point for such an experiment.
-Gradual, progressive soundness may prove itself to be an effective general strategy for
-gradual types in practice.
+An optionally language such as TypeScript would be an ideal starting
+point.
+%% Gradual, progressive soundness may prove itself to be an effective general strategy for
+%% gradual types in practice.
 

### 2022-01-28 pj-7.1/paper.tex
update

@@
 
-The modelling process helped uncover several critical issues with the
-implementation of \SP{}---an important payoff given that \SP{} is running in
-production at Instagram.
-Overall, we reported 25 bug reports to the \SP{} team via GitHub~(listed in
-\cref{a:github-issues}).
-Four of these were soundness issues, one of which could be easily exploited to raise
+The modelling process helped uncover several critical issues in \SP{}.
+This is a very important payoff given that \SP{} is running in production at
+Instagram.
+Overall, we made 25 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
+Four of these were soundness issues, one of which we could exploit to raise
 a segmentation fault.
 All but one of these soundness bugs have been fixed.
-Five issues were relatively minor; these dealt with confusing error messages
+Of the remaining issues, five were relatively minor; these dealt with confusing error messages
 and incorrect tests.
 The remaining 16 issues report bugs in language design and implementation.
-The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label: \lstinline{sp-correctness}.
+The \SP{} team has acknowledged these as bugs by applying a specific GitHub label: \lstinline{sp-correctness}.
 

### 2022-01-28 pj-7.1/paper.tex
ack
pj: more acks, more author

@@
 \affiliation[b]{Meta, Menlo Park, CA, USA}
+\author[b]{Aniket Panse}
+\authorinfo{(\email{aniketpanse@fb.com}) } %% FILL
 \author[b]{Dino Viehland}
@@
 \acks{
-  \SP{} was developed by Sinan Cepel, William Meehan, Carl Meyer, Aniket Panse,
-  and Dino Viehland.
+  \SP{} and the Cinder JIT were developed by
+  Max Bernstein,
+  John Biesnecker,
+  Jacob Bower,
+  Sinan Cepel,
+  Tiansi Hu,
+  Orvid King,
+  Vladimir Matveev,
+  William Meehan,
+  Carl Meyer,
+  Matt Page,
+  Aniket Panse,
+  Brett Simmers,
+  Andrei Talaba,
+  Dino Viehland,
+  and Shiyu Wang.
   The Brown authors contributed to the language by modeling it independently; they
@@
 
-  Special thanks to Guido van Rossum for helping the authors meet via Twitter
-  and for his ongoing contributions to Python.
+  Special thanks to Twitter, thanks to which the two teams
+  (specifically, Meyer and Krishnamurthi) met.
   This work was partly supported by the US National Science Foundation.

### 2022-01-31 pj-7.1/paper.tex
pj: author order
pj: add deltablue commit
pj: hiding for submission
- remove nbody
- hide acks
- add threats to the microbenchmark section

--- 2022-01-28 pj-7.1/paper.tex
@@
 \authorinfo{(\email{dinoviehland@fb.com}) } %% FILL
+\author[b]{Aniket Panse}
+\authorinfo{(\email{aniketpanse@fb.com}) } %% FILL
 \author[a]{Shriram Krishnamurthi}
@@
 \begin{table}[t]
-  \caption{Microbenchmark performance ratios. NB: data for nbody is pending.}
+  \caption{Microbenchmark performance ratios}
   %% FILL don't forget to update \numbenchmark{}
@@
     \bmname{deltablue} &         0.59 &           0.30 \\
-    \bmname{nbody}     &         FILL &           FILL \\
+    %% \bmname{nbody}     &         FILL &           FILL \\
     \bmname{richards}  &         0.53 &           0.22 \\
@@
   \end{tabular}
+
 \end{table}
@@
 
-%% FILL threats to validity
+%% \textbf{Note to Programming Reviewers:}
+%%  We plan to add a fourth benchmark to \cref{t:microbenchmark}.
+
+\paragraph{Threats to Validity}
+
+\Cref{t:microbenchmark} does not analyze benchmark configurations that mix
+typed and untyped code. %% ---only fully-typed and untyped configurations.
+Mixed configurations may reveal performance bugs similar to those found
+by an evaluation of Retifulated Python~\cite{gm-pepm-2018}.
+
+%% Although mixed configurations are unlikely to have disastrous
+%% overhead~\cite{tfgnvf-popl-2016} due to \SP's efficient strategy for enforcing
+%% types and Instagram's positive experience with mixed-typed code in
+%% production~(\cref{s:migration-path}), they could reveal performance bugs
+%% similar to those found by a prior evaluation of Retifulated
+%% Python~\cite{gm-pepm-2018}.
 
@@
 
-\acks{
-  \SP{} was developed by Sinan Cepel, William Meehan, Carl Meyer, Aniket Panse,
-  and Dino Viehland.
-  The Brown authors contributed to the language by modeling it independently; they
-  are not part of the \SP{} team.
-
-  Special thanks to Guido van Rossum for helping the authors meet via Twitter
-  and for his ongoing contributions to Python.
-  This work was partly supported by the US National Science Foundation.
-  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
+%\acks{
+%  \SP{} and the Cinder JIT were developed by
+%  Max Bernstein,
+%  John Biesnecker,
+%  Jacob Bower,
+%  Sinan Cepel,
+%  Tiansi Hu,
+%  Orvid King,
+%  Vladimir Matveev,
+%  William Meehan,
+%  Carl Meyer,
+%  Matt Page,
+%  Aniket Panse,
+%  Brett Simmers,
+%  Andrei Talaba,
+%  Dino Viehland,
+%  and Shiyu Wang.
+%  The Brown authors contributed to the language by modeling it independently; they
+%  are not part of the \SP{} team.
+%
+%  Special thanks to Twitter, thanks to which the two teams
+%  (specifically, Meyer and Krishnamurthi) met.
+%  This work was partly supported by the US National Science Foundation.
+%  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+%}
 
@@
 
-Benchmarks run on the following commit:
+Benchmarks \bmname{richards} and \bmname{fannkuch} ran on the following commit:
 
@@
 \footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}
+\end{center}
+
+\noindent\bmname{deltablue} ran on a later commit:
+
+\begin{center}
+\footnotesize\url{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}
 \end{center}

### 2022-01-31 pj-7.1/paper.tex
pj: cleanup

@@
 % > be making it work -- Shriram 2021-09-30
-
-%% TODO before submission
-%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
-%%   trying to avoid nonstatic? staticifying?
-%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
-%% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
-%% - threats to validity: only 1 migration path, no systematic evaluation (lattice)
 
@@
 \newcommand{\mffont}[1]{\mathit{#1}}
-\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}} % {\mfapply{\mffont{T}_{\downarrow}\!\mffont{S}}{#1}}
+\newcommand{\mftypeF}[1]{\mfapply{\mffont{R}}{#1}} % {\mfapply{\mffont{T}_{\downarrow}\!\mffont{S}}{#1}}
 \newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
@@
 
-\keywords{gradual typing, migratory typing}
+\keywords{gradual typing, migratory typing, language design, concrete types, shallow types}
 
@@
 in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
-and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
-compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
+and proposed implementation methods~\cite{kas-pldi-2019,fgsfs-oopsla-2018},
+compiler technology~\cite{vsc-dls-2019,bbst-oopsla-2017},
 and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
@@
 This paper reports an exception to the rule among industry-made gradual type systems.
-The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
+The \SP{} team at Instagram has developed a sound type system for a subset of
 Python along with a runtime system that uses soundness to drive optimizations.
@@
     A second level of \emph{concrete types\/} describes generic data structures
-    that guarantee the types of their elements.
+    that check the types of their elements.
     If programmers modify their code to build concrete structures instead of Python
@@
     and ran both property-based soundness tests and over 200 tests adapted from the \SP{} regression suite.
-    The model brings across the essence of \SP{} and lends support to the claim
-    that core \SP{} is indeed sound.
-    Perhaps more importantly, the \emph{process} of modeling revealed 20 significant
-    issues in \SP{}.
-    Four of these were soundness bugs.
+    The modeling effort revealed 20 significant issues in \SP{},
+    four of which were soundness bugs.
 
@@
 The interface that it offers to programmers, however, replicates the
-Python experience.
-\SP{} runs untyped Python 3.8 programs with minimal changes to their
+standard Python experience.
+\SP{} runs Python 3.8 programs with minimal changes to their
 behavior~(\cref{s:not-python}),
-and it compiles code on-the-fly to enable the same
-developer tools as standard Python.
+and it compiles code on-the-fly to support existing
+IDEs and developer tools.
 The advanced features of the \SP{} type system are offered on an opt-in basis
@@
 \SP{} can run this program as-is and even JIT-compile the function.
-The other two versions import dictionary types from the \SP{} standard library:
+The other two versions use dictionary types:
 \begin{itemize}
@@
     The $\sptrawpydict$ type describes normal Python dictionaries in a \emph{shallow} sense.
-    At compile-time, \SP{} uses this type to check the program for logical errors.
+    At compile-time, \SP{} uses this type to find basic logical errors.
     At run-time, \SP{} enforces the type by checking that all inputs to \code{f} are
@@
 \end{itemize}
-These typed versions demonstrate the multi-level nature of the \SP{} interface.
+These typed versions demonstrate the multi-level nature of \SP{}.
 At one level, there are types that describe standard Python values that can be added
 to a program with little-to-no code changes.
-Beyond that, \SP{} offers special-purpose types with stronger guarantees that impose
-nonlocal maintenance costs (i.e., on constructor calls).
+At the next level, \SP{} offers special-purpose types with stronger guarantees that impose
+nonlocal maintenance costs, such as requiring edits to constructor calls.
+% (i.e., programmers may need to modify constructor calls).
 
@@
     locally-defined functions and classes are thus a low priority; instead, the
-    dynamic type is available as a coarse approximation.
+    dynamic type serves as a coarse approximation.
   \item
@@
 \code{y} is a string, then the expression \code{x or y} has a union type.
-They also narrow down via type tests as a kind of occurrence
+Unions also narrow down via type tests as a kind of occurrence
 typing~\cite{tf-icfp-2010,gsk-esop-2011}.
 Union types are not, however, generally supported at run-time~(\cref{s:eval-types}).
-The only exception is that binary unions of a type and the Python \code{None}
-value get enforced with a run-time check ($\sptunion{\sptype, \sptnone}$, or
-$\sptoptional{\sptype}$).
+The only exception is binary unions with the Python \code{None}
+type ($\sptunion{\sptype, \sptnone}$, or
+$\sptoptional{\sptype}$); these unions are enforced at run-time.
 
@@
 recursive types.
-Pyre and mypy offer some support for the first two kinds, but not for recursive types.
+Pyre and mypy have some support for the first two kinds, but no support for recursive types.
 %% mypy rec.types = https://github.com/python/mypy/issues/731
@@
 \PEP{} types that \SP{} does not yet understand.
-If existing code uses the type $\sptset{\sptint}$ for sets of integers,
+For example, if existing code uses the type $\sptset{\sptint}$ for sets of integers,
 then \SP{} replaces it with dynamic.
@@
     function that expects dictionaries with dynamic values.
-    At runtime, the function rejects this argument because its type is not an
-    exact match.
+    At runtime, the function rejects this argument because the two types are
+    not an exact match.
 \end{itemize}
@@
 The dynamic error comes only after adopting a concrete type for checked dictionaries.
-More to the point, \SP{} is less interested in guarantees about \emph{removing
+In general, \SP{} is less interested in guarantees about \emph{removing
 an arbitrary annotation} and more interested in making sure that \emph{an
@@
 
-First off, the casts for non-generic nominal types are straightforward.
+First off, casts for non-generic nominal types are straightforward.
 Every type has a name and every value has a tag that corresponds to the name.
-A cast must examine the parents of a class to find a match in the worst case,
+In the worst case, a cast must examine the parents of a class to find a match,
 but these traversals are bounded in length and typically short.
@@
 the validity of its elements.
-These two dicts have distinct types:
+These different values have different types:
 \begin{enumerate}
@@
   \item
-    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
-    keys and values, but is incompatible with built-in Python dictionaries.
-    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
+    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a constant-time,
+    compound tag check for concrete dict values that were initialized with
+    exactly the same key and value types.
+    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
 \end{enumerate}
@@
 These options let programmers decide whether to edit their code to support concrete types
-or leave it as-is with shallow types.\footnote{Not all generic \SP{} types have shallow
-  and concrete versions at present.
-  In particular, user-defined classes do not have concrete support.
-  The shallow check for a user-defined class does, however, guarantee the types of all
-  non-generic fields and methods.}
+or leave it as-is with shallow types.
+
 Often, the shallow types are more attractive because concrete types impose
@@
 check all writes from untyped contexts.
-\Cref{t:shallow-vs-concrete} presents in more detail the worst-case costs of
+For comparison, \cref{t:shallow-vs-concrete} presents the worst-case costs of
 shallow and concrete dict types.
@@
 
+Not all generic \SP{} types have shallow and concrete versions at present,
+though they are coming soon.
+In particular, user-defined classes do not have concrete support.
+The shallow check for a user-defined class does, however, guarantee the types
+of all non-generic fields and methods.
+
 \begin{table}[t]
@@
 Gradual class hierarchies are rarely studied in the literature, especially
-for a language where truly untyped classes can enter the mix.\footnote{Two
-languages that provide separate hierarchies for untyped and (gradually) typed
-classes are Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}.}
+for a language where truly untyped classes can enter the mix.
+Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}, for example,
+provide separate hierarchies for untyped and (gradually) typed classes.
 The closest related work, for Nom~\cite{mt-oopsla-2017}, implements
@@
 There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
-
 The purpose of these types is to enable unboxed values and arithmetic at
@@
 refinement over the semantics of basic Python types.
-Primitive types cannot be implicitly cast to non-primitive types,
-including the dynamic type---another violation of the static gradual guarantee.
-Operations such as \code{and} (boolean conjunction) do not accept a mix of primitive and non-primitive data.
-These restrictions give programmers a static separation between fast-running
-primitives and standard data.
+
+Primitive types are restricted in two ways to enforce a static separation
+between fast-running primitives and standard data.
+First, primitives cannot be implicitly cast to non-primitive types,
+including the dynamic type.
+This is another violation of the static gradual guarantee.
+Second, operations such as \code{and} (boolean conjunction) do not accept a mix
+of primitive and non-primitive data.
 
@@
 module-level variables cannot be mutated; and
-multiple inheritance among typed classes (or untyped classes with
-a typed ancestor) is forbidden.
+multiple inheritance of either typed classes or untyped children
+of typed classes is forbidden.
 
@@
 Cinder vtables are further specialized to check untyped overrides of typed methods
-using a wrapper implemented in C rather than in Python, to reduce performance costs.
+using a wrapper implemented in C (rather than in Python) to reduce performance costs.
 
@@
 Any method can be JIT-compiled whether or not it uses \SP{} types,
-although types are needed to enable the most-effective optimizations.
+even though fewer types usually leads to fewer optimizations.
 
@@
 casts either succeed or fail immediately.
-In other words, \SP{} provides \emph{immediate accountability} same as
+In other words, \SP{} provides the same \emph{immediate accountability} as
 Nom~\cite{mt-oopsla-2017}.
@@
     \sptobject & \mbox{Accepts any value, except primitives~(\cref{s:c-types})} \\
-    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
+    \sptclass_0 & \mbox{Accepts instances of the class $\sptclass_0$} \\
     \sptnone & \mbox{Accepts the Python none value} \\
@@
 \medskip
-Consequently, this derivation calls for two casts to match the \trule{Matr} applications.
+Consequently, this derivation calls for two casts at the \trule{Matr} applications.
 
@@
   \item
-    concrete-typed dictionaries check all writes; and
+    concrete-typed dictionaries check all key and value writes; and
   \item
@@
 There are two cases in which \SP{} types impose dynamic constraints on untyped Python code.
-First, concrete generic types (e.g., $\sptrawchkdict$)
+First, concrete types (e.g., $\sptrawchkdict$)
 reject inputs that were not created via a checked constructor.
@@
 Calls to \code{eval} and \code{exec} have the dynamic type.
-The type system makes no finer prediction about what kind of value such calls will
-return.
-
+%The type system makes no finer prediction about what kind of value such calls will
+%return.
+%
 Studies of JavaScript and R have shown that many uses of dynamic execution
@@
     \bcinst{STORE\_FIELD} & Optimization & Write to an object field \\
-    \bcinst{FAST\_LEN} & Optimization & Get length of a built-in value \\
+    \bcinst{FAST\_LEN} & Optimization & Get the length of a built-in value \\
     \bcinst{REFINE\_TYPE} & Optimization & Type declaration for the JIT \\
@@
 \SP{} can generate efficient code because it targets
-the Cinder runtime, which has instructions to check types at run-time, to
-construct \SP{} data structures, and to perform actions in a way that benefits
-from static information.
+the Cinder runtime, which has bytecode instructions to check types at run-time, to
+construct \SP{} data structures, and to perform type-directed actions.
 \Cref{t:bytecode} lists a few representative bytecode instructions.
@@
 against their declared types.
-
 The instructions \bcinst{BUILD\_CHECKED\_MAP} and \bcinst{TP\_ALLOC}
@@
 Cinder comes with a similar instruction to build checked lists.
-%% The developers will add more instructions along the same lines in the future.
-
 The remaining instructions are for optimization.
@@
 The load and store instructions for fields improve upon Python's generic attribute lookup.
-In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
-%% The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
-%% of using the Python \code{\_\_len\_\_} attribute.
-Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
-type is not clear from the context.
+In the JIT, these instructions can be further optimized to a single assembly instruction.
+Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value
+when it is not clear from the context.
 
@@
 
-To measure the effect of a code change on these hosts, profiling tools select
+To measure the effect of a code change on these hosts, an internal profiling tool selects
 two representative sets of machines to run as experimental and control groups.
@@
 the control group gets the previous version.
-Next, the profiling tools slowly increase the number of requests that these
+Next, the profiling tool slowly increases the number of requests that these
 machines receive until each is running at maximum load.
-Once the machines are fully allocated, the tools measure how many requests
+Once the machines are fully allocated, the tool measures how many requests
 per second each group is able to handle.
@@
 Taking the sum of several CPU efficiency changes over a long
-timespan is therefore only a rough measure of their net effect.
+timespan (as we have done) is therefore only a rough
+measure of their net effect.
 %% In particular, although we report a \CPUchange{} net improvement
@@
 We therefore reviewed the $\sim 30$ conversion diffs that significantly improved
-performance to get a sense of the code changes at hand.
-
-In general, the patches have minor code changes and most of these changes
-affect tests rather than production code.
-Digging deeper, there are nine common kinds of changes: five address
-type errors and the other four are for performance.
+performance to assess the extent of the code changes.
+In general, the patches have minor changes that
+affect tests more than production code.
+There are nine common kinds of changes: five for types and four for speed.
 
@@
     \ipara{Fix type errors due to mock wrappers}
-    When test code introduces a mock wrapper, it may inadvertantly change the
-    return type of a function.
-    For example, the wrapped version of a function that returns \code{None}
-    will return a \code{MagicMock} object unless the programmer explicitly
-    specifies \code{None} as the desired return value.
+    When test code uses a mock wrapper, it may change the
+    return type of a function because wrappers return a \code{MagicMock} object
+    by default.
+    The fix is to specify a return value.
     %% Carl: This is an inherent and desired
@@
     \SP{} does not.
-    Programmers have to decide where an attribute belongs so that \SP{} can optimize
-    reads from instance attributes.
+    %% Programmers have to decide where an attribute belongs so that \SP{} can optimize
+    %% reads from instance attributes.
     %% Carl: (efficient via \code{LOAD\_FIELD})
@@
     must be changed or moved to an untyped module.
-    So far, these changes have not required major rewrites.
+    %% So far, these changes have not required major rewrites.
 
@@
     Functions calls that use only keyword arguments are not yet optimized by \SP{}.
-    Rewriting these functions and their call-sites to use positional arguments improves performance.
+    %% Rewriting these functions and their call-sites to use positional arguments improves performance.
 
@@
 The second compares a hand-refined benchmark to untyped Python.
-The refinements are modest, mostly adding primitive types to hot loops;
-they do not require significant expertise to reproduce.
-
+The refinements are modest, mostly adding primitive types to hot loops.
+%% they did not require significant expertise to create.
 All runtimes for this table came from a 64-bit CentOS Linux system using a build
@@
 Mixed configurations may reveal performance bugs similar to those found
-by an evaluation of Retifulated Python~\cite{gm-pepm-2018}.
-
-%% Although mixed configurations are unlikely to have disastrous
-%% overhead~\cite{tfgnvf-popl-2016} due to \SP's efficient strategy for enforcing
-%% types and Instagram's positive experience with mixed-typed code in
-%% production~(\cref{s:migration-path}), they could reveal performance bugs
-%% similar to those found by a prior evaluation of Retifulated
-%% Python~\cite{gm-pepm-2018}.
+by an evaluation of Reticulated Python~\cite{gm-pepm-2018}.
 

### 2022-02-01 pj-7.1/paper.tex
pj: threats -> threat
pj: nbody data

@@
     \bmname{deltablue} &         0.59 &           0.30 \\
-    %% \bmname{nbody}     &         FILL &           FILL \\
-    \bmname{richards}  &         0.53 &           0.22 \\
-    \bmname{fannkuch}  &         1.03 &           0.46
+    \bmname{fannkuch}  &         1.03 &           0.46 \\
+    \bmname{nbody}     &         1.09 &           0.24 \\
+    \bmname{richards}  &         0.53 &           0.22
     %% fannkuch typedbasic2 = 0.90
@@
 Python despite the cost of enforcing soundness.
-The one exception is the basic typed version of \bmname{fannkuch}, which sees a minor
-slowdown.
-This benchmark is primarily about number-crunching and the basic version uses standard
-numeric types that \SP{} does not optimize.
-Switching to primitive types without otherwise changing the code
-gives an immediate speedup (ratio = $0.90$).
+The exceptions are the simply-typed version of \bmname{fannkuch} and
+\bmname{nbody}.
+Both these benchmarks focus on numeric operations, which \SP{} does not
+optimize because it offers primitive types.
+Indeed, changing \bmname{fannkuch} to use a few primitive types without
+otherwise changing the code gives an immediate speedup (ratio = $0.90$).
 
@@
 
-\paragraph{Threats to Validity}
+\paragraph{Threat to Validity}
 
@@
 \begin{table}[tp]
+  \caption{Nbody microbenchmark data}
+  \label{t:mb:nbody}
+  \footnotesize\centering
+
+  \begin{tabular}{rrrrrr}
+    \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+    $0.17$ & $0.18$ & $3.61$ & $0.92$ & $0.88$ & $2.12$ \\
+    $0.17$ & $0.18$ & $3.57$ & $0.91$ & $0.88$ & $2.09$ \\
+    $0.17$ & $0.18$ & $3.68$ & $0.96$ & $0.88$ & $2.24$ \\
+    $0.17$ & $0.17$ & $3.82$ & $0.93$ & $0.9$ & $2.15$ \\
+    $0.17$ & $0.17$ & $3.72$ & $0.89$ & $0.9$ & $2.17$ \\
+    $0.18$ & $0.18$ & $3.63$ & $0.89$ & $0.9$ & $2.14$ \\
+    $0.18$ & $0.16$ & $3.5$ & $0.89$ & $0.89$ & $2.23$ \\
+    $0.18$ & $0.17$ & $3.45$ & $0.89$ & $0.93$ & $2.32$ \\
+    $0.2$ & $0.17$ & $3.61$ & $0.92$ & $0.88$ & $2.16$ \\
+    $0.2$ & $0.18$ & $3.63$ & $0.9$ & $0.92$ & $2.06$ \\
+  \end{tabular}
+
+  \begin{tabular}{rrrrrr}
+    \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+    $0.8$ & $0.81$ & $1.3$ & $0.83$ & $0.74$ & $1.2$ \\
+    $0.8$ & $0.8$ & $1.28$ & $0.78$ & $0.74$ & $1.22$ \\
+    $0.81$ & $0.8$ & $1.31$ & $0.75$ & $0.77$ & $1.22$ \\
+    $0.87$ & $0.86$ & $1.29$ & $0.74$ & $0.76$ & $1.18$ \\
+    $0.85$ & $0.8$ & $1.26$ & $0.76$ & $0.77$ & $1.19$ \\
+    $0.81$ & $0.81$ & $1.3$ & $0.77$ & $0.75$ & $1.22$ \\
+    $0.86$ & $0.79$ & $1.25$ & $0.76$ & $0.75$ & $1.17$ \\
+    $0.87$ & $0.83$ & $1.24$ & $0.74$ & $0.77$ & $1.16$ \\
+    $0.81$ & $0.81$ & $1.3$ & $0.74$ & $0.74$ & $1.17$ \\
+    $0.82$ & $0.83$ & $1.23$ & $0.76$ & $0.76$ & $1.16$ \\
+  \end{tabular}
+
+  \begin{tabular}{rrr}
+    \colname{Orig} & \colname{Orig} & \colname{Orig} \\
+    (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
+    $0.78$ & $0.75$ & $1.15$ \\
+    $0.74$ & $0.73$ & $1.19$ \\
+    $0.75$ & $0.72$ & $1.19$ \\
+    $0.73$ & $0.73$ & $1.18$ \\
+    $0.83$ & $0.76$ & $1.16$ \\
+    $0.82$ & $0.72$ & $1.24$ \\
+    $0.75$ & $0.72$ & $1.15$ \\
+    $0.72$ & $0.76$ & $1.2$ \\
+    $0.72$ & $0.72$ & $1.15$ \\
+    $0.74$ & $0.73$ & $1.2$ \\
+  \end{tabular}
+\end{table}
+
+\begin{table}[tp]
   \caption{Exact commands to invoke each benchmark (1/2)}
-  \label{t:mb:commands}
+  \label{t:mb:commands1}
   \tiny\centering
@@
     () &  \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
     (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
     (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
     (\colname{SP}) &  \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
     (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
     (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min-2} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
+    Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
     () &  \\
@@
     Fannkuch \colname{Orig} & \texttt{time ./python Tools/benchmarks/fannkuch.py 5} \\
-    () &
+    () &  \\
   \end{tabular}
@@
   \caption{Exact commands to invoke each benchmark (2/2)}
-  \label{t:mb:commands}
+  \label{t:mb:commands2}
   \tiny\centering
@@
   \begin{tabular}{ll}
+    \colname{Benchmark} & \colname{Command} \\\hline
     DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
@@
     () &  \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
     (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
     (\colname{SP} \colname{JIT}) & \texttt{ildcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
     (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
     (\colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
     (\colname{JIT}) & \texttt{ildcards Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min-2} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
+    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
     () &  \\
@@
     () &  \\
+    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
+    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
+    (\colname{SP} \colname{JIT}) & \texttt{ install-strict-loader Tools/benchmarks/nbody\_static.py} \\
+    NBody \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
+    (\colname{SP}) &  \\
+    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
+    (\colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame Tools/benchmarks/nbody\_static.py} \\
+    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
+    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
+    NBody \colname{T-Max} & \texttt{time ./python Tools/benchmarks/nbody\_static.py} \\
+    () &  \\
+    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
+    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
+    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
+    (\colname{SP} \colname{JIT}) & \texttt{rds -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
+    NBody \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
+    (\colname{SP}) &  \\
+    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
+    (\colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame Tools/benchmarks/nbody\_static\_basic.py} \\
+    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
+    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
+    NBody \colname{T-Min} & \texttt{time ./python Tools/benchmarks/nbody\_static\_basic.py} \\
+    () &  \\
+    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
+    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/nbody.py} \\
+    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
+    (\colname{JIT}) & \texttt{hmarks/nbody.py} \\
+    NBody \colname{Orig} & \texttt{time ./python Tools/benchmarks/nbody.py} \\
+    () &  \\
   \end{tabular}
+
 \end{table}

### 2022-02-01 pj-7.1/paper.tex
pj: kc / grammarly fixes
oh dear 'fuction' how can I mis-spell that

@@
 The premise is simple: let programmers introduce types in part of a
-codebase and while leaving the rest untyped.
+codebase while leaving the rest untyped.
 Run-time casts and checks enforce the assumptions that typed code makes about
@@
     despite the costs of enforcing soundness~(avg. \pctnum{28}).
-    With fine-tuned types, \SP{} consistently out-performs the Python baseline~(avg. \pctnum{67}).
+    With fine-tuned types, \SP{} consistently outperforms the Python baseline~(avg. \pctnum{67}).
 
@@
 Second, we want to entice researchers.
-The \SP{} type system has many restrictions that are suprising
-at first glance.
+The \SP{} type system has many noteworthy restrictions.
 For example, functions are supported only by shallow types
@@
 function against dicts with ill-typed elements.
-The \SP{} solution is provide a second kind of dictionary that ensures
+The \SP{} solution is to provide a second kind of dictionary that ensures
 the validity of its elements.
@@
 and zero cost for reads.
-If a value travels to untyped code, gets written to many times, and gets read few times,
+If a value travels to untyped codes and receives many writes but few reads,
 then the shallow types may run faster in the end.
@@
 \Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
-Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
+Expressions use lowercase names (\code{none}, \code{chkdict}) and types use capitalized
 names (\code{None}, \code{CheckedDict}).
@@
 checked dict values that were initialized with exactly the same key and value type.
-As noted in~\cref{s:type-dynamic}, this exact-match rules applies even when $\spteval_0$
+As noted in~\cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
 or $\spteval_1$ is the dynamic type.
@@
 type; and \trule{M-App} shows that typed methods impose similar constraints as
-typed fuctions.
+typed functions.
 
@@
 particular named static class.
-Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
+Flatt et al.~\cite{fkf-popl-1998} propose a more flexible approach, %(specifically for mixins)
 but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
@@
 two representative sets of machines to run as experimental and control groups.
-The experimental groups gets the latest version of the server code;
+The experimental group gets the latest version of the server code;
 the control group gets the previous version.
@@
 few compiled modules spread across an interpreted codebase.
-Replacing these modules with \SP{} let programmers return to a conventional
+Replacing these modules with \SP{} lets programmers return to a conventional
 Python workflow.
@@
     Integers and booleans in hot code paths run fastest with
-    primitive type~(\cref{s:c-types}).
+    primitive types~(\cref{s:c-types}).
     %%As noted above, primitives were required for \SP{} to match the performance of
@@
     Static methods can get invoked directly as functions, bypassing the class vtable.
-    Twenty-four modules use static methods, though many of these uses pre-date \SP{}.
+    Twenty-four modules use static methods (many of these pre-date \SP{}).
     %% 2022-01-12 Carl:
@@
 Second, the \SP{} approach should be adapted to other languages.
-An optionally language such as TypeScript would be an ideal starting
+An optionally-typed language such as TypeScript would be an ideal starting
 point.

### 2022-02-01 pj-7.1/paper.tex
pj: sk gmail
pj: fix sk confusions, remove adjectives

@@
 \author[a]{Shriram Krishnamurthi}
-\authorinfo{(\email{shriram@brown.edu}) } %% {is the Vice President of Programming Languages (no, not really) at Brown University.}
+\authorinfo{(\email{shriram@gmail.com}) } %% {is the Vice President of Programming Languages (no, not really) at Brown University.}
 
@@
 Almost all ratios are less than one, showing that \SP{} achieves a speedup over
-Python despite the cost of enforcing soundness.
-The exceptions are the simply-typed version of \bmname{fannkuch} and
-\bmname{nbody}.
-Both these benchmarks focus on numeric operations, which \SP{} does not
-optimize because it offers primitive types.
-Indeed, changing \bmname{fannkuch} to use a few primitive types without
-otherwise changing the code gives an immediate speedup (ratio = $0.90$).
-
-%% \textbf{Note to Programming Reviewers:}
-%%  We plan to add a fourth benchmark to \cref{t:microbenchmark}.
+Python.
+The exceptions are the simply-typed versions of \bmname{fannkuch} and
+\bmname{nbody}, which is unsurprising because these benchmarks focus
+on number-crunching and \SP{} optimizes arithmetic only for primitive types.
+Indeed, changing a few types in \bmname{fannkuch} to primitives
+gives an immediate speedup (ratio = $0.90$).
 
@@
 typed and untyped code. %% ---only fully-typed and untyped configurations.
-Mixed configurations may reveal performance bugs similar to those found
-by an evaluation of Reticulated Python~\cite{gm-pepm-2018}.
+A study of mixed configurations (i.e.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
+may reveal performance bugs that have not yet shown up in production.
 
@@
 and plans to convert thousands more in the future.
-And yet, \SP{} is carefully restricted to set attainable goals:
+And yet, \SP{} is restricted to set attainable goals:
 %% for its implementors and for adopters:
@@
 First, \SP{} must be applied to additional projects in a module-at-a-time manner.
-A careful evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
+A systematic evaluation of a few small projects would be
 especially useful to find performance bottlenecks.

### 2022-02-03 pj-7.1/paper.tex
pj: ie -> eg
draft header

@@
 
-\titlerunning{DRAFT do not distribute}
-
 \author[a]{Kuang-Chen Lu}
@@
 
-\authorrunning{DRAFT do not distribute}
+\titlerunning{Draft document---please do not distribute or cite!}
+\authorrunning{Draft document---please do not distribute or cite!}
 
@@
 typed and untyped code. %% ---only fully-typed and untyped configurations.
-A study of mixed configurations (i.e.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
+A study of mixed configurations (e.g.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
 may reveal performance bugs that have not yet shown up in production.

### 2022-04-04 pj-7.1/paper.tex
pj: note for later, ack Meta funding
pj: update speedup %s, follow up in microbench section

@@
     Using only shallow types, \SP{} runs slightly faster than Python
-    despite the costs of enforcing soundness~(avg. \pctnum{28}).
-    With fine-tuned types, \SP{} consistently outperforms the Python baseline~(avg. \pctnum{67}).
+    despite the costs of enforcing soundness~(avg. \pctnum{20}).
+    With fine-tuned types, \SP{} consistently outperforms the Python
+    baseline~(avg. \pctnum{70}).
 
@@
 Almost all ratios are less than one, showing that \SP{} achieves a speedup over
-Python.
-The exceptions are the simply-typed versions of \bmname{fannkuch} and
-\bmname{nbody}, which is unsurprising because these benchmarks focus
+Python (typed avg. $0.8$, refined avg. $0.3$).
+The exceptions are the unrefined versions of \bmname{fannkuch} and
+\bmname{nbody}, which is because these benchmarks focus
 on number-crunching and \SP{} optimizes arithmetic only for primitive types.
-Indeed, changing a few types in \bmname{fannkuch} to primitives
+Changing a few types in \bmname{fannkuch} to primitives
 gives an immediate speedup (ratio = $0.90$).
@@
 %  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+%  TODO partial funded by Meta
 %}

### 2022-04-19 pj-7.1/paper.tex
pj: comment, list of soundness issues
pj: checkpoint for artifact
pj: typed attr restriction

@@
 %% TODO after submission, before camera-ready
-%% -
+%% - reword 'fields' to attribute/descriptor/property?
+
+%% TODO 2022-04-08 from Pyre team
+%% Hi Carl! The small notational question I mentioned about in the paper was in
+%% Figure 5 -- there's this "M-APP" rule that describes how classmethod invocation
+%% is handled. I was trying to figure out how inherited classmethod would work
+%% with this rule. Say I have a class B that defines a method f0(B, S0) -> S1, and
+%% then I have another class D that inherits B which does not define additional
+%% method. Now given Env0 \vdash expr0: S0, how can Env0 \vdash D.f0(expr0): S1 be
+%% proven? If I understand correctly, the rule only looks up the type of f0 in
+%% Env0(D), i.e. the body of the derived class. But the real definition of f0 is
+%% syntactically located in the base class B. Does the notation system used here
+%% assume that every function definition in base class gets automatically copied
+%% into the derived class? Or should there be an additional preconditions added to
+%% the "M-APP" rule saying that f0 is allowed to exist in any of the superclass of
+%% D?
 
@@
 In fact, \SP{} has no short-term plans to support all \PEP{} types.
-The main priority is to implement a small language of types efficiently.
 
@@
 
+  \item
+    Typed fields cannot be overridden by descriptors or properties (\code{@property}).
+
 \end{enumerate}
@@
 and use checked dispatch for other combinations.
+\SP{} can also optimize reads from typed fields.
 One extra step arises when an untyped method overrides a typed method.
@@
 ensure predictable layouts (all \PEP{} checkers agree on this);
-module-level variables cannot be mutated; and
+typed attributes are resolved statically, and thus cannot be overridden with
+descriptors or properties;
+module-level variables cannot be mutated;
+typed fields 
+and
 multiple inheritance of either typed classes or untyped children
@@
 
+%% Carl 2022-04-07
+%% Ok, I completed my review of open issues. Turns out #55 was fixed back in
+%% December already, we just never closed it :) I found and closed a couple others
+%% in the same situation. We are left with a total of 9 issues opened by you all,
+%% 8 by KC and 1 by Ben:
+%%
+%%  https://github.com/facebookincubator/cinder/issues/created_by/LuKC1024
+%%
+%%  https://github.com/facebookincubator/cinder/issues/created_by/bennn
+%%
+%% #61 is trivial, I have a fix up and it should land today.
+%%
+%% Of the rest, I don’t think any are soundness issues. Most are cases where we
+%% could just be stricter or issue better errors statically (#40, #49, #50, #51,
+%% #52, #60 – #60 might already be fixed, #50 is about to land). #63 is a small
+%% improvement to type refinement. #71 is a pretty nasty issue that prevents use
+%% of CheckedDict with dynamic key or value type in certain positions that might
+%% take us a while to fix. But I’m comfortable saying that none of these are
+%% soundness issues. Let me know if your analysis differs.
+
 The modelling process helped uncover several critical issues in \SP{}.
@@
 The \SP{} team has acknowledged these as bugs by applying a specific GitHub label: \lstinline{sp-correctness}.
+
+%% critical soundness issues:
+%% https://github.com/facebookincubator/cinder/issues/55
+%% https://github.com/facebookincubator/cinder/issues/53
+%% https://github.com/facebookincubator/cinder/issues/39
+%% https://github.com/facebookincubator/cinder/issues/36
 
@@
 
-%\acks{
-%  \SP{} and the Cinder JIT were developed by
-%  Max Bernstein,
-%  John Biesnecker,
-%  Jacob Bower,
-%  Sinan Cepel,
-%  Tiansi Hu,
-%  Orvid King,
-%  Vladimir Matveev,
-%  William Meehan,
-%  Carl Meyer,
-%  Matt Page,
-%  Aniket Panse,
-%  Brett Simmers,
-%  Andrei Talaba,
-%  Dino Viehland,
-%  and Shiyu Wang.
-%  The Brown authors contributed to the language by modeling it independently; they
-%  are not part of the \SP{} team.
-%
-%  Special thanks to Twitter, thanks to which the two teams
-%  (specifically, Meyer and Krishnamurthi) met.
-%  This work was partly supported by the US National Science Foundation.
-%  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-%  TODO partial funded by Meta
-%}
+\acks{
+  \SP{} and the Cinder JIT were developed by
+  Max Bernstein,
+  John Biesnecker,
+  Jacob Bower,
+  Sinan Cepel,
+  Tiansi Hu,
+  Orvid King,
+  Vladimir Matveev,
+  William Meehan,
+  Carl Meyer,
+  Matt Page,
+  Aniket Panse,
+  Brett Simmers,
+  Andrei Talaba,
+  Dino Viehland,
+  and Shiyu Wang.
+  The Brown authors contributed to the language by modeling it independently; they
+  are not part of the \SP{} team.
+
+  Special thanks to Twitter, thanks to which the two teams
+  (specifically, Meyer and Krishnamurthi) met.
+  This work was partly supported by the US National Science Foundation.
+  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+  TODO partial funded by Meta
+}
 

### 2022-04-23 pj-7.1/paper.tex
p edits keep coming

@@
 %% - reword 'fields' to attribute/descriptor/property?
+%% - 
+%% - 
+%% -
 
@@
   \item
-    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
+    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
     for dict values that ignores elements.
@@
   \item
-    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a constant-time,
-    compound tag check for concrete dict values that were initialized with
-    exactly the same key and value types.
+    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
+    three-part check for concrete dicts with exactly the same key and value
+    types.
+    The overall cost of such a check depends on the nesting of concrete types,
+    but (crucially) does not depend on the size of incoming values.
     %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
@@
 %
-These options let programmers decide whether to edit their code to support concrete types
-or leave it as-is with shallow types.
-
+These options keep run-time costs low and let programmers decide between
+concrete and shallow types.
 Often, the shallow types are more attractive because concrete types impose
@@
 \begin{table}[t]
-  \caption{Worst-case costs for shallow and concrete types (T=typed, U=untyped)}
+  \newcommand{\nocost}{\textrm{---}}
+  \caption{Costs of using shallow and concrete dictionaries in typed (T) and untyped (U) contexts relative to the number (N) of dict elements. A dash (\nocost{}) means zero cost.}
   \label{t:shallow-vs-concrete}
@@
                    & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
-    \sptrawpydict         &           0 &        O(1) &          O(1) &              0 &             0 &              0 \\
-    \sptchkdict{K}{V}     &        O(N) &        O(1) &             0 &              0 &             0 &           O(1)
+    \sptrawpydict         &   \nocost{} &        O(1) &          O(1) &      \nocost{} &     \nocost{} &      \nocost{} \\
+    \sptchkdict{K}{V}     &        O(N) &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1)
   \end{array}\)
@@
 The closest related work, for Nom~\cite{mt-oopsla-2017}, implements
-a rather flexible point in the design space.
+a flexible design inspired by the gradual guarantees.
+In particular, Nom subclasses can override any part of a signature with the
+dynamic type.
 \SP{} implements a simpler design that restricts types and
@@
 There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
-The purpose of these types is to enable unboxed values and arithmetic at
+The purpose of these types is to allow unboxed values and arithmetic at
 runtime.
 Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
-refinement over the semantics of basic Python types.
-
-Primitive types are restricted in two ways to enforce a static separation
-between fast-running primitives and standard data.
-First, primitives cannot be implicitly cast to non-primitive types,
-including the dynamic type.
-This is another violation of the static gradual guarantee.
-Second, operations such as \code{and} (boolean conjunction) do not accept a mix
-of primitive and non-primitive data.
+refinement over the semantics of basic Python types:
+they permit fewer behaviors statically to enable the unboxing.
+Operations such as \code{and} (boolean conjunction), for example, do not accept
+a mix of primitive and non-primitive arguments.
+
+Because performance is the top priority for primitive types, they
+are intentionally incompatible with the dynamic type.
+Casts between a primitive type and the dynamic type are rejected
+statically---another violation of the gradual guarantee---to encourage the
+spread of primitives in typed code.
+At the boundaries to untyped Python code, however, \SP{} inserts casts.
+These casts let programmers introduce primitive types in one module
+without having to rewrite its untyped dependents.
 
@@
 own expertise---to look for non-conformance.  Second, 
-we used Redex's random testing tools~\cite{kf-sfp-2009} to check
+we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
 type soundness on thousands of examples (1,600 expressions and 11,000
 programs).
-% \footnote{Random testing helped us find and fix several issues in the
+% \footnote{Property-based testing helped us find and fix several issues in the
 % model.  None of issues had implications for \SP{}'s soundness.}

### 2022-04-24 pj-7.1/paper.tex
pj: review edits v2

@@
 
+%% 2022-04-24 More model issues:
+%% - C.f(e) uses a class type, not an instance ... how to invoke method on instance?!
+%% ... we should really apologize for a too-simple model
+%% - Object is not clearly in C
+%% - introduce c again as variable name? gee why not expr.method()?
+
 \newcommand{\shorturl}[2]{\href{#1#2}{#2}}
@@
 \newcommand{\paramtype}[2]{#1[#2]}
-\newcommand{\sptype}{\typefont{T}}
-\newcommand{\spteval}{\typefont{S}}
+\newcommand{\sptype}{\typefont{S}}
+\newcommand{\spteval}{\typefont{T}}
 \newcommand{\sptclass}{\typefont{C}}
@@
 \newcommand{\spbool}{\codefont{bool}}
-\newcommand{\spstr}{\typefont{str}}
+\newcommand{\spstr}{\codefont{str}}
 \newcommand{\spfloat}{\codefont{float}}
@@
 \newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
+\newcommand{\spcnew}[2]{\spapp{#1}{#2}}
 \newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
@@
     The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
-    three-part check for concrete dicts with exactly the same key and value
+    check for concrete dicts that have exactly the same key and value
     types.
-    The overall cost of such a check depends on the nesting of concrete types,
-    but (crucially) does not depend on the size of incoming values.
+    To implement such checks efficiently, \SP{} stores a run-time registry
+    of instantiated concrete types.
+    Every concrete dict carries a pointer into this registry to enable
+    constant-time type validation, even for nested concrete types.
     %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
@@
   \item
-    The dictionary value stores a compound type tag 
-    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
-    to support casts from the dynamic type.
+    The dictionary value stores a type tag to support casts from the dynamic
+    type.
+    A tag represents a three-part type $\sptchkdict{K}{V}$ and is implemented
+    as a pointer into a global registry of instantiated types.
     When a concrete dictionary enters typed code from an untyped context,
@@
 In general, other checked datatypes have the same three components:
-a constructor, checked update functions, and a compound tag.
+a constructor, checked update functions, and a tag.
+The tag is supported by a global registry of instantiated concrete types.
 
@@
 Functions must have one positional argument.
-Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
+Classes must declare one parent (either $\sptobject$ or another class), one field, and one method.
 Expressions describe values and simple computations.
@@
 user-defined class, and union types.
+We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
+\SP{} does not have a surface syntax for its dynamic
+type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
     \spexpr & \defeq &
-      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid \spobject \langmid
+      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid
   \\ & &
@@
       \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
+      \spcnew{\sptclass}{\spexpr, \ldots} \langmid
   \\ & &
@@
     \spx, \spf & \defeq & \mbox{variable names}
+    %% TODO should C be a claass name? ... should c be variable name?
   \end{langarray}
@@
 
-  \caption{Surface Syntax and Types}
+  \caption{Surface syntax and types}
   \label{f:surface-types}
@@
 
-  \caption{Evaluation Types, Surface-to-Evaluation Mapping}
+  \caption{Evaluation types, Surface-to-Evaluation mapping}
   \label{f:surface-to-eval-types}
@@
 
-Evaluation types $\spteval$ are a subset of the surface types.
+Evaluation types $\spteval$ are a simplification of the surface types.
 These are the types that \SP{} promises to soundly enforce.
@@
 \begin{table}[t]
-  \caption{How to Enforce the Evaluation Types}
+  \begin{threeparttable}
+  \centering
+  \caption{Description of run-time checks for the evaluation types}
   \label{t:cast}
 
-  \centering
-  \(\begin{array}{l@{\quad}l}
-    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
-    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
-    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
-    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
-    \sptdyn & \mbox{No check needed, accepts any value} \\
-    \sptobject & \mbox{Accepts any value, except primitives~(\cref{s:c-types})} \\
-    \sptclass_0 & \mbox{Accepts instances of the class $\sptclass_0$} \\
-    \sptnone & \mbox{Accepts the Python none value} \\
-    \sptint & \mbox{Accepts integers} \\
-    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
-    \sptstr & \mbox{Accepts strings}
-  \end{array}\)
+    \begin{tabular}{l@{\quad}l}
+    {Eval. Type $\spteval$} & {Run-time check} \\\hline
+    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
+    $\sptrawpydict$ & {Accepts any Python dictionary} \\
+    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
+    $\sptdyn$ & {No check needed, accepts any value} \\
+    $\sptobject$ & {Accepts any value\tnote{*}} \\
+    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$} \\
+    $\sptnone$ & {Accepts the Python none value} \\
+    $\sptint$ & {Accepts integers} \\
+    $\sptbool$ & {Accepts booleans (the integers \code{0} and \code{1})} \\
+    $\sptstr$ & {Accepts strings}
+    \end{tabular}
+
+    \begin{tablenotes}
+    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ rejects primitive values~(\cref{s:c-types})}
+    \end{tablenotes}
+\end{threeparttable}
 \end{table}
@@
     \inferrule*{
-      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
-      \\
-      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
+      \sptclassdef{\sptclass_0}{\sptclass_1}{\ldots}{\ldots} \in \sptenv_0
     }{
@@
 
-  \caption{Expression Typing, Consistent Subtyping, and Materialization}
+  \caption{Expression typing, consistent subtyping, and materialization}
   \label{f:eval-types}
@@
 the behavior of complex expressions---and this job is
-better left to the Redex mechanization.
+better left to the Redex mechanization, which at any rate supports a much
+larger subset of Python~(\cref{s:surface}).
 
@@
 %
-%  \caption{Program Typing}
+%  \caption{Program typing}
 %  \label{f:program-typing}

### 2022-04-26 pj-7.1/response.txt
pj: first-draft response + updated paper

TODO
- write a quick thank-you and summary of changes
- add section + page numbers to the responses

> Review #45A
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The paper makes a contribution to gradual typing in theory and practice. The
> subject language is Static Python, which adds a compile-time type discipline to
> Python.
>
> On the theory side, the paper describes a formal model created with an
> explorative tool for semantics modeling and prototyping. This model includes a
> static semantics and a dynamic semantics. The dynamic semantics is tested
> against a regression test suite.
>
> On the practical side, the paper describes the design of Static Python as well
> as experiments with an implementation of Static Python. This implementation
> allegedly improves the throughput of a closed source implementation of a web
> server by 3.6%. It improves the run time of several microbenchmarks
> significantly.
>
> Assessment and comments (for authors)
> -------------------------------------
> ## strengths
>
> * The main strength of the paper is clearly the favorable evaluation of a
>   particular datapoint in the design space of gradual typing.
> * The formal model enabled the discovery of problems in a industrial strength
>   implementation of Static Python, at least in its type checker.
>
> ## weaknesses
>
> * the most significant evaluation is on a closed source program
> * the microbenchmarks potentially gain more efficiency from progressive
>   primitive types as most of them are computationally intensive.
> * the theoretical evaluation is only probabilistic with unclear probability
>
> ## comments
>
> I like the paper, it is generally well readable. There are a few places, where
> I'd expect more guidance from the authors, rather than expecting the reader to
> interpolate.
>
> ### Table 1
>
> The caption is misleading. The data is the cost of creating and using
> shallow/concrete dictionary types in typed and untyped contexts.  I was
> initially confused by the 0 entries. I'd prefer you put --- to indicate that no
> casts is needed.  You should indicate that $N$ is the length of the dictionary.

Thanks, we've updated the caption, explained N, and used --- for no cast.

> ### enforcement of CheckedDict[K,V]
>
> *... gets enforced in constant-time* What if V is also a CheckedDict?

We've added an explanation of nested types (they do run in constant time).

> ### p8, 2
>
> *Nom [...] flexible point in the design space*
> Explain this point in two sentences.

Done.

> ### p8, se 2.5
>
> *...static semantics is progressive...*
> It seems like the second paragraph *Primitive types...* explains the
> progressive aspect. You should affirm that in the text.

We've expanded the first paragraph and revised the second paragraph. The first
paragraph focuses on the "normal" progressive aspects. The second focuses on
gradual typing.

> ### p9, -3
>
> *one token*
> Is this a variable size token or fixed size (if V is CheckedDict? see above)

It is always a fixed-size pointer.

> ### p10, 2
>
> *only if*
> Is that an `iff`?

No, the "only if" is intentional. An "iff" would be correct, but the paragraph
in which this sentence appears is about one direction in particular.

> ### p10, -17
>
> *Redex's random testing ... to check type soundness*
>
> That seems a bit shaky. What's the baseline / ground truth / oracle for these tests?
> In the end, all that testing only gives you a probabilistic guarantee.
> Can you give an estimate of this probability? (I mean, quantitative)

These are property-based tests. Our model defines a soundness property for terms
and asks Redex to generate random terms in the hope of finding a counterexample.
So, the soundness property is the ground truth; these random tests are all
checking soundness.

One could imagine a probabilistic guarantee about what kinds of terms the tests
explore, but the guarantee would depend on two complicated parameters: our
model's grammar and Redex's algorithm for exploring grammars (Klein & Findler,
SFP 2009). These parameters make it difficult to formulate a guarantee. On top
of that, **explaining** the guarantee to readers would be a huge challenge.

Property-based testing is shaky compared to formal verification, but it's still
a helpful tool for finding bugs.  Indeed, we found several bugs in our model
thanks to these property-based tests (none of which had implications for Static
Python itself).

> ### p10, -2
>
> *All but one...*
>
> Which soundness bug was **not** fixed and why?

All soundness bugs have been fixed at this time.

> ### p11, -3
>
> `Dyn` should be `Any` in Python.

The revised paper explains why we use the name `Dyn`.

> ### p12, fig 3, 4
>
> * `str` should be italic
> * why is `object` part of the expression language?

Fixed `str` and removed `object`.

(`object` was there as a simple instance of the `Object` type. Since it's not
needed to explain type boundary issues, we can remove it.)

> Line 6: `C.f(expr)`
> * looks like a static method call to me, is that right? That's usually trivial
>   compared with the dynamic dispatch of a real method call.
> * how do you invoke a method on an object?

Oh dear. This was not meant to be a static method call, but we were far too
aggressive when simplifying the model.

The revision uses `expr.f(expr)` instead and looks up the class type of the
receiver to type check the call.

> * how do you create an object? Is there a syntax like `C(expr)`?

Syntax for making an object was missing from the paper (not from SP or Redex!).
We've added it, thanks!

> ### fig 4
>
> Some evaluation types don't make sense
>
> * can Optional types be nested? E.g. Optional[Optional[int]]   doesn't make
>   sense to me
> * what about Optional[None] ??
> * or Optional[Dyn]??

Yes, those types can all be written. Just like surface unions (section 4.1), we
assume that Optional evaluation types get normalized in a standard way. The
normal forms for these examples are: `Optional[int]`, `None`, and `Dyn`.

> BTW, I find it confusing that surface types are `T`, but evaluation types are
> `S`, given that surface starts with s.

Good point, we've switched `S` and `T`.

> ### table 2
>
> Misleading caption: this is about the cast from Any to an evaluation type.

We have reworded the caption.

> ### p14, fig 5
>
> * rule `M-App` seems too limiting. Literally read, it's not possible to invoke
>   inherited methods because the method `f0` must be defined in class `C0` in
>   which it is defined. This rule affirms the class-method conjecture from
>   above.

We had envisioned the lookup `Env(C0)` as searching both `C0` and its parents
for all visible fields and methods.

The revised figure writes `Env*(C0)` instead and includes a note to this effect.
(The revision also uses `\Gamma` instead of `Env`)

> * the `object` rule for subtyping contradicts Table 2, which says that *objects
>   accepts any value expect primitives*. So `S0` must not be a primitive; but
>   primitives are not part of the model. This is not consistent!

We've moved the Table 2 comment about primitives to a footnote (because, as you
rightly point out, they are not part of the model).

> * given transitivity, the last rule could be simplified by omitting the
>   assumption `C2 <: C1` and changing the conclusion to `C0 <: C1`.

Thanks!

> ### p15, -16
>
> *better left to the Redex mechanization*
>
> It would be useful to show the excerpt for method calls here, as they seem to
> be nonstandard.

Method calls in Redex are standard; they work like calls in (Static) Python.
The confusion here is due to our over-simplified model in the submission, which
gave the impression that only a class can receive a method call.

> ### p24
>
> references [23] and [24] and [41] have broken DOIs

Fixed.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45B
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The contribution of the paper is the following:
>
> - The paper describes Static Python, a project developed at Instagram to add
>   types to Python code in a gradual manner, with the main objective of
>   improving performance. The project has already shown its effectiveness. The
>   authors (informally) describe both the language (the gradual type system) and
>   how this is implemented to achieve better performance.
>
> - The authors have developed an implementation in PLT redex of a subset of
>   Static Python, focusing on features related to gradual typing. Using this
>   implementation, they did an extended testing to show:
>   1) that this implementation is sound
>   2) that this implementation gives the same results of Static Python.
>
> Assessment and comments (for authors)
> -------------------------------------
> Assessment: The contributions of the paper summarized above are certainly
> useful for the PL community, so I support acceptance. The empirical evaluation
> seems to me (but I am definitely not an expert in such kind of contributions)
> well-driven and adequate. I have some reservations on the formalization,
> explained in the comments below, which should be solved/improved in the final
> version.
>
> Detailed comments:
>
> Page 3 Mechanization: you should explain better what do you mean by
> "soundness". I thought of standard soundness in the sense that "well-typed
> programs do not go stuck", but here you say "property-based soundness tests",
> what do you mean more precisely?

Section 4 in the revised paper gives more details.

In short, we started with the property "well-typed programs do not get stuck"
and tested it on thousands of terms generated by Redex.

> Section 4.1 line 4: In the formal syntax in Figure 3, Object seems not to be an
> instance of C, so class C(C) seems not to include the case C = Object.
> In the "Other expressions" you mention object fields reads and writes, but in
> the formal syntax in Figure 3 you have C.x and C.x = expr, I was expecting
> expr.x and expr.x = expr' instead.

We've made several changes in response:

- `Object` is no longer part of the grammar
- The grammar comes with a note that the set `C` must contain `Object`
- Field reads and writes use `expr` instead of `C` (example: C.x is now expr.x)

> Figure 3: it would be better to use a more abstract style with metavariables as
> it is customary now, e.g.
> prog = stmt_1, ..., stmt_n
> expr = ... [x_1:expr_1, ..., x_n:expr_n] | expr[expr'] |

We appreciate the suggestion, but prefer to keep the current grammar because it has
no substripts and uses a uniform style for the "sequence" terminals `prog` and `Env`.

> Page 13 line -7: A dynamically-typed variable -> A dynamically-typed function

No, "variable" is there on purpose because we don't know statically what a
dynamically typed variable points to. It might not be a function.

> Figure 5 rule D-App: in Figure 3 a function f seems not to be allowed to have
> type Dyn in the enviroment

This is allowed, but only indirectly. The definition of a function `f_0` can't
directly say that `f_0` has type Dyn, but it can declare the type `Dyn -> Dyn`
and then apply `f_0(f_0)` to assign the dynamic type to a function.

(Note that `x` and `f` are undifferentiated variable names. The model uses `x`
 in some definitions and `f` in others as hints to readers, but it's fine to
 name a function `x_2` and so on.)

> Also, Object seems to be the top type, hence also Bool and Int are seen as
> classes, right?

Yes, Bool and Int are technically classes. The paper says little about this,
but the Redex model (and Static Python) give them full support as classes.

> In rules D-Set and CD-Set, the type None seems to be used as a "Unit" type. The
> fact that these two types are identified seems strange.

Yes it is a bit strange, but it helps keep the model small for the paper.
(Really, the best choice would be to move these rules into a typing judgment
for **statements** rather than expressions. That's what the Redex model does.)

> In subtyping rules, which is the relation between two Optional types?

Fixed, thank you for catching this omission.

> Why do you use metavariable Env_0 rather than just Env in the rules?

We've removed the subscript.

> page 15 line 4: I would not say "upcast" since there is no explicit cast here.

Ok. We have removed this explanatory sentence.

> My main criticism about your formalization of the typing rules is that, in my
> opinion, they should include generation of code annotated with casts.
> Moreover,it should be made clear that they are algorithmic. I mean, what I
> expect is that there is an algorithm which, taken an expression expr and an
> environment Env, produces a type (or a typing error) and the corresponding
> annotated code, inserting casts. I am sure that this is what happens in your
> PLT Redex implementation, and I think that this should be reported.

The typing judgment in Redex is algorithmic, but the simple judgment in the
paper is not. Section 4.2.2 in the revised paper makes this clear.

> Section 4.3: I would perhaps eliminate this section. As it is, it says almost
> nothing.

We disagree. The section is a proof sketch that ill-typed values cannot "sneak
in" through the dynamic type.

> Section 6.3 I miss a conclusion/summary of the analysis here.

We've added a conclusion section.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45C
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> This paper presents the design of Static Python, which introduces
> gradual typing for Python.  It is already in use by engineers at
> Instagram.  One of the notable features in the design is that a
> programmer can have some control over the choice of what kind of
> invariant can be enfored by types: untyped (no guarantee), shallow (in
> the sense that types guarantee only the top-level type constructor),
> and concrete (where full type information is relevant).  The use of
> concrete types promotes code optimization, which can be also an
> incentive to programmers.
>
> The paper presents an overview of the language and its run-time
> system, followed by the description of the formal type system.  The
> type system is formalized with PLT Redex.  Although no formal
> metatheory is developed, the model has been checked by Redex's random
> testing tools and test suites.  Subtleties in the language features
> that are not modeled but implemented are discussed, too.  The section
> on evaluation reports that 3.7% imprevement in CPU efficiency.
>
> Assessment and comments (for authors)
> -------------------------------------
> The language design mixing concrete and shallow types is interesting.
> It seems that the language has been accepted by engineers and used for
> a large software project.  Nevertheless I have a few reservations to
> accept the paper in its current form.
>
> The paper puts emphasis on "gradual soundness" at the beginning but,
> after reading the paper, I failed to see what it really was.  The
> phrase is never mentioned on page 3 and afterwards.  This is
> unfortunate because this is the phrase used even in the title!

We've updated the paper to mention "gradual soundness" in the following
sections: 2.3, 2.5, 6.3, and 8.

> It's not clear how faithfully the formal model presented in Section 4
> models the implementation.  In particular, C-Sub allows the type of an
> expression can be changed to Dyn (using the consistent subtyping) and
> Matr allows conversion from Dyn to any non-Dyn type.  So, the type
> system seems to give any expression (that references only declared
> variables, functions, classes) any type.  However, apparently, some
> expressions are rejected by the implementation.  I have no idea what's
> really going on.  (I tried to access the Redex model but it wasn't
> accessible.)

We've added a disclaimer near the top of section 4.2.2 to say that the
paper presents declarative typing rules whereas the Redex model implements a
typing algorithm.

> The distinction between surface types and evaluation types is
> interesting but its implication is not so clear.  Are type arguments
> to Dict completely ignored by the type system?  If so, what is the
> point of writing type arguments in a program?  I may misunderstand
> something and am wondering if my question is related to the paragraph at
> the bottom of p.11, "we omit the surface typing judgment, which is a
> kind of linter that merely scans typed code for logical errors".  What
> are the logical errors mentioned here?  If type arguments to Dict are
> completely ignored, an alternative design might have been like this:
> instead of providing two classes Dict and CheckedDict, a programmer
> could use the raw type Dict and fully parameterized types Dict[S0,S1].

The discussion is intentionally vague because Static Python does not have firm
guidelines for what the surface type checker (linter) should do. For Dict,
it currently ignores the type arguments so the only reasons to write them are
for documentation and for other type checkers (Pyre, mypy) to check them.
In the future, though, Static Python may do more surface checks.

Using the raw type Dict and parameterized type Dict[S0, S1] seems problematic
because these types describe different data structures. The types look similar,
but a value with type Dict[S0, S1] cannot be used where a Dict is expected.

> The related work section is light.  It cites relevant papers and
> describes what they discuss but doesn't really compare with the
> present work in depth.

We have expanded the related work.

> In summary, I'd recommend that the paper be revised by:
> * clarifying gradual soundness,
> * clarifying the relationship between the model and the implementation,
> * clarifying the role of surface types,
> * enhancing the discussion on related work.
> It would be appreciated if you make the code of the Redex model accessible.

We apologize that the Redex model was inaccessible. For convenience, here is
the link:

  <https://github.com/brownplt/insta-model>

> Detailed comments:
>
> Table 1: Why does a constructor call for CheckedDict pay the cost in
> O(N)? (What does N stand for?)

N is the number of keys and values in the table (we have updated the table
caption to explain). The constructor needs to validate each element.

> * It seems that efficient run-time checking depends on the fact that
>   the set of type instantiations CheckedDict[S0, S1] is finite and
>   known at compile-time.  How is it achieved?  (Is it because there
>   are no user-defined generic classes?)  My guess above may be wrong.
>   In that case, it's not very clear how tags are implemented.

Section 2.3 and 3 in the revision say more about the "how." Static Python
implements these tags as pointers into a run-time table of all generic
instantiations that have appeared.

> p.12: "Evaluation types S are a subset of the surface types".
> Is "Dict", which is an evaluation type, a surface type?

No, Dict is not a surface type in the model. We have rephrased "subset" to
avoid this confusion.

### 2022-04-26 pj-7.1/paper.tex
pj: first-draft response + updated paper

@@
 %% -
-
-%% TODO 2022-04-08 from Pyre team
-%% Hi Carl! The small notational question I mentioned about in the paper was in
-%% Figure 5 -- there's this "M-APP" rule that describes how classmethod invocation
-%% is handled. I was trying to figure out how inherited classmethod would work
-%% with this rule. Say I have a class B that defines a method f0(B, S0) -> S1, and
-%% then I have another class D that inherits B which does not define additional
-%% method. Now given Env0 \vdash expr0: S0, how can Env0 \vdash D.f0(expr0): S1 be
-%% proven? If I understand correctly, the rule only looks up the type of f0 in
-%% Env0(D), i.e. the body of the derived class. But the real definition of f0 is
-%% syntactically located in the base class B. Does the notation system used here
-%% assume that every function definition in base class gets automatically copied
-%% into the derived class? Or should there be an additional preconditions added to
-%% the "M-APP" rule saying that f0 is allowed to exist in any of the superclass of
-%% D?
-
-%% 2022-04-24 More model issues:
-%% - C.f(e) uses a class type, not an instance ... how to invoke method on instance?!
-%% ... we should really apologize for a too-simple model
-%% - Object is not clearly in C
-%% - introduce c again as variable name? gee why not expr.method()?
 
@@
 \newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
-\newcommand{\sptenv}{\typefont{Env}}
+\newcommand{\sptenv}{\Gamma} %{\typefont{Env}}
 \newcommand{\sptenvnil}{\cdot}
@@
 \newcommand{\spcnew}[2]{\spapp{#1}{#2}}
-\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
+\newcommand{\spenvapp}[2]{\spapp{#1{}^*}{#2}}
 \newcommand{\spdictref}[2]{#1[#2]}
@@
 %
-These options keep run-time costs low and let programmers decide between
+These options keep run-time costs low and let programmers choose between
 concrete and shallow types.
+This choice is one aspect of gradual soundness in \SP{}.
+
 Often, the shallow types are more attractive because concrete types impose
@@
 a mix of primitive and non-primitive arguments.
+These primitive types are another main aspect of gradual soundness.
 
@@
 we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
-type soundness on thousands of examples (1,600 expressions and 11,000
-programs).
+type soundness.
+Specifically, we asked Redex to generate well-formed terms (using its default
+algorithm) and checked whether the well-typed terms that terminated within a
+fixed time budget (1,200 expressions and 11,000 programs in a typical run)
+produced only well-typed values and allowed errors.
 % \footnote{Property-based testing helped us find and fix several issues in the
@@
 conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
-In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
+We translated \numSPtest{} tests from the \SP{} regression suite to the
 syntax of the model and confirmed that the results
@@
 use because they fall outside the scope of the model.
-\Cref{s:impl} summarizes the out-of-scope features; \cref{a:banned-test} gives
+\Cref{s:impl} summarizes the out-of-scope features and \cref{a:banned-test} gives
 a detailed categorization.
@@
 
-The rest of this section illustrates our Redex model using
-a small formalization.
+%% TODO gaps between paper and redex
+%% (if these are small enough, explain here ... maybe with minor notes further down
+%%  if bigger, consider a "threats" or "gaps" section)
+%% - typing is not algorithmic
+%% - few python features (well that's obvious)
+%% - few expressions (very subtle to introduce here/now)
+
+The rest of this section illustrates our Redex model using a small
+formalization.
+%% TODO EXAMPLE simplification
 \Cref{s:surface} presents a surface syntax to highlight the boundaries between
@@
       \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid
+      \spapp{\spf}{\spexpr} \langmid
   \\ & &
@@
       \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
-      \spcnew{\sptclass}{\spexpr, \ldots} \langmid
   \\ & &
-      \spapp{\spf}{\spexpr} \langmid
       \spdictref{\spexpr}{\spexpr} \langmid
       \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
-      \spobjref{\spc}{\spx} \langmid
-      \spobjset{\spc}{\spx}{\spexpr} \langmid
-      \spobjapp{\spc}{\spf}{\spexpr}
+  \\ & &
+      \spcnew{\sptclass}{\spexpr, \ldots} \langmid
+      \spobjref{\spexpr}{\spx} \langmid
+      \spobjset{\spexpr}{\spx}{\spexpr} \langmid
+      \spobjapp{\spexpr}{\spf}{\spexpr}
     \\[1.5ex]
@@
       \sptstr \langmid
-      \sptobject \langmid
       \sptclass \langmid
@@
     \spx, \spf & \defeq & \mbox{variable names}
-    %% TODO should C be a claass name? ... should c be variable name?
   \end{langarray}
@@
   \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}
+
+  \medskip
+  \mbox{Convention: the set of all class types $\sptclass$ contains the top type $\sptobject$}
 
@@
     $\sptdyn$ & {No check needed, accepts any value} \\
-    $\sptobject$ & {Accepts any value\tnote{*}} \\
-    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$} \\
+    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$\tnote{*}} \\
     $\sptnone$ & {Accepts the Python none value} \\
@@
     \begin{tablenotes}
-    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ rejects primitive values~(\cref{s:c-types})}
+    \item {*} {\footnotesize In \SP{}, the check for $\sptobject$ accepts all values except for primitives~(\cref{s:c-types})}
     \end{tablenotes}
@@
     \inferrule*[left=S-App]{
-      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
+      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv
       \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+      \sptenv \wtexpr \spexpr_0 : \spteval_0
     }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
+      \sptenv \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
     }
@@
     \inferrule*[left=D-App]{
-      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
+      \sptvardef{\spf_0}{\sptdyn} \in \sptenv
       \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \sptobject
+      \sptenv \wtexpr \spexpr_0 : \sptobject
     }{
-      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
+      \sptenv \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
     }
@@
 %    \inferrule*[left=D-Ref]{
-%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+%      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
 %      \\\\
-%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+%      \sptenv \wtexpr \spexpr_1 : \sptobject
 %    }{
-%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
+%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
 %    }
@@
 %    \inferrule*[left=CD-Ref]{
-%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+%      \sptenv \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
 %      \\\\
-%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+%      \sptenv \wtexpr \spexpr_1 : \spteval_0
 %    }{
-%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
+%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
 %    }
@@
     \inferrule*[lab=D-Set]{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
+      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
+      \sptenv \wtexpr \spexpr_1 : \sptobject
       \\\\
-      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
+      \sptenv \wtexpr \spexpr_2 : \sptobject
     }{
-      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
+      \sptenv \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
     }
@@
     \inferrule*[lab=CD-Set]{
-      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
+      \sptenv \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
       \\\\
-      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
+      \sptenv \wtexpr \spexpr_1 : \spteval_0
       \\\\
-      \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
+      \sptenv \wtexpr \spexpr_2 : \spteval_1
     }{
-      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
+      \sptenv \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
     }
@@
     %\inferrule*[left=F-Ref]{
-    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
+    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv}{\spc_0}
     %}{
-    %  \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
+    %  \sptenv \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
     %}
@@
     \inferrule*[left=F-Set]{
-      \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
+      \sptenv \wtexpr \spexpr_0 : \spc_0
       \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+      \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv}{\spc_0}
+      \\\\
+      \sptenv \wtexpr \spexpr_1 : \spteval_0
     }{
-      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
+      \sptenv \wtexpr \spobjset{\spexpr_0}{\spx_0}{\spexpr_1} : \sptnone
     }
@@
     \inferrule*[left=M-App]{
-      \sptfundef{\spf_0}{\spc_0, \spteval_0}{\spteval_1} \in \spenvapp{\sptenv_0}{\spc_0}
+      \sptenv \wtexpr \spexpr_0 : \spc_0
       \\\\
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+      \sptfundef{\spf_0}{\spc_0, \spteval_0}{\spteval_1} \in \spenvapp{\sptenv}{\spc_0}
+      \\\\
+      \sptenv \wtexpr \spexpr_1 : \spteval_0
     }{
-      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \spteval_1
+      \sptenv \wtexpr \spobjapp{\spexpr_0}{\spf_0}{\spexpr_1} : \spteval_1
     }
@@
     \inferrule*[left=C-Sub]{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \sptenv \wtexpr \spexpr_0 : \spteval_1
       \\\\
-      \sptenv_0 \wtsub \spteval_1 \spcompat \spteval_0
+      \sptenv \wtsub \spteval_1 \spcompat \spteval_0
     }{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+      \sptenv \wtexpr \spexpr_0 : \spteval_0
     }
@@
     \inferrule*[left=Matr]{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
+      \sptenv \wtexpr \spexpr_0 : \spteval_1
       \\\\
@@
     }{
-      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
+      \sptenv \wtexpr \spexpr_0 : \spteval_0
     }
 
+  \mbox{Where $\spenvapp{\sptenv}{\spc}$ collects field and method types from $\spc$ and from its parents}
+
   \end{mathpar}
 
+  \bigskip
   \begin{minipage}[t]{0.6\columnwidth}
@@
       \inferrule*{
-        \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
+        \sptenv \wtsub \spteval_0 \spsubt \spteval_1
       }{
-        \sptenv_0 \wtsub \spteval_0 \spcompat \spteval_1
+        \sptenv \wtsub \spteval_0 \spcompat \spteval_1
       }
@@
       }{
-        \sptenv_0 \wtsub \spteval_0 \spcompat \sptdyn
+        \sptenv \wtsub \spteval_0 \spcompat \sptdyn
       }
@@
     }{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \sptobject
+      \sptenv \wtsub \spteval_0 \spsubt \sptobject
     }
@@
     }{
-      \sptenv_0 \wtsub \sptbool \spsubt \sptint
+      \sptenv \wtsub \sptbool \spsubt \sptint
     }
@@
     \inferrule*{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
+      \sptenv \wtsub \spteval_0 \spsubt \spteval_1
     }{
-      \sptenv_0 \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
+      \sptenv \wtsub \sptoptional{\spteval_0} \spsubt \sptoptional{\spteval_1}
     }
@@
     \inferrule*{
-      \sptclassdef{\sptclass_0}{\sptclass_1}{\ldots}{\ldots} \in \sptenv_0
+      \sptenv \wtsub \spteval_0 \spsubt \spteval_1
     }{
-      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
+      \sptenv \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
+    }
+
+    \inferrule*{
+      \sptclassdef{\sptclass_0}{\sptclass_1}{\ldots}{\ldots} \in \sptenv
+    }{
+      \sptenv \wtsub \sptclass_0 \spsubt \sptclass_1
     }
@@
 
+By contrast to our Redex model, the typing judgment in~\cref{f:eval-type} is
+not syntax directed.
+The rules do not describe a typechecking algorithm.
+Instead these rules declare various conclusions and their preconditions to
+illustrate how \SP{} reasons about types---dynamic or otherwise.
+
 The first two rules are for function application.
@@
 $\spteval_1$ is the dynamic type.
-When this relation holds, it is safe to upcast a value from type $\spteval_0$
-to type $\spteval_1$.
 Materialization relates the dynamic type to any non-dynamic type.
@@
       \inferrule*[left=Matr]{
-        \sptenv_0 \wtexpr \spx : \sptdyn
+        \sptenv \wtexpr \spx : \sptdyn
         \\
@@
       }{
-        \sptenv_0 \wtexpr \spx : \sptint
+        \sptenv \wtexpr \spx : \sptint
       }
     }{
-      \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptdyn
+      \sptenv \wtexpr \spapp{\spf}{\spx} : \sptdyn
     }
@@
   }{
-    \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptstr
+    \sptenv \wtexpr \spapp{\spf}{\spx} : \sptstr
   }
 
-  \mbox{Where $\sptenv_0 = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
+  \mbox{Where $\sptenv = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
 \end{mathpar}
@@
 migrations, the question arises as to whether the changes themselves
-are responsible for more speedups than the types.
+are responsible for more speedups than the types and gradual soundness.
 %% Carl: "I reviewed the ~30 conversion diffs"
-We therefore reviewed the $\sim 30$ conversion diffs that significantly improved
-performance to assess the extent of the code changes.
-In general, the patches have minor changes that
-affect tests more than production code.
-There are nine common kinds of changes: five for types and four for speed.
+We therefore reviewed conversion diffs that significantly improved
+performance (\nnum{30}) to assess the extent of the code changes.
+There are nine common kinds of changes: five that satisfy the type checker
+and four that aim to improve performance.
 
@@
 
+\subsubsection{Conclusion}
+
+In general, the patches responsible for performance gains contain minor
+code changes and 
+affect test code rather than production code.
+We conclude that the addition of \SP{} types to a codebase can significantly
+improve performance without major refactoring.
+
 %% More Migration Notes
@@
 % 2022-01-24 bg: hidden from submission to save space
-%\subsubsection{Developer Experience}
+%\subsection{Developer Experience}
 %
@@
 Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
-and StrongScript~\cite{rzv-ecoop-2015},
-and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
-%% It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
-%% in the original Nom paper inspired the \SP{} team to pursue an implementation.
-In addition to concrete types, Thorn and StrongScript have optional (unsound) \emph{like} types.
-The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
-types.
+and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
+and untyped objects.
+Thorn is a new scripting language for the JVM; StrongScript is designed to
+interoperate with JavaScript.
+Concrete types came to \SP{} by way of Nom~\cite{mt-oopsla-2017} and its
+successor MonNom~\cite{mt-oopsla-2021}.
+The impressive performance figures in the original Nom paper inspired the \SP{}
+team to adapt the method.
+But whereas MonNom has been exploring designs for generic concrete types that
+support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
+\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).
+
+With its mix of shallow and concrete types, \SP{} is one of a few gradual
+language that lets programmers choose between multiple type-enforcement strategies.
+Thorn and StrongScript offer a choice of concrete types and optional (unsound)
+\emph{like} types.
+Typed Racket supports deep (structural) types, shallow types, and optional
+types~\cite{g-pldi-2022}.
+By contrast to these others, \SP{} asks programmers to change their code
+in way that reflects changes to the types~(\cref{s:checked-type}).
+The changes impose a maintenance burden, but enable performance improvements.
 
@@
 eliminate the costs of shallow types.
-This finding complements prior work on shallow types;
-e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
-and in Grace with the Graal partial-evaluation-based JIT~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
-Similarly, the Pycket team successfully used a tracing JIT to reduce the costs of higher-order
-checks~\cite{bbst-oopsla-2017}.
-%% It remains to be seen whether a non-tracing JIT is effective for higher-order
-%% casts.
+This finding complements prior work.
+In the context of shallow types,
+the pypy tracing JIT improved Reticulated~\cite{vsc-dls-2019}
+and the Graal partial-evaluation-based JIT improved Grace~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
+In the context of concrete types, the HiggsCheck VM improved an implementation of
+SafeTypeScript~\cite{rsfbv-popl-2015,rat-oopsla-2017}.
+And in the context of deep/guarded/structural types, the Pycket tracing JIT greatly reduced the costs of higher-order
+checks in Typed Racket~\cite{bbst-oopsla-2017}.
+% In sum, there is much evidence to suggest that gradual languages go well with
+% JIT technology.
 
@@
 mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}.
+These checkers use types only for static analysis; optional types have no
+effect (positive or negative) at run-time.
 Another sound type system is Reticulated, which enforces structural types and
 pioneered the transient semantics~\cite{vss-popl-2017,v-thesis-2019}.
-Optimizing type systems include Reticulated and mypyc~\cite{mypyc}.
-%% The latter compiles Python source to C extension modules; unlike \SP{},
-%% it does not offer a pure-Python developer experience.
+A rigorous comparison of Reticulated and \SP{} is an important topic for future work.
+We conjecture that adapting code to Reticulated requires fewer edits than \SP{}
+but delivers slower performance.
+Optimizing type systems for Python include Reticulated and mypyc~\cite{mypyc}.
+The latter compiles Python source to C extension modules;
+it does not offer a pure-Python developer experience.
 
@@
 using a catch-all \code{external} type~\cite{jspp}.
+None of these four implement gradual types as envisioned by researchers~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006};
+instead, they offer a pragmatic solution to interoperability problems.
 %% Programmers can manually cast external values to a more precise type.
@@
 %% also, how does JS++ protect arrays sent to JS?
-
-%% Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
-%% of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
-%% Primitive types are progressive types for base values.
-%% Concrete types are essentially a progressive form of shallow types.
 
@@
 
-In conclusion, the \SP{} approach to gradual typing seems to be a promising
+In conclusion, the gradual soundness approach of \SP{} seems to be a promising
 way to realize the vision of gradual typing.
@@
 
-  Special thanks to Twitter, thanks to which the two teams
+  Special thanks to Twitter, through which the two teams
   (specifically, Meyer and Krishnamurthi) met.
-  This work was partly supported by the US National Science Foundation.
-  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-  TODO partial funded by Meta
+  This work was partly supported by Meta and by the US National Science Foundation.
+  Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
 }

### 2022-04-26 pj-7.1/response.txt
pj: response, wrap to 80 cols

@@
 > The paper makes a contribution to gradual typing in theory and practice. The
-> subject language is Static Python, which adds a compile-time type discipline to
-> Python.
+> subject language is Static Python, which adds a compile-time type discipline
+> to Python.
 >
@@
 >
-> On the practical side, the paper describes the design of Static Python as well
-> as experiments with an implementation of Static Python. This implementation
-> allegedly improves the throughput of a closed source implementation of a web
-> server by 3.6%. It improves the run time of several microbenchmarks
-> significantly.
+> On the practical side, the paper describes the design of Static Python as
+> well as experiments with an implementation of Static Python. This
+> implementation allegedly improves the throughput of a closed source
+> implementation of a web server by 3.6%. It improves the run time of several
+> microbenchmarks significantly.
 >
@@
 >
-> I like the paper, it is generally well readable. There are a few places, where
-> I'd expect more guidance from the authors, rather than expecting the reader to
-> interpolate.
+> I like the paper, it is generally well readable. There are a few places,
+> where I'd expect more guidance from the authors, rather than expecting the
+> reader to interpolate.
 >
@@
 > shallow/concrete dictionary types in typed and untyped contexts.  I was
-> initially confused by the 0 entries. I'd prefer you put --- to indicate that no
-> casts is needed.  You should indicate that $N$ is the length of the dictionary.
+> initially confused by the 0 entries. I'd prefer you put --- to indicate that
+> no casts is needed.  You should indicate that $N$ is the length of the
+> dictionary.
 
@@
 >
-> That seems a bit shaky. What's the baseline / ground truth / oracle for these tests?
+> That seems a bit shaky. What's the baseline / ground truth / oracle for these
+> tests?
 > In the end, all that testing only gives you a probabilistic guarantee.
@@
 > Line 6: `C.f(expr)`
-> * looks like a static method call to me, is that right? That's usually trivial
->   compared with the dynamic dispatch of a real method call.
+> * looks like a static method call to me, is that right? That's usually
+>   trivial compared with the dynamic dispatch of a real method call.
 > * how do you invoke a method on an object?
@@
 >
-> * rule `M-App` seems too limiting. Literally read, it's not possible to invoke
->   inherited methods because the method `f0` must be defined in class `C0` in
->   which it is defined. This rule affirms the class-method conjecture from
->   above.
+> * rule `M-App` seems too limiting. Literally read, it's not possible to
+>   invoke inherited methods because the method `f0` must be defined in class
+>   `C0` in which it is defined. This rule affirms the class-method conjecture
+>   from above.
 
@@
 
-> * the `object` rule for subtyping contradicts Table 2, which says that *objects
->   accepts any value expect primitives*. So `S0` must not be a primitive; but
->   primitives are not part of the model. This is not consistent!
+> * the `object` rule for subtyping contradicts Table 2, which says that
+>   *objects accepts any value expect primitives*. So `S0` must not be a
+>   primitive; but primitives are not part of the model. This is not
+>   consistent!
 
@@
 >   improving performance. The project has already shown its effectiveness. The
->   authors (informally) describe both the language (the gradual type system) and
->   how this is implemented to achieve better performance.
+>   authors (informally) describe both the language (the gradual type system)
+>   and how this is implemented to achieve better performance.
 >
@@
 
-> Section 4.1 line 4: In the formal syntax in Figure 3, Object seems not to be an
-> instance of C, so class C(C) seems not to include the case C = Object.
+> Section 4.1 line 4: In the formal syntax in Figure 3, Object seems not to be
+> an instance of C, so class C(C) seems not to include the case C = Object.
 > In the "Other expressions" you mention object fields reads and writes, but in
@@
 
-> Figure 3: it would be better to use a more abstract style with metavariables as
-> it is customary now, e.g.
+> Figure 3: it would be better to use a more abstract style with metavariables
+> as it is customary now, e.g.
 > prog = stmt_1, ..., stmt_n
@@
 
-> In rules D-Set and CD-Set, the type None seems to be used as a "Unit" type. The
-> fact that these two types are identified seems strange.
+> In rules D-Set and CD-Set, the type None seems to be used as a "Unit" type.
+> The fact that these two types are identified seems strange.
 

### 2022-04-26 pj-7.1/paper.tex
edit cleanup

@@
 type soundness.
-Specifically, we asked Redex to generate well-formed terms (using its default
+Specifically, we used Redex to generate well-formed terms (using its default
 algorithm) and checked whether the well-typed terms that terminated within a
@@
 
-By contrast to our Redex model, the typing judgment in~\cref{f:eval-type} is
+By contrast to our Redex model, the typing judgment in~\cref{f:eval-types} is
 not syntax directed.
@@
 Almost all ratios are less than one, showing that \SP{} achieves a speedup over
-Python (typed avg. $0.8$, refined avg. $0.3$).
+Python (typed avg.~$0.8$, refined avg.~$0.3$).
 The exceptions are the unrefined versions of \bmname{fannkuch} and
@@
 and untyped objects.
-Thorn is a new scripting language for the JVM; StrongScript is designed to
+Thorn is a scripting language for the JVM; StrongScript is designed to
 interoperate with JavaScript.

### 2022-04-26 pj-7.1/response.txt
pj: edit pass paper + response

@@
-TODO
-- write a quick thank-you and summary of changes
-- add section + page numbers to the responses
+Thank you all for the detailed reviews. We have made the major changes that
+we promised in our edit plan and followed many of your other suggestions.
+
+For reference, here are the major changes:
+
+ 1. (Section 4) Clearly explain the purpose of the paper's model and the Redex
+    model, and explain how they relate to one another and to Static Python. In
+    particular, address the following specific points from the reviews:
+    - (#45A) Explain how Redex's random testing helps to check type soundness.
+    - (#45B) Explain "soundness" and "property-based soundness tests" more
+      clearly.
+    - (#45B, #45C) Clearly say that the model's typing is not algorithmic and
+      that Redex's typing is algorithmic. Explain how and why the model is
+      different.
+    - (#45A, #45B) Explain the semantics of objects and method calls in more
+      detail.
+
+  2. Explain gradual soundness in more detail throughout the paper.
+
+  3. (Section 7) Extend the related work and be sure to compare this work to
+     prior work.
+
+Below, please find point-by-point replies with page and section numbers to
+show where the revised paper has changed.
 
@@
 >
-> Paper summary (for authors)
-> ---------------------------
-> The paper makes a contribution to gradual typing in theory and practice. The
-> subject language is Static Python, which adds a compile-time type discipline
-> to Python.
->
-> On the theory side, the paper describes a formal model created with an
-> explorative tool for semantics modeling and prototyping. This model includes a
-> static semantics and a dynamic semantics. The dynamic semantics is tested
-> against a regression test suite.
->
-> On the practical side, the paper describes the design of Static Python as
-> well as experiments with an implementation of Static Python. This
-> implementation allegedly improves the throughput of a closed source
-> implementation of a web server by 3.6%. It improves the run time of several
-> microbenchmarks significantly.
->
-> Assessment and comments (for authors)
-> -------------------------------------
-> ## strengths
->
-> * The main strength of the paper is clearly the favorable evaluation of a
->   particular datapoint in the design space of gradual typing.
-> * The formal model enabled the discovery of problems in a industrial strength
->   implementation of Static Python, at least in its type checker.
->
-> ## weaknesses
->
-> * the most significant evaluation is on a closed source program
-> * the microbenchmarks potentially gain more efficiency from progressive
->   primitive types as most of them are computationally intensive.
-> * the theoretical evaluation is only probabilistic with unclear probability
+> [... clipped ...]
 >
@@
 These are property-based tests. Our model defines a soundness property for terms
-and asks Redex to generate random terms in the hope of finding a counterexample.
+and uses Redex to generate random terms in the hope of finding a counterexample.
 So, the soundness property is the ground truth; these random tests are all
-checking soundness.
+checking soundness. The basic outline is: if a random term is well-typed and
+terminates within a finite number of steps, then it must either reduce to a
+well-typed value or an error. (To get thousands of well-typed and terminating
+terms, we have Redex generate tens of thousands of random terms.)
 
@@
 
-Property-based testing is shaky compared to formal verification, but it's still
-a helpful tool for finding bugs.  Indeed, we found several bugs in our model
-thanks to these property-based tests (none of which had implications for Static
-Python itself).
+In the end, although property-based testing is not as convincing as formal
+verification, it is still a helpful tool for finding bugs. Indeed, we found
+several bugs in our model thanks to these property-based tests (none of which
+had implications for Static Python itself).
 
@@
 >
-> Paper summary (for authors)
-> ---------------------------
-> The contribution of the paper is the following:
->
-> - The paper describes Static Python, a project developed at Instagram to add
->   types to Python code in a gradual manner, with the main objective of
->   improving performance. The project has already shown its effectiveness. The
->   authors (informally) describe both the language (the gradual type system)
->   and how this is implemented to achieve better performance.
->
-> - The authors have developed an implementation in PLT redex of a subset of
->   Static Python, focusing on features related to gradual typing. Using this
->   implementation, they did an extended testing to show:
->   1) that this implementation is sound
->   2) that this implementation gives the same results of Static Python.
->
-> Assessment and comments (for authors)
-> -------------------------------------
-> Assessment: The contributions of the paper summarized above are certainly
-> useful for the PL community, so I support acceptance. The empirical evaluation
-> seems to me (but I am definitely not an expert in such kind of contributions)
-> well-driven and adequate. I have some reservations on the formalization,
-> explained in the comments below, which should be solved/improved in the final
-> version.
+> [... clipped ...]
 >
@@
 >
-> Paper summary (for authors)
-> ---------------------------
-> This paper presents the design of Static Python, which introduces
-> gradual typing for Python.  It is already in use by engineers at
-> Instagram.  One of the notable features in the design is that a
-> programmer can have some control over the choice of what kind of
-> invariant can be enfored by types: untyped (no guarantee), shallow (in
-> the sense that types guarantee only the top-level type constructor),
-> and concrete (where full type information is relevant).  The use of
-> concrete types promotes code optimization, which can be also an
-> incentive to programmers.
->
-> The paper presents an overview of the language and its run-time
-> system, followed by the description of the formal type system.  The
-> type system is formalized with PLT Redex.  Although no formal
-> metatheory is developed, the model has been checked by Redex's random
-> testing tools and test suites.  Subtleties in the language features
-> that are not modeled but implemented are discussed, too.  The section
-> on evaluation reports that 3.7% imprevement in CPU efficiency.
+> [... clipped ...]
 >

### 2022-04-26 pj-7.1/paper.tex
pj: edit pass paper + response

@@
 %% TODO 2022-01-25 recount tests
-\newcommand{\numbenchmark}{3}
+\newcommand{\numbenchmark}{4}
 \newcommand{\numSPtest}{265}
@@
     A third set of progressive types enable further optimizations.
-    In particular, \emph{primitive types\/} allow for unboxed arithmetic,
-    which has been critical in practice.
+    In particular, \emph{primitive types\/} allow for unboxed arithmetic.
 \end{enumerate}
@@
     To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
-    and ran both property-based soundness tests and over 200 tests adapted from the \SP{} regression suite.
+    and ran both property-based soundness tests and over 250 tests adapted from the \SP{} regression suite.
     The modeling effort revealed 20 significant issues in \SP{},
@@
 \Cref{s:model} uses a small formal model to introduce our Redex mechanization
-and to convince readers that \SP{} is based on a sound core.
+and to demonstrate that \SP{} is based on a sound core.
 \Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
@@
 The advanced features of the \SP{} type system are offered on an opt-in basis
+(\emph{gradual soundness})
 and arranged so that programmers can begin adding types one module at a time.
@@
     including expressions that have the dynamic type~(\cref{s:type-dynamic}).
-    If this behavior is a problem for certain expressions, it is easy enough to move
-    them to an untyped Python module.
+    If this behavior is problematic for certain expressions, they must move
+    to an untyped Python module.
 \end{itemize}
@@
 With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
-Each typed class can furthermore assume that if a method has a precise (non-dynamic)
+Typed classes can assume that if a method has a precise (non-dynamic)
 type, then all typed overrides are subtypes of this type.
-\SP{} can therefore optimize dispatch from typed callers to typed methods
-and use checked dispatch for other combinations.
-\SP{} can also optimize reads from typed fields.
-One extra step arises when an untyped method overrides a typed method.
+This property enables optimized dispatch from typed callers to typed methods
+and optimized field reads.
+Other method and field accesses typically incur one run-time check.
+An extra step arises when an untyped method overrides a typed method.
 In this case, \SP{} creates a wrapper around the overriding method to check
@@
 a mix of primitive and non-primitive arguments.
-These primitive types are another main aspect of gradual soundness.
+These primitive types are another central aspect of gradual soundness.
 
@@
 sound types and strong optimizations:
+module-level variables cannot be mutated;
 class attributes must be set in an \code{\_\_init\_\_} method to
@@
 descriptors or properties;
-module-level variables cannot be mutated;
 typed fields 
@@
 As an example of the third kind, the \code{FAST\_LEN} instruction quickly
-computes the length of a built-in value, which helps to optimize for loops.
+computes the length of a built-in value, which helps to optimize loops.
 \SP{} uses type information to decide where this instruction is appropriate
@@
 implementation.
-The implementation provides the same interface as the built-in but uses a
-type tag to reject certain operations.
+The implementation provides the same interface as the built-in, but uses
+type tags to reject certain operations.
 
@@
 Any method can be JIT-compiled whether or not it uses \SP{} types,
-even though fewer types usually leads to fewer optimizations.
+though fewer types usually result in fewer optimizations.
 
@@
 Instagram.
-Overall, we made 25 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
-Four of these were soundness issues, one of which we could exploit to raise
+Overall, we made 26 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
+Five of these were soundness issues, one of which we could exploit to raise
 a segmentation fault.
-All but one of these soundness bugs have been fixed.
+All of these soundness bugs have been fixed.
 Of the remaining issues, five were relatively minor; these dealt with confusing error messages
@@
 %% critical soundness issues:
+%% https://github.com/facebookincubator/cinder/issues/62
 %% https://github.com/facebookincubator/cinder/issues/55
@@
 
-The rest of this section illustrates our Redex model using a small
-formalization.
-%% TODO EXAMPLE simplification
-\Cref{s:surface} presents a surface syntax to highlight the boundaries between
-typed and untyped code.
+The rest of this section illustrates our Redex model using a tiny
+formalization to highlight the boundaries between typed and dynamic-typed
+code.
+\Cref{s:surface} presents a surface syntax.
 \Cref{s:eval-types} explains how types get enforced at run-time.
@@
 The basic values are the none value, integers,
-booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
+booleans (which are the integers \code{0} and \code{1}), and strings.
 There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
-Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
-and method calls.
+Other expressions describe function calls, dictionary reads and writes,
+object creation, object field reads and writes, and method calls.
 
@@
 and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
-\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
+\Cref{f:surface-types} does not follow these standards.
+Instead, it keeps the two syntaxes distinct.
 Expressions use lowercase names (\code{none}, \code{chkdict}) and types use capitalized
@@
 
-Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
+Surface types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
 user-defined class, and union types.
@@
 \SP{} does not have a surface syntax for its dynamic
-type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
+type and allows missing and unknown types~(\cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
 The sole exception is optional types, which require a tag check and a test for the none value.
-Even checked dictionaries rely on tag checks.
-A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
+Checked dictionaries also rely on tag checks;
+a checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
 checked dict values that were initialized with exactly the same key and value type.
@@
 the behavior of complex expressions---and this job is
-better left to the Redex mechanization, which at any rate supports a much
+better left to the Redex mechanization, which supports a much
 larger subset of Python~(\cref{s:surface}).
@@
 $\sptoptional{\sptdyn}$ normalizes to the dynamic type.
-Therefore casts at static occurrences of the materialization rule are
+Therefore, inserting casts at static occurrences of the materialization rule is
 enough to protect all boundaries.
@@
 
-The model intentionally does not cover all of Python.
+The Redex model intentionally does not cover all of Python.
 Some aspects of \SP{} are left out because they are straightforward to
@@
 \SP{} can generate efficient code because it targets
-the Cinder runtime, which has bytecode instructions to check types at run-time, to
-construct \SP{} data structures, and to perform type-directed actions.
-\Cref{t:bytecode} lists a few representative bytecode instructions.
+the Cinder runtime.
+Cinder provides bytecode instructions that check types at run-time,
+construct \SP{} data structures, and perform type-directed actions.
+\Cref{t:bytecode} lists a few representative instructions.
 The two instructions that express run-time checks are \bcinst{CAST}
@@
 Only 9 modules use concrete types and only 12 modules use primitive types,
-but these features delivered critical performance in key modules.
+but these features delivered critical performance gains in key modules.
 
@@
 few compiled modules spread across an interpreted codebase.
-Replacing these modules with \SP{} lets programmers return to a conventional
+Replacing these modules with \SP{} let programmers return to a conventional
 Python workflow.
@@
 They also modified a code-generating module to output \SP{} code;
-this one change brought over four hundred generated modules into the typed fold.
+this one change resulted in over four hundred generated typed modules.
 Overall, the second-half changes added 2\% to the number of requests
@@
     Well-typed Pyre code is not always well-typed \SP{} code.
-    \Cref{a:pyre-limits} lists the main friction points.
+    \Cref{a:pyre-limits} lists the main points of friction.
 
@@
 
-\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
+\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks~\cite{spbench}.
 These microbenchmarks are admittedly small and quite different than typical application code,
@@
 
-The 25 GitHub issues that arose from our work may be found at the following three links:
+The 26 GitHub issues that arose from our work may be found at the following three links.
+A list of issue numbers follows each link.
 
@@
   \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
-  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
+    \subitem \code{\#37, 39, 40, 41, 42, 43, 45, 46, 47, 49, 50, 51, 52, 53, 59, 60, 61, 62, 63, 65}
+  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=5)
+    \subitem \code{\#35, 36, 54, 55, 71}
   \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
+    \subitem \code{\#64}
   %% extra, can't remember if we inspired Dino or not: https://github.com/facebookincubator/cinder/issues/32
 \end{itemize}
+
+\noindent{
+  \newcommand{\spsoundness}[1]{\href{https://github.com/facebookincubator/cinder/issues/#1}{\code{#1}}}
+The critical soundness issues were:
+\code{\#}\spsoundness{36},
+         \spsoundness{39},
+         \spsoundness{53},
+         \spsoundness{55}, and
+         \spsoundness{62}.
+}
 

### 2022-04-27 pj-7.1/paper.tex
pj: table 1, remove dict constructor
as Carl says, constructor costs are implied by write costs --- they're free in
typed code, but the constructor column doesn't bring that across

@@
 shallow and concrete dict types.
-Shallow types pay a constant-time cost whenever an untyped value enters typed code
+Shallow types pay a constant-time cost for casts
 and whenever a typed context reads from a shallow value.
-Concrete types pay a potentially-high cost for constructors,
-a constant-time cost for casts and for writes in untyped code,
-and zero cost for reads.
-If a value travels to untyped codes and receives many writes but few reads,
-then the shallow types may run faster in the end.
+Concrete types pay nothing for reads, but incur constant-time costs
+for casts and for writes from untyped code.
+If untyped code frequently writes to a value
+(or equivalently, if untyped code initializes a huge value),
+then shallow types may run faster.
 
@@
   \centering
-  \(\begin{array}{l@{}rr@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
-                   & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
-    \sptrawpydict         &   \nocost{} &        O(1) &          O(1) &      \nocost{} &     \nocost{} &      \nocost{} \\
-    \sptchkdict{K}{V}     &        O(N) &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1)
+  \(\begin{array}{l@{\qquad}r@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
+                                        & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
+    \sptrawpydict                       &        O(1) &          O(1) &      \nocost{} &     \nocost{} &      \nocost{} \\
+    \sptchkdict{K}{V}                   &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1)
   \end{array}\)

### 2022-04-29 pj-7.1/response.txt
pj: kc + r3 fixes
- use Dyn instead of Object in rules
- explain the "magic" with C-Sub and Matr
- add None <: Opt[T]
- move comment about booleans 0 and 1

@@
 
-We've added a disclaimer near the top of section 4.2.2 to say that the
-paper presents declarative typing rules whereas the Redex model implements a
-typing algorithm.
+You're absolutely right, mixing C-Sub and Matr is bad because it lets any
+term have any type. (Run-time checks would catch the bad casts, but still,
+we want to detect errors statically.)
+
+We've revised section 4.2.2 to explain how the typing rules differ from
+the Redex model's algorithm and in particular how the model uses subtyping
+and materialization.
+
+TODO
 

### 2022-04-29 pj-7.1/paper.tex
pj: kc + r3 fixes
- use Dyn instead of Object in rules
- explain the "magic" with C-Sub and Matr
- add None <: Opt[T]
- move comment about booleans 0 and 1

@@
 %% TODO after submission, before camera-ready
+%% - double-check section 4, can we be more convincing about soundness / what it is?
 %% - reword 'fields' to attribute/descriptor/property?
 %% -
-%% - 
+
+%% TODO for camera-ready
 %% -
@@
     $\sptnone$ & {Accepts the Python none value} \\
-    $\sptint$ & {Accepts integers} \\
-    $\sptbool$ & {Accepts booleans (the integers \code{0} and \code{1})} \\
+    $\sptint$ & {Accepts integers and booleans} \\
+    $\sptbool$ & {Accepts booleans} \\
     $\sptstr$ & {Accepts strings}
@@
       \\\\
-      \sptenv \wtexpr \spexpr_0 : \sptobject
+      \sptenv \wtexpr \spexpr_0 : \sptdyn
     }{
@@
 %      \\\\
-%      \sptenv \wtexpr \spexpr_1 : \sptobject
+%      \sptenv \wtexpr \spexpr_1 : \sptdyn
 %    }{
@@
       \\\\
-      \sptenv \wtexpr \spexpr_1 : \sptobject
+      \sptenv \wtexpr \spexpr_1 : \sptdyn
       \\\\
-      \sptenv \wtexpr \spexpr_2 : \sptobject
+      \sptenv \wtexpr \spexpr_2 : \sptdyn
     }{
@@
 
-    \inferrule*[left=C-Sub]{
+    \inferrule*[left=C-Sub*]{
       \sptenv \wtexpr \spexpr_0 : \spteval_1
@@
 
-    \inferrule*[left=Matr]{
+    \inferrule*[left=Matr*]{
       \sptenv \wtexpr \spexpr_0 : \spteval_1
@@
       \sptenv \wtsub \sptoptional{\spteval_0} \spsubt \sptoptional{\spteval_1}
+    }
+
+    \inferrule*{
+    }{
+      \sptenv \wtsub \sptnone \spsubt \sptoptional{\spteval_0}
     }
@@
 
-By contrast to our Redex model, the typing judgment in~\cref{f:eval-types} is
-not syntax directed.
-The rules do not describe a typechecking algorithm.
-Instead these rules declare various conclusions and their preconditions to
-illustrate how \SP{} reasons about types---dynamic or otherwise.
+By contrast to our Redex model, the typing judgment outlined in~\cref{f:eval-types}
+\emph{does not describe} a typechecking algorithm.
+It is rather a declarative set of rules chosen to illustrate two points:
+how \SP{} reasons about data structure types and how it enforces promises made
+via the dynamic type.
+The key steps that the Redex model takes to turn these rules into an algorithm
+is to remove the two coercion rules
+(both are marked with asterisks: \trule{C-Sub*} and \trule{Matr*})
+and to allow restricted coercions at elimination forms.
+For example, a function application may apply either consistent subtyping
+or materialization to its argument.
+It cannot apply both kinds of coecion because that can be abused to delay
+any static error until run-time.
 
@@
 in which case it computes a value that matches its codomain type.
-A dynamically-typed variable may be applied to any input (that matches the top
-type $\sptobject$) to yield a dynamically-typed result.
-The next rules illustrate writes to Python dictionaries and checked dictionaries.
+A dynamically-typed variable may be applied to any input
+to yield a dynamically-typed result.
+The following rules illustrate writes to Python dictionaries and checked dictionaries.
 A shallow Python dict may be updated with any kind of key and value.
@@
 
-The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
+The final two rules, \trule{C-Sub*} and \trule{Matr*}, depend on auxiliary judgments
 for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
@@
 Materialization relates the dynamic type to any non-dynamic type.
-Occurrences of the \trule{Matr} rule are downcasts that require a
+Occurrences of the \trule{Matr*} rule are downcasts that require a
 run-time check.
 
-The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
+The name \emph{materialization} and the rule \trule{Matr*} are inspired by prior
 work~\cite{clps-popl-2019}.
@@
 \begin{mathpar}
-  \inferrule*[left=Matr]{
+  \inferrule*[left=Matr*]{
     \inferrule*{
-      \inferrule*[left=Matr]{
+      \inferrule*[left=Matr*]{
         \sptenv \wtexpr \spx : \sptdyn
@@
 \medskip
-Consequently, this derivation calls for two casts at the \trule{Matr} applications.
+Consequently, this derivation calls for two casts at the \trule{Matr*} applications.
 
@@
 \end{itemize}
-Second, the type-directed optimizer skips casts wherever a well-typed context
-interacts with a typed value.
+Second, the type-directed optimizer skips these casts wherever
+a well-typed context interacts with a typed value.
 \Cref{s:impl} explains the optimizer in more detail.
@@
 Dart 2 is a nominally-typed language with a dynamic type that is similar
-to a top (object) type but statically allows all method calls~\cite{dart-types}.
+to a top type but statically allows all method calls~\cite{dart-types}.
 C\# has a dynamic type that delays type checking until runtime~\cite{bmt-ecoop-2010}.

### 2022-04-29 pj-7.1/response.txt
pj: page and section numbers for review

@@
 Thanks, we've updated the caption, explained N, and used --- for no cast.
+[p.7 s.2.3]
 
@@
 We've added an explanation of nested types (they do run in constant time).
+[p.7 s.2.3]
 
@@
 Done.
+[p.8 s.2.4]
 
@@
 gradual typing.
+[p.8 s.2.5]
 
@@
 It is always a fixed-size pointer.
+[p.10 s.3]
 
@@
 in which this sentence appears is about one direction in particular.
+[p.10 s.3]
 
@@
 
+We have added more details about the testing.
+[p.10 s.4]
+
 > ### p10, -2
@@
 All soundness bugs have been fixed at this time.
+[p.11 s.4]
 
@@
 The revised paper explains why we use the name `Dyn`.
+[p.12 s.4.1]
 
@@
 Fixed `str` and removed `object`.
+[p.12 s.4.1]
 
@@
 
-Oh dear. This was not meant to be a static method call, but we were far too
-aggressive when simplifying the model.
+Oh dear. This was not meant to be a static method call. We were far too
+aggressive when simplifying from the Redex model.
 
@@
 receiver to type check the call.
+[p.12 s.4.1]
 
@@
 We've added it, thanks!
+[p.12 s.4.1]
 
@@
 normal forms for these examples are: `Optional[int]`, `None`, and `Dyn`.
+[p.12 s.4.1]
 
@@
 Good point, we've switched `S` and `T`.
+[p.12 s.4.1 and p.13 s.4.2]
 
@@
 We have reworded the caption.
+[p.15 s.4.2.1]
 
@@
 (The revision also uses `\Gamma` instead of `Env`)
+[p.14 s.4.2.2]
 
@@
 rightly point out, they are not part of the model).
+[p.15 s.4.2.2]
 
@@
 Thanks!
+[p.14 s.4.2.2]
 
@@
 
+The revised syntax for the model addresses this.
+[p.12 s.4.1]
+
 > ### p24
@@
 Fixed.
+[p.26 and p.27]
 
@@
 Section 4 in the revised paper gives more details.
+[p.10 s.4]
 
@@
 - Field reads and writes use `expr` instead of `C` (example: C.x is now expr.x)
+
+[p.12 s.4.1]
 
@@
 
-This is allowed, but only indirectly. The definition of a function `f_0` can't
+This is allowed, but only indirectly. The definition of a function `f_0` cannot
 directly say that `f_0` has type Dyn, but it can declare the type `Dyn -> Dyn`
@@
 
-Yes, Bool and Int are technically classes. The paper says little about this,
+Yes, Bool and Int are technically classes. The paper does not discuss this,
 but the Redex model (and Static Python) give them full support as classes.
@@
 Fixed, thank you for catching this omission.
+[p.14 s.4.2.2]
 
@@
 We've removed the subscript.
+[p.14 s.4.2.2]
 
@@
 Ok. We have removed this explanatory sentence.
+[p.15 s.4.2.2]
 
@@
 paper is not. Section 4.2.2 in the revised paper makes this clear.
+[p.13 s.4.2.2]
 
@@
 We've added a conclusion section.
+[p.21 s.6.3.3]
 
@@
 sections: 2.3, 2.5, 6.3, and 8.
+[p.6 and p.8 and p.20 and p.24]
 
@@
 and materialization.
-
-TODO
+[p.15]
 
@@
 We have expanded the related work.
+[p.22 s.7]
 
@@
 caption to explain). The constructor needs to validate each element.
+[p.7 s.2.3]
 
@@
 instantiations that have appeared.
+[p.7 and p.10]
 
@@
 avoid this confusion.
+[p.13 s.2.4]

### 2022-05-16 pj-7.1/paper.tex
pj: add section on gradual soundness

@@
 %% TODO for camera-ready
-%% -
+%% - reviews-r2: clarify gradual soundness again, what do we mean?
+%%    (1) two layers of types+code for some algorithms ... esp. for data structures ... key to practical concrete
+%%    (2) progressive types for other code
+%%    ... in short, progressive types adapted to GT
+%%   - try section 2 for this
 
@@
 The advanced features of the \SP{} type system are offered on an opt-in basis
-(\emph{gradual soundness})
+(\emph{gradual soundness},~\cref{s:gradual-soundness})
 and arranged so that programmers can begin adding types one module at a time.
@@
 The \SP{} team plans to lift these restrictions in the future.
+
+\subsection{Gradual Soundness}
+\label{s:gradual-soundness}
+
+\SP{} represents a new approach to the design of gradually-typed
+languages.
+Whereas prior works choose to either force changes upon untyped code
+with \emph{concrete} types~\cite{wnlov-popl-2010,mt-oopsla-2017}
+or accommodate untyped code as-is with \emph{migratory}
+types~\cite{st-sfp-2006,tfffgksst-snapl-2017}, \SP{} provides a multi-level
+type system that supports both styles without expensive run-time checks.
+Shallow types accommodate Python code, but offer little in the way
+of type-based reasoning and type-directed optimization.
+Concrete types and primitive types give stronger type guarantees, but have
+drawbacks:
+concrete types force changes upon untyped code and primitive types
+add restrictions to typed code.
+The overall system, in which programmers can choose between untyped code
+and several increasingly-restrictive typed languages,
+is a combination of gradual typing and progressive
+types~\cite{pqk-onward-2012} that we call \emph{gradual soundness}.
+
+To be clear, gradual soundness has a different focus than gradual typing.
+Gradual typing in the traditional, broad sense is about finding a best-of-both worlds
+combination of typed and untyped
+code~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006}.
+Gradual soundness fits under this broad umbrella, and contributes the idea that multiple
+type systems may be needed for an optimal mix.
+Gradual typing in the refined sense calls for a dynamic type
+that satisfies the gradual guarantees~\cite{svcb-snapl-2015}.
+A gradually-sound language may choose to satisfy these guarantees.
+\SP{}, however, does not~(\cref{s:type-dynamic}).
 

### 2022-05-17 pj-7.1/paper.tex
pj: reword gt comparison, remove contradiction

@@
 
-To be clear, gradual soundness has a different focus than gradual typing.
-Gradual typing in the traditional, broad sense is about finding a best-of-both worlds
-combination of typed and untyped
-code~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006}.
+Compared to traditional gradual typing, gradual soundness is a refinement.
+The original gradual typing vision is to find a best-of-both-worlds combination
+of typed and untyped code~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006}.
 Gradual soundness fits under this broad umbrella, and contributes the idea that multiple
 type systems may be needed for an optimal mix.
-Gradual typing in the refined sense calls for a dynamic type
-that satisfies the gradual guarantees~\cite{svcb-snapl-2015}.
+Gradual typing in the refined sense~\cite{svcb-snapl-2015} calls for a dynamic
+type that satisfies the gradual guarantees.
 A gradually-sound language may choose to satisfy these guarantees.

### 2022-05-17 pj-7.1/paper.tex
pj: rewording, prep for camera-ready

@@
 % > be making it work -- Shriram 2021-09-30
-
-%% TODO after submission, before camera-ready
-%% - double-check section 4, can we be more convincing about soundness / what it is?
-%% - reword 'fields' to attribute/descriptor/property?
-%% -
-
-%% TODO for camera-ready
-%% - reviews-r2: clarify gradual soundness again, what do we mean?
-%%    (1) two layers of types+code for some algorithms ... esp. for data structures ... key to practical concrete
-%%    (2) progressive types for other code
-%%    ... in short, progressive types adapted to GT
-%%   - try section 2 for this
 
@@
 \author[a]{Kuang-Chen Lu}
-\authorinfo{(\email{LuKuangchen1024@gmail.com}) } %% { is a PhD student at Brown University.}
+\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
 \affiliation[a]{Brown University, Providence, RI, USA}
 \author[a]{Ben Greenman}
-\authorinfo{(\email{benjamin.l.greenman@gmail.com}) } %% {is a PLT member and a postdoc at Brown University.}
+\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a postdoc at Brown University.}
 \author[b]{Carl Meyer}
-\authorinfo{(\email{carljm@fb.com}) } %% FILL
+\authorinfo{(\email{carljm@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
 \affiliation[b]{Meta, Menlo Park, CA, USA}
 \author[b]{Dino Viehland}
-\authorinfo{(\email{dinoviehland@fb.com}) } %% FILL
+\authorinfo{(\email{dinoviehland@fb.com}) is a software engineer at Meta working on the Cinder Python runtime.
+Previously he has worked on other Python related efforts including Python Tools for Visual Studio and IronPython.}
 \author[b]{Aniket Panse}
-\authorinfo{(\email{aniketpanse@fb.com}) } %% FILL
+\authorinfo{(\email{aniketpanse@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
 \author[a]{Shriram Krishnamurthi}
-\authorinfo{(\email{shriram@gmail.com}) } %% {is the Vice President of Programming Languages (no, not really) at Brown University.}
-
-\titlerunning{Draft document---please do not distribute or cite!}
-\authorrunning{Draft document---please do not distribute or cite!}
+\authorinfo{(\email{shriram@gmail.com}) is the Vice President of Programming Languages (no, not really) at Brown University.}
+
+%\titlerunning{Draft document---please do not distribute or cite!}
+%\authorrunning{Draft document---please do not distribute or cite!}
 
@@
 \begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007</concept_id>
+       <concept_desc>Software and its engineering</concept_desc>
+       <concept_significance>500</concept_significance>
+   </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011039.10011311</concept_id>
+       <concept_desc>Software and its engineering~Semantics</concept_desc>
+       <concept_significance>100</concept_significance>
+   </concept>
+</ccs2012>
 \end{CCSXML}
-% \ccsdesc[100]{FILL}
-%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\ccsdesc[100]{Software and its engineering~Semantics}
 
@@
   %% License options include: cc-by (default), cc-by-nc
-  % license=cc-by,
+  %license=cc-by,
 }
@@
 
-Gradual typing has attracted significant interest as a solution to
+Gradual typing is an enticing solution to
 the debate between static and dynamic typing.
@@
 codebase while leaving the rest untyped.
-Run-time casts and checks enforce the assumptions that typed code makes about
+Run-time casts and checks can enforce the assumptions that typed code makes about
 untyped components, thereby ensuring that the types are sound no matter how
@@
 gradual types, formulating correctness properties and studying ever-more-descriptive
-types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
+type systems~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021}).
 A few academics have studied the cost of run-time checks
@@
     and ran both property-based soundness tests and over 250 tests adapted from the \SP{} regression suite.
-    The modeling effort revealed 20 significant issues in \SP{},
-    four of which were soundness bugs.
+    The modeling effort revealed 21 significant issues in \SP{},
+    five of which were soundness bugs.
 
@@
 may wish to focus on shallow types and JIT compilation as a
-first step toward sound gradual types.
+first step toward sound and optimized gradual types.
 Second, we want to entice researchers.
 The \SP{} type system has many noteworthy restrictions.
-For example, functions are supported only by shallow types
+For example, functions are supported only by shallow types,
 and method overrides that use the dynamic type are more constrained
@@
     dictionary values.
-    These checks enable optimizations within the function body.
+    These checks enable additional optimizations within the function body.
 
@@
 type ($\sptunion{\sptype, \sptnone}$, or
-$\sptoptional{\sptype}$); these unions are enforced at run-time.
+$\sptoptional{\sptype}$); these unions are enforced with run-time checks.
 
@@
 
-The behavior of the \SP{} dynamic type is subtle because dynamic code
-faces more restrictions than untyped Python code.
-Replacing part of a type with dynamic can lead to either a static error or
-a run-time error.
+The behavior of the \SP{} dynamic type is subtle.
+Dynamic-typed code faces more restrictions than untyped Python code.
+Because of the restrictions, replacing part of a type with dynamic can lead to
+either a static error or a run-time error.
 \Cref{fig:gg-failure} presents two examples, one for each kind of failure:
@@
 nonlocal changes.
-If one type changes to concrete, then several changes may need to happen:
+If one type changes to concrete, then several other changes may need to happen:
 all values that reach this type must be initialized as concrete,
@@
 descriptors or properties;
-typed fields 
 and
@@
 
-The implementation of vtables happens to be built on the Python \code{vectorcall}
+The implementation of vtables is built on the Python \code{vectorcall}
 protocol; it is not a from-scratch development.
@@
   \item
-    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
+    The constructor requires two type parameters and a Python dictionary (a $\sptrawpydict$).
     It checks that the elements of the dictionary match the key and value types.
@@
 In general, other checked datatypes have the same three components:
-a constructor, checked update functions, and a tag.
+a constructor, a set of checked update functions, and a tag.
 The tag is supported by a global registry of instantiated concrete types.
@@
 Cinder includes a JIT (just-in-time) compiler for its bytecode.
-At the moment, programmers enable the JIT by supplying a list of method names
+Programmers can enable the JIT by supplying a list of method names
 to the compiler.
@@
 \SP{} does not have a surface syntax for its dynamic
-type and allows missing and unknown types~(\cref{s:type-dynamic}).
+type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
     \begin{tablenotes}
-    \item {*} {\footnotesize In \SP{}, the check for $\sptobject$ accepts all values except for primitives~(\cref{s:c-types})}
+    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ accepts all values except for primitives~(\cref{s:c-types})}
     \end{tablenotes}
@@
 
-The first two rules are for function application.
+The first two typing rules are for function application.
 A typed function may be applied to an argument that matches its domain type,
@@
 existing codebase.
-Consequently, \SP{} supports interactions with Python modules in the only viable
-way: by letting the Python code run with zero static constraints and minimal
+Consequently, \SP{} supports interactions with Python modules
+by letting the Python code run with zero static constraints and minimal
 dynamic constraints.
 
-%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
 There are two cases in which \SP{} types impose dynamic constraints on untyped Python code.
@@
 structural types as defined by PEP~544~\cite{pep544}.
-MonNom interfaces may be a promising way to support these types.
+Interfaces may be a promising way to support these types.
 
@@
 few critical Cython modules.
-Cython improved performance by compiling these modules to C,
-but its partial adoption had led to an awkward workflow with a
+Cython had improved performance by compiling these modules to C,
+but its partial adoption also led to an awkward workflow with a
 few compiled modules spread across an interpreted codebase.
@@
 We conclude that the addition of \SP{} types to a codebase can significantly
-improve performance without major refactoring.
+improve performance without major refactoring, and that refactoring is not
+the main reason for Instagram's speedups.
 
@@
   \caption{Microbenchmark performance ratios}
-  %% FILL don't forget to update \numbenchmark{}
   \label{t:microbenchmark}
@@
 A study of mixed configurations (e.g.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
-may reveal performance bugs that have not yet shown up in production.
+may reveal performance bugs that have not appeared in production.
 
@@
 in way that reflects changes to the types~(\cref{s:checked-type}).
-The changes impose a maintenance burden, but enable performance improvements.
+The changes impose a maintenance burden, but enable performance improvements and low-cost checks.
 
@@
 using a catch-all \code{external} type~\cite{jspp}.
-None of these four implement gradual types as envisioned by researchers~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006};
+None of these four implement gradual types as envisioned by researchers~\cite{svcb-snapl-2015,tfffgksst-snapl-2017,g-snapl-2019};
 instead, they offer a pragmatic solution to interoperability problems.
@@
 
-\noindent{
+{
   \newcommand{\spsoundness}[1]{\href{https://github.com/facebookincubator/cinder/issues/#1}{\code{#1}}}
-The critical soundness issues were:
+\noindent{}The critical soundness issues were:
 \code{\#}\spsoundness{36},
@@
       \item
-        Mutable class attributes are covariant in Pyre and other
+        Mutable class attributes are unsoundly covariant in Pyre and other
         Python type checkers, including mypy~\cite{mypy}.
@@
     including byte strings, memory management, specific decorators
-    (\code{@staticmethod}), and optimizations.
+    (such as \code{@staticmethod}), and optimizations.
 
@@
 via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
-initial warmup run to ensure there is an up-to-date bytecode cache file and
-compilation time will not be included), and all individual data points are
+initial warmup run to update the bytecode cache file and ignoring compilation time),
+and all individual data points are
 recorded. All data points are in seconds.
@@
 JIT runs use a JIT list that includes only the benchmark code itself, to
-minimize overhead of compiling unrelated standard library code not used in the
+minimize the overhead of compiling standard library code not used in the
 benchmark.
@@
 \footnotesize\url{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}
+\end{center}
+
+\noindent\bmname{nbody} ran on an even later commit:
+
+\begin{center}
+\footnotesize\url{https://github.com/facebookincubator/cinder/commit/144a21e66aeae5cde08847482249396b2e16eccc}
 \end{center}

### 2022-05-18 pj-7.1/paper.tex
pj: bookkeeping
- add access dates to ALL urls
- doi or url for all bib entries

@@
 
-\overfullrule=1mm
+%\overfullrule=1mm
 
@@
 \author[a]{Shriram Krishnamurthi}
-\authorinfo{(\email{shriram@gmail.com}) is the Vice President of Programming Languages (no, not really) at Brown University.}
+\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}
 
@@
 Whereas prior works choose to either force changes upon untyped code
-with \emph{concrete} types~\cite{wnlov-popl-2010,mt-oopsla-2017}
+with \emph{concrete} types~\cite{wzlov-popl-2010,mt-oopsla-2017}
 or accommodate untyped code as-is with \emph{migratory}
@@
 It follows \SP{}'s approach to typing these features.
-The model is available at:
+The model is on GitHub (accessed 2022-05-17):
 \begin{center}
@@
 
-Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
+Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wzlov-popl-2010}
 and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
@@
 The 26 GitHub issues that arose from our work may be found at the following three links.
-A list of issue numbers follows each link.
+A list of issue numbers follows each link (all accessed 2022-05-17).
 
@@
 
-Benchmarks \bmname{richards} and \bmname{fannkuch} ran on the following commit:
-
-\begin{center}
-\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}
-\end{center}
-
-\noindent\bmname{deltablue} ran on a later commit:
-
-\begin{center}
-\footnotesize\url{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}
-\end{center}
-
-\noindent\bmname{nbody} ran on an even later commit:
-
-\begin{center}
-\footnotesize\url{https://github.com/facebookincubator/cinder/commit/144a21e66aeae5cde08847482249396b2e16eccc}
-\end{center}
+\begin{itemize}
+  \item
+    Benchmarks \bmname{richards} and \bmname{fannkuch} ran on commit
+    \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{d0d071}
+  \item
+    \bmname{deltablue} ran on commit
+    \href{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}{c9d14c}
+  \item
+    \bmname{nbody} ran on commit
+    \href{https://github.com/facebookincubator/cinder/commit/144a21e66aeae5cde08847482249396b2e16eccc}{144a21}
+\end{itemize}
 

### 2022-05-19 pj-7.1/programming-camera/paper.tex
pj: prepare src bundle for tobias

--- 2022-05-18 pj-7.1/paper.tex
@@
     \emph{Evidence}.
-    We present evidence that \SP{} improves performance with few code changes.
-    At Instagram, the web server has become significantly more efficient
-    after the application of \SP{} to high-profile modules.
+    We present evidence that \SP{} improves performance with few code changes
+    / refactorings.
+    The Instagram web server has become significantly more efficient
+    thanks to a gradual application of \SP{} to high-profile modules.
     Because the server code is closed-source, we additionally present data for
-    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
+    four public microbenchmarks~(\cref{s:microbenchmarks}).
     Using only shallow types, \SP{} runs slightly faster than Python
@@
 \SP{} does not have a surface syntax for its dynamic
-type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
+type and instead allows missing types~(\cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
 Cython had improved performance by compiling these modules to C,
-but its partial adoption also led to an awkward workflow with a
+but its partial adoption led to an awkward workflow with a
 few compiled modules spread across an interpreted codebase.
@@
 Because the \SP{} team has been changing code as well as types during its
-migrations, the question arises as to whether the changes themselves
-are responsible for more speedups than the types and gradual soundness.
+migrations, the question arises as to whether the refactoring caused
+more speedups than the types and gradual soundness.
 %% Carl: "I reviewed the ~30 conversion diffs"
-We therefore reviewed conversion diffs that significantly improved
-performance (\nnum{30}) to assess the extent of the code changes.
-There are nine common kinds of changes: five that satisfy the type checker
-and four that aim to improve performance.
+We therefore reviewed 30 conversion diffs that significantly improved
+performance to assess the extent of the code changes.
+Across the diffs, there are nine common kinds of changes: five that satisfy the
+type checker and four that aim to improve performance.
 
@@
 in way that reflects changes to the types~(\cref{s:checked-type}).
-The changes impose a maintenance burden, but enable performance improvements and low-cost checks.
+The changes impose a maintenance burden, but enable performance improvements
+and an efficient implementation of run-time checks.
 
@@
 via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
-initial warmup run to update the bytecode cache file and ignoring compilation time),
+initial warmup run to update the bytecode cache file and thereby ignore compilation time),
 and all individual data points are

### 2022-05-27 pj-7.1/paper.tex
pj: conference-specific edits
motivated by the long list of frequently-noticed issues from this page:

https://www.conference-publishing.com/Instructions.php?Event=PROGJ23MAIN&Paper=662b383ad35248c699a34f136f1eaeeb76909f9e

@@
-\documentclass[english,cleveref,submission]{programming}
+\documentclass[english,cleveref,crc]{programming}
 
@@
 %\titlerunning{Draft document---please do not distribute or cite!}
-%\authorrunning{Draft document---please do not distribute or cite!}
+\authorrunning{KC. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, and S. Krishnamurthi}
 
@@
 
+%% 2022-05-27: commented out, from submission
+%%\paperdetails{
+%%  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
+%%  perspective=scienceempirical,
+%%  %% State one or more areas, separated by a comma. (see 2.2)
+%%  %% Please see list of areas in http://programming-journal.org/cfp/
+%%  %% The list is open-ended, so use other areas if yours is/are not listed.
+%%  area={General-purpose programming},
+%%  %% License options include: cc-by (default), cc-by-nc
+%%  %license=cc-by,
+%%}
+
+%%% The following is specific to Programming Journal, Volume 7, Issue 1 and the paper
+%%% 'Gradual Soundness: Lessons from Static Python'
+%%% by Kuang-Chen Lu, Ben Greenman, Carl Meyer, Dino Viehland, Aniket Panse, and Shriram Krishnamurthi.
+%%%
 \paperdetails{
-  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
-  perspective=scienceempirical,
-  %% State one or more areas, separated by a comma. (see 2.2)
-  %% Please see list of areas in http://programming-journal.org/cfp/
-  %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={General-purpose programming},
-  %% License options include: cc-by (default), cc-by-nc
-  %license=cc-by,
+  submitted=2022-04-29,
+  published=2022-06-15,
+  year=2023,
+  volume=7,
+  issue=1,
+  articlenumber=2,
 }
@@
     \emph{Evidence}.
-    We present evidence that \SP{} improves performance with few code changes.
-    At Instagram, the web server has become significantly more efficient
-    after the application of \SP{} to high-profile modules.
+    We present evidence that \SP{} improves performance with few code changes
+    / refactorings.
+    The Instagram web server has become significantly more efficient
+    thanks to a gradual application of \SP{} to high-profile modules.
     Because the server code is closed-source, we additionally present data for
-    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
+    four public microbenchmarks~(\Cref{s:microbenchmarks}).
     Using only shallow types, \SP{} runs slightly faster than Python
@@
 This paper begins with an informal description of \SP{} in two parts.
-First, we present a user-oriented summary of the language~(\cref{s:tour}).
+First, we present a user-oriented summary of the language~(\Cref{s:tour}).
 Second, we present the key ingredients of the runtime system, Cinder,
-that supports \SP{}~(\cref{s:runtime}).
+that supports \SP{}~(\Cref{s:runtime}).
 \Cref{s:model} uses a small formal model to introduce our Redex mechanization
@@
 production and on public microbenchmarks.
-The paper concludes with related work~(\cref{s:related})
-and a final summary~(\cref{s:conclusion}).
+The paper concludes with related work~(\Cref{s:related})
+and a final summary~(\Cref{s:conclusion}).
 
@@
 \SP{} runs Python 3.8 programs with minimal changes to their
-behavior~(\cref{s:not-python}),
+behavior~(\Cref{s:not-python}),
 and it compiles code on-the-fly to support existing
@@
 The advanced features of the \SP{} type system are offered on an opt-in basis
-(\emph{gradual soundness},~\cref{s:gradual-soundness})
+(\emph{gradual soundness},~\Cref{s:gradual-soundness})
 and arranged so that programmers can begin adding types one module at a time.
@@
     Once \SP{} is enabled for a module, it compiles all code in that module
-    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
+    including expressions that have the dynamic type~(\Cref{s:type-dynamic}).
     If this behavior is problematic for certain expressions, they must move
@@
 typing~\cite{tf-icfp-2010,gsk-esop-2011}.
-Union types are not, however, generally supported at run-time~(\cref{s:eval-types}).
+Union types are not, however, generally supported at run-time~(\Cref{s:eval-types}).
 The only exception is binary unions with the Python \code{None}
@@
 Three other unsupported types bear special mention:
-first-class class and object types~(\cref{s:inheritance});
+first-class class and object types~(\Cref{s:inheritance});
 first-class function (callable) types; and
@@
 %%   %% tailored = describes exactly the migrations that SP allows
-%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
+%% \footnote{In addition, C types~(\Cref{s:c-types}) are incompatible with type dynamic.}
 \begin{itemize}
@@
 check all writes from untyped contexts.
-For comparison, \cref{t:shallow-vs-concrete} presents the worst-case costs of
+For comparison, \Cref{t:shallow-vs-concrete} presents the worst-case costs of
 shallow and concrete dict types.
@@
     Nested class declarations, first-class classes, and classes with
-    multiple parents default to un-optimized Python behavior~(\cref{s:impl}).
+    multiple parents default to un-optimized Python behavior~(\Cref{s:impl}).
 
@@
     For example, a method that returns a number cannot be overridden by a method
-    that returns the dynamic type; \cref{f:gg-failure-stat} shows the static error
+    that returns the dynamic type; \Cref{f:gg-failure-stat} shows the static error
     that arises from such an override.
@@
 % \SP{}
-The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).
+The wrappers are handled efficiently by a tailored implementation of vtables~(\Cref{s:vtable}).
 
@@
 A gradually-sound language may choose to satisfy these guarantees.
-\SP{}, however, does not~(\cref{s:type-dynamic}).
+\SP{}, however, does not~(\Cref{s:type-dynamic}).
 
@@
 to a direct function-call jump.
-%% Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
+%% Just-in-time compilation~(\Cref{s:jit}) can also upgrade vtable lookups to
 %% function calls.
@@
 use because they fall outside the scope of the model.
-\Cref{s:impl} summarizes the out-of-scope features and \cref{a:banned-test} gives
+\Cref{s:impl} summarizes the out-of-scope features and \Cref{a:banned-test} gives
 a detailed categorization.
@@
 Instagram.
-Overall, we made 26 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
+Overall, we made 26 bug reports~(\Cref{a:github-issues}) to the \SP{} team via GitHub.
 Five of these were soundness issues, one of which we could exploit to raise
@@
 booleans (which are the integers \code{0} and \code{1}), and strings.
-There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
+There are two data structures: dictionaries and checked dictionaries~(\Cref{s:checked-type}).
 Other expressions describe function calls, dictionary reads and writes,
@@
 \SP{} does not have a surface syntax for its dynamic
-type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
+type and instead allows missing types~(\Cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
 \SP{} does not enforce all surface types against untyped code, only the evaluation
-types presented in the next subsection~(\cref{s:eval-types}).
+types presented in the next subsection~(\Cref{s:eval-types}).
 For this reason, we omit the surface typing judgment, which is a kind of linter
@@
 
-Relative to \SP{} and our Redex mechanization, the formalization in \cref{f:surface-types} omits
+Relative to \SP{} and our Redex mechanization, the formalization in \Cref{f:surface-types} omits
 many details of Python including class variables, imports, conditionals, and exception handlers.
@@
 checked dict values that were initialized with exactly the same key and value type.
-As noted in~\cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
+As noted in~\Cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
 or $\spteval_1$ is the dynamic type.
@@
     \begin{tablenotes}
-    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ accepts all values except for primitives~(\cref{s:c-types})}
+    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ accepts all values except for primitives~(\Cref{s:c-types})}
     \end{tablenotes}
@@
 To show where \SP{} needs to insert casts, we present a selection of typing rules.
-Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
+Recall that a program declares variables, functions, and classes~(\Cref{f:surface-types}).
 These declarations fill a type environment ($\sptenv$).
 Relative to the current environment, the expressions within each statement must
-satisfy the typing judgment in~\cref{f:eval-types}.
-
-By contrast to our Redex model, the typing judgment outlined in~\cref{f:eval-types}
+satisfy the typing judgment in~\Cref{f:eval-types}.
+
+By contrast to our Redex model, the typing judgment outlined in~\Cref{f:eval-types}
 \emph{does not describe} a typechecking algorithm.
@@
 better left to the Redex mechanization, which supports a much
-larger subset of Python~(\cref{s:surface}).
+larger subset of Python~(\Cref{s:surface}).
 
@@
 a detailed proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
-Given a program written in the syntax of \cref{f:surface-types}, one can derive
+Given a program written in the syntax of \Cref{f:surface-types}, one can derive
 a Nom program with the same type boundaries by replacing every checked
@@
 First, observe that every evaluation type comes with a decidable
-cast~(\cref{t:cast}).
+cast~(\Cref{t:cast}).
 For basic types such as strings, a tag check clearly suffices.
@@
 handle.
-These include primitive types~(\cref{s:c-types}), module-level variables,
-and the boundary to untyped Python~(\cref{s:py-boundary}).
+These include primitive types~(\Cref{s:c-types}), module-level variables,
+and the boundary to untyped Python~(\Cref{s:py-boundary}).
 %% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
@@
 Other Python features are left out because their \SP{} semantics is identical
-to standard Python~(\cref{s:py-boundary}).
-Our model also does not cover the optimizer~(\cref{s:optimize}) because
+to standard Python~(\Cref{s:py-boundary}).
+Our model also does not cover the optimizer~(\Cref{s:optimize}) because
 a formal account of type-directed optimization would require substantial additional
@@
 For this reason, \SP{} compiles every typed function to check its inputs.
-The optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
+The optimizer can bypass these checks for typed-to-typed calls~(\Cref{s:optimize}).
 Similarly, concrete types such as $\sptrawchkdict$ must check writes that
@@
 The first creates a checked dictionary; the second allocates an uninitialized instance
-of an object~(\cref{s:vtable}).
+of an object~(\Cref{s:vtable}).
 Cinder comes with a similar instruction to build checked lists.
@@
 Instagram's internal profiling tools attribute a \CPUchange{} improvement
-in CPU efficiency~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
+in CPU efficiency~(\Cref{s:what-is-cpu}) to \SP{} conversions~(\Cref{s:migration-path}).
 This is a big improvement at Instagram scale.
@@
 Despite some initial regressions, none of the converted modules
-ran slower after small rewrites~(\cref{s:code-changes})
+ran slower after small rewrites~(\Cref{s:code-changes})
 and/or enabling the JIT on certain functions.
@@
 0.7\% improvement in web server CPU efficiency.
-Primitive types~(\cref{s:c-types}) were essential for matching Cython.
+Primitive types~(\Cref{s:c-types}) were essential for matching Cython.
 
@@
 During the first half of 2021, the \SP{} team identified hot
-modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
+modules and proposed types plus small code changes to the maintainers of these modules~(\Cref{s:code-changes}).
 The accepted proposals resulted in a 1\% improvement in CPU efficiency.
@@
     \ipara{Move unsupported Python features}
-    Code that uses the unsupported features listed in~\cref{s:not-python}
+    Code that uses the unsupported features listed in~\Cref{s:not-python}
     must be changed or moved to an untyped module.
@@
     Integers and booleans in hot code paths run fastest with
-    primitive types~(\cref{s:c-types}).
+    primitive types~(\Cref{s:c-types}).
     %%As noted above, primitives were required for \SP{} to match the performance of
@@
 support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
-\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).
+\SP{} opts for a simpler and more rigid design~(\Cref{s:type-dynamic}).
 
@@
 By contrast to these others, \SP{} asks programmers to change their code
-in way that reflects changes to the types~(\cref{s:checked-type}).
-The changes impose a maintenance burden, but enable performance improvements and low-cost checks.
+in way that reflects changes to the types~(\Cref{s:checked-type}).
+The changes impose a maintenance burden, but enable performance improvements
+and an efficient implementation of run-time checks.
 
@@
     Higher-order types are absent.
-    The focus is instead on first-order objects and standard data structures~(\cref{s:ts-context}).
+    The focus is instead on first-order objects and standard data structures~(\Cref{s:ts-context}).
   \item
@@
     and although it supports a dynamic type, dynamic type-checked code
-    does not have the same freedom as untyped code~(\cref{s:type-dynamic}).
+    does not have the same freedom as untyped code~(\Cref{s:type-dynamic}).
   \item
@@
     %% These types have been the top priority for \SP{}.
-    Concrete types are deeply sound, but add nonlocal constraints~(\cref{s:checked-type}).
+    Concrete types are deeply sound, but add nonlocal constraints~(\Cref{s:checked-type}).
     %% \SP{} supports few concrete types at the moment and plans to add more.
@@
     Gradual class hierarchies are permitted, but restricted to single inheritance.
-    Static-to-static method overrides cannot reduce the precision of types~(\cref{s:inheritance}).
+    Static-to-static method overrides cannot reduce the precision of types~(\Cref{s:inheritance}).
     %% ---a restriction
@@
   \item
-    Progressive primitive types are available to maximize performance~(\cref{{s:c-types}}).
+    Progressive primitive types are available to maximize performance~(\Cref{{s:c-types}}).
     %%These types are intentionally incompatible with the dynamic type to help programmers
@@
   Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
-
-{\sloppy
-\printbibliography
 }
@@
 
+\clearpage
+
+{\sloppy
+\printbibliography
+}
+
 \end{document}

### 2022-05-31 pj-7.1/programming-camera/paper.tex
pj: update camera-ready sources

@@
 
-\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
+\newcommand{\shorturl}[2]{\href{#1#2}{\texttt{#1#2}}}
 \newcommand{\SP}{Static Python}
@@
 \newcommand{\totalnum}[1]{$#1$ total}
-\newcommand{\pctnum}[1]{$#1\%$}
+\newcommand{\pctnum}[1]{\SI{#1}{\percent}}
 
@@
 \newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
-\newcommand{\CPUchange}{3.7\%}
+\newcommand{\CPUchange}{\SI{3.7}{\percent}}
 
@@
 \newcommand{\spann}[2]{#1\!:\!#2}
-\newcommand{\bmname}[1]{\textbf{#1}}
+\newcommand{\bmname}[1]{\textsf{#1}}
 \newcommand{\typefont}[1]{\mathsf{#1}}
@@
 
+\usepackage[shortcuts]{extdash}
+\usepackage{longtable}
+\usepackage{placeins}
 \usepackage{tikz}
@@
 %\titlerunning{Draft document---please do not distribute or cite!}
-%\authorrunning{Draft document---please do not distribute or cite!}
+\authorrunning{K.-C. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, S. Krishnamurthi}
 
@@
 
+%%\paperdetails{
+%%  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
+%%  perspective=scienceempirical,
+%%  %% State one or more areas, separated by a comma. (see 2.2)
+%%  %% Please see list of areas in http://programming-journal.org/cfp/
+%%  %% The list is open-ended, so use other areas if yours is/are not listed.
+%%  area={General-purpose programming},
+%%  %% License options include: cc-by (default), cc-by-nc
+%%  %license=cc-by,
+%%}
+
+%%% The following is specific to Programming Journal, Volume 7, Issue 1 and the paper
+%%% 'Gradual Soundness: Lessons from Static Python'
+%%% by Kuang-Chen Lu, Ben Greenman, Carl Meyer, Dino Viehland, Aniket Panse, and Shriram Krishnamurthi.
+%%%
 \paperdetails{
-  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
-  perspective=scienceempirical,
-  %% State one or more areas, separated by a comma. (see 2.2)
-  %% Please see list of areas in http://programming-journal.org/cfp/
-  %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={General-purpose programming},
-  %% License options include: cc-by (default), cc-by-nc
-  %license=cc-by,
+  submitted=2022-04-29,
+  published=2022-06-15,
+  year=2023,
+  volume=7,
+  issue=1,
+  articlenumber=2,
 }
@@
 
-  \vglue10pt
-
+  % \vglue10pt
   \paragraph{Context}
@@
   approach to gradual typing.
-
-  \vglue10pt
+  %
+  % \vglue10pt
 
@@
   to bridge the gap between untyped code and typed concrete code.
-
-  \vglue10pt
+  %
+  % \vglue10pt
 
@@
   regression test suite to develop this model.
-
-  \vglue10pt
+  %
+  % \vglue10pt
 
@@
   found several errors in the implementation, including fatal errors.
-
-  \vglue10pt
+  %
+  % \vglue10pt
 
@@
   requests handled per second at maximum CPU load.
-
-  \vglue10pt
+  %
+  % \vglue10pt
 
@@
 without overly restricting the promise of gradual typing.
-Industry researchers are sidestepping the problem with unsound types---for the most part.
+Industry researchers are sidestepping the problem with unsound types\--- for the most part.
 
@@
     Because the server code is closed-source, we additionally present data for
-    four public microbenchmarks~(\cref{s:microbenchmarks}).
+    four public microbenchmarks~(\Cref{s:microbenchmarks}).
     Using only shallow types, \SP{} runs slightly faster than Python
-    despite the costs of enforcing soundness~(avg. \pctnum{20}).
+    despite the costs of enforcing soundness~(avg.~\pctnum{20}).
     With fine-tuned types, \SP{} consistently outperforms the Python
-    baseline~(avg. \pctnum{70}).
+    baseline~(avg.~\pctnum{70}).
 
@@
   \(\begin{array}{l@{\qquad}r@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
-                                        & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
+    \toprule
+                                        & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\midrule
     \sptrawpydict                       &        O(1) &          O(1) &      \nocost{} &     \nocost{} &      \nocost{} \\
-    \sptchkdict{K}{V}                   &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1)
+    \sptchkdict{K}{V}                   &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1) \\
+    \bottomrule
   \end{array}\)
@@
 Casts between a primitive type and the dynamic type are rejected
-statically---another violation of the gradual guarantee---to encourage the
+statically\--- another violation of the gradual guarantee\--- to encourage the
 spread of primitives in typed code.
@@
 % Neither a module-level nor a closure-level variable may have a primitive
-% type---because untyped code can mutate such variables.
+% type\--- because untyped code can mutate such variables.
 
@@
 three methods to give confidence that our model matches reality.
-First, we manually reviewed the model in depth---using our
-own expertise---to look for non-conformance.  Second, 
+First, we manually reviewed the model in depth\--- using our
+own expertise\--- to look for non-conformance.  Second, 
 we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
@@
 
-\medskip
-\begin{minipage}[t]{0.25\columnwidth}~
-\end{minipage}\begin{minipage}[t]{0.4\columnwidth}
+% \medskip
+% \begin{minipage}[t]{0.25\columnwidth}~
+% \end{minipage}\begin{minipage}[t]{0.4\columnwidth}
+\noindent\null\hspace*{-.5em} % tobias pape.
+\begin{minipage}{\linewidth}
 \begin{lstlisting}
@@
 \SP{} does not have a surface syntax for its dynamic
-type and instead allows missing types~(\cref{s:type-dynamic}).
+type and instead allows missing types~(\Cref{s:type-dynamic}).
 We assume that all unions are written in a flat and simplified form, e.g., that
@@
     \begin{tabular}{l@{\quad}l}
-    {Eval. Type $\spteval$} & {Run-time check} \\\hline
+    \toprule
+    {Eval. Type $\spteval$} & {Run-time check} \\\midrule
     $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
@@
     $\sptbool$ & {Accepts booleans} \\
-    $\sptstr$ & {Accepts strings}
+    $\sptstr$ & {Accepts strings} \\
+    \bottomrule
     \end{tabular}
@@
 After all, the main benefit of a full formal semantics is to validate
-the behavior of complex expressions---and this job is
+the behavior of complex expressions\--- and this job is
 better left to the Redex mechanization, which supports a much
-larger subset of Python~(\cref{s:surface}).
+larger subset of Python~(\Cref{s:surface}).
 
@@
 %% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
-%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness
+%% 2022-01-13 don't think so \---  leave prims alone, we don't want to frighten reviews with unsoundness
 
@@
   \begin{tabular}{lll}
-    {Instruction} & {Purpose} & {Description} \\\hline
+    \toprule
+    {Instruction} & {Purpose} & {Description} \\\midrule
     \bcinst{CAST} & Soundness & Assert that a value matches a type \\
@@
     \bcinst{REFINE\_TYPE} & Optimization & Type declaration for the JIT \\
-
+    \bottomrule
 %%% Other bytecodes ... either from Carl's talk or the SP repo
@@
 Instagram's internal profiling tools attribute a \CPUchange{} improvement
-in CPU efficiency~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
+in CPU efficiency~(\Cref{s:what-is-cpu})~to \SP{} conversions~(\Cref{s:migration-path}).
 This is a big improvement at Instagram scale.
@@
 Furthermore, \SP{} types combined with Cinder resulted in a
-0.7\% improvement in web server CPU efficiency.
-Primitive types~(\cref{s:c-types}) were essential for matching Cython.
+\SI{0.7}{\percent} improvement in web server CPU efficiency.
+Primitive types~(\Cref{s:c-types}) were essential for matching Cython.
 
@@
 During the first half of 2021, the \SP{} team identified hot
-modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
-The accepted proposals resulted in a 1\% improvement in CPU efficiency.
+modules and proposed types plus small code changes to the maintainers of these modules~(\Cref{s:code-changes}).
+The accepted proposals resulted in a \SI{1}{\percent} improvement in CPU efficiency.
 During the second half of 2021, the \SP{} team applied the same
@@
 this one change resulted in over four hundred generated typed modules.
-Overall, the second-half changes added 2\% to the number of requests
+Overall, the second-half changes added \SI{2}{\percent} to the number of requests
 that servers could handle per second at maximum CPU load.
@@
   \centering
-  \begin{tabular}{lrr}
-    Name               & Typed / Python & Refined / Python \\\midrule
+  \begin{tabular}{lSS}\toprule
+    Name               & {Typed / Python} & {Refined / Python} \\\midrule
     \bmname{deltablue} &         0.59 &           0.30 \\
@@
     \bmname{nbody}     &         1.09 &           0.24 \\
-    \bmname{richards}  &         0.53 &           0.22
+    \bmname{richards}  &         0.53 &           0.22 \\
+    \bottomrule    
     %% fannkuch typedbasic2 = 0.90
@@
 Almost all ratios are less than one, showing that \SP{} achieves a speedup over
-Python (typed avg.~$0.8$, refined avg.~$0.3$).
+Python (typed avg.~\num{0.8}, refined avg.~\num{0.3}).
 The exceptions are the unrefined versions of \bmname{fannkuch} and
@@
 Changing a few types in \bmname{fannkuch} to primitives
-gives an immediate speedup (ratio = $0.90$).
+gives an immediate speedup (ratio = \num{0.90}).
 
@@
 \Cref{t:microbenchmark} does not analyze benchmark configurations that mix
-typed and untyped code. %% ---only fully-typed and untyped configurations.
+typed and untyped code. %% \--- only fully-typed and untyped configurations.
 A study of mixed configurations (e.g.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
@@
 By contrast to these others, \SP{} asks programmers to change their code
-in way that reflects changes to the types~(\cref{s:checked-type}).
+in way that reflects changes to the types~(\Cref{s:checked-type}).
 The changes impose a maintenance burden, but enable performance improvements
@@
     Gradual class hierarchies are permitted, but restricted to single inheritance.
-    Static-to-static method overrides cannot reduce the precision of types~(\cref{s:inheritance}).
-    %% ---a restriction
+    Static-to-static method overrides cannot reduce the precision of types~(\Cref{s:inheritance}).
+    %% \--- a restriction
     %% that enables optimizations but breaks the static gradual guarantee.
   \item
-    Progressive primitive types are available to maximize performance~(\cref{{s:c-types}}).
+    Progressive primitive types are available to maximize performance~(\Cref{{s:c-types}}).
     %%These types are intentionally incompatible with the dynamic type to help programmers
@@
 
-\acks{
+\acks
   \SP{} and the Cinder JIT were developed by
@@
   (specifically, Meyer and Krishnamurthi) met.
+  Thanks also to Tobias Pape for splendid \TeX{} assistance.
   This work was partly supported by Meta and by the US National Science Foundation.
   Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
-
-{\sloppy
-\printbibliography
-}
-
+
+\clearpage
 \appendix
@@
 We also observe significant untapped opportunity to improve Static Python's
-performance; for example, $20\%$ of Typed Fannkuch SP+JIT time is spent on
+performance; for example, \SI{20}{\percent} of Typed Fannkuch SP+JIT time is spent on
 bounds-checking array accesses; most of this runtime bounds-checking cost could
@@
 
+% For the remaining tables:
+\sisetup{mode = math, detect-mode = false, table-format=2.2}
+
 \begin{table}[tp]
@@
 
-\begin{tabular}{rrrrrr}
+\begin{tabular}{SSSSSS}
+  \toprule
   \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ \\
-  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ \\
-  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ \\
-  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ \\
-  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ \\
-  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ \\
-  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ \\
-  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ \\
-  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ \\
-  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ \\
+  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  0.57 & 1.29 & 7.86 & 8.36 & 10.24 & 16 \\
+  0.59 & 1.32 & 7.92 & 8.84 & 9.97 & 15.73 \\
+  0.58 & 1.33 & 7.89 & 8.9 & 10.57 & 16.2 \\
+  0.58 & 1.33 & 7.92 & 8.53 & 10.71 & 16.71 \\
+  0.56 & 1.33 & 7.83 & 9.15 & 9.89 & 16.62 \\
+  0.56 & 1.33 & 7.97 & 8.25 & 10.44 & 15.65 \\
+  0.55 & 1.32 & 8.18 & 8.78 & 10.07 & 16.22 \\
+  0.55 & 1.31 & 7.85 & 8.67 & 10.07 & 15.73 \\
+  0.56 & 1.31 & 7.96 & 8.52 & 10.14 & 15.82 \\\
+  0.57 & 1.32 & 8 & 8.35 & 10.57 & 16.66 \\
+  \bottomrule
 \end{tabular}
 
-\begin{tabular}{rrrrrr}
+\begin{tabular}{SSSSSS}
+  \toprule
   \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ \\
-  $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ \\
-  $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ \\
-  $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ \\
-  $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ \\
-  $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ \\
-  $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ \\
-  $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ \\
-  $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ \\
-  $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ \\
+  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  1.41 & 2.68 & 7.67 & 5.58 & 7.97 & 13.32 \\
+  1.38 & 2.6 & 7.9 & 5.92 & 7.25 & 13.59 \\
+  1.39 & 2.64 & 7.74 & 5.54 & 7.64 & 13.83 \\
+  1.38 & 2.65 & 7.8 & 6.06 & 7.24 & 13.95 \\
+  1.37 & 2.62 & 7.68 & 5.52 & 7.2 & 13.5 \\
+  1.39 & 2.58 & 7.8 & 5.63 & 7.27 & 13.52 \\
+  1.36 & 2.65 & 7.79 & 5.74 & 7.17 & 13.34 \\
+  1.43 & 2.66 & 7.82 & 5.56 & 7.54 & 13.68 \\
+  1.53 & 2.63 & 7.9 & 5.59 & 6.89 & 13.79 \\
+  1.37 & 2.71 & 7.95 & 5.74 & 6.91 & 13.5 \\
+  \bottomrule
 \end{tabular}
 
-\begin{tabular}{rrr}
+\begin{tabular}{SSS}
+  \toprule
   \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $2.6$ & $3.88$ & $10.17$ \\
-  $2.6$ & $4.01$ & $10.18$ \\
-  $2.61$ & $4.17$ & $10.73$ \\
-  $2.9$ & $3.92$ & $9.96$ \\
-  $2.63$ & $3.79$ & $10.04$ \\
-  $2.64$ & $3.95$ & $9.99$ \\
-  $2.62$ & $3.9$ & $10.16$ \\
-  $2.6$ & $3.84$ & $9.69$ \\
-  $2.6$ & $4.04$ & $9.82$ \\
-  $2.62$ & $3.85$ & $9.85$ \\
+  {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  2.6 & 3.88 & 10.17 \\
+  2.6 & 4.01 & 10.18 \\
+  2.61 & 4.17 & 10.73 \\
+  2.9 & 3.92 & 9.96 \\
+  2.63 & 3.79 & 10.04 \\
+  2.64 & 3.95 & 9.99 \\
+  2.62 & 3.9 & 10.16 \\
+  2.6 & 3.84 & 9.69 \\
+  2.6 & 4.04 & 9.82 \\
+  2.62 & 3.85 & 9.85 \\
+  \bottomrule
 \end{tabular}
@@
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}
+  \toprule
   \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ \\
-  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ \\
-  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ \\
-  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ \\
-  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ \\
-  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ \\
-  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ \\
-  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ \\
-  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ \\
-  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ \\
+  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  1.27 & 1.39 & 34.69 & 40.74 & 41 & 48.23 \\
+  1.32 & 1.25 & 34.35 & 40.23 & 40.94 & 49.03 \\
+  1.28 & 1.26 & 34.67 & 40.74 & 40.65 & 49.42 \\
+  1.23 & 1.31 & 35.71 & 40.94 & 39.7 & 49.35 \\
+  1.28 & 1.38 & 34.85 & 39.59 & 40.46 & 49.18 \\
+  1.26 & 1.34 & 33.96 & 40.45 & 39.99 & 47.79 \\
+  1.26 & 1.23 & 34.65 & 40.78 & 40.97 & 49.7 \\
+  1.27 & 1.35 & 33.83 & 40.96 & 41.78 & 47.92 \\
+  1.23 & 1.4 & 36.23 & 40.49 & 40.29 & 48.68 \\
+  1.23 & 1.27 & 36.05 & 40.72 & 40.11 & 48.29 \\
+  \bottomrule
 \end{tabular}
 
-\begin{tabular}{rrrrrr}
+\begin{tabular}{SSSSSS}
+  \toprule
   \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ \\
-  $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ \\
-  $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ \\
-  $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ \\
-  $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ \\
-  $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ \\
-  $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ \\
-  $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ \\
-  $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ \\
-  $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ \\
+  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  2.77 & 2.83 & 6.18 & 2.89 & 2.86 & 4.07 \\
+  2.85 & 2.8 & 6.12 & 2.94 & 2.9 & 4.06 \\
+  2.79 & 2.77 & 6.38 & 2.87 & 2.86 & 4.1 \\
+  2.87 & 2.81 & 6.23 & 2.87 & 2.85 & 4.29 \\
+  2.85 & 2.84 & 6.17 & 2.92 & 2.81 & 4.14 \\
+  2.87 & 2.8 & 6.21 & 2.94 & 2.8 & 4.06 \\
+  2.92 & 2.76 & 6.04 & 3.11 & 3.05 & 4.08 \\
+  2.78 & 2.76 & 6.13 & 3.1 & 2.86 & 4.19 \\
+  2.84 & 2.76 & 6.13 & 2.87 & 2.96 & 4.31 \\
+  2.92 & 2.79 & 6.17 & 2.91 & 2.96 & 4.1 \\
+  \bottomrule
 \end{tabular}
 
-\begin{tabular}{rrrrrr}
+\begin{tabular}{SSSSSS}
+  \toprule
   \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} \\
-  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $2.48$ & $2.52$ & $7.28$ & $3.16$ & $3.08$ & $4.59$ \\
-  $2.52$ & $2.48$ & $7.28$ & $3.26$ & $3.08$ & $4.51$ \\
-  $2.47$ & $2.53$ & $7.24$ & $3.11$ & $3.18$ & $4.59$ \\
-  $2.48$ & $2.57$ & $7.35$ & $3.18$ & $3.14$ & $4.73$ \\
-  $2.5$ & $2.56$ & $7.34$ & $3.14$ & $3.12$ & $4.63$ \\
-  $2.43$ & $2.47$ & $7.17$ & $3.23$ & $3.09$ & $4.53$ \\
-  $2.47$ & $2.69$ & $7.18$ & $3.41$ & $3.21$ & $4.48$ \\
-  $2.47$ & $2.45$ & $7.14$ & $3.43$ & $3.14$ & $4.53$ \\
-  $2.43$ & $2.55$ & $7.23$ & $3.08$ & $3.24$ & $4.53$ \\
-  $2.52$ & $2.49$ & $7.2$ & $3.16$ & $3.18$ & $4.45$ \\
+  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  2.48 & 2.52 & 7.28 & 3.16 & 3.08 & 4.59 \\
+  2.52 & 2.48 & 7.28 & 3.26 & 3.08 & 4.51 \\
+  2.47 & 2.53 & 7.24 & 3.11 & 3.18 & 4.59 \\
+  2.48 & 2.57 & 7.35 & 3.18 & 3.14 & 4.73 \\
+  2.5 & 2.56 & 7.34 & 3.14 & 3.12 & 4.63 \\
+  2.43 & 2.47 & 7.17 & 3.23 & 3.09 & 4.53 \\
+  2.47 & 2.69 & 7.18 & 3.41 & 3.21 & 4.48 \\
+  2.47 & 2.45 & 7.14 & 3.43 & 3.14 & 4.53 \\
+  2.43 & 2.55 & 7.23 & 3.08 & 3.24 & 4.53 \\
+  2.52 & 2.49 & 7.2 & 3.16 & 3.18 & 4.45 \\
+  \bottomrule
 \end{tabular}
 
-\begin{tabular}{rrr}
+\begin{tabular}{SSS}
+  \toprule
   \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-  $2.81$ & $2.94$ & $4.02$ \\
-  $2.66$ & $2.85$ & $4.22$ \\
-  $2.69$ & $2.78$ & $4.01$ \\
-  $2.75$ & $2.8$ & $4.01$ \\
-  $2.75$ & $2.92$ & $4.07$ \\
-  $2.78$ & $2.83$ & $4.36$ \\
-  $2.84$ & $2.81$ & $4.19$ \\
-  $2.7$ & $2.86$ & $4.25$ \\
-  $2.76$ & $2.78$ & $4.08$ \\
-  $2.89$ & $2.76$ & $4.16$ \\
+  {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+  2.81 & 2.94 & 4.02 \\
+  2.66 & 2.85 & 4.22 \\
+  2.69 & 2.78 & 4.01 \\
+  2.75 & 2.8 & 4.01 \\
+  2.75 & 2.92 & 4.07 \\
+  2.78 & 2.83 & 4.36 \\
+  2.84 & 2.81 & 4.19 \\
+  2.7 & 2.86 & 4.25 \\
+  2.76 & 2.78 & 4.08 \\
+  2.89 & 2.76 & 4.16 \\
+  \bottomrule
 \end{tabular}
@@
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}\toprule
     \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.3$ & $0.4$ & $1.32$ & $0.51$ & $0.64$ & $1.02$ \\
-    $0.32$ & $0.39$ & $1.23$ & $0.54$ & $0.64$ & $1$ \\
-    $0.31$ & $0.4$ & $1.24$ & $0.54$ & $0.64$ & $0.99$ \\
-    $0.3$ & $0.42$ & $1.31$ & $0.56$ & $0.65$ & $1.02$ \\
-    $0.28$ & $0.4$ & $1.24$ & $0.52$ & $0.65$ & $0.96$ \\
-    $0.28$ & $0.39$ & $1.24$ & $0.55$ & $0.65$ & $0.98$ \\
-    $0.29$ & $0.4$ & $1.27$ & $0.55$ & $0.68$ & $0.97$ \\
-    $0.3$ & $0.42$ & $1.29$ & $0.53$ & $0.67$ & $1$ \\
-    $0.29$ & $0.39$ & $1.28$ & $0.53$ & $0.69$ & $0.99$ \\
-    $0.28$ & $0.42$ & $1.32$ & $0.51$ & $0.64$ & $0.99$ \\
+    {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.3 & 0.4 & 1.32 & 0.51 & 0.64 & 1.02 \\
+    0.32 & 0.39 & 1.23 & 0.54 & 0.64 & 1 \\
+    0.31 & 0.4 & 1.24 & 0.54 & 0.64 & 0.99 \\
+    0.3 & 0.42 & 1.31 & 0.56 & 0.65 & 1.02 \\
+    0.28 & 0.4 & 1.24 & 0.52 & 0.65 & 0.96 \\
+    0.28 & 0.39 & 1.24 & 0.55 & 0.65 & 0.98 \\
+    0.29 & 0.4 & 1.27 & 0.55 & 0.68 & 0.97 \\
+    0.3 & 0.42 & 1.29 & 0.53 & 0.67 & 1 \\
+    0.29 & 0.39 & 1.28 & 0.53 & 0.69 & 0.99 \\
+    0.28 & 0.42 & 1.32 & 0.51 & 0.64 & 0.99 \\
+    \bottomrule
   \end{tabular}
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}
+    \toprule
     \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.6$ & $0.74$ & $1.32$ & $0.69$ & $0.82$ & $1.27$ \\
-    $0.59$ & $0.74$ & $1.32$ & $0.67$ & $0.84$ & $1.27$ \\
-    $0.57$ & $0.71$ & $1.31$ & $0.7$ & $0.89$ & $1.23$ \\
-    $0.58$ & $0.74$ & $1.23$ & $0.73$ & $0.82$ & $1.26$ \\
-    $0.6$ & $0.71$ & $1.32$ & $0.71$ & $0.84$ & $1.26$ \\
-    $0.59$ & $0.73$ & $1.4$ & $0.71$ & $0.84$ & $1.23$ \\
-    $0.56$ & $0.7$ & $1.37$ & $0.71$ & $0.85$ & $1.35$ \\
-    $0.58$ & $0.69$ & $1.3$ & $0.71$ & $0.82$ & $1.31$ \\
-    $0.58$ & $0.69$ & $1.27$ & $0.69$ & $0.81$ & $1.34$ \\
-    $0.6$ & $0.73$ & $1.39$ & $0.68$ & $0.8$ & $1.33$ \\
+    {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.6 & 0.74 & 1.32 & 0.69 & 0.82 & 1.27 \\
+    0.59 & 0.74 & 1.32 & 0.67 & 0.84 & 1.27 \\
+    0.57 & 0.71 & 1.31 & 0.7 & 0.89 & 1.23 \\
+    0.58 & 0.74 & 1.23 & 0.73 & 0.82 & 1.26 \\
+    0.6 & 0.71 & 1.32 & 0.71 & 0.84 & 1.26 \\
+    0.59 & 0.73 & 1.4 & 0.71 & 0.84 & 1.23 \\
+    0.56 & 0.7 & 1.37 & 0.71 & 0.85 & 1.35 \\
+    0.58 & 0.69 & 1.3 & 0.71 & 0.82 & 1.31 \\
+    0.58 & 0.69 & 1.27 & 0.69 & 0.81 & 1.34 \\
+    0.6 & 0.73 & 1.39 & 0.68 & 0.8 & 1.33 \\
+    \bottomrule
   \end{tabular}
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}
+    \toprule
     \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.54$ & $0.66$ & $1.38$ & $0.61$ & $0.76$ & $1.17$ \\
-    $0.51$ & $0.62$ & $1.44$ & $0.61$ & $0.75$ & $1.25$ \\
-    $0.55$ & $0.69$ & $1.41$ & $0.6$ & $0.82$ & $1.17$ \\
-    $0.54$ & $0.69$ & $1.41$ & $0.63$ & $0.75$ & $1.16$ \\
-    $0.54$ & $0.7$ & $1.41$ & $0.65$ & $0.75$ & $1.17$ \\
-    $0.55$ & $0.67$ & $1.4$ & $0.66$ & $0.75$ & $1.15$ \\
-    $0.52$ & $0.67$ & $1.43$ & $0.61$ & $0.75$ & $1.15$ \\
-    $0.52$ & $0.66$ & $1.4$ & $0.62$ & $0.75$ & $1.18$ \\
-    $0.52$ & $0.64$ & $1.51$ & $0.65$ & $0.76$ & $1.18$ \\
-    $0.5$ & $0.65$ & $1.41$ & $0.62$ & $0.79$ & $1.15$ \\
+    {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.54 & 0.66 & 1.38 & 0.61 & 0.76 & 1.17 \\
+    0.51 & 0.62 & 1.44 & 0.61 & 0.75 & 1.25 \\
+    0.55 & 0.69 & 1.41 & 0.6 & 0.82 & 1.17 \\
+    0.54 & 0.69 & 1.41 & 0.63 & 0.75 & 1.16 \\
+    0.54 & 0.7 & 1.41 & 0.65 & 0.75 & 1.17 \\
+    0.55 & 0.67 & 1.4 & 0.66 & 0.75 & 1.15 \\
+    0.52 & 0.67 & 1.43 & 0.61 & 0.75 & 1.15 \\
+    0.52 & 0.66 & 1.4 & 0.62 & 0.75 & 1.18 \\
+    0.52 & 0.64 & 1.51 & 0.65 & 0.76 & 1.18 \\
+    0.5 & 0.65 & 1.41 & 0.62 & 0.79 & 1.15 \\
+    \bottomrule
   \end{tabular}
 
-  \begin{tabular}{rrr}
+  \begin{tabular}{SSS}
+    \toprule
     \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-    (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.64$ & $0.78$ & $1.22$ \\
-    $0.64$ & $0.81$ & $1.25$ \\
-    $0.63$ & $0.81$ & $1.23$ \\
-    $0.62$ & $0.79$ & $1.2$ \\
-    $0.67$ & $0.81$ & $1.23$ \\
-    $0.7$ & $0.8$ & $1.23$ \\
-    $0.68$ & $0.78$ & $1.34$ \\
-    $0.66$ & $0.79$ & $1.22$ \\
-    $0.69$ & $0.85$ & $1.23$ \\
-    $0.68$ & $0.79$ & $1.24$ \\
+    {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.64 & 0.78 & 1.22 \\
+    0.64 & 0.81 & 1.25 \\
+    0.63 & 0.81 & 1.23 \\
+    0.62 & 0.79 & 1.2 \\
+    0.67 & 0.81 & 1.23 \\
+    0.7 & 0.8 & 1.23 \\
+    0.68 & 0.78 & 1.34 \\
+    0.66 & 0.79 & 1.22 \\
+    0.69 & 0.85 & 1.23 \\
+    0.68 & 0.79 & 1.24 \\
+    \bottomrule
   \end{tabular}
@@
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}
+    \toprule
     \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.17$ & $0.18$ & $3.61$ & $0.92$ & $0.88$ & $2.12$ \\
-    $0.17$ & $0.18$ & $3.57$ & $0.91$ & $0.88$ & $2.09$ \\
-    $0.17$ & $0.18$ & $3.68$ & $0.96$ & $0.88$ & $2.24$ \\
-    $0.17$ & $0.17$ & $3.82$ & $0.93$ & $0.9$ & $2.15$ \\
-    $0.17$ & $0.17$ & $3.72$ & $0.89$ & $0.9$ & $2.17$ \\
-    $0.18$ & $0.18$ & $3.63$ & $0.89$ & $0.9$ & $2.14$ \\
-    $0.18$ & $0.16$ & $3.5$ & $0.89$ & $0.89$ & $2.23$ \\
-    $0.18$ & $0.17$ & $3.45$ & $0.89$ & $0.93$ & $2.32$ \\
-    $0.2$ & $0.17$ & $3.61$ & $0.92$ & $0.88$ & $2.16$ \\
-    $0.2$ & $0.18$ & $3.63$ & $0.9$ & $0.92$ & $2.06$ \\
+    {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.17 & 0.18 & 3.61 & 0.92 & 0.88 & 2.12 \\
+    0.17 & 0.18 & 3.57 & 0.91 & 0.88 & 2.09 \\
+    0.17 & 0.18 & 3.68 & 0.96 & 0.88 & 2.24 \\
+    0.17 & 0.17 & 3.82 & 0.93 & 0.9 & 2.15 \\
+    0.17 & 0.17 & 3.72 & 0.89 & 0.9 & 2.17 \\
+    0.18 & 0.18 & 3.63 & 0.89 & 0.9 & 2.14 \\
+    0.18 & 0.16 & 3.5 & 0.89 & 0.89 & 2.23 \\
+    0.18 & 0.17 & 3.45 & 0.89 & 0.93 & 2.32 \\
+    0.2 & 0.17 & 3.61 & 0.92 & 0.88 & 2.16 \\
+    0.2 & 0.18 & 3.63 & 0.9 & 0.92 & 2.06 \\
+    \bottomrule
   \end{tabular}
 
-  \begin{tabular}{rrrrrr}
+  \begin{tabular}{SSSSSS}
+    \toprule
     \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.8$ & $0.81$ & $1.3$ & $0.83$ & $0.74$ & $1.2$ \\
-    $0.8$ & $0.8$ & $1.28$ & $0.78$ & $0.74$ & $1.22$ \\
-    $0.81$ & $0.8$ & $1.31$ & $0.75$ & $0.77$ & $1.22$ \\
-    $0.87$ & $0.86$ & $1.29$ & $0.74$ & $0.76$ & $1.18$ \\
-    $0.85$ & $0.8$ & $1.26$ & $0.76$ & $0.77$ & $1.19$ \\
-    $0.81$ & $0.81$ & $1.3$ & $0.77$ & $0.75$ & $1.22$ \\
-    $0.86$ & $0.79$ & $1.25$ & $0.76$ & $0.75$ & $1.17$ \\
-    $0.87$ & $0.83$ & $1.24$ & $0.74$ & $0.77$ & $1.16$ \\
-    $0.81$ & $0.81$ & $1.3$ & $0.74$ & $0.74$ & $1.17$ \\
-    $0.82$ & $0.83$ & $1.23$ & $0.76$ & $0.76$ & $1.16$ \\
+    {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.8 & 0.81 & 1.3 & 0.83 & 0.74 & 1.2 \\
+    0.8 & 0.8 & 1.28 & 0.78 & 0.74 & 1.22 \\
+    0.81 & 0.8 & 1.31 & 0.75 & 0.77 & 1.22 \\
+    0.87 & 0.86 & 1.29 & 0.74 & 0.76 & 1.18 \\
+    0.85 & 0.8 & 1.26 & 0.76 & 0.77 & 1.19 \\
+    0.81 & 0.81 & 1.3 & 0.77 & 0.75 & 1.22 \\
+    0.86 & 0.79 & 1.25 & 0.76 & 0.75 & 1.17 \\
+    0.87 & 0.83 & 1.24 & 0.74 & 0.77 & 1.16 \\
+    0.81 & 0.81 & 1.3 & 0.74 & 0.74 & 1.17 \\
+    0.82 & 0.83 & 1.23 & 0.76 & 0.76 & 1.16 \\
+    \bottomrule
   \end{tabular}
 
-  \begin{tabular}{rrr}
+  \begin{tabular}{SSS}
+    \toprule
     \colname{Orig} & \colname{Orig} & \colname{Orig} \\
-    (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
-    $0.78$ & $0.75$ & $1.15$ \\
-    $0.74$ & $0.73$ & $1.19$ \\
-    $0.75$ & $0.72$ & $1.19$ \\
-    $0.73$ & $0.73$ & $1.18$ \\
-    $0.83$ & $0.76$ & $1.16$ \\
-    $0.82$ & $0.72$ & $1.24$ \\
-    $0.75$ & $0.72$ & $1.15$ \\
-    $0.72$ & $0.76$ & $1.2$ \\
-    $0.72$ & $0.72$ & $1.15$ \\
-    $0.74$ & $0.73$ & $1.2$ \\
+    {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
+    0.78 & 0.75 & 1.15 \\
+    0.74 & 0.73 & 1.19 \\
+    0.75 & 0.72 & 1.19 \\
+    0.73 & 0.73 & 1.18 \\
+    0.83 & 0.76 & 1.16 \\
+    0.82 & 0.72 & 1.24 \\
+    0.75 & 0.72 & 1.15 \\
+    0.72 & 0.76 & 1.2 \\
+    0.72 & 0.72 & 1.15 \\
+    0.74 & 0.73 & 1.2 \\
+    \bottomrule
   \end{tabular}
@@
 
-\begin{table}[tp]
-  \caption{Exact commands to invoke each benchmark (1/2)}
-  \label{t:mb:commands1}
-  \tiny\centering
-
-  \begin{tabular}{ll}
-    \colname{Benchmark} & \colname{Command} \\\hline
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    (\colname{SP}) &  \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT}) & \texttt{ Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python Tools/benchmarks/richards\_static.py 100} \\
-    () &  \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    (\colname{SP}) &  \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT}) & \texttt{dcards Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python Tools/benchmarks/richards\_static\_basic.py 100} \\
-    () &  \\
-    Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/richards.py 100} \\
-    Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/richards.py 100} \\
-    Richards \colname{Orig} & \texttt{time ./python Tools/benchmarks/richards.py 100} \\
-    () &  \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT}) & \texttt{ Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python Tools/benchmarks/fannkuch\_static.py 5} \\
-    () &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT}) & \texttt{dcards Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    () &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    () &  \\
-    Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/fannkuch.py 5} \\
-    Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/fannkuch.py 5} \\
-    Fannkuch \colname{Orig} & \texttt{time ./python Tools/benchmarks/fannkuch.py 5} \\
-    () &  \\
-  \end{tabular}
-\end{table}
-
-\begin{table}[tp]
-  \caption{Exact commands to invoke each benchmark (2/2)}
-  \label{t:mb:commands2}
-  \tiny\centering
-
-  \begin{tabular}{ll}
-    \colname{Benchmark} & \colname{Command} \\\hline
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{SP} \colname{JIT}) & \texttt{s -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{JIT}) & \texttt{s Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python Tools/benchmarks/deltablue\_static.py 100} \\
-    () &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    () &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ildcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{JIT}) & \texttt{ildcards Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    () &  \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/deltablue.py 100} \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/deltablue.py 100} \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python Tools/benchmarks/deltablue.py 100} \\
-    () &  \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    (\colname{SP}) &  \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
-    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python Tools/benchmarks/nbody\_static.py} \\
-    () &  \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{SP} \colname{JIT}) & \texttt{rds -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    (\colname{SP}) &  \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
-    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python Tools/benchmarks/nbody\_static\_basic.py} \\
-    () &  \\
-    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/nbody.py} \\
-    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/nbody.py} \\
-    NBody \colname{Orig} & \texttt{time ./python Tools/benchmarks/nbody.py} \\
-    () &  \\
-  \end{tabular}
-
-\end{table}
+\FloatBarrier
+{
+% tobiaspape: this is the equivalent of \begin{wide}..\end{wide}, with a longtable
+\setlength{\LTleft}{-1.5cm plus 1fil}
+\setlength{\LTright}{-1.5cm plus 1fil}
+\setlength{\LTpre}{.5\baselineskip}
+\setlength{\LTpost}{.5\baselineskip}
+
+% tobiaspape: reduce the prominence of the factors and adjust the printing of the commands
+\let\origcolname\colname
+\renewcommand*\colname[1]{\textsmaller{\textbf{#1}}}
+\lstset{basicstyle=\footnotesize\ttfamily}
+
+% tobiaspape: similar to the  hook that listings/lstlistings uses for breaks:
+\newcommand*\postbreak{\raisebox{0ex}[0ex][0ex]{\ensuremath{\color{P@Blue}\hookleftarrow}}}
+\footnotesize
+\begin{longtable}{ll}
+\caption{Exact commands to invoke each benchmark}\\
+\toprule
+\origcolname{Benchmark} & \origcolname{Command} \\\midrule
+\endfirsthead
+\caption[]{Exact commands to invoke each benchmark \emph{(continued)}} \\
+\toprule
+\origcolname{Benchmark} & \origcolname{Command} \\\midrule
+\endhead
+\midrule
+\endfoot
+\bottomrule
+\endlastfoot
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    {(\colname{SP})} &  \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT})} & \lstinline! Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/richards_static.py 100! \\
+    {()} &  \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!dcards -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    {(\colname{SP})} &  \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!dcards Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/richards_static_basic.py 100! \\
+    {()} &  \\
+    Richards \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/richards.py 100! \\
+    Richards \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/richards.py 100! \\
+    Richards \colname{Orig} & \lstinline!time ./python Tools/benchmarks/richards.py 100! \\
+    {()} &  \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT})} & \lstinline! Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/fannkuch_static.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!dcards -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!dcards Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ldcards -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ldcards Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/fannkuch.py 5! \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/fannkuch.py 5! \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python Tools/benchmarks/fannkuch.py 5! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!s -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!s -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!s -X jit-shadow-frame Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!s Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/deltablue_static.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ldcards -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ldcards Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/deltablue_static_basic.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ildcards -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ildcards -X jit-shadow-frame Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ildcards Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/deltablue.py 100! \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/deltablue.py 100! \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python Tools/benchmarks/deltablue.py 100! \\
+    {()} &  \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    {(\colname{SP})} &  \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! jit-shadow-frame Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards To!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ols/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/nbody_static.py! \\
+    {()} &  \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!rds -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!rds -X install-strict-loader Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/nbody_static.py!\postbreak \\*
+    {(\colname{SP})} &  \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!rds -X jit-shadow-frame Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards To!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ols/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/nbody_static_basic.py! \\
+    {()} &  \\
+    NBody \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/nbody.py! \\
+    NBody \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/nbody.py! \\
+    NBody \colname{Orig} & \lstinline!time ./python Tools/benchmarks/nbody.py! \\
+    {()} &  \\
+\end{longtable}
+}
+
+\FloatBarrier
+\clearpage
+\printbibliography
 

### 2022-05-31 pj-7.1/paper.tex
pj: update ack, fix bib capitalization

@@
 %\titlerunning{Draft document---please do not distribute or cite!}
-\authorrunning{KC. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, and S. Krishnamurthi}
+\authorrunning{K.-C. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, S. Krishnamurthi}
 
@@
 
-%% 2022-05-27: commented out, from submission
 %%\paperdetails{
@@
     Using only shallow types, \SP{} runs slightly faster than Python
-    despite the costs of enforcing soundness~(avg. \pctnum{20}).
+    despite the costs of enforcing soundness~(avg.~\pctnum{20}).
     With fine-tuned types, \SP{} consistently outperforms the Python
-    baseline~(avg. \pctnum{70}).
+    baseline~(avg.~\pctnum{70}).
 
@@
 After all, the main benefit of a full formal semantics is to validate
-the behavior of complex expressions---and this job is
+the behavior of complex expressions\--- and this job is
 better left to the Redex mechanization, which supports a much
@@
 Instagram's internal profiling tools attribute a \CPUchange{} improvement
-in CPU efficiency~(\Cref{s:what-is-cpu}) to \SP{} conversions~(\Cref{s:migration-path}).
+in CPU efficiency~(\Cref{s:what-is-cpu})~to \SP{} conversions~(\Cref{s:migration-path}).
 This is a big improvement at Instagram scale.
@@
 Furthermore, \SP{} types combined with Cinder resulted in a
-0.7\% improvement in web server CPU efficiency.
+\SI{0.7}{\percent} improvement in web server CPU efficiency.
 Primitive types~(\Cref{s:c-types}) were essential for matching Cython.
@@
 modules and proposed types plus small code changes to the maintainers of these modules~(\Cref{s:code-changes}).
-The accepted proposals resulted in a 1\% improvement in CPU efficiency.
+The accepted proposals resulted in a \SI{1}{\percent} improvement in CPU efficiency.
 During the second half of 2021, the \SP{} team applied the same
@@
     Static-to-static method overrides cannot reduce the precision of types~(\Cref{s:inheritance}).
-    %% ---a restriction
+    %% \--- a restriction
     %% that enables optimizations but breaks the static gradual guarantee.
@@
   (specifically, Meyer and Krishnamurthi) met.
+  Thanks also to Tobias Pape for going above and beyond to offer splendid assistance with the paper typesetting.
   This work was partly supported by Meta and by the US National Science Foundation.
   Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-}
-
+
+\clearpage
 \appendix
@@
 
-\begin{table}[tp]
-  \caption{Exact commands to invoke each benchmark (1/2)}
-  \label{t:mb:commands1}
-  \tiny\centering
-
-  \begin{tabular}{ll}
-    \colname{Benchmark} & \colname{Command} \\\hline
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
-    (\colname{SP}) &  \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT}) & \texttt{ Tools/benchmarks/richards\_static.py 100} \\
-    Richards \colname{T-Max} & \texttt{time ./python Tools/benchmarks/richards\_static.py 100} \\
-    () &  \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
-    (\colname{SP}) &  \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT}) & \texttt{dcards Tools/benchmarks/richards\_static\_basic.py 100} \\
-    Richards \colname{T-Min} & \texttt{time ./python Tools/benchmarks/richards\_static\_basic.py 100} \\
-    () &  \\
-    Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/richards.py 100} \\
-    Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/richards.py 100} \\
-    Richards \colname{Orig} & \texttt{time ./python Tools/benchmarks/richards.py 100} \\
-    () &  \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
-    (\colname{JIT}) & \texttt{ Tools/benchmarks/fannkuch\_static.py 5} \\
-    Fannkuch \colname{T-Max} & \texttt{time ./python Tools/benchmarks/fannkuch\_static.py 5} \\
-    () &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
-    (\colname{JIT}) & \texttt{dcards Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
-    () &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    (\colname{SP}) &  \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
-    () &  \\
-    Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/fannkuch.py 5} \\
-    Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/fannkuch.py 5} \\
-    Fannkuch \colname{Orig} & \texttt{time ./python Tools/benchmarks/fannkuch.py 5} \\
-    () &  \\
-  \end{tabular}
-\end{table}
-
-\begin{table}[tp]
-  \caption{Exact commands to invoke each benchmark (2/2)}
-  \label{t:mb:commands2}
-  \tiny\centering
-
-  \begin{tabular}{ll}
-    \colname{Benchmark} & \colname{Command} \\\hline
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{SP} \colname{JIT}) & \texttt{s -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{JIT} \colname{SF}) & \texttt{s -X jit-shadow-frame Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static.txt -X jit-enable-jit-list-wildcard} \\
-    (\colname{JIT}) & \texttt{s Tools/benchmarks/deltablue\_static.py 100} \\
-    DeltaBlue \colname{T-Max} & \texttt{time ./python Tools/benchmarks/deltablue\_static.py 100} \\
-    () &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic.txt -X jit-enable-jit-list-wi} \\
-    (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic.py 100} \\
-    () &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ildcards -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    (\colname{SP}) &  \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ildcards -X jit-shadow-frame Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_deltablue\_static\_basic2.txt -X jit-enable-jit-list-w} \\
-    (\colname{JIT}) & \texttt{ildcards Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    DeltaBlue \colname{T-Min} & \texttt{time ./python Tools/benchmarks/deltablue\_static\_basic2.py 100} \\
-    () &  \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/deltablue.py 100} \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/deltablue.py 100} \\
-    DeltaBlue \colname{Orig} & \texttt{time ./python Tools/benchmarks/deltablue.py 100} \\
-    () &  \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{SP} \colname{JIT}) & \texttt{ install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    (\colname{SP}) &  \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards -X} \\
-    (\colname{JIT} \colname{SF}) & \texttt{ jit-shadow-frame Tools/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
-    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Max} & \texttt{time ./python Tools/benchmarks/nbody\_static.py} \\
-    () &  \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{SP} \colname{JIT}) & \texttt{rds -X install-strict-loader Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/nbody\_static.py} \\
-    (\colname{SP}) &  \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static\_basic.txt -X jit-enable-jit-list-wildca} \\
-    (\colname{JIT} \colname{SF}) & \texttt{rds -X jit-shadow-frame Tools/benchmarks/nbody\_static\_basic.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_nbody\_static.txt -X jit-enable-jit-list-wildcards To} \\
-    (\colname{JIT}) & \texttt{ols/benchmarks/nbody\_static.py} \\
-    NBody \colname{T-Min} & \texttt{time ./python Tools/benchmarks/nbody\_static\_basic.py} \\
-    () &  \\
-    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
-    (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/nbody.py} \\
-    NBody \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
-    (\colname{JIT}) & \texttt{hmarks/nbody.py} \\
-    NBody \colname{Orig} & \texttt{time ./python Tools/benchmarks/nbody.py} \\
-    () &  \\
-  \end{tabular}
-
-\end{table}
-
+\FloatBarrier
+{
+% tobiaspape: this is the equivalent of \begin{wide}..\end{wide}, with a longtable
+\setlength{\LTleft}{-1.5cm plus 1fil}
+\setlength{\LTright}{-1.5cm plus 1fil}
+\setlength{\LTpre}{.5\baselineskip}
+\setlength{\LTpost}{.5\baselineskip}
+
+% tobiaspape: reduce the prominence of the factors and adjust the printing of the commands
+\let\origcolname\colname
+\renewcommand*\colname[1]{\textsmaller{\textbf{#1}}}
+\lstset{basicstyle=\footnotesize\ttfamily}
+
+% tobiaspape: similar to the  hook that listings/lstlistings uses for breaks:
+\newcommand*\postbreak{\raisebox{0ex}[0ex][0ex]{\ensuremath{\color{P@Blue}\hookleftarrow}}}
+\footnotesize
+\begin{longtable}{ll}
+\caption{Exact commands to invoke each benchmark}\\
+\toprule
+\origcolname{Benchmark} & \origcolname{Command} \\\midrule
+\endfirsthead
+\caption[]{Exact commands to invoke each benchmark \emph{(continued)}} \\
+\toprule
+\origcolname{Benchmark} & \origcolname{Command} \\\midrule
+\endhead
+\midrule
+\endfoot
+\bottomrule
+\endlastfoot
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/richards_static.py 100! \\
+    {(\colname{SP})} &  \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT})} & \lstinline! Tools/benchmarks/richards_static.py 100! \\
+    Richards \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/richards_static.py 100! \\
+    {()} &  \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!dcards -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/richards_static_basic.py 100! \\
+    {(\colname{SP})} &  \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_richards_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!dcards Tools/benchmarks/richards_static_basic.py 100! \\
+    Richards \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/richards_static_basic.py 100! \\
+    {()} &  \\
+    Richards \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/richards.py 100! \\
+    Richards \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/richards.py 100! \\
+    Richards \colname{Orig} & \lstinline!time ./python Tools/benchmarks/richards.py 100! \\
+    {()} &  \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! -X jit-shadow-frame Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static.txt -X jit-enable-jit-list-wildcards!\postbreak \\*
+    {(\colname{JIT})} & \lstinline! Tools/benchmarks/fannkuch_static.py 5! \\
+    Fannkuch \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/fannkuch_static.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!dcards -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dcards -X jit-shadow-frame Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic.txt -X jit-enable-jit-list-wil!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!dcards Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/fannkuch_static_basic.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ldcards -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    {(\colname{SP})} &  \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_fannkuch_static_basic2.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ldcards Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    Fannkuch \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/fannkuch_static_basic2.py 5! \\
+    {()} &  \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/fannkuch.py 5! \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/fannkuch.py 5! \\
+    Fannkuch \colname{Orig} & \lstinline!time ./python Tools/benchmarks/fannkuch.py 5! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!s -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!s -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!s -X jit-shadow-frame Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static.txt -X jit-enable-jit-list-wildcard!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!s Tools/benchmarks/deltablue_static.py 100! \\
+    DeltaBlue \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/deltablue_static.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ldcards -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static_basic.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ldcards -X jit-shadow-frame Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic.txt -X jit-enable-jit-list-wi!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ldcards Tools/benchmarks/deltablue_static_basic.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/deltablue_static_basic.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!ildcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!ildcards -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    {(\colname{SP})} &  \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!ildcards -X jit-shadow-frame Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_deltablue_static_basic2.txt -X jit-enable-jit-list-w!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ildcards Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    DeltaBlue \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/deltablue_static_basic2.py 100! \\
+    {()} &  \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/deltablue.py 100! \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/deltablue.py 100! \\
+    DeltaBlue \colname{Orig} & \lstinline!time ./python Tools/benchmarks/deltablue.py 100! \\
+    {()} &  \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline! jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline! install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/nbody_static.py! \\
+    {(\colname{SP})} &  \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards -X!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline! jit-shadow-frame Tools/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards To!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ols/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Max} & \lstinline!time ./python Tools/benchmarks/nbody_static.py! \\
+    {()} &  \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{SP} \colname{JIT} \colname{SF})} & \lstinline!rds -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{SP} \colname{JIT})} & \lstinline!rds -X install-strict-loader Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X install-strict-loader Tools/benchmarks/nbody_static.py!\postbreak \\*
+    {(\colname{SP})} &  \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static_basic.txt -X jit-enable-jit-list-wildca!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!rds -X jit-shadow-frame Tools/benchmarks/nbody_static_basic.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_nbody_static.txt -X jit-enable-jit-list-wildcards To!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!ols/benchmarks/nbody_static.py! \\
+    NBody \colname{T-Min} & \lstinline!time ./python Tools/benchmarks/nbody_static_basic.py! \\
+    {()} &  \\
+    NBody \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards -X jit-sha!\postbreak \\*
+    {(\colname{JIT} \colname{SF})} & \lstinline!dow-frame Tools/benchmarks/nbody.py! \\
+    NBody \colname{Orig} & \lstinline!time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist_main.txt -X jit-enable-jit-list-wildcards Tools/benc!\postbreak \\*
+    {(\colname{JIT})} & \lstinline!hmarks/nbody.py! \\
+    NBody \colname{Orig} & \lstinline!time ./python Tools/benchmarks/nbody.py! \\
+    {()} &  \\
+\end{longtable}
+}
+
+\FloatBarrier
 \clearpage
-
-{\sloppy
 \printbibliography
-}
 
